---
title: "Assignment 1: Getting Started with R"
date: "2026-04-05"
execute:
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
```

::: {.callout-note}
## Assignment Details
**Assigned:** Wednesday, April 1 (Session 2)
**Due:** Sunday, April 5 at 11:59 PM
**Submit:** R script (`.R` file) on Canvas
:::

::: {.callout-tip}
## New to R?
See the step-by-step guides: [Setting Up an R Project](/resource/r-projects.qmd) | [Using R Scripts](/resource/r-scripts.qmd)
:::

## Overview

This assignment gets you comfortable with RStudio and introduces you to creating visualizations with ggplot2. You'll work with the built-in `mpg` dataset to create your first plots.

## Setup

1. Open RStudio and your `psy410` project
2. Create a new R script: File → New File → R Script
3. Save it as `assignment-01-YOURNAME.R`
4. At the top of your script, add:

```r
# Assignment 1: Getting Started with R
# Your Name
# Date

library(tidyverse)
```

## Tasks

### Task 1: Explore the data (10 points)

Use the `mpg` dataset (loaded automatically with tidyverse).

a) Use `glimpse()` to look at the structure of `mpg`. In a comment, note how many rows and columns there are.

b) Use `?mpg` to read the documentation. In a comment, explain what `displ` and `hwy` represent.

::: {.callout-note collapse="true"}
## Answer

**Part a:** `glimpse()` gives you a quick overview of the structure.

```{r}
glimpse(mpg)
# 234 rows, 11 columns
```

**Part b:** `?mpg` opens the help page in RStudio — run this in your console, it won't produce output in a document.

```{r}
#| eval: false
?mpg
```

```r
# displ: engine displacement, in litres — a measure of engine size.
#         Larger values = bigger engine.
# hwy: highway miles per gallon — fuel efficiency on the highway.
#       Larger values = better fuel economy.
```
:::

### Task 2: Create a basic scatterplot (15 points)

Create a scatterplot showing the relationship between engine displacement (`displ`) and highway fuel efficiency (`hwy`).

Your plot should:

- Have displacement on the x-axis
- Have highway mpg on the y-axis
- Include a clear title
- Have labeled axes (not just variable names)

::: {.callout-note collapse="true"}
## Answer

```{r}
#| fig-width: 7
#| fig-height: 4
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  labs(
    title = "Engine Size vs. Highway Fuel Efficiency",
    x = "Engine Displacement (liters)",
    y = "Highway MPG"
  )
```
:::

### Task 3: Add color (20 points)

Modify your scatterplot to color the points by vehicle class (`class`).

- The legend should appear automatically
- Give the legend a clear title (hint: use `labs()`)

::: {.callout-note collapse="true"}
## Answer

```{r}
#| fig-width: 7
#| fig-height: 4
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  labs(
    title = "Engine Size vs. Highway Fuel Efficiency by Vehicle Class",
    x = "Engine Displacement (liters)",
    y = "Highway MPG",
    color = "Vehicle Class"
  )
```
:::

### Task 4: Create faceted plots (20 points)

Create the same scatterplot (displacement vs highway mpg, colored by class), but now faceted by drive type (`drv`).

- Use `facet_wrap()`
- Each panel should be clearly labeled

::: {.callout-note collapse="true"}
## Answer

```{r}
#| fig-width: 9
#| fig-height: 4
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  facet_wrap(~drv) +
  labs(
    title = "Engine Size vs. Highway MPG by Drive Type",
    x = "Engine Displacement (liters)",
    y = "Highway MPG",
    color = "Vehicle Class"
  )
```

The three panels are `4` (four-wheel drive), `f` (front-wheel drive), and `r` (rear-wheel drive). Rear-wheel drive vehicles cluster at high displacement and low mpg — mostly sports cars and trucks.
:::

### Task 5: Save your plot (15 points)

Save your final faceted plot as a PNG file.

```r
ggsave("mpg_faceted_plot.png", width = 10, height = 6)
```

Verify the file was created in your project folder.

::: {.callout-note collapse="true"}
## Answer

`ggsave()` saves the most recently created plot. Run this immediately after your `ggplot()` code.

```{r}
#| eval: false
ggsave("mpg_faceted_plot.png", width = 10, height = 6)
```

If you want to be explicit about which plot to save, store it in an object first:

```{r}
#| eval: false
my_plot <- ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  facet_wrap(~drv) +
  labs(
    title = "Engine Size vs. Highway MPG by Drive Type",
    x = "Engine Displacement (liters)",
    y = "Highway MPG",
    color = "Vehicle Class"
  )

ggsave("mpg_faceted_plot.png", plot = my_plot, width = 10, height = 6)
```

Check your Files pane in RStudio — `mpg_faceted_plot.png` should appear in your project folder.
:::

### Task 6: Reflection (10 points)

At the bottom of your script, answer these questions in comments:

```r
# Reflection Questions:
# 1. What relationship do you observe between engine size and fuel efficiency?
# 2. Do all vehicle classes follow the same pattern?
# 3. What was the most challenging part of this assignment?
```

::: {.callout-note collapse="true"}
## Sample Answers

Reflection questions don't have a single right answer, but here's what a strong response looks like:

```r
# 1. What relationship do you observe between engine size and fuel efficiency?
# There is a clear negative relationship: as engine displacement increases,
# highway fuel efficiency decreases. Larger engines tend to get fewer miles
# per gallon.

# 2. Do all vehicle classes follow the same pattern?
# Mostly yes, but there are differences by class. SUVs and pickups cluster
# in the high-displacement, low-mpg region. Compact and midsize cars cluster
# in the low-displacement, high-mpg region. When faceted by drive type,
# rear-wheel drive vehicles are almost entirely large-engine, low-efficiency
# vehicles (sports cars and pickups). Front-wheel drive vehicles span a wider
# range of engine sizes but have better fuel economy overall.

# 3. What was the most challenging part of this assignment?
# (Individual answer — common responses: file paths for ggsave, figuring out
# where the PNG was saved, understanding the difference between color inside
# vs. outside aes().)
```
:::

## Submission Checklist

Before submitting, make sure:

- [ ] Your script runs without errors from start to finish
- [ ] All tasks are completed with code and comments
- [ ] Your PNG file was created successfully
- [ ] Your name is at the top of the script

Submit your `.R` file on Canvas.

## Grading Rubric

| Task | Points |
|------|--------|
| Task 1: Data exploration | 10 |
| Task 2: Basic scatterplot | 15 |
| Task 3: Color mapping | 20 |
| Task 4: Faceting | 20 |
| Task 5: Saving plot | 15 |
| Task 6: Reflection | 10 |
| Code runs without errors | 10 |
| **Total** | **100** |
