---
title: "Assignment 2: Data Transformation"
date: "2026-04-12"
execute:
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(nycflights13)
```

::: {.callout-note}
## Assignment Details
**Assigned:** Wednesday, April 8 (Session 4)
**Due:** Sunday, April 12 at 11:59 PM
**Submit:** R script (`.R` file) on Canvas
:::

::: {.callout-tip}
## Getting started
See the step-by-step guides: [Setting Up an R Project](/resource/r-projects.qmd) | [Using R Scripts](/resource/r-scripts.qmd)
:::

## Overview

This assignment practices data transformation using dplyr verbs and the pipe operator. You'll work with the `nycflights13` dataset to answer questions about flight delays.

## Setup

```r
# Assignment 2: Data Transformation
# Your Name
# Date

library(tidyverse)
library(nycflights13)

# If you haven't installed nycflights13:
# install.packages("nycflights13")
```

## Part 1: Filtering and Arranging (25 points)

### Task 1.1
Find all flights that departed from JFK in June and were delayed by more than 1 hour. How many flights meet these criteria?

::: {.callout-note collapse="true"}
## Answer

```{r}
jfk_june_delayed <- flights |>
  filter(origin == "JFK", month == 6, dep_delay > 60)

jfk_june_delayed
nrow(jfk_june_delayed)
```
:::

### Task 1.2
Of those flights, which one had the longest departure delay? Use `arrange()` to find out. Report the carrier, destination, and delay time.

::: {.callout-note collapse="true"}
## Answer

```{r}
flights |>
  filter(origin == "JFK", month == 6, dep_delay > 60) |>
  arrange(desc(dep_delay))

# The first row is the worst offender — look at carrier, dest, and dep_delay.
```
:::

### Task 1.3
Find all flights to Los Angeles (LAX) or San Francisco (SFO) that departed on time or early. How many were there?

::: {.callout-note collapse="true"}
## Answer

```{r}
flights |>
  filter(dest %in% c("LAX", "SFO"), dep_delay <= 0) |>
  nrow()
```

`dep_delay <= 0` captures both on-time (0) and early (negative) departures. A common mistake here is using `dest == "LAX" | dest == "SFO"` instead of `dest %in% c("LAX", "SFO")`. Both work, but `%in%` is cleaner when checking one variable against multiple values.
:::

## Part 2: Selecting and Mutating (35 points)

### Task 2.1
Create a new dataset that contains only:

- `carrier`
- `origin`
- `dest`
- `dep_delay`
- `arr_delay`

Call this dataset `flights_subset`.

::: {.callout-note collapse="true"}
## Answer

```{r}
flights_subset <- flights |>
  select(carrier, origin, dest, dep_delay, arr_delay)

flights_subset
```
:::

### Task 2.2
Using `flights_subset`, create two new variables:

- `delay_diff`: the difference between arrival delay and departure delay
- `dep_delay_hrs`: the departure delay in hours (not minutes)

::: {.callout-note collapse="true"}
## Answer

::: {.callout-tip}
This builds on Task 2.1 — in your own script, make sure you've created `flights_subset` first.
:::

```{r}
flights_subset <- flights_subset |>
  mutate(
    delay_diff = arr_delay - dep_delay,
    dep_delay_hrs = dep_delay / 60
  )

flights_subset
```
:::

### Task 2.3
What does a positive `delay_diff` mean? What does a negative value mean? Answer in a comment, then find a flight that "made up time" (arrived less delayed than it departed).

::: {.callout-note collapse="true"}
## Answer

::: {.callout-tip}
This builds on Tasks 2.1 and 2.2 — `delay_diff` must exist in `flights_subset` before running this.
:::

```{r}
# A positive delay_diff means the flight arrived MORE delayed than it departed —
# it lost additional time in the air (e.g., circling, headwinds).
#
# A negative delay_diff means the flight "made up time" — it arrived
# less delayed than it departed (e.g., favorable tailwinds, shorter route).

flights_subset |>
  filter(dep_delay > 0, delay_diff < 0) |>
  arrange(delay_diff)
```
:::

## Part 3: Grouped Summaries (30 points)

### Task 3.1
Calculate the average departure delay for each carrier. Which carrier has the worst average delay?

::: {.callout-note collapse="true"}
## Answer

```{r}
flights |>
  group_by(carrier) |>
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE)) |>
  arrange(desc(avg_dep_delay))

# Frontier Airlines (F9) has the worst average departure delay.
# na.rm = TRUE is essential — without it, any carrier with even one
# missing value returns NA for the whole mean.
```
:::

### Task 3.2
Calculate the average departure delay for each origin airport, by month. Which month is the worst for delays at each airport?

::: {.callout-note collapse="true"}
## Answer

```{r}
flights |>
  group_by(origin, month) |>
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE)) |>
  arrange(origin, desc(avg_dep_delay))

# Summer months (June/July) tend to be worst across all three airports.
```

`group_by(origin, month)` groups by the combination of both variables, giving one row per airport-month pair.
:::

### Task 3.3
Create a summary table showing, for each carrier:

- Average departure delay
- Average arrival delay
- Number of flights
- Number of destinations served (`n_distinct()`)

Arrange by number of flights (descending).

::: {.callout-note collapse="true"}
## Answer

```{r}
flights |>
  group_by(carrier) |>
  summarize(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    n_flights = n(),
    n_dest = n_distinct(dest)
  ) |>
  arrange(desc(n_flights))
```

`n()` counts rows (total flights) while `n_distinct(dest)` counts unique values — these serve very different purposes.
:::

## Putting It Together

Write a single piped sequence that:

1. Filters to United Airlines (UA) flights
2. Removes rows with missing arrival delay
3. Groups by destination
4. Calculates mean arrival delay and number of flights
5. Filters to destinations with at least 100 flights
6. Arranges by mean delay (worst first)

What are the top 3 worst destinations for United delays?

::: {.callout-note collapse="true"}
## Answer

```{r}
flights |>
  filter(carrier == "UA") |>
  filter(!is.na(arr_delay)) |>
  group_by(dest) |>
  summarize(
    mean_arr_delay = mean(arr_delay),
    n_flights = n()
  ) |>
  filter(n_flights >= 100) |>
  arrange(desc(mean_arr_delay))
```

A common mistake is putting `filter(n_flights >= 100)` *before* `summarize()` — at that point, `n_flights` doesn't exist yet. The filter on a summary variable always has to come after `summarize()`.

Note: we don't need `na.rm = TRUE` inside `mean()` here because `filter(!is.na(arr_delay))` already removed the missing rows. Both approaches work, but the explicit filter step makes the cleaning visible.
:::

## Grading Rubric

| Component | Points |
|-----------|--------|
| Part 1: Filtering & Arranging | 25 |
| Part 2: Selecting & Mutating | 35 |
| Part 3: Grouped Summaries | 30 |
| Code runs without errors | 10 |
| **Total** | **100** |

## Submission

- [ ] Script runs without errors
- [ ] All tasks completed with clear comments
- [ ] Answers to questions in comments
- [ ] Your name at the top

Submit your `.R` file on Canvas.
