---
title: "Assignment 3: Tidying & Importing Data"
date: "2026-04-26"
execute:
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(readxl)

anxiety_wide <- tibble(
  participant_id = 1:5,
  condition = c("treatment", "treatment", "control", "control", "treatment"),
  anxiety_t1 = c(45, 52, 48, 55, 42),
  anxiety_t2 = c(38, 45, 47, 54, 35),
  anxiety_t3 = c(32, 40, 46, 52, 30)
)

emotions_long <- tibble(
  participant = rep(1:4, each = 3),
  emotion = rep(c("happy", "sad", "anxious"), 4),
  rating = c(7, 2, 3, 5, 4, 6, 8, 1, 2, 6, 5, 4)
)
```

::: {.callout-note}
## Assignment Details
**Assigned:** Monday, April 20 (Session 7)
**Due:** Sunday, April 26 at 11:59 PM
**Submit:** Quarto document (`.qmd`) AND rendered HTML on Canvas
:::

::: {.callout-tip}
## First Quarto assignment!
This is the first assignment submitted as a Quarto document. See the guides: [Setting Up an R Project](/resource/r-projects.qmd) | [Using Quarto Documents](/resource/quarto-docs.qmd)
:::

## Overview

This assignment practices reshaping data with `pivot_longer()` and `pivot_wider()`, and importing data from CSV files. You'll work with psychology-style datasets.

## Setup

```r
# Assignment 3: Tidying & Importing Data
# Your Name
# Date

library(tidyverse)
```

## Part 1: Pivoting Wide to Long (30 points)

A researcher collected anxiety scores at three time points. The data is in "wide" format:

```r
anxiety_wide <- tibble(
  participant_id = 1:5,
  condition = c("treatment", "treatment", "control", "control", "treatment"),
  anxiety_t1 = c(45, 52, 48, 55, 42),
  anxiety_t2 = c(38, 45, 47, 54, 35),
  anxiety_t3 = c(32, 40, 46, 52, 30)
)
```

### Task 1.1
Pivot this data to long format so you have columns for:

- `participant_id`
- `condition`
- `time` (values: "t1", "t2", "t3")
- `anxiety` (the score)

::: {.callout-note collapse="true"}
## Answer

```{r}
anxiety_long <- anxiety_wide |>
  pivot_longer(
    cols = starts_with("anxiety_"),
    names_to = "time",
    values_to = "anxiety",
    names_prefix = "anxiety_"
  )

anxiety_long
```

`names_prefix = "anxiety_"` strips the prefix so you get clean labels (`t1`, `t2`, `t3`) instead of `anxiety_t1`, etc.
:::

### Task 1.2
Using your long dataset, calculate the mean anxiety score at each time point, separately for each condition.

::: {.callout-note collapse="true"}
## Answer

::: {.callout-tip}
This builds on Task 1.1 — `anxiety_long` must exist first.
:::

```{r}
anxiety_means <- anxiety_long |>
  group_by(condition, time) |>
  summarise(mean_anxiety = mean(anxiety), .groups = "drop")

anxiety_means
```
:::

### Task 1.3
Create a line plot showing anxiety over time, with separate lines for each condition. Add points for the individual observations.

::: {.callout-note collapse="true"}
## Answer

The cleanest version uses `stat_summary()` so the line is computed from the data and the individual observations are still visible as points.

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Line plot of anxiety scores at three time points (t1, t2, t3), with separate lines for treatment and control conditions and individual participant observations shown as points. Treatment shows a clear decline; control stays roughly flat."
ggplot(anxiety_long, aes(x = time, y = anxiety,
                         colour = condition, group = condition)) +
  geom_point(alpha = 0.6) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  labs(
    title = "Anxiety over time by condition",
    x = "Time point",
    y = "Anxiety score",
    colour = "Condition"
  )
```
:::

## Part 2: Pivoting Long to Wide (25 points)

A survey measured different emotions. The data is in long format:

```r
emotions_long <- tibble(
  participant = rep(1:4, each = 3),
  emotion = rep(c("happy", "sad", "anxious"), 4),
  rating = c(7, 2, 3, 5, 4, 6, 8, 1, 2, 6, 5, 4)
)
```

### Task 2.1
Pivot this to wide format so each emotion is its own column.

::: {.callout-note collapse="true"}
## Answer

```{r}
emotions_wide <- emotions_long |>
  pivot_wider(names_from = emotion, values_from = rating)

emotions_wide
```

The column order (`happy / sad / anxious`) reflects the order the emotions appear in the data. That's fine.
:::

### Task 2.2
Using the wide format, create a new variable that is the sum of all three emotion ratings for each participant.

::: {.callout-note collapse="true"}
## Answer

::: {.callout-tip}
This builds on Task 2.1 — `emotions_wide` must exist first.
:::

```{r}
emotions_wide |>
  mutate(total = happy + sad + anxious)
```
:::

## Part 3: Importing Data (35 points)

Download the provided data files from Canvas and save them in your project's `data/raw/` folder:

- `survey_data.csv` — Survey responses with some messy formatting
- `demographics.xlsx` — Participant demographics in Excel format

### Task 3.1
Import `survey_data.csv` using `read_csv()`. Note any warnings or issues.

::: {.callout-note collapse="true"}
## Answer

```{r}
survey_first <- read_csv("../files/data/survey_data.csv")
glimpse(survey_first)
```

Notice that `post_anxiety` and `post_wellbeing` come in as **character** instead of numeric. That's because the literal string `"N/A"` appears in those columns, and `read_csv()` doesn't know it's meant to represent a missing value. We'll fix that in Task 3.2.
:::

### Task 3.2
The CSV has some problems:

- Missing values are coded as "N/A" instead of blank
- Some columns have wrong types

Re-import the data handling these issues using the appropriate `read_csv()` arguments (hint: `na = ` and `col_types = `).

::: {.callout-note collapse="true"}
## Answer

Setting `na = "N/A"` is enough on its own — once the strings are treated as missing, the columns parse as numeric automatically.

```{r}
survey <- read_csv("../files/data/survey_data.csv", na = "N/A")
glimpse(survey)
```

If you want to be explicit about column types, you can add `col_types`:

```{r}
#| eval: false
read_csv(
  "../files/data/survey_data.csv",
  na = "N/A",
  col_types = cols(
    participant_id  = col_integer(),
    condition       = col_character(),
    pre_anxiety     = col_double(),
    post_anxiety    = col_double(),
    pre_wellbeing   = col_double(),
    post_wellbeing  = col_double()
  )
)
```
:::

### Task 3.3
Import `demographics.xlsx` using the `readxl` package. The data is on the second sheet.

```r
library(readxl)
# Your code here
```

::: {.callout-note collapse="true"}
## Answer

```{r}
demographics <- read_excel("../files/data/demographics.xlsx", sheet = 2)
demographics
```

You can also reference the sheet by name (`sheet = "demographics"`), which is more readable. **Watch out:** sheet 1 is the codebook (5 rows describing each variable), not the demographics data. Make sure you're pulling sheet 2.
:::

## Grading Rubric

| Task | Points |
|------|--------|
| 1.1: Pivot to long format | 10 |
| 1.2: Calculate means by time and condition | 10 |
| 1.3: Create line plot | 10 |
| 2.1: Pivot to wide format | 15 |
| 2.2: Create sum of emotion ratings | 10 |
| 3.1: Import CSV with `read_csv()` | 10 |
| 3.2: Re-import handling NA values and column types | 15 |
| 3.3: Import Excel with `readxl` | 10 |
| QMD file knits without errors | 10 |
| **Total** | **100** |

**Grading scale per task:** Full points = correct. Full - 1 = one minor error (typo, small formatting issue). Half credit = one major error. 0 = not attempted or multiple major errors. "QMD file knits" refers to whether the Quarto document renders successfully, not whether individual code produces correct output.

## Submission

- [ ] All pivot operations produce correct output
- [ ] CSV and Excel files imported successfully
- [ ] Document renders without errors

Submit your `.qmd` file **and** your rendered `.html` file on Canvas.

---

::: {.callout-note}
## PSY 510 (Graduate Students)
Students enrolled in PSY 510 must complete the following extension in addition to all tasks above.
:::

## Graduate Extension: Qualtrics API

Most psychology researchers collect data through Qualtrics. Instead of logging in and downloading a CSV by hand, you can pull data directly into R using the `qualtRics` package — a workflow that scales to repeated data collection and eliminates manual steps that introduce error.

For this extension you'll pull data from the anonymous start-of-term survey that your classmates completed in Week 1. Because it was collected anonymously, the data are safe to work with directly — no masking needed.

### Setup

```r
# install.packages("qualtRics")
library(qualtRics)
```

### Task G.1

Authenticate with the Qualtrics API using your UO API key and data center ID. You'll find both in Qualtrics under Account Settings → Qualtrics IDs.

```r
qualtrics_api_credentials(
  api_key  = "YOUR_API_KEY",
  base_url = "YOUR_DATACENTER.qualtrics.com"
)
```

### Task G.2

Pull responses from the start-of-term survey using `fetch_survey()`. The survey ID is posted on Canvas.

```r
survey_id <- "SV_XXXXXXXXXXXXXXX"  # replace with actual ID from Canvas
raw <- fetch_survey(surveyID = survey_id)
glimpse(raw)
```

### Task G.3

The raw API output includes a lot of Qualtrics metadata columns (timing, location, status flags) alongside the actual responses. Select only the columns that correspond to the survey questions and give them informative names.

### Task G.4

Compare your cleaned API output to the manually exported CSV version of the same survey (posted on Canvas). Note at least two differences in column names, data types, or structure. What would you need to do to make them match exactly?

### Task G.5

Write a short reflection (~half a page) answering: When would you use the API rather than a manual export in your own research? What are the tradeoffs in terms of effort, reliability, and reproducibility?

**Submission:** Add your code and reflection to your `.qmd` file under a clearly marked `## Graduate Extension` section.
