---
title: "Assignment 4: Visualization Deep Dive"
date: "2026-05-03"
execute:
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(palmerpenguins)

# Simulated partisan_bias data used to render the Part 4 answer key.
# Students download the real file from Canvas to data/raw/partisan_bias.csv.
set.seed(2026)
n_per_team <- 30
partisan_data <- tibble(
  participant_id = 1:(3 * n_per_team),
  team = rep(1:3, each = n_per_team),
  SP_c = c(rnorm(n_per_team, -0.20, 0.30),
           rnorm(n_per_team,  0.30, 0.30),
           rnorm(n_per_team,  0.05, 0.30)),
  GR_c = c(rnorm(n_per_team,  0.30, 0.30),
           rnorm(n_per_team, -0.20, 0.30),
           rnorm(n_per_team,  0.05, 0.30))
)
```

::: {.callout-note}
## Assignment Details
**Assigned:** Monday, April 27 (Session 9)
**Due:** Sunday, May 3 at 11:59 PM
**Submit:** Quarto document (`.qmd`) AND rendered HTML on Canvas
:::

::: {.callout-tip}
## Getting started
See the guides: [Setting Up an R Project](/resource/r-projects.qmd) | [Using Quarto Documents](/resource/quarto-docs.qmd)
:::

## Overview

This assignment focuses on creating publication-quality visualizations. You'll practice using different geoms, scales, themes, and design principles.

## Setup

```r
# Assignment 4: Visualization Deep Dive
# Your Name
# Date

library(tidyverse)
library(palmerpenguins)  # install.packages("palmerpenguins") if needed
```

## Part 1: Bar Charts with Error Bars (25 points)

Use the `palmerpenguins` dataset.

### Task 1.1
Create a summary table showing mean flipper length, standard deviation (SD), and sample size (N) for each species.

```r
# Hint: use group_by() + summarize() with mean(), sd(), and n().
# Don't forget na.rm = TRUE for the mean and sd.
```

::: {.callout-note collapse="true"}
## Answer

```{r}
penguins |>
  group_by(species) |>
  summarize(
    mean_flipper = mean(flipper_length_mm, na.rm = TRUE),
    sd_flipper   = sd(flipper_length_mm,   na.rm = TRUE),
    n            = n()
  )
```

`n()` counts every row in the group, including rows where `flipper_length_mm` is `NA`. If you want only the rows that contributed to the mean, use `sum(!is.na(flipper_length_mm))` instead.
:::

### Task 1.2
Create a bar chart showing mean flipper length by species with error bars representing the **95% confidence interval**. Use `stat_summary()` for both the bar and the error bars — this is the pattern from the Session 8 slides.

Your plot should:

- Have clear axis labels (not variable names)
- Use a clean theme
- Have a descriptive title that names what the error bars represent (or include it in a caption)
- Use a colorblind-friendly palette (hint: `scale_fill_viridis_d()`)

::: {.callout-tip}
## Hint
You'll need `stat_summary()` twice — once for the bar and once for the error bars:

```r
ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  stat_summary(fun = mean, geom = "bar") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2)
```

`mean_cl_normal` returns the 95% confidence interval (the Session 8 slides note this is the most common error bar in psychology).
:::

::: {.callout-note collapse="true"}
## Answer

```{r}
#| fig-alt: "Bar chart of mean flipper length in millimeters by penguin species (Adelie, Chinstrap, Gentoo) with 95% confidence interval error bars. Gentoo penguins have the longest flippers by a clear margin."
ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  stat_summary(fun = mean, geom = "bar") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2) +
  scale_fill_viridis_d() +
  labs(
    title   = "Mean flipper length by species",
    caption = "Error bars represent 95% confidence intervals.",
    x       = "Species",
    y       = "Flipper length (mm)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The legend duplicates the x-axis labels, so it's safe to drop it with `theme(legend.position = "none")`. Always say what the error bars represent — in the title, caption, or a figure note.
:::

## Part 2: Distributions (25 points)

### Task 2.1
Create a histogram of penguin body mass. Try at least three different bin widths in your scratch work to see how the shape changes, then submit a single histogram with the bin width you think works best. In a comment, say what bin width you chose and why.

::: {.callout-note collapse="true"}
## Answer

```{r}
#| fig-alt: "Histogram of penguin body mass in grams using 250 g bins, showing a bimodal distribution."
# Bin width of 250 g shows the bimodal shape clearly without being so jagged
# that the structure gets lost — wider bins (e.g., 500 g) smooth over the
# second mode, narrower bins (e.g., 100 g) are too noisy.
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of penguin body mass",
    x     = "Body mass (g)",
    y     = "Count"
  ) +
  theme_minimal()
```

There's no single correct bin width — what matters is that you tried a few and made a deliberate choice. Anything in the 200–400 g range here is defensible.
:::

### Task 2.2
Create overlapping density plots of body mass, colored by species. Set transparency so all distributions are visible.

::: {.callout-tip}
## Hint
To color the densities by species, set `fill = species` inside `aes()`, then add `alpha = 0.5` (or similar) inside `geom_density()` to make the overlap visible.
:::

::: {.callout-note collapse="true"}
## Answer

```{r}
#| fig-alt: "Overlapping density plots of body mass by penguin species. Adelie and Chinstrap distributions overlap heavily in the 3,000-4,000 g range; Gentoo sits cleanly above them around 5,000 g."
ggplot(penguins, aes(x = body_mass_g, fill = species)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d() +
  labs(
    title = "Body mass by species",
    x     = "Body mass (g)",
    y     = "Density",
    fill  = "Species"
  ) +
  theme_minimal()
```

The histogram in 2.1 showed a bimodal shape; the density plot reveals why — Gentoo penguins are noticeably larger than the other two species. Adelie and Chinstrap overlap almost entirely.
:::

### Task 2.3
Create a boxplot of body mass by species. Then recreate it as a violin plot. In a comment, discuss what each shows that the other doesn't.

::: {.callout-note collapse="true"}
## Answer

```{r}
#| fig-alt: "Boxplot of body mass by penguin species. Gentoo's median sits above the upper whisker of both Adelie and Chinstrap."
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(x = "Species", y = "Body mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| fig-alt: "Violin plot of body mass by penguin species, showing the full distribution shape for each."
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_violin() +
  scale_fill_viridis_d() +
  labs(x = "Species", y = "Body mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")
```

The boxplot is great for spotting the median, IQR, and outliers at a glance — five-number summary, no shape. The violin shows the full distribution, so you can see whether the data are bimodal, skewed, or symmetric, but you lose the explicit quartile markers. Layering both (`geom_violin() + geom_boxplot(width = 0.1)`) is a common compromise.
:::

## Part 3: The "Bad" and "Good" Graph (25 points)

Using any dataset of your choice (`mpg`, `penguins`, `diamonds`, etc.):

### Task 3.1: The Bad Graph
Create an intentionally bad visualization that violates at least 4 design principles. In comments, list what makes it bad.

Examples of "bad" choices (pick things you can actually do in ggplot2):

- Rainbow color scheme (`scale_color_gradientn(colors = rainbow(7))`)
- Missing or unhelpful axis labels (variable names instead of plain English)
- Truncated y-axis that exaggerates differences (`coord_cartesian(ylim = ...)`)
- Way too many categories shown with the same palette
- Overlapping text or unreadable font sizes
- Chartjunk: garish background colors, redundant gridlines, distracting fills
- A geom that doesn't fit the data (e.g., a line plot connecting unordered categories)

::: {.callout-note collapse="true"}
## Answer

There's no single right answer — credit goes to plots that violate at least four principles *intentionally* and name each violation. Here's one example.

```{r}
#| fig-alt: "Intentionally bad bar chart of mean highway mpg by car class, with a rainbow palette, truncated y-axis, missing axis labels, a hot pink background, and an overly busy title."
# Bad choices made on purpose:
# 1. Rainbow palette (no perceptual ordering, bad for colorblind viewers)
# 2. Truncated y-axis exaggerating tiny differences
# 3. Variable names instead of plain-English labels
# 4. Garish background and 3-D-style gridlines (chartjunk)
# 5. Redundant legend (color and x-axis encode the same thing)
ggplot(mpg, aes(x = class, y = hwy, fill = class)) +
  stat_summary(fun = mean, geom = "bar") +
  scale_fill_manual(values = rainbow(7)) +
  coord_cartesian(ylim = c(15, 30)) +
  labs(title = "AVERAGE HWY MPG!!!", x = "class", y = "hwy") +
  theme(
    panel.background = element_rect(fill = "hotpink"),
    panel.grid.major = element_line(color = "yellow", linewidth = 1.2),
    axis.text.x      = element_text(angle = 0, size = 6)
  )
```
:::

### Task 3.2: The Good Graph
Create a good version of the same data. In comments, explain your design choices.

::: {.callout-note collapse="true"}
## Answer

```{r}
#| fig-alt: "Bar chart of mean highway mpg by car class, reordered from highest to lowest, on a y-axis starting at zero with plain-English labels and a single muted fill color."
# Good choices:
# 1. Y-axis starts at zero — no exaggeration of differences
# 2. Classes reordered by mean mpg so the ranking is the story
# 3. Single, muted fill color (no decorative palette)
# 4. Plain-English labels and a descriptive title
# 5. Minimal theme with no chartjunk
mpg |>
  group_by(class) |>
  summarize(mean_hwy = mean(hwy)) |>
  mutate(class = fct_reorder(class, mean_hwy)) |>
  ggplot(aes(x = mean_hwy, y = class)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Compact and subcompact cars get the best highway mileage",
    x     = "Mean highway mpg",
    y     = NULL
  ) +
  theme_minimal()
```

A horizontal bar chart with reordered categories is almost always easier to read than a vertical chart with seven labels crammed along the x-axis. Stating the takeaway in the title turns the chart from "data display" into "argument."
:::

## Part 4: Recreate a Figure (15 points)

This task uses data from a real psychology study on partisan bias in news evaluation. Participants were assigned to one of three teams (Team Spain, Team Greece, or No Team) and read news statements slanted toward one side or the other. The outcome is an acceptance threshold (*c*) from signal detection theory — higher values mean the participant required stronger evidence before accepting a statement as true.

Download `partisan_bias.csv` from Canvas and save it in your `data/raw/` folder. The file has four columns:

| Column | Description |
|--------|-------------|
| `participant_id` | Unique participant identifier |
| `team` | Group assignment: 1 = Team Spain, 2 = Team Greece, 3 = No Team |
| `SP_c` | Acceptance threshold for Pro-Spain statements |
| `GR_c` | Acceptance threshold for Pro-Greece statements |

### Task 4.1
Import the data and convert `team` from a number to a factor with labels "Spain", "Greece", and "No Team" (in that order).

::: {.callout-tip}
## Hint
The safest way to recode is to specify both `levels` and `labels` so the mapping is explicit:

```r
mutate(team = factor(team,
                     levels = c(1, 2, 3),
                     labels = c("Spain", "Greece", "No Team")))
```
:::

::: {.callout-note collapse="true"}
## Answer

```{r}
#| eval: false
partisan_data <- read_csv("data/raw/partisan_bias.csv")
```

```{r}
partisan_data <- partisan_data |>
  mutate(team = factor(team,
                       levels = c(1, 2, 3),
                       labels = c("Spain", "Greece", "No Team")))

partisan_data
```

Specifying both `levels` and `labels` is safer than relying on the default ordering — if a value of 4 or 0 ever sneaks into the data, it becomes `NA` rather than getting silently mis-labeled.
:::

### Task 4.2
Recreate Figure 3 from the paper (provided on Canvas) as closely as you can. Pay attention to:

- Geom type and positioning
- Error bars
- Axis labels and legend title
- Theme

You won't match it exactly — plain text axis labels are fine.

::: {.callout-tip}
## Hint
The data has `SP_c` and `GR_c` as two separate columns (one row per participant). To plot acceptance threshold by team and statement type, you'll need to **pivot to long format first** so that statement type becomes one column and the *c* value becomes another. Look back at Session 5 (Data Tidying) for `pivot_longer()`.
:::

::: {.callout-note collapse="true"}
## Answer

```{r}
#| fig-alt: "Bar chart of mean acceptance threshold c by team (Spain, Greece, No Team), with separate bars for Pro-Spain and Pro-Greece statements and 95% confidence interval error bars. Team members are more accepting of statements that favor their own side."
partisan_long <- partisan_data |>
  pivot_longer(
    cols      = c(SP_c, GR_c),
    names_to  = "statement",
    values_to = "c"
  ) |>
  mutate(statement = recode(statement,
                            SP_c = "Pro-Spain",
                            GR_c = "Pro-Greece"))

ggplot(partisan_long, aes(x = team, y = c, fill = statement)) +
  stat_summary(fun = mean, geom = "bar",
               position = position_dodge(width = 0.9)) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar",
               width = 0.2,
               position = position_dodge(width = 0.9)) +
  labs(
    x    = "Team",
    y    = "Acceptance threshold (c)",
    fill = "Statement type"
  ) +
  theme_minimal()
```

The pivot is the critical step — until `statement` and `c` are their own columns, ggplot has no way to map statement type to `fill`. `position_dodge()` must be set to the *same* width on the bars and the error bars, or the error bars float to the wrong place.
:::

## Grading Rubric

| Task | Points |
|------|:------:|
| 1.1: Summary table (mean, SD, N by species) | 10 |
| 1.2: Bar chart with error bars | 15 |
| 2.1: Histogram with bin width exploration | 8 |
| 2.2: Overlapping density plots by species | 8 |
| 2.3: Boxplot vs. violin plot comparison | 9 |
| 3.1: Intentionally bad visualization | 12 |
| 3.2: Good visualization with design justification | 13 |
| 4.1: Import data and convert team to factor | 5 |
| 4.2: Recreate Figure 3 | 10 |
| QMD file knits without errors | 10 |
| **Total** | **100** |

Each task is graded on a 4-level scale: full credit, full minus one (small slip), half credit (one major error), or none (not attempted or multiple major errors). "QMD file knits" refers to whether the Quarto document renders successfully, not whether individual code produces correct output.

::: {.callout-tip}
## If a chunk won't run
If you can't get a particular chunk to work, set `#| eval: false` at the top of that chunk so the rest of the document still renders. You'll lose points on the task itself, but you'll keep the 10 points for the file knitting cleanly.
:::

## Submission

Submit:

- [ ] Your `.qmd` file with all code and narrative
- [ ] Your rendered HTML file
- [ ] Plots should be embedded in the document (they render automatically from code chunks)

---

::: {.callout-note}
## PSY 510 (Graduate Students)
Students enrolled in PSY 510 must complete the following extension in addition to all tasks above.
:::

## Graduate Extension: Publication-Ready Figure

Creating figures for a course assignment and creating figures for a manuscript are different tasks. Journals specify file format, resolution, and dimensions, and reviewers expect a caption that stands on its own without the surrounding text.

### Task G.1

Choose the figure you're most proud of from Parts 1–3. Look up the author guidelines for one real journal in your area (*Psychological Science*, *JPSP*, *Emotion*, or similar) and save the figure at **300 dpi** with dimensions matching their specifications.

```r
# Example — adjust width/height to your target journal's specs:
ggsave("publication_figure.tiff",
       plot  = my_plot,
       width = 3.5, height = 3,
       units = "in", dpi = 300)
```

### Task G.2

Write a figure caption in APA style — the kind you would include in a manuscript. A good APA caption:

- Starts with a bold figure number and title (e.g., **Figure 1.** *Mean Flipper Length by Species*)
- Describes what the figure shows in one or two sentences
- Defines any abbreviations
- States what the error bars represent (if applicable) and the *N*
- Stands alone — a reader who hasn't read the paper should understand the figure

Include the caption as a comment directly above your `ggsave()` call. (We'll cover APA figure conventions in more depth in Session 16; for now, a thoughtful caption is all that's needed.)

### Task G.3

In 3–5 sentences (as a comment), justify your design choices: Why this geom? Why this color palette? What did you deliberately remove or simplify? Connect your reasoning to the perception and design principles from Session 9.

**Submission:** Add `ggsave()`, the caption, and the justification to your `.qmd` file under a clearly marked `## Graduate Extension` section. Submit the saved figure file alongside your `.qmd` and HTML.
