---
title: "Your First Visualization"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 1, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 2"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
# Load packages for all slides
library(tidyverse)
```

# Why visualize? {background-color="#2c3e50"}

## Anscombe's Quartet

Four datasets with identical summary statistics:

```{r}
#| echo: false
anscombe_long <- anscombe |>
  pivot_longer(everything(),
               names_to = c(".value", "set"),
               names_pattern = "(.)(.)")

anscombe_long |>
  group_by(set) |>
  summarise(
    mean_x = round(mean(x), 2),
    mean_y = round(mean(y), 2),
    sd_x = round(sd(x), 2),
    sd_y = round(sd(y), 2),
    cor = round(cor(x, y), 2)
  ) |>
  knitr::kable()
```

## But look at the plots!

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
ggplot(anscombe_long, aes(x, y)) +
  geom_point(size = 3, color = "#3498db") +
  geom_smooth(method = "lm", se = FALSE, color = "#e74c3c") +
  facet_wrap(~set, labeller = labeller(set = c("1" = "Set 1", "2" = "Set 2",
                                                "3" = "Set 3", "4" = "Set 4"))) +
  theme_minimal(base_size = 16) +
  labs(title = "Same statistics, very different data!")
```

**Always visualize your data before running statistics.**

## The datasaurus dozen

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
# Simulating the concept - in practice you'd use datasauRus package
set.seed(42)
# Just showing the concept with a simple example
tibble(
  x = c(rnorm(50, 50, 10), rnorm(50, 50, 10)),
  y = c(rnorm(50, 50, 10), 50 + 0.5*(1:50 - 25)^2/25 + rnorm(50, 0, 5)),
  dataset = rep(c("Random", "Parabola"), each = 50)
) |>
  ggplot(aes(x, y)) +
  geom_point(color = "#9b59b6", size = 2) +
  facet_wrap(~dataset) +
  theme_minimal(base_size = 16) +
  labs(title = "Different patterns can have similar summary stats")
```

# Introduction to ggplot2 {background-color="#2c3e50"}

## ggplot2 builds plots in layers, like a grammar

- Created by Hadley Wickham (2005)
- Based on the **Grammar of Graphics** by Leland Wilkinson
- Most popular R visualization package
- Part of the tidyverse

. . .

The "gg" stands for "Grammar of Graphics"

## The grammar of graphics

Every ggplot has three essential components:

1. **Data** — what you want to visualize
2. **Aesthetics (aes)** — how variables map to visual properties
3. **Geoms** — what geometric shapes represent the data

. . .

```{r}
#| eval: false
# The basic template
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +
  <GEOM_FUNCTION>()
```

## Our dataset: mpg

```{r}
# Fuel economy data for 234 cars
glimpse(mpg)
```

## Your first plot

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
# Relationship between engine size and highway mpg
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

## Breaking it down

```{r}
#| eval: false
#| code-line-numbers: "1|2|3"
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  # ↑ Data and aesthetic mappings
  geom_point()
  # ↑ Geometric object (points = scatterplot)
```

- `data = mpg` — use the mpg dataset
- `aes(x = displ, y = hwy)` — map displacement to x, highway mpg to y
- `geom_point()` — represent data as points

## A cleaner way to write it

You can drop `data =` and `mapping =`:

```{r}
#| eval: false
# These are equivalent
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point()

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()

ggplot(mpg, aes(displ, hwy)) + 
  geom_point()  # x and y are first args
```

I'll use the middle version — clear but not overly verbose.

# Aesthetic mappings {background-color="#2c3e50"}

## What are aesthetics?

**Aesthetics** are visual properties of geoms:

- `x`, `y` — position
- `color` — outline color
- `fill` — interior color
- `size` — how big
- `shape` — what shape
- `alpha` — transparency

## Mapping color to a variable

What if we want to see which points are which car class?

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

## Mapping size to a variable

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
# Size by number of cylinders
ggplot(mpg, aes(x = displ, y = hwy, size = cyl)) +
  geom_point()
```

## Mapping alpha (transparency)

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy, alpha = cyl)) +
  geom_point()
```

## Mapping shape

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
# Shape by drive type (front, rear, 4wd)
ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) +
  geom_point()
```

## Combining multiple aesthetics

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy, color = class, shape = drv)) +
  geom_point(size = 3)
```

## Setting vs. mapping

**Mapping** — aesthetic varies with data (inside `aes()`)
```{r}
#| eval: false
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

**Setting** — aesthetic is constant (outside `aes()`)
```{r}
#| eval: false
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue", size = 3)
```

## Setting aesthetics manually

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.6)
```

## Common mistake!

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
# What happens if you put a constant inside aes()?
ggplot(mpg, aes(x = displ, y = hwy, color = "blue")) +
  geom_point()
```

ggplot thinks "blue" is a category name!

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

With a partner, create a scatterplot using the `mpg` dataset:

1. Plot `cty` (x-axis) vs `hwy` (y-axis)
2. Color points by **fuel type** (`fl`)
3. Add a smooth trend line
4. Give it a title and axis labels
5. Who can make theirs look the best?

::: {.callout-tip}
You have everything you need from the last few slides. Start with the basic template and build from there.
:::

---

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
ggplot(mpg, aes(x = cty, y = hwy, color = fl)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "City vs Highway MPG by Fuel Type",
    x = "City MPG",
    y = "Highway MPG",
    color = "Fuel type"
  )
```



## Before we move on

📤 **Upload your code to Canvas** for participation credit. Paste what you have into today's in-class submission — it doesn't need to work perfectly.

# Different data types need different geoms {background-color="#2c3e50"}

## Common geoms

| Geom | What it makes |
|------|---------------|
| `geom_point()` | Scatterplot |
| `geom_line()` | Line graph |
| `geom_bar()` | Bar chart |
| `geom_histogram()` | Histogram |
| `geom_boxplot()` | Box plot |
| `geom_smooth()` | Smoothed line |

## geom_smooth()

Add a trend line to your scatterplot:

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()  # Adds a smoothed trend line
```

## Linear trend line

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm")  # lm = linear model
```

## Layering geoms

Each `+` adds a layer:

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)  # se = FALSE removes confidence band
```

## geom_bar() — categorical data

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
# Counts of each car class
ggplot(mpg, aes(x = class)) +
  geom_bar()
```

## geom_histogram() — distributions

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
# Distribution of highway mpg
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white")
```

## geom_boxplot() — comparing groups

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
# Highway mpg by car class
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot()
```

# Facets {background-color="#2c3e50"}

## What are facets?

**Facets** split your plot into small multiples based on a variable.

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~class)
```

## facet_wrap()

Creates a ribbon of panels:

```{r}
#| eval: false
# Syntax: facet_wrap(~variable)
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~class)
```

- Use `ncol` or `nrow` to control layout
- `scales = "free"` allows different axis ranges

## facet_grid()

Creates a grid of panels with two variables:

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 6
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

## When to use facets

Facets are great when:

- You have too many colors/shapes to distinguish
- You want to compare patterns across groups
- You want each group to "stand alone"

# Making it look good {background-color="#2c3e50"}

## Adding labels

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  labs(
    title = "Fuel Efficiency vs Engine Size",
    subtitle = "Larger engines tend to be less efficient",
    x = "Engine displacement (liters)",
    y = "Highway fuel efficiency (mpg)",
    color = "Vehicle class"
  )
```

## Themes

Themes control the overall look:

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  theme_minimal()  # Clean, minimal theme
```

## Built-in themes

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
library(patchwork)

p1 <- ggplot(mpg, aes(displ, hwy)) + geom_point() + theme_gray() + ggtitle("theme_gray() [default]")
p2 <- ggplot(mpg, aes(displ, hwy)) + geom_point() + theme_bw() + ggtitle("theme_bw()")
p3 <- ggplot(mpg, aes(displ, hwy)) + geom_point() + theme_minimal() + ggtitle("theme_minimal()")
p4 <- ggplot(mpg, aes(displ, hwy)) + geom_point() + theme_classic() + ggtitle("theme_classic()")

(p1 + p2) / (p3 + p4)
```

## Saving plots

Use `ggsave()` to save your plot:

```{r}
#| eval: false
# Create the plot
my_plot <- ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  theme_minimal()

# Save it
ggsave("my_plot.png", my_plot, width = 8, height = 6)
```

# Putting it together {background-color="#2c3e50"}

## A complete example

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 6
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class), size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", se = TRUE) +
  labs(
    title = "Fuel Efficiency Decreases with Engine Size",
    subtitle = "Data from 234 vehicles (1999-2008)",
    x = "Engine displacement (liters)",
    y = "Highway MPG",
    color = "Vehicle class",
    caption = "Source: EPA fuel economy data"
  ) +
  theme_minimal(base_size = 14)
```

## Get a head start: Assignment 1

Open a new R script in your project. Try these on your own:

1. Make a scatterplot of `displ` vs `hwy` from `mpg`
2. Color it by `class`
3. Facet by `drv` (drive type)
4. Save it with `ggsave()`

This is the first part of Assignment 1 — finish it before next class.

# Wrapping up {background-color="#2c3e50"}

## The ggplot2 template

```{r}
#| eval: false
ggplot(<DATA>, aes(<MAPPINGS>)) +
  <GEOM_FUNCTION>() +
  <FACET_FUNCTION>() +
  labs(<LABELS>) +
  theme_<THEME>()
```

You can build almost any visualization with this template!

## Before next class

📖 **Read:**

- [R4DS Ch 3: Data transformation](https://r4ds.hadley.nz/data-transform) (sections 3.1–3.4)

✅ **Practice:**

- Create 3 different plots with `mpg` or `diamonds`
- Try different geoms, aesthetics, and facets
- Save your favorite plot with `ggsave()`

## Key takeaways

1. **Always visualize your data** — summary stats can hide patterns
2. **ggplot2 uses layers** — data + aesthetics + geoms
3. **Aesthetics can be mapped or set** — `aes()` vs direct values
4. **Facets are powerful** — split plots by categories
5. **Labels and themes matter** — make your plots readable

## The one thing to remember

Never trust a summary statistic you haven't plotted.

Next time: Data Transformation with dplyr
