---
title: "Data Transformation I"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 6, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 3"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---


```{r}
#| include: false
library(tidyverse)
library(nycflights13)  # For the flights dataset
```

# Data transformation {background-color="#2c3e50"}

## A question you can't answer yet

You collected survey data from 500 participants. You only want women over 25 who passed the attention check.

. . .

How do you get to *just those rows?*

. . .

That's what `filter()` does. And it's just the beginning — today we learn four verbs that turn raw data into exactly what you need.

## The dplyr verbs

| Verb | What it does |
|------|--------------|
| `filter()` | Pick **rows** by their values |
| `arrange()` | Reorder **rows** |
| `select()` | Pick **columns** by name |
| `mutate()` | Create new **columns** |
| `summarize()` | Collapse to a **summary** |

Today: `filter()`, `arrange()`, `select()`, `mutate()`

## Our dataset: flights

```{r}
# All 336,776 flights departing NYC in 2013
glimpse(flights)
```

## Understanding the data

- `year`, `month`, `day` — departure date
- `dep_time`, `arr_time` — actual times (HHMM format)
- `dep_delay`, `arr_delay` — delays in minutes (negative = early)
- `carrier` — airline code
- `origin`, `dest` — airport codes
- `air_time`, `distance` — in minutes and miles

# filter() {background-color="#2c3e50"}

## filter() picks rows

```{r}
#| output-location: fragment
# All flights on January 1st
filter(flights, month == 1, day == 1)
```

## Comparison operators

| Operator | Meaning |
|----------|---------|
| `==` | equal to |
| `!=` | not equal to |
| `<`, `>` | less than, greater than |
| `<=`, `>=` | less/greater than or equal |

::: {.callout-warning}
Use `==` for comparison, not `=`!
:::

## Filter examples

```{r}
#| eval: false
# Flights to Los Angeles
filter(flights, dest == "LAX")

# Flights with arrival delay over 2 hours
filter(flights, arr_delay > 120)

# Flights by United Airlines
filter(flights, carrier == "UA")
```

## Multiple conditions

Conditions separated by `,` are combined with AND:

```{r}
#| eval: false
# January 1st flights (both must be true)
filter(flights, month == 1, day == 1)
```

Equivalent to:

```{r}
#| eval: false
filter(flights, month == 1 & day == 1)
```

## Logical operators

| Operator | Meaning |
|----------|---------|
| `&` | AND (both true) |
| `|` | OR (either true) |
| `!` | NOT (negation) |

## Using OR

```{r}
# Flights in November OR December
filter(flights, month == 11 | month == 12)
```

## A useful shortcut: %in%

```{r}
# Same as above, but cleaner
filter(flights, month %in% c(11, 12))
```

`%in%` checks if the value is in the vector.

## filter() with psychology data

Imagine a survey dataset:

```{r}
#| eval: false
# Exclude participants who failed attention check
filter(survey, attention_check == "correct")

# Keep only complete responses
filter(survey, !is.na(total_score))

# Adults only
filter(survey, age >= 18)

# Specific conditions
filter(survey, condition %in% c("treatment", "control"))
```

## Missing values: NA

`NA` means "not available" — a missing value.

```{r}
x <- c(1, 2, NA, 4)
x > 2
```

Any operation with `NA` returns `NA` (it's unknown!).

## Checking for NA

Use `is.na()` to check for missing values:

```{r}
x <- c(1, 2, NA, 4)
is.na(x)
```

```{r}
#| eval: false
# Keep rows where dep_delay is NOT missing
filter(flights, !is.na(dep_delay))
```

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

With a partner, using the `flights` dataset:

1. Find all **United Airlines** (`"UA"`) flights
2. that were **more than 2 hours late** arriving
3. and were flying **to Los Angeles** (`"LAX"`)

How many flights match? Which origin airport had the most?

::: {.callout-tip}
You'll need `filter()` with multiple conditions. Think about which operators you need.
:::

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
flights |>
  filter(carrier == "UA", arr_delay > 120, dest == "LAX")

# Bonus: count by origin
flights |>
  filter(carrier == "UA", arr_delay > 120, dest == "LAX") |>
  count(origin)
```

---

## Before we move on

📤 **Upload your code to Canvas** for participation credit. Paste what you have into today's in-class submission — it doesn't need to work perfectly.

# arrange() {background-color="#2c3e50"}

## arrange() reorders rows

```{r}
# Sort by departure delay (smallest first)
arrange(flights, dep_delay)
```

## Descending order

```{r}
# Most delayed flights first
arrange(flights, desc(dep_delay))
```

## Multiple sort columns

```{r}
# Sort by month, then by day, then by departure time
arrange(flights, month, day, dep_time)
```

## arrange() use cases

```{r}
#| eval: false
# Psychology examples:
# Find participants with highest scores
arrange(data, desc(total_score))

# Sort by condition, then by participant ID
arrange(data, condition, participant_id)

# Find earliest responses
arrange(data, response_time)
```

# select() {background-color="#2c3e50"}

## select() picks columns

```{r}
# Just these three columns
select(flights, year, month, day)
```

## Select a range

```{r}
# All columns from year to day
select(flights, year:day)
```

## Select helpers

| Helper | What it does |
|--------|--------------|
| `starts_with("x")` | Columns starting with "x" |
| `ends_with("x")` | Columns ending with "x" |
| `contains("x")` | Columns containing "x" |
| `everything()` | All remaining columns |

## Using select helpers

```{r}
# All delay-related columns
select(flights, contains("delay"))
```

## Using select helpers

```{r}
# All time columns
select(flights, ends_with("time"))
```

## Reorder columns

```{r}
# Move air_time and distance to the front
select(flights, air_time, distance, everything())
```

## Exclude columns

Use `-` to remove columns:

```{r}
# Remove year column
select(flights, -year)
```

## Exclude columns

```{r}
#| eval: false
# Remove multiple columns
select(flights, -year, -month, -day)
select(flights, -(year:day))  # same thing
```

## select() in psychology

```{r}
#| eval: false
# Select demographic columns
select(survey, starts_with("demo_"))

# Select all BDI items
select(survey, contains("bdi"))

# Remove identifying information
select(survey, -name, -email, -ip_address)

# Reorder for analysis
select(survey, participant_id, condition, starts_with("outcome"))
```

# mutate() {background-color="#2c3e50"}

## mutate() creates new columns

```{r}
# Calculate total delay
mutate(flights, total_delay = dep_delay + arr_delay)
```

## Multiple new columns

```{r}
mutate(flights,
       # Create new variables
       total_delay = dep_delay + arr_delay,
       speed = distance / air_time * 60,  # mph
       # Can reference columns you just created!
       delay_per_mile = total_delay / distance
)
```

## mutate() keeps all columns

`mutate()` adds new columns while keeping existing ones.

Use `transmute()` to *only* keep new columns:

```{r}
transmute(flights,
          total_delay = dep_delay + arr_delay,
          speed = distance / air_time * 60
)
```

## Useful functions in mutate()

**Arithmetic:** `+`, `-`, `*`, `/`, `^`, `%%` (modulo), `%/%` (integer division)

**Logs:** `log()`, `log2()`, `log10()`

**Offsets:** `lead()`, `lag()` (for time series)

**Cumulative:** `cumsum()`, `cummean()`, `cummax()`

**Ranking:** `min_rank()`, `dense_rank()`, `row_number()`

## Common transformations

```{r}
#| eval: false
# Z-score (standardize)
mutate(data, score_z = (score - mean(score)) / sd(score))

# Log transform (for skewed data)
mutate(data, rt_log = log(reaction_time))

# Create categories from continuous
mutate(data, age_group = case_when(
  age < 30 ~ "young",
  age < 60 ~ "middle",
  TRUE ~ "older"
))
```

## mutate() for psychology

```{r}
#| eval: false
# Calculate scale scores (mean of items)
mutate(survey,
       bdi_total = (bdi_1 + bdi_2 + bdi_3 + bdi_4) / 4,
       # Or rowwise if you have many items:
       anxiety = rowMeans(select(., anx_1:anx_20), na.rm = TRUE)
)

# Create dummy codes
mutate(survey,
       female = if_else(gender == "female", 1, 0),
       treatment = if_else(condition == "treatment", 1, 0)
)

# Reverse code items
mutate(survey,
       item_5r = 8 - item_5  # For 1-7 scale
)
```

# The pipe makes your code read like a sentence {background-color="#2c3e50"}

## The problem with nested functions

:::: {.columns}
::: {.column width="50%"}
What if we want to:

1. Filter to January flights
2. Select departure time and delay
3. Arrange by delay
:::

::: {.column width="50%"}
Nested approach:

```{r}
#| eval: false
arrange(
  select(
    filter(flights, month == 1),
    dep_time, dep_delay
  ),
  dep_delay
)
```
:::
::::

This is hard to read!

## Intermediate objects approach

```{r}
#| eval: false
# Save each step
jan_flights <- filter(flights, month == 1)
jan_selected <- select(jan_flights, dep_time, dep_delay)
jan_arranged <- arrange(jan_selected, dep_delay)
```

Works, but clutters your environment with objects.

## The pipe: |>

The pipe takes the result of one function and passes it as the first argument to the next:

:::: {.columns}
::: {.column width="50%"}
```{r}
#| label: pipe-example
#| eval: false
# Same result, much cleaner!
flights |>
  filter(month == 1) |>
  select(dep_time, dep_delay) |>
  arrange(dep_delay)
```

:::

::: {.column width="50%"}
```{r}
#| echo: false
# Same result, much cleaner!
flights |>
  filter(month == 1) |>
  select(dep_time, dep_delay) |>
  arrange(dep_delay)
```

:::
::::

## Reading piped code

Read `|>` as **"then"**:

```{r}
#| eval: false
flights |>                     # Start with flights, THEN
  filter(month == 1) |>        # filter to January, THEN
  select(dep_time, dep_delay) |>  # select these columns, THEN
  arrange(dep_delay)           # arrange by delay
```

## Keyboard shortcut

The pipe is so common, there's a shortcut:

- **Windows/Linux:** Ctrl + Shift + M
- **Mac:** Cmd + Shift + M

. . .

::: {.callout-tip}
In RStudio settings, make sure "Use native pipe operator" is enabled (Tools → Global Options → Code)
:::

## Note: |> vs %>%

You'll see both:

- `|>` — the **native** pipe (built into R 4.1+)
- `%>%` — the **magrittr** pipe (older, from tidyverse)

They work almost identically. We'll use `|>` since it's now standard.

## Piping into ggplot

The pipe works beautifully with ggplot:

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
flights |>
  filter(month == 1, carrier == "UA") |>
  ggplot(aes(x = dep_delay, y = arr_delay)) +
  geom_point(alpha = 0.3) +
  labs(title = "United Airlines Delays (January)")
```

# Putting it together {background-color="#2c3e50"}

## A complete workflow

```{r}
#| output-location: slide
# Which airlines have the longest delays in summer?
flights |>
  filter(month %in% c(6, 7, 8)) |>           # Summer months
  filter(!is.na(arr_delay)) |>               # Remove NAs
  mutate(delay_hours = arr_delay / 60) |>    # Convert to hours
  select(carrier, delay_hours) |>            # Keep relevant columns
  arrange(desc(delay_hours))                 # Longest delays first
```

## Your turn!

Using the `flights` dataset:

1. Filter to American Airlines ("AA") flights to Los Angeles ("LAX")
2. Create a new variable `speed` (distance / air_time * 60)
3. Select carrier, origin, dest, and speed
4. Arrange by speed (highest first)
5. What's the fastest AA flight to LAX?

```{r}
#| echo: false
#| eval: false
# Solution
flights |>
  filter(carrier == "AA", dest == "LAX") |>
  mutate(speed = distance / air_time * 60) |>
  select(carrier, origin, dest, speed) |>
  arrange(desc(speed))
```

# Wrapping up {background-color="#2c3e50"}

## The dplyr workflow

```{r}
#| eval: false
data |>
  filter(<conditions>) |>    # Pick rows
  select(<columns>) |>       # Pick columns
  mutate(<new vars>) |>      # Create columns
  arrange(<order>)           # Sort rows
```

The pipe (`|>`) connects these verbs into a readable workflow.

## Before next class

📖 **Read:**

- [R4DS Ch 3: Data transformation](https://r4ds.hadley.nz/data-transform) (section 3.5)
- [R4DS Ch 4: Workflow: code style](https://r4ds.hadley.nz/workflow-style)

✅ **Practice:**

- Transform `flights` in different ways
- Create new variables with `mutate()`
- Build multi-step pipelines

## Key takeaways

1. **filter()** picks rows by condition
2. **arrange()** reorders rows (use `desc()` for descending)
3. **select()** picks columns (helpers like `contains()` are useful)
4. **mutate()** creates new columns
5. **The pipe |>** makes code readable and elegant

## The one thing to remember

The pipe turns a wall of nested code into a sentence you can read aloud.

Next time: group_by() and summarize()
