---
title: "Data Transformation II"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 8, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 4"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(nycflights13)
```

# Review {background-color="#2c3e50"}

## The dplyr verbs so far

| Verb | What it does |
|------|--------------|
| `filter()` | Pick rows by condition |
| `arrange()` | Sort rows |
| `select()` | Pick columns |
| `mutate()` | Create new columns |

Today we add the most powerful duo: **group_by()** + **summarize()**

# summarize() {background-color="#2c3e50"}

## summarize() collapses data

Reduces a data frame to a single row of summary statistics:

```{r}
flights |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    max_delay = max(dep_delay, na.rm = TRUE),
    n_flights = n()
  )
```

## Key summary functions

| Function | What it computes |
|----------|-----------------|
| `mean()`, `median()` | Central tendency |
| `sd()`, `var()` | Spread |
| `min()`, `max()` | Extremes |
| `sum()` | Total |
| `n()` | Count of rows |
| `n_distinct()` | Count of unique values |

## The na.rm argument

Most summary functions need `na.rm = TRUE` to handle missing values:

```{r}
x <- c(1, 2, NA, 4, 5)

mean(x)                  # Returns NA
mean(x, na.rm = TRUE)    # Returns 3
```

::: {.callout-tip}
Always use `na.rm = TRUE` unless you specifically want NA propagation.
:::

## summarize() alone isn't very useful

One row of summary stats? That's what `mean()` already does.

```{r}
mean(flights$dep_delay, na.rm = TRUE)
```

The power comes when we combine it with **group_by()**.

# group_by() {background-color="#2c3e50"}

## group_by() + summarize() = magic

```{r}
flights |>
  group_by(month) |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n_flights = n()
  )
```

## What happened?

1. `group_by(month)` split the data into 12 groups (one per month)
2. `summarize()` calculated statistics **within each group**
3. Result: one row per group

## Visualizing group_by()

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4
tibble(
  group = rep(c("A", "B", "C"), each = 4),
  value = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12)
) |>
  ggplot(aes(x = group, y = value, fill = group)) +
  geom_col(position = "dodge") +
  theme_minimal(base_size = 18) +
  labs(title = "Data split by group", subtitle = "Each group gets summarized separately") +
  theme(legend.position = "none")
```

## Group by multiple variables

```{r}
# Average delay by month AND carrier
flights |>
  group_by(month, carrier) |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n_flights = n()
  )
```

## The .groups argument

You may see this message:

```
`summarise()` has grouped output by 'month'. You can override using the `.groups` argument.
```

Control this with:

```{r}
#| eval: false
# Keep grouping (default with multiple groups)
summarize(..., .groups = "keep")

# Drop all grouping
summarize(..., .groups = "drop")

# Drop last group level
summarize(..., .groups = "drop_last")
```

## Group by for psychology

```{r}
#| eval: false
# Means by condition
data |>
  group_by(condition) |>
  summarize(
    mean_score = mean(score, na.rm = TRUE),
    sd_score = sd(score, na.rm = TRUE),
    n = n()
  )

# Descriptives by condition AND gender
data |>
  group_by(condition, gender) |>
  summarize(
    m = mean(outcome),
    se = sd(outcome) / sqrt(n())
  )
```

## Multiple summaries at once

```{r}
flights |>
  group_by(carrier) |>
  summarize(
    # Central tendency
    mean_delay = mean(arr_delay, na.rm = TRUE),
    median_delay = median(arr_delay, na.rm = TRUE),
    # Spread
    sd_delay = sd(arr_delay, na.rm = TRUE),
    # Counts
    n_flights = n(),
    n_destinations = n_distinct(dest)
  ) |>
  arrange(mean_delay)  # Best to worst carriers
```

# count() — a shortcut {background-color="#2c3e50"}

## Counting is common

```{r}
# How many flights per carrier?
flights |>
  group_by(carrier) |>
  summarize(n = n())
```

## count() does this automatically

```{r}
flights |>
  count(carrier)
```

Same result, less typing!

## count() with multiple variables

```{r}
flights |>
  count(carrier, origin)
```

## Adding weights

```{r}
#| eval: false
# Total distance flown by each carrier
flights |>
  count(carrier, wt = distance, name = "total_miles")
```

# Combining operations {background-color="#2c3e50"}

## The full dplyr toolkit

Now you can combine everything:

```{r}
#| eval: false
data |>
  filter()    |>   # Subset rows
  select()    |>   # Pick columns
  mutate()    |>   # Create columns
  group_by()  |>   # Define groups
  summarize() |>   # Calculate summaries
  arrange()        # Sort results
```

## A complete analysis pipeline

```{r}
# Which airlines were most delayed in summer?
flights |>
  filter(month %in% c(6, 7, 8)) |>    # Summer months only
  filter(!is.na(arr_delay)) |>         # Remove missing
  group_by(carrier) |>                 # Group by airline
  summarize(
    avg_delay = mean(arr_delay),
    pct_delayed = mean(arr_delay > 0) * 100,  # % of flights delayed
    n_flights = n()
  ) |>
  filter(n_flights > 1000) |>          # Only airlines with many flights
  arrange(desc(avg_delay))             # Worst first
```

## Piping into ggplot

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
flights |>
  filter(!is.na(arr_delay)) |>
  group_by(month) |>
  summarize(avg_delay = mean(arr_delay)) |>
  ggplot(aes(x = factor(month), y = avg_delay)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Arrival Delay by Month",
    x = "Month",
    y = "Average delay (minutes)"
  ) +
  theme_minimal(base_size = 14)
```

## group_by() with mutate()

`group_by()` also affects `mutate()` and `filter()`:

```{r}
# Add a column for mean delay PER CARRIER
flights |>
  group_by(carrier) |>
  mutate(
    carrier_mean_delay = mean(arr_delay, na.rm = TRUE),
    delay_vs_carrier = arr_delay - carrier_mean_delay
  ) |>
  select(carrier, arr_delay, carrier_mean_delay, delay_vs_carrier)
```

## Filtering within groups

```{r}
# Find the 2 most delayed flights per month
flights |>
  group_by(month) |>
  filter(!is.na(arr_delay)) |>
  slice_max(arr_delay, n = 2) |>
  select(month, carrier, arr_delay)
```

## slice() variants

| Function | What it does |
|----------|--------------|
| `slice_head(n)` | First n rows |
| `slice_tail(n)` | Last n rows |
| `slice_max(var, n)` | n rows with highest var |
| `slice_min(var, n)` | n rows with lowest var |
| `slice_sample(n)` | Random n rows |

All work within groups when preceded by `group_by()`.

## ungroup()

Sometimes you need to remove grouping:

```{r}
flights |>
  group_by(carrier) |>
  mutate(carrier_mean = mean(arr_delay, na.rm = TRUE)) |>
  ungroup() |>                              # Remove grouping
  mutate(grand_mean = mean(arr_delay, na.rm = TRUE)) |>  # Now uses ALL flights
  select(carrier, arr_delay, carrier_mean, grand_mean)
```

# Code style {background-color="#2c3e50"}

## Good style matters

Code is read more often than it's written. Make it readable!

```{r}
#| eval: false
# Bad: crammed together
flights|>filter(month==1)|>group_by(carrier)|>summarize(n=n())

# Good: spaced and readable
flights |>
  filter(month == 1) |>
  group_by(carrier) |>
  summarize(n = n())
```

## Spacing rules

- Spaces around operators: `x == 1`, not `x==1`
- Space after commas: `c(1, 2, 3)`, not `c(1,2,3)`
- No space before parentheses: `mean(x)`, not `mean (x)`
- Spaces around `|>` with line breaks

## Line breaks

- One verb per line for readability
- Keep lines under ~80 characters
- Indent continuation lines consistently

```{r}
#| eval: false
# Long summarize calls
data |>
  group_by(condition) |>
  summarize(
    mean_score = mean(score, na.rm = TRUE),
    sd_score = sd(score, na.rm = TRUE),
    n = n()
  )
```

## Naming conventions

- Use `snake_case` for object names: `my_data`, `mean_score`
- Use meaningful names: `delay_by_carrier`, not `x`
- Be consistent!

```{r}
#| eval: false
# Good
participant_data <- read_csv("participants.csv")
mean_reaction_time <- mean(participant_data$rt)

# Bad
pD <- read_csv("participants.csv")
m <- mean(pD$rt)
```

## Comments

Use comments to explain **why**, not **what**:

```{r}
#| eval: false
# Bad: restates the obvious
# Filter to January
flights |> filter(month == 1)

# Good: explains the reasoning
# January only because weather patterns differ by season
flights |> filter(month == 1)
```

# Practice: Psychology example {background-color="#2c3e50"}

## Simulated experiment data

```{r}
set.seed(42)

# Create a simulated experiment dataset
experiment <- tibble(
  participant_id = rep(1:60, each = 20),
  condition = rep(rep(c("control", "treatment"), each = 10), 60),
  trial = rep(1:20, 60),
  reaction_time = rnorm(1200, mean = 500, sd = 100) +
                  ifelse(rep(rep(c(0, 1), each = 10), 60) == 1, -30, 0),
  accuracy = rbinom(1200, 1, prob = 0.85 +
                    ifelse(rep(rep(c(0, 1), each = 10), 60) == 1, 0.05, 0))
)
```

## Examine the data

```{r}
glimpse(experiment)
```

## Participant-level summaries

```{r}
# First, summarize BY PARTICIPANT
participant_summaries <- experiment |>
  group_by(participant_id, condition) |>
  summarize(
    mean_rt = mean(reaction_time),
    accuracy = mean(accuracy) * 100,
    n_trials = n(),
    .groups = "drop"
  )

participant_summaries
```

## Condition-level summaries

```{r}
# Now summarize ACROSS PARTICIPANTS by condition
condition_summaries <- participant_summaries |>
  group_by(condition) |>
  summarize(
    M_rt = mean(mean_rt),
    SD_rt = sd(mean_rt),
    M_acc = mean(accuracy),
    SD_acc = sd(accuracy),
    n = n()
  )

condition_summaries
```

## Visualize the results

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
participant_summaries |>
  ggplot(aes(x = condition, y = mean_rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(
    title = "Reaction Time by Condition",
    x = "Condition",
    y = "Mean RT (ms)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## Your turn!

Using the `experiment` data:

1. Filter to only accurate trials (`accuracy == 1`)
2. Calculate mean RT per participant per condition
3. Find the 5 participants with the fastest mean RT in each condition
4. Create a bar plot showing mean RT by condition with error bars

# Wrapping up {background-color="#2c3e50"}

## Today's additions

| Verb | What it does |
|------|--------------|
| `group_by()` | Define groups for operations |
| `summarize()` | Calculate summary statistics |
| `count()` | Quick frequency counts |
| `ungroup()` | Remove grouping |
| `slice_*()` | Select rows by position/rank |

## The analysis workflow

```{r}
#| eval: false
raw_data |>
  filter(valid_data) |>           # Clean
  select(relevant_vars) |>        # Focus
  mutate(new_vars) |>             # Transform
  group_by(conditions) |>         # Split
  summarize(statistics) |>        # Aggregate
  arrange(order) |>               # Sort
  ggplot() + ...                  # Visualize
```

## Before next class

📖 **Read:**

- [R4DS Ch 5: Data tidying](https://r4ds.hadley.nz/data-tidy)

✅ **Practice:**

- Calculate descriptive statistics by group
- Build complete analysis pipelines
- Focus on writing clean, readable code

## Key takeaways

1. **group_by() + summarize()** is incredibly powerful
2. **count()** is a quick shortcut for frequencies
3. **Grouped operations** work with mutate() and filter() too
4. **ungroup()** when you need to work with all data again
5. **Code style matters** — write for your future self

## Questions?

Next time: **Data Tidying with tidyr**

We'll learn to reshape data between wide and long formats!
