---
title: "Data Transformation II"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 8, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 4"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(nycflights13)
```

# From rows to statistics {background-color="#2c3e50"}

## Your advisor's first question

Last time, you learned to pick rows, sort them, and create new columns.

. . .

But your advisor doesn't want rows. They want a number:

*"What's the average depression score by condition?"*

. . .

Today we learn to answer that — with **group_by()** + **summarize()**.

::: {.notes}
Start by connecting to Session 3. Ask students what verbs they learned last time (filter, arrange, mutate, select). Then pivot: "Those are great for exploring, but your advisor doesn't want 1,000 rows — they want a single number." The depression score example resonates because it sounds like something they'd actually need to do. Spend ~2 minutes here.
:::

# summarize() {background-color="#2c3e50"}

## summarize() collapses data

Reduces a data frame to a single row of summary statistics:

```{r}
flights |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    max_delay = max(dep_delay, na.rm = TRUE),
    n_flights = n()
  )
```

## Key summary functions

| Function | What it computes |
|----------|-----------------|
| `mean()`, `median()` | Central tendency |
| `sd()`, `var()` | Spread |
| `min()`, `max()` | Extremes |
| `sum()` | Total |
| `n()` | Count of rows |
| `n_distinct()` | Count of unique values |

::: {.notes}
Don't dwell on this table — it's a reference slide. Point out that `n()` and `n_distinct()` are new (no base R equivalent students would know), while the rest are familiar from stats classes. Mention that `n()` takes no arguments, which trips students up later.
:::

## The na.rm argument

Most summary functions need `na.rm = TRUE` to handle missing values:

```{r}
x <- c(1, 2, NA, 4, 5)

mean(x)                  # Returns NA
mean(x, na.rm = TRUE)    # Returns 3
```

::: {.callout-tip}
Always use `na.rm = TRUE` unless you specifically want NA propagation.
:::

::: {.notes}
This is one of the most common early frustrations. Run the code live and let them see the NA result before showing the fix. Ask: "Why does R do this?" The answer is that R is being cautious — if data is missing, the true mean is unknown. Emphasize that forgetting `na.rm = TRUE` is the #1 reason students get unexpected NAs in their output. They will hit this on Assignment 2.
:::

## summarize() alone isn't very useful

One row of summary stats? That's what `mean()` already does.

```{r}
mean(flights$dep_delay, na.rm = TRUE)
```

The power comes when we combine it with **group_by()**.

::: {.notes}
This slide sets up the "aha" moment. Pause here and say something like: "Okay, so summarize alone is kind of boring. But watch what happens when we add one line." Then move to the next section.
:::

# group_by() {background-color="#2c3e50"}

## group_by() + summarize() = magic

```{r}
flights |>
  group_by(month) |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n_flights = n()
  )
```

::: {.notes}
This is the key moment of the deck. Run the code and let the output land. Point to the 12-row result and ask: "How many rows? Why 12?" Make sure students see the connection between `group_by(month)` and getting one row per month. Spend ~3 minutes here. If you can, compare it to the single-row result from the previous slide.
:::

## What happened?

1. `group_by(month)` split the data into 12 groups (one per month)
2. `summarize()` calculated statistics **within each group**
3. Result: one row per group

## Visualizing group_by()

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| fig-alt: "Bar chart of average departure delay by month, showing the result of the group_by(month) + summarize() pipeline from the previous slide."
flights |>
  group_by(month) |>
  summarize(avg_delay = mean(dep_delay, na.rm = TRUE)) |>
  ggplot(aes(x = factor(month), y = avg_delay)) +
  geom_col(fill = "steelblue") +
  theme_minimal(base_size = 18) +
  labs(
    title = "Average departure delay by month",
    x = "Month",
    y = "Average delay (minutes)"
  )
```

## Group by multiple variables

```{r}
# Average delay by month AND carrier
flights |>
  group_by(month, carrier) |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n_flights = n()
  )
```

::: {.notes}
Point out that the output now has `Groups: month [12]` printed at the top — this is R telling you grouping is still partially active. Ask: "How many rows would you expect?" (12 months x 16 carriers = up to 192). This naturally leads to the `.groups` message on the next slide.
:::

## The .groups argument

You may see this message:

```
`summarise()` has grouped output by 'month'. You can override using the `.groups` argument.
```

Control this with:

```{r}
#| eval: false
# Keep grouping (default with multiple groups)
summarize(..., .groups = "keep")

# Drop all grouping
summarize(..., .groups = "drop")

# Drop last group level
summarize(..., .groups = "drop_last")
```

::: {.notes}
Students will see this message and panic — reassure them it's informational, not an error. For this course, recommend `.groups = "drop"` as the safest default. The message appears because after summarizing two grouping variables, R keeps the first one grouped. You don't need to belabor `"keep"` vs `"drop_last"` — just tell them `"drop"` is the safe bet and move on. Spend ~2 minutes max.
:::

## Multiple summaries at once

```{r}
#| label: test
#| eval: false
flights |>
  group_by(carrier) |>
  summarize(
    # Central tendency
    mean_delay = mean(arr_delay, na.rm = TRUE),
    median_delay = median(arr_delay, na.rm = TRUE),
    # Spread
    sd_delay = sd(arr_delay, na.rm = TRUE),
    # Counts
    n_flights = n(),
    n_dest = n_distinct(dest)
  ) |>
  arrange(mean_delay)  # Best to worst carriers
```

## Multiple summaries at once

```{r}
#| ref.label: test
#| echo: false
```

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

With a partner, using the `flights` dataset:

1. Calculate the **average departure delay** for each carrier
2. Which airline has the **worst** average delay?
3. **Bonus:** Also calculate the number of flights per carrier. Does the worst airline just have fewer flights?

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
flights |>
  group_by(carrier) |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n_flights = n()
  ) |>
  arrange(desc(avg_delay))
```

::: {.notes}
Set a timer for 10 minutes. Circulate and watch for: (1) forgetting `na.rm = TRUE` and getting NA results, (2) using `mean(dep_delay)` without `summarize()`, (3) not knowing how to sort descending (`desc()`). The bonus question is intentionally a "think about your data" prompt — Frontier (F9) has the worst average delay but relatively few flights. This is a good moment to talk about sample size and whether an average is trustworthy. Reveal the solution after time is up.
:::

---

## Before we move on

📤 **[Submit your code on Canvas](https://canvas.uoregon.edu/courses/287793/pages/pair-coding-check-ins)** for participation credit. Paste what you have — it doesn't need to work perfectly.

## What you can already do

With `summarize()` + `group_by()`, you can answer questions like:

- *"What's the average depression score by condition?"*
- *"Which group had the most variability?"*
- *"How many participants completed each phase?"*

. . .

These are the building blocks of a results section. Next: some shortcuts.

::: {.notes}
This is a pivot slide. Use it to let the energy from the pair coding settle and reset for part 2. Emphasize that they've already learned the single most important dplyr pattern. Everything from here is refinements and shortcuts. ~1 minute.
:::

# count() — a shortcut {background-color="#2c3e50"}

## Counting is common

```{r}
# How many flights per carrier?
flights |>
  group_by(carrier) |>
  summarize(n = n())
```

## count() does this automatically

```{r}
flights |>
  count(carrier)
```

Same result, less typing!

::: {.notes}
Show both versions side by side (scroll back if needed). Emphasize that `count()` is just syntactic sugar for `group_by() |> summarize(n = n())`. Students often ask: "When do I use `count()` vs `group_by() + summarize()`?" Answer: use `count()` when you only need frequencies; use the full pattern when you need means, SDs, or other statistics alongside the count.
:::

## count() with multiple variables

```{r}
flights |>
  count(carrier, origin)
```

## Adding weights

```{r}
# Total distance flown by each carrier
flights |>
  count(carrier, wt = distance, name = "total_miles")
```

# Combining operations {background-color="#2c3e50"}

## The full dplyr toolkit

Now you can combine everything:

```{r}
#| eval: false
data |>
  filter()    |>   # Subset rows
  select()    |>   # Pick columns
  mutate()    |>   # Create columns
  group_by()  |>   # Define groups
  summarize() |>   # Calculate summaries
  arrange()        # Sort results
```

::: {.notes}
This is a "big picture" slide. Emphasize that the order listed here is a common workflow, not a rigid rule. Students often ask whether the order matters — it does for some combinations (e.g., `filter()` before `summarize()` reduces what gets summarized), but `select()` can go anywhere. The key insight: each verb does one thing, and the pipe chains them together.
:::

## A complete analysis pipeline

```{r}
#| label: complete
#| eval: false
# Which airlines were most delayed in summer?
flights |>
  filter(month %in% c(6, 7, 8)) |>    # Summer months only
  filter(!is.na(arr_delay)) |>         # Remove missing
  group_by(carrier) |>                 # Group by airline
  summarize(
    avg_delay = mean(arr_delay),
    pct_delayed = mean(arr_delay > 0) * 100,  # % of flights delayed
    n_flights = n()
  ) |>
  filter(n_flights > 1000) |>          # Only airlines with many flights
  arrange(desc(avg_delay))             # Worst first
```

## A complete analysis pipeline

```{r}
#| ref.label: complete
#| echo: false
```

::: {.notes}
Walk through the pipeline step by step, narrating what each line does. Point out the `mean(arr_delay > 0)` trick — since TRUE = 1 and FALSE = 0, taking the mean of a logical gives you the proportion. This is a useful pattern students haven't seen before. Also highlight the post-summarize `filter(n_flights > 1000)` — filtering after summarizing is a common and powerful pattern for removing small-sample groups.
:::

## Piping into ggplot

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Bar chart of average arrival delay by month showing delays peak in summer (June and July) and are lowest in fall, demonstrating how a dplyr pipeline can feed directly into ggplot."
flights |>
  filter(!is.na(arr_delay)) |>
  group_by(month) |>
  summarize(avg_delay = mean(arr_delay)) |>
  ggplot(aes(x = factor(month), y = avg_delay)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Arrival Delay by Month",
    x = "Month",
    y = "Average delay (minutes)"
  ) +
  theme_minimal(base_size = 14)
```

::: {.notes}
This slide previews a pattern they'll use heavily later: pipe straight from data wrangling into ggplot. Point out the switch from `|>` to `+` once you enter ggplot — students frequently mix these up. You don't need to explain the ggplot code in detail here; just show that the summary data flows directly into the visualization. Sessions 8 and 9 will cover ggplot in depth.
:::

## group_by() with mutate()

`group_by()` also affects `mutate()` and `filter()`:

```{r}
#| label: mix-group-mutate
#| eval: false
# Add a column for mean delay PER CARRIER
flights |>
  group_by(carrier) |>
  mutate(
    carrier_mean_delay = mean(arr_delay, na.rm = TRUE),
    delay_vs_carrier = arr_delay - carrier_mean_delay
  ) |>
  select(carrier, arr_delay, carrier_mean_delay, delay_vs_carrier)
```

## group_by() with mutate()

`group_by()` also affects `mutate()` and `filter()`:

```{r}
#| ref.label: mix-group-mutate
#| echo: false

```

::: {.notes}
This is a more advanced use of `group_by()` that students find surprising. Emphasize the difference: with `summarize()`, you collapse to one row per group; with `mutate()`, you keep all rows but the calculation is done within each group. The psychology use case: centering scores within each participant (person-mean centering) for repeated-measures designs. Don't dwell too long — this is a "you'll need this later" preview. ~2 minutes.
:::

## Filtering within groups

```{r}
# Find the 2 most delayed flights per month
flights |>
  group_by(month) |>
  filter(!is.na(arr_delay)) |>
  slice_max(arr_delay, n = 2) |>
  select(month, carrier, arr_delay)
```

## slice() variants

| Function | What it does |
|----------|--------------|
| `slice_head(n)` | First n rows |
| `slice_tail(n)` | Last n rows |
| `slice_max(var, n)` | n rows with highest var |
| `slice_min(var, n)` | n rows with lowest var |
| `slice_sample(n)` | Random n rows |

All work within groups when preceded by `group_by()`.

::: {.notes}
Present `slice_*()` as a reference — students don't need to memorize all five variants today. The most useful ones for this course are `slice_max()` and `slice_min()`, which appear in the end-of-deck exercise. Mention that `slice_sample()` is handy for quickly peeking at random rows when exploring data.
:::

## ungroup()

Sometimes you need to remove grouping:

```{r}
flights |>
  group_by(carrier) |>
  mutate(carrier_mean = mean(arr_delay, na.rm = TRUE)) |>
  ungroup() |>                              # Remove grouping
  mutate(grand_mean = mean(arr_delay, na.rm = TRUE)) |>  # Now uses ALL flights
  select(carrier, arr_delay, carrier_mean, grand_mean)
```

::: {.notes}
The key teaching point: without `ungroup()`, the `grand_mean` would be calculated per carrier (same as `carrier_mean`). With `ungroup()`, it uses all rows. Students often ask "How do I know if my data is still grouped?" — point them to the `Groups:` line printed at the top of tibble output. A good rule of thumb: if you're done with grouped operations, add `ungroup()` to be safe.
:::

# Code style matters {background-color="#2c3e50"}

## Spot the difference

```{r}
#| eval: false
# Version A
pD<-read_csv("participants.csv")
m<-pD|>filter(age>25)|>group_by(condition)|>summarize(m=mean(score,na.rm=TRUE),s=sd(score,na.rm=TRUE),n=n())
```

. . .

```{r}
#| eval: false
# Version B
participant_data <- read_csv("participants.csv")

# Adults only — age cutoff matches preregistration
participant_data |>
  filter(age > 25) |>
  group_by(condition) |>
  summarize(
    mean_score = mean(score, na.rm = TRUE),
    sd_score = sd(score, na.rm = TRUE),
    n = n()
  )
```

Same code. One you can debug at midnight. One you can't.

::: {.notes}
Let students look at Version A for a moment before revealing Version B. Ask: "Which one would you rather debug at midnight before the assignment is due?" This gets a laugh and makes the point viscerally. Version A is deliberately awful but realistic — students write code like this when they're rushing. Emphasize that both versions produce identical output; the difference is entirely for the human reader.
:::

## The rules behind the difference

::: {.columns}
::: {.column width="50%"}
**Spacing**

- Spaces around `<-`, `==`, `|>`
- Spaces after commas
- One verb per line
:::
::: {.column width="50%"}
**Naming**

- `snake_case`: `mean_score`, not `m`
- Meaningful: `participant_data`, not `pD`
- Comments explain **why**, not what
:::
:::

# Practice: Psychology example {background-color="#e67e22"}

## Simulated experiment data

::: {.notes}
Transition: "Now let's put it all together with data that looks like what you'd actually collect." Walk through the `set.seed()` and `tibble()` call briefly — students don't need to understand the simulation code, just the resulting data structure. Point out that this is a 2-condition experiment with 60 participants doing 20 trials each (1,200 rows). This mirrors a real cognitive psychology experiment.
:::

```{r}
set.seed(42)

# Create a simulated experiment dataset
experiment <- tibble(
  participant_id = rep(1:60, each = 20),
  condition = rep(rep(c("control", "treatment"), each = 10), 60),
  trial = rep(1:20, 60),
  reaction_time = rnorm(1200, mean = 500, sd = 100) +
                  ifelse(rep(rep(c(0, 1), each = 10), 60) == 1, -30, 0),
  accuracy = rbinom(1200, 1, prob = 0.85 +
                    ifelse(rep(rep(c(0, 1), each = 10), 60) == 1, 0.05, 0))
)
```

## Examine the data

```{r}
experiment
```

## Participant-level summaries

```{r}
#| label: part-sum
#| eval: false
# First, summarize BY PARTICIPANT
participant_summaries <- experiment |>
  group_by(participant_id, condition) |>
  summarize(
    mean_rt = mean(reaction_time),
    accuracy = mean(accuracy) * 100,
    n_trials = n(),
    .groups = "drop"
  )

participant_summaries
```

## Participant-level summaries

```{r}
#| ref.label: part-sum
#| echo: false
```

## Condition-level summaries

```{r}
# Now summarize ACROSS PARTICIPANTS by condition
condition_summaries <- participant_summaries |>
  group_by(condition) |>
  summarize(
    M_rt = mean(mean_rt),
    SD_rt = sd(mean_rt),
    M_acc = mean(accuracy),
    SD_acc = sd(accuracy),
    n = n()
  )

condition_summaries
```

::: {.notes}
This two-step summarization (trials to participants, then participants to conditions) is the most important methodological pattern in the deck. Emphasize why you can't just do `group_by(condition) |> summarize(mean(reaction_time))` on the raw data — that would weight participants with more valid trials more heavily. Students who've taken stats will recognize this as the distinction between treating trials vs. participants as the unit of analysis. Spend ~3 minutes walking through both steps.
:::

## Visualize the results

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Boxplot with overlaid data points comparing mean reaction time between control and treatment conditions, showing the treatment group has slightly faster reaction times with similar spread."
participant_summaries |>
  ggplot(aes(x = condition, y = mean_rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(
    title = "Reaction Time by Condition",
    x = "Condition",
    y = "Mean RT (ms)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## Get a head start: Assignment 2

Using the `experiment` data we just created:

1. Filter to only accurate trials (`accuracy == 1`)
2. Calculate mean RT **per participant** per condition
3. Find the 5 participants with the **fastest** mean RT in each condition

::: {.callout-tip}
This uses everything from today — `filter()`, `group_by()`, `summarize()`, and `slice_min()`. Work through it step by step.
:::

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
experiment |>
  filter(accuracy == 1) |>
  group_by(participant_id, condition) |>
  summarize(mean_rt = mean(reaction_time), .groups = "drop") |>
  group_by(condition) |>
  slice_min(mean_rt, n = 5)
```

::: {.notes}
Give students ~10 minutes of individual work time. This exercise requires chaining four operations and doing a second `group_by()` after the first summarize — that's the hard part. Hint if students are stuck: "Build the pipeline one line at a time. First just filter, run it, then add the group_by, run it, then add summarize..." This incremental approach is a skill in itself. Assignment 2 is assigned today, so mention that it builds on the same skills. Reveal the solution with ~2 minutes left.
:::

# Wrapping up {background-color="#2c3e50"}

## Today's additions

| Verb | What it does |
|------|--------------|
| `group_by()` | Define groups for operations |
| `summarize()` | Calculate summary statistics |
| `count()` | Quick frequency counts |
| `ungroup()` | Remove grouping |
| `slice_*()` | Select rows by position/rank |

## The analysis workflow

```{r}
#| eval: false
raw_data |>
  filter(valid_data) |>           # Clean
  select(relevant_vars) |>        # Focus
  mutate(new_vars) |>             # Transform
  group_by(conditions) |>         # Split
  summarize(statistics) |>        # Aggregate
  arrange(order) |>               # Sort
  ggplot() + ...                  # Visualize
```

## Before next class

📖 **Read:**

- [R4DS Ch 5: Data tidying](https://r4ds.hadley.nz/data-tidy)

✅ **Practice:**

- Calculate descriptive statistics by group
- Build complete analysis pipelines
- Focus on writing clean, readable code

## Key takeaways

1. **group_by() + summarize()** is incredibly powerful
2. **count()** is a quick shortcut for frequencies
3. **Grouped operations** work with mutate() and filter() too
4. **ungroup()** when you need to work with all data again
5. **Code style matters** — write for your future self

## The one thing to remember

`group_by()` + `summarize()` is how you go from raw data to the descriptive statistics table in every published paper.

Next time: Data Tidying with tidyr

::: {.notes}
End on this line — it connects today's skills directly to something students will encounter in every journal article they read. Remind them that Assignment 2 is now posted and due Sunday at 11:59 PM. Next session is Data Tidying (pivot_longer, pivot_wider), which is conceptually harder but builds on what they learned today.
:::
