---
title: "Data Tidying"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 13, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 5"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
```

# What is tidy data? {background-color="#2c3e50"}

## The tidyverse philosophy

> "Tidy datasets are all alike, but every messy dataset is messy in its own way."
> — Hadley Wickham

The tools we've learned (ggplot2, dplyr) expect data in a specific format: **tidy data**.

## The three rules of tidy data

1. Each **variable** is a column
2. Each **observation** is a row
3. Each **value** is a cell

. . .

Simple in theory, surprisingly complex in practice.

## Tidy data visualized

```{r}
#| echo: false
tibble(
  participant = c("P1", "P2", "P3"),
  pre_test = c(45, 52, 48),
  post_test = c(62, 58, 71)
) |>
  knitr::kable(caption = "Is this tidy?")
```

. . .

**No!** Time (pre/post) is a variable, but it's spread across columns.

## The tidy version

```{r}
#| echo: false
tibble(
  participant = rep(c("P1", "P2", "P3"), each = 2),
  time = rep(c("pre", "post"), 3),
  score = c(45, 62, 52, 58, 48, 71)
) |>
  knitr::kable(caption = "Now it's tidy!")
```

Now: participant, time, and score are all columns.

## Why does it matter?

**Tidy data works with tidyverse tools:**

```{r}
#| eval: false
# Easy to visualize
ggplot(data, aes(x = time, y = score, color = participant)) +
  geom_point()

# Easy to analyze
data |>
  group_by(time) |>
  summarize(mean = mean(score))
```

## Wide vs. Long

:::: {.columns}
::: {.column width="50%"}
**Wide format**

- Variables spread across columns
- One row per subject
- Humans like to read this
:::

::: {.column width="50%"}
**Long format**

- Variables in a single column
- Multiple rows per subject
- R likes to work with this
:::
::::

Most real data needs reshaping.

# Common untidy patterns {background-color="#2c3e50"}

## Pattern 1: Column headers are values

```{r}
wide_scores <- tibble(
  student = c("Alice", "Bob", "Carol"),
  fall_2024 = c(85, 78, 92),
  spring_2025 = c(88, 82, 95),
  fall_2025 = c(91, 85, 94)
)
wide_scores
```

The semester names are **values**, not variable names.

## Pattern 2: Multiple variables in one column

```{r}
messy_data <- tibble(
  id = 1:3,
  age_sex = c("25_M", "32_F", "28_F")
)
messy_data
```

Age and sex are crammed into one column.

## Pattern 3: Variables in rows and columns

```{r}
weather <- tibble(
  id = c("MX001", "MX001", "MX002", "MX002"),
  year = c(2020, 2020, 2020, 2020),
  month = c(1, 2, 1, 2),
  element = c("tmax", "tmax", "tmin", "tmin"),
  value = c(85, 87, 32, 35)
)
weather
```

`element` contains variable names (tmax, tmin).

## Psychology-specific patterns

Surveys often look like:

```{r}
survey_wide <- tibble(
  participant = 1:3,
  bdi_1 = c(2, 1, 3),
  bdi_2 = c(1, 0, 2),
  bdi_3 = c(3, 2, 2),
  bdi_4 = c(2, 1, 1)
)
survey_wide
```

Each item is a column — wide format.

# pivot_longer() {background-color="#2c3e50"}

## The most common tidying operation

`pivot_longer()` takes wide data and makes it long:

```{r}
wide_scores |>
  pivot_longer(
    cols = fall_2024:fall_2025,  # Which columns to pivot
    names_to = "semester",        # New column for old column names
    values_to = "score"           # New column for values
  )
```

## Breaking it down

```{r}
#| eval: false
pivot_longer(
  cols = ...,        # Columns to reshape (use select helpers!)
  names_to = "...",  # Name for the new "names" column
  values_to = "..."  # Name for the new "values" column
)
```

## Selecting columns to pivot

Use any of the `select()` helpers:

```{r}
#| eval: false
# By name
pivot_longer(cols = c(fall_2024, spring_2025, fall_2025))

# By range
pivot_longer(cols = fall_2024:fall_2025)

# By pattern
pivot_longer(cols = starts_with("fall"))
pivot_longer(cols = contains("202"))

# Everything except
pivot_longer(cols = -student)
```

## Psychology example: Survey items

```{r}
survey_wide |>
  pivot_longer(
    cols = starts_with("bdi"),
    names_to = "item",
    values_to = "response"
  )
```

Now each response is its own row!

## Extracting information from names

What if column names contain useful info? We want to extract the time point (t1, t2, t3).

```{r}
# Scores at different time points
experiment_wide <- tibble(
  id = 1:3,
  score_t1 = c(100, 95, 110),
  score_t2 = c(105, 100, 115),
  score_t3 = c(108, 102, 120)
)
experiment_wide
```

## names_prefix argument

```{r}
experiment_wide |>
  pivot_longer(
    cols = starts_with("score"),
    names_to = "time",
    names_prefix = "score_",  # Remove this prefix from names
    values_to = "score"
  )
```

## names_pattern argument

For more complex parsing:

```{r}
# Column names like "bdi_1", "anxiety_1", etc.
multi_scale <- tibble(
  id = 1:2,
  bdi_1 = c(2, 1), bdi_2 = c(1, 2),
  anxiety_1 = c(3, 2), anxiety_2 = c(2, 3)
)
multi_scale
```

## names_pattern argument

For more complex parsing:
```{r}
multi_scale |>
  pivot_longer(
    cols = -id,
    names_to = c("scale", "item"),
    names_pattern = "(.+)_(.+)",  # Regex: anything_anything
    values_to = "response"
  )
```

## Pivoting to calculate scale scores

```{r}
# Calculate BDI total from long format
survey_wide |>
  pivot_longer(
    cols = starts_with("bdi"),
    names_to = "item",
    values_to = "response"
  ) |>
  group_by(participant) |>
  summarize(bdi_total = sum(response))
```

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

We defined `survey_wide` earlier in this deck:

```{r}
survey_wide
```

With a partner, figure out: Which participant has the highest mean BDI score?

::: {.callout-tip}
You'll need `pivot_longer()` followed by `group_by()` + `summarize()`. The column names start with `"bdi"` — that's a hint.
:::

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
survey_wide |>
  pivot_longer(
    cols = starts_with("bdi"),
    names_to = "item",
    values_to = "response"
  ) |>
  group_by(participant) |>
  summarize(mean_bdi = mean(response))
```

---

## Before we move on

📤 **Upload your code to Canvas** for participation credit. Paste what you have into today's in-class submission — it doesn't need to work perfectly.

# pivot_wider() {background-color="#2c3e50"}

## The opposite operation

Sometimes you need to go from long to wide:

```{r}
long_data <- tibble(
  participant = rep(1:3, each = 2),
  time = rep(c("pre", "post"), 3),
  score = c(45, 62, 52, 58, 48, 71)
)
long_data
```

## pivot_wider() syntax

```{r}
long_data |>
  pivot_wider(
    names_from = time,   # Column to get new column names from
    values_from = score  # Column to get values from
  )
```

## When to use pivot_wider()

- Creating summary tables for reports
- Some analyses need wide format
- Merging data that was collected differently
- Human-readable output

## Multiple value columns

```{r}
# Long data with multiple measures
long_multi <- tibble(
  id = rep(1:2, each = 2),
  time = rep(c("pre", "post"), 2),
  score = c(45, 62, 52, 58),
  rt = c(500, 480, 520, 490)
)
long_multi
```

## Multiple value columns

```{r}
long_multi |>
  pivot_wider(
    names_from = time,
    values_from = c(score, rt)  # Multiple columns!
  )
```


# Separating and uniting {background-color="#2c3e50"}

## separate_wider_delim()

Split one column into multiple:

```{r}
tibble(
  id = 1:3,
  age_sex = c("25_M", "32_F", "28_F"))|>
  separate_wider_delim(
    cols = age_sex,
    delim = "_",
    names = c("age", "sex")
  )
```

## separate_wider_regex()

For complex patterns:

```{r}
tibble(
  code = c("A123", "B456", "C789")
) |>
  separate_wider_regex(
    cols = code,
    patterns = c(letter = "[A-Z]", number = "[0-9]+")
  )
```

## unite()

The opposite — combine columns:

```{r}
tibble(
  year = c(2024, 2024, 2025),
  month = c(1, 6, 1),
  day = c(15, 20, 10)
) |>
  unite(
    col = "date",       # New column name
    year, month, day,   # Columns to combine
    sep = "-"           # Separator
  )
```

# Real-world examples {background-color="#2c3e50"}

## Example 1: Repeated measures experiment

```{r}
# Data as you might receive it from SPSS
wide_rm <- tibble(
  subject = 1:4,
  cond_a_time1 = c(450, 520, 480, 510),
  cond_a_time2 = c(420, 490, 460, 480),
  cond_b_time1 = c(480, 540, 500, 530),
  cond_b_time2 = c(440, 510, 470, 500)
)
wide_rm
```

## Tidying repeated measures

```{r}
tidy_rm <- wide_rm |>
  pivot_longer(
    cols = -subject,
    names_to = c("condition", "time"),
    names_pattern = "cond_(.+)_time(.+)",
    values_to = "rt"
  )
tidy_rm
```

## Now we can analyze it!

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
tidy_rm |>
  ggplot(aes(x = time, y = rt, color = condition, group = condition)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1) +
  labs(
    title = "Reaction Time by Condition and Time",
    x = "Time Point",
    y = "Reaction Time (ms)"
  ) +
  theme_minimal(base_size = 14)
```

## Example 2: Questionnaire with subscales

```{r}
# Raw questionnaire data
quest <- tibble(
  pid = 1:3,
  anx_1 = c(3, 2, 4), anx_2 = c(2, 3, 3), anx_3 = c(4, 2, 5),
  dep_1 = c(2, 1, 3), dep_2 = c(3, 2, 4), dep_3 = c(2, 1, 3)
)

quest
```

## Example 2: Questionnaire with subscales

```{r}
#| label: quest-sub
#| eval: false
# Tidy and calculate subscales
quest |>
  pivot_longer(
    cols = -pid,
    names_to = c("scale", "item"),
    names_pattern = "(.+)_(.+)",
    values_to = "response"
  ) |>
  group_by(pid, scale) |>
  summarize(subscale_mean = mean(response), .groups = "drop") |>
  pivot_wider(names_from = scale, values_from = subscale_mean)
```

## Example 2: Questionnaire with subscales

```{r}
#| ref.label: quest-sub
#| echo: false
```

## Example 3: Multilevel/nested data

```{r}
# Students nested in classrooms
students <- tibble(
  classroom = rep(c("A", "B"), each = 3),
  student = 1:6,
  pretest = c(70, 75, 72, 68, 71, 69),
  posttest = c(80, 82, 78, 75, 79, 77)
)

students
```

## Example 3: Multilevel/nested data

```{r}
# Tidy for analysis
students |>
  pivot_longer(
    cols = c(pretest, posttest),
    names_to = "time",
    values_to = "score"
  )
```

# Common pitfalls {background-color="#2c3e50"}

## Pitfall 1: Forgetting what's a variable

Ask yourself: What are my **variables**?

- Participant ID? ✓ Variable
- Time point? ✓ Variable (not separate columns!)
- Score? ✓ Variable
- Item number? Depends on your analysis

## Pitfall 2: Over-pivoting

Not everything needs to be long:

```{r}
#| eval: false
# Maybe this is fine as-is?
tibble(
  id = 1:3,
  age = c(25, 32, 28),
  gender = c("M", "F", "F"),
  score = c(85, 92, 88)
)
```

Age, gender, and score are **different variables** — keep them as columns.

## General tidying strategy

1. **Identify** the variables (what are you measuring?)
2. **Look** at your current structure (what's a row? column?)
3. **Determine** what operations you need
4. **Test** with a small subset first
5. **Verify** you haven't lost data

# Get a head start {background-color="#2c3e50"}

## Try it yourself

We created `wide_rm` earlier — a repeated measures dataset:

On your own:

1. Pivot it to long format, extracting **condition** and **time** from the column names
2. Calculate **mean RT by condition and time**
3. Sketch (on paper or in ggplot) what you'd expect the plot to look like

This is very close to what Assignment 3 will ask you to do.

```{r}
#| echo: false
#| eval: false
# Solution (reveal if needed)
wide_rm |>
  pivot_longer(
    cols = -subject,
    names_to = c("condition", "time"),
    names_pattern = "cond_(.+)_time(.+)",
    values_to = "rt"
  ) |>
  group_by(condition, time) |>
  summarize(mean_rt = mean(rt))
```

# Wrapping up {background-color="#2c3e50"}

## The tidyr toolkit

| Function | What it does |
|----------|--------------|
| `pivot_longer()` | Wide → Long |
| `pivot_wider()` | Long → Wide |
| `separate_*()` | Split columns |
| `unite()` | Combine columns |

## The tidy data mantra

1. Each **variable** is a column
2. Each **observation** is a row
3. Each **value** is a cell

When in doubt, ask: "What would make this easiest to plot/analyze?"

## Before next class

📖 **Read:**

- [R4DS Ch 7: Data import](https://r4ds.hadley.nz/data-import)
- [R4DS Ch 20: Spreadsheets](https://r4ds.hadley.nz/spreadsheets)

✅ **Practice:**

- Reshape a dataset you've worked with
- Try tidying some messy example data
- Practice `pivot_longer()` — it's the most common

## Key takeaways

1. **Tidy data** has a specific structure that works with tidyverse
2. **pivot_longer()** is your most-used tidying function
3. **pivot_wider()** is useful for tables and some analyses
4. **Think about your variables** before reshaping
5. **Column names contain information** — extract it with `names_pattern`

## Questions?

Next time: **Data Import**

We'll learn to read CSV files, Excel spreadsheets, and more!
