---
title: "Data Import"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 15, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 6"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
```

# Getting real data into R {background-color="#2c3e50"}

## Your data isn't inside a package

So far we've used datasets that were already loaded in R — `mpg`, `flights`, `bfi`.

. . .

But your thesis data isn't sitting inside a package. It's a CSV on your desktop, an Excel file from your advisor, or a Qualtrics export with 200 columns named `Q1`, `Q2`, `Q3`...

. . .

Today we learn how to get it into R — and what to check when things don't import cleanly.

# Reading CSV files {background-color="#2c3e50"}

## read_csv() — your workhorse

```{r}
#| eval: false
# The basic pattern
my_data <- read_csv("path/to/file.csv")
```

- Part of the `readr` package (loaded with tidyverse)
- Returns a **tibble**, not a base R data frame
- Smarter about guessing column types than `read.csv()`

## Why not read.csv()?

| | `read_csv()` | `read.csv()` |
|---|---|---|
| Package | readr (tidyverse) | base R |
| Returns | tibble | data.frame |
| Speed | Much faster | Slower |
| Type guessing | Better | Converts strings to factors |
| Consistent | Yes | Depends on options |

Always use `read_csv()` in this class.

## What read_csv() does automatically

```{r}
#| eval: false
# It figures out column types from the first 1000 rows
my_data <- read_csv("survey.csv")

# It tells you what it decided:
glimpse(my_data)
```

Types it recognizes: `dbl` (number), `chr` (text), `lgl` (TRUE/FALSE), `date`, `dttm` (date-time)

## When the guess is wrong

Sometimes R guesses wrong — especially with:

- ID numbers (should be character, not numeric)
- Zip codes, phone numbers
- Dates in unusual formats
- Columns that are mostly empty

```{r}
#| eval: false
# Check for problems after reading
my_data <- read_csv("survey.csv")
problems(my_data)  # Shows any parsing issues
```

## Specifying column types

```{r}
#| eval: false
# Tell read_csv() what types to use
my_data <- read_csv("survey.csv",
  col_types = cols(
    participant_id = col_character(),  # Don't treat as number
    age = col_integer(),
    gender = col_character(),
    score = col_double()
  )
)
```

## col_types shortcuts

| Shortcut | Type |
|----------|------|
| `"c"` | character |
| `"d"` | double |
| `"i"` | integer |
| `"l"` | logical |
| `"D"` | date |
| `"?"` | let R guess |

```{r}
#| eval: false
# Compact version
read_csv("survey.csv", col_types = "cdiccc?")
```

## Handling messy headers

Real data often has ugly column names:

```{r}
#| eval: false
# Skip rows at the top (metadata, notes)
read_csv("messy.csv", skip = 2)

# Only read specific rows
read_csv("messy.csv", n_max = 100)

# Rename columns after import
my_data <- read_csv("messy.csv") |>
  rename(age = `Age (years)`, score = `Total Score`)
```

## Missing values

By default, `read_csv()` recognizes these as NA:

`NA`, `N/A`, `NaN`, `""` (empty), `"."`, `"NULL"`

```{r}
#| eval: false
# Add your own NA strings
read_csv("survey.csv", na = c("NA", "N/A", "-999", ""))
```

Psychology datasets often use `-999` or `99` as missing value codes!

## Writing CSV files

```{r}
#| eval: false
# Save your cleaned data
write_csv(my_cleaned_data, "data/cleaned_survey.csv")
```

Save intermediate cleaned files — don't re-clean every time you start R.

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

I've created a messy CSV file for you. [Download it here.](https://github.com/sjweston/datasci410/blob/main/messy_survey.csv) With a partner:

1. Read it back in with `read_csv()` — what types does R guess?
2. Fix the missing values: `-999` and `"N/A"` should be `NA`
3. Rename the columns to clean snake_case names
4. What does `problems()` show you?

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
messy <- read_csv("messy_survey.csv", na = c("N/A", "-999", ""))
mess = messy |>
  rename(
    participant_id = `Participant ID`,
    age = `Age (years)`,
    gender = Gender,
    score_1 = Score_1,
    score_2 = Score_2,
    notes = Notes
  )
```

---

## Before we move on

📤 **Upload your code to Canvas** for participation credit. Paste what you have into today's in-class submission — it doesn't need to work perfectly.

# Excel files {background-color="#2c3e50"}

## readxl::read_excel()

Excel is everywhere in psychology. The `readxl` package handles it:

```{r}
#| eval: false
library(readxl)  # Or just use readxl:: directly

# Read the first sheet
my_data <- read_excel("survey_results.xlsx")

# Read a specific sheet
my_data <- read_excel("survey_results.xlsx", sheet = "Demographics")

# Read a specific range
my_data <- read_excel("survey_results.xlsx",
  sheet = "Scores",
  range = "A1:F50"
)
```

## Excel pitfalls

Things that look fine in Excel but break in R:

- **Merged cells** — R sees them as one value + empty cells
- **Colors as data** — R can't read cell colors
- **Multiple tables on one sheet** — use `range` to isolate them
- **Formatted numbers** — "1,234" might read as text

::: {.callout-tip}
The best fix is usually to clean the Excel file first, or use `range` to read just the clean part.
:::

## Listing sheet names

```{r}
#| eval: false
# See what sheets are in a file
excel_sheet_names("survey_results.xlsx")
```

Useful when you don't know the file structure yet.

# Other file formats {background-color="#2c3e50"}

## SPSS and SAS files

Psychology labs often have data in SPSS (`.sav`) or SAS formats:

```{r}
#| eval: false
library(haven)

# Read SPSS
spss_data <- read_sav("study.sav")

# Read SAS
sas_data <- read_sas("study.sas")
```

These return tibbles, just like `read_csv()`.

## A note about SPSS labels

SPSS files often have **variable labels** and **value labels**:

```{r}
#| eval: false
# Labels are stored as attributes
attr(spss_data$gender, "label")   # "Participant Gender"
attr(spss_data$gender, "levels")  # 1="Male", 2="Female"

# To drop labels and work with plain values:
library(haven)
spss_data <- zap_labels(spss_data)
```

## Getting data out of Qualtrics

The most common workflow in psychology:

1. In Qualtrics: **Download** → **CSV** (or use the API — advanced)
2. The export includes metadata rows at the top — skip them:

```{r}
#| eval: false
# Qualtrics CSVs typically have 3 header rows
qualtrics_data <- read_csv("my_survey_export.csv", skip = 2)
```

3. Clean up the automatic column names (they're ugly)
4. Handle the timestamp columns if you need them

::: {.callout-note}
The first two rows after the header contain Qualtrics' own descriptions. `skip = 2` gets past them. Always check with `glimpse()` after importing.
:::

# Putting it together {background-color="#2c3e50"}

## A realistic import workflow

```{r}
#| eval: false
# 1. Read the file
raw_data <- read_csv("qualtrics_export.csv",
  skip = 2,
  na = c("", "N/A", "-999")
)

# 2. Check what you got
glimpse(raw_data)
problems(raw_data)

# 3. Rename columns
clean_data <- raw_data |>
  rename(
    id = ResponseId,
    age = Q1,
    condition = Q2
  )

# 4. Fix types if needed
clean_data <- clean_data |>
  mutate(
    id = as.character(id),
    age = as.numeric(age)
  )

# 5. Save the cleaned version
write_csv(clean_data, "data/clean_survey.csv")
```

## Get a head start: Assignment 3

Try this mini-challenge before next class:

```{r}
#| eval: false
# 1. Import the messy CSV from the pair exercise
messy <- read_csv("data/messy_survey.csv", na = c("", "NA", "-999", "N/A"))

# 2. Check for problems
problems(messy)

# 3. Fix at least one column type issue with col_types or
#    readr::parse_number()

# 4. Save your cleaned version
write_csv(clean, "data/clean_survey.csv")
```

Assignment 3 will have you do this end to end with real messy data.

## Solution

```{r}
#| echo: false
#| eval: false
# Solution to end-of-deck exercise
messy <- read_csv("data/messy_survey.csv",
                  na = c("", "NA", "-999", "N/A"),
                  col_types = cols(
                    id = col_character(),
                    age = col_double(),
                    score = col_double()
                  ))

problems(messy)  # Check what readr flagged

clean <- messy |>
  mutate(age = parse_number(age))

write_csv(clean, "data/clean_survey.csv")
```

# Wrapping up {background-color="#2c3e50"}

## The import toolkit

| Function | What it does |
|----------|--------------|
| `read_csv()` | Read CSV files |
| `write_csv()` | Save CSV files |
| `read_excel()` | Read Excel files |
| `read_sav()` | Read SPSS files |
| `problems()` | Check for import issues |
| `col_types` | Control column types |

## Before next class

📖 **Read:**

- [R4DS Ch 9: Layers](https://r4ds.hadley.nz/layers)

✅ **Practice:**

- Try importing a CSV or Excel file you have lying around
- Check `problems()` — does it flag anything?
- Practice renaming messy column names

## Key takeaways

1. **`read_csv()` is your default** — it's faster and smarter than `read.csv()`
2. **Check your imports** — `glimpse()` and `problems()` are your friends
3. **Name your missing values** — psychology data loves `-999`
4. **Excel is tricky** — use `readxl` and be suspicious of formatting
5. **Save cleaned files** — don't re-clean every time

## The one thing to remember

Your analysis is only as good as your import. Check it before you trust it.

Next time: Layers & Aesthetics
