---
title: "Data Import"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 15, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 6"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
    include-in-header:
      text: |
        <link rel="stylesheet" href="toc-nav.css">
    include-after-body:
      text: |
        <script src="toc-nav.js"></script>
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(here)
```

# Getting real data into R {background-color="#2c3e50"}

## Your data isn't inside a package

So far we've used datasets that were already loaded in R — `mpg`, `flights`, `bfi`.

. . .

But your thesis data isn't sitting inside a package. It's a CSV on your desktop, an Excel file from your advisor, or a Qualtrics export with 200 columns named `Q1`, `Q2`, `Q3`...

. . .

Today we learn how to get it into R — and what to check when things don't import cleanly.

::: {.notes}
This opening connects to students' real frustration — they know R can do things, but they haven't been able to use their own data yet. Ask: "How many of you have data sitting in a CSV or Excel file right now?" Most hands should go up. That's the motivation. This is a shorter, more practical deck than Session 5, so the pace can be a bit faster.
:::

# Reading CSV files {background-color="#2c3e50"}

## read_csv() — your workhorse

```{r}
#| eval: false
# The basic pattern
my_data <- read_csv("path/to/file.csv")
```

- Part of the `readr` package (loaded with tidyverse)
- Returns a **tibble**, not a base R data frame
- Smarter about guessing column types than `read.csv()`

::: {.notes}
Emphasize the file path — this is where students get stuck most often. Common issues: wrong working directory, file not in the project folder, using backslashes on Windows. If you're doing a live demo, show what happens when the path is wrong (a clear error message). Remind them about the RStudio Files pane for navigating to files.
:::

## The file path problem

`read_csv()` needs to know *where* your file is:

```{r}
#| eval: false
# This works on your laptop...
my_data <- read_csv("/Users/sara/Desktop/project/data/survey.csv")

# ...but not on anyone else's. And not if you move the folder.
```

. . .

**Hardcoded paths break.** Share your code with a collaborator, move your project folder, or switch computers — and it's broken.

::: {.notes}
This is the number one "it worked yesterday" problem. Walk through the scenario: student emails code to their advisor, advisor opens it on their laptop, and it immediately errors because `/Users/sara/Desktop` doesn't exist. Make this feel real — it happens constantly. The fix is to never use absolute paths and instead anchor paths to the project root.
:::

## here::here() — paths that always work

```{r}
#| eval: false
library(here)

# Build paths relative to your project root
my_data <- read_csv(here("data", "survey.csv"))

# Or use a single string with slashes
my_data <- read_csv(here("data/survey.csv"))

# See where here() thinks your root is:
here()
```

- `here()` finds your **project root** — where the `.Rproj` file lives
- Works on any machine, any operating system
- No more "works on my laptop" problems

::: {.notes}
`here` uses the presence of an `.Rproj` file (or a few other markers) to locate the project root, then builds paths from there. This means the same code runs on any collaborator's machine as long as they have the same project folder structure. Show students how to run `here()` alone first — it prints the root path so they can confirm it's pointing at the right place. The two-argument vs. single-string forms are equivalent; the multi-argument form is slightly more readable. This pairs naturally with the RStudio Projects content from Session 1.
:::

## Why not read.csv()?

| | `read_csv()` | `read.csv()` |
|---|---|---|
| Package | readr (tidyverse) | base R |
| Returns | tibble | data.frame |
| Speed | Much faster | Slower |
| Type guessing | Better | Converts strings to factors |
| Consistent | Yes | Depends on options |

Always use `read_csv()` in this class.

::: {.notes}
Students will encounter read.csv() in StackOverflow answers and older textbooks. The key practical difference: read.csv() silently converts character columns to factors, which causes confusing errors later. read_csv() (underscore) doesn't do this. Emphasize the underscore vs. period distinction — it's an easy typo. If a student's code uses read.csv(), it will still work, but they may hit unexpected behavior.
:::

## What read_csv() does automatically

```{r}
#| eval: false
# It figures out column types from the first 1000 rows
my_data <- read_csv("survey.csv")

# It tells you what it decided:
glimpse(my_data)
```

Types it recognizes: `dbl` (number), `chr` (text), `lgl` (TRUE/FALSE), `date`, `dttm` (date-time)

## When the guess is wrong

Sometimes R guesses wrong — especially with:

- ID numbers (should be character, not numeric)
- Zip codes, phone numbers
- Dates in unusual formats
- Columns that are mostly empty

```{r}
#| eval: false
# Check for problems after reading
my_data <- read_csv("survey.csv")
problems(my_data)  # Shows any parsing issues
```

::: {.notes}
The participant ID example resonates most with psychology students. If IDs are "001", "002", "003", R reads them as numbers 1, 2, 3 — losing the leading zeros. Ask: "Why does that matter?" It matters for merging datasets later, and it changes how the ID behaves in plots. Always run problems() after importing — make this a habit students build now.
:::

## Specifying column types

```{r}
#| eval: false
# Tell read_csv() what types to use
my_data <- read_csv("survey.csv",
  col_types = cols(
    participant_id = col_character(),  # Don't treat as number
    age = col_integer(),
    major = col_character(),
    score = col_double()
  )
)
```

::: {.notes}
Frame this as: "When R guesses wrong, you override it." Students don't need to memorize every col_* function — the important ones are col_character() and col_double(). Mention that you can specify types for just the problem columns and let R guess the rest (using .default = col_guess() inside cols()). Spend ~2 minutes here.
:::

## col_types shortcuts

| Shortcut | Type |
|----------|------|
| `"c"` | character |
| `"d"` | double |
| `"i"` | integer |
| `"l"` | logical |
| `"D"` | date |
| `"?"` | let R guess |

```{r}
#| eval: false
# Compact version
read_csv("survey.csv", col_types = "cdiccc?")
```

## Handling messy headers

Real data often has ugly column names:

```{r}
#| eval: false
# Skip rows at the top (metadata, notes)
read_csv("messy.csv", skip = 2)

# Only read specific rows
read_csv("messy.csv", n_max = 100)

# Rename columns after import
my_data <- read_csv("messy.csv") |>
  rename(age = `Age (years)`, score = `Total Score`)
```

::: {.notes}
Point out the backticks around column names with spaces or special characters. Students will see this a lot with Qualtrics exports and Excel files. The backtick syntax confuses people — explain that R needs them because "Age (years)" isn't a valid R name. The janitor::clean_names() function is a nice shortcut here if you want to mention it briefly, though it's not required for this course.
:::

## Missing values

By default, `read_csv()` recognizes these as NA:

`NA`, `N/A`, `NaN`, `""` (empty), `"."`, `"NULL"`

```{r}
#| eval: false
# Add your own NA strings
read_csv("survey.csv", na = c("NA", "N/A", "-999", ""))
```

Psychology datasets often use `-999` or `99` as missing value codes!

::: {.notes}
This is critical for psychology data. Many students will have worked with SPSS datasets that use -999 as a missing value code. If you don't catch these at import, they'll silently corrupt your means and standard deviations. Ask: "What happens to your mean anxiety score if one participant has -999 instead of NA?" The answer is devastating — and that's the point. Handling missing values at import prevents downstream errors.
:::

## Writing CSV files

```{r}
#| eval: false
# Save your cleaned data
write_csv(my_cleaned_data, here("data/cleaned_survey.csv"))
```

Save intermediate cleaned files — don't re-clean every time you start R.

::: {.notes}
This is a workflow habit, not a coding lesson. Emphasize the practical benefit: cleaning data takes time, so save the cleaned version and read that in future scripts. This also connects to reproducibility (Session 7) — your cleaning script documents exactly what you did to the raw data. Transition to the pair exercise: "Now let's practice the full import-check-fix workflow."
:::

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

I've created a messy CSV file for you. [Download it here.](https://github.com/sjweston/datasci410/blob/main/messy_survey.csv) With a partner:

1. Read it back in with `read_csv()` — what types does R guess?
2. Fix the missing values: `-999` and `"N/A"` should be `NA`
3. Rename the columns to clean snake_case names
4. What does `problems()` show you?

::: {.notes}
Set a 10-minute timer. Make sure students have downloaded the CSV file before starting — this is a common bottleneck. Circulate and check that students are using read_csv() (underscore, not period). The most common mistake is forgetting to include all the NA strings in the na argument.
:::

---

## Solution

```{r}
#| echo: true
#| eval: false
messy <- read_csv(
  here("data/messy_survey.csv"), #the path depends on where you saved your data
  na = c("N/A", "-999", ""),
  col_types = "cdciic")

problems(messy)

clean <- messy |>
  rename(
    participant_id = `Participant ID`,
    age = `Age (years)`,
    major = Major,
    score_1 = Score_1,
    score_2 = Score_2,
    notes = Notes
  )
```

# Excel files {background-color="#2c3e50"}

## readxl::read_excel()

Excel is everywhere in psychology. The `readxl` package handles it:

```{r}
#| eval: false
library(readxl)  # Or just use readxl:: directly

# Read the first sheet
my_data <- read_excel("survey_results.xlsx")

# Read a specific sheet
my_data <- read_excel("survey_results.xlsx", sheet = "Demographics")

# Read a specific range
my_data <- read_excel("survey_results.xlsx",
  sheet = "Scores",
  range = "A1:F50"
)
```

::: {.notes}
Note that readxl is installed with tidyverse but not loaded by library(tidyverse) — you need to load it separately or use the readxl:: prefix. Students often ask "Why not just save as CSV first?" That's a reasonable approach, but read_excel() handles it directly and avoids data loss from CSV conversion (e.g., Excel date formatting). Spend ~3 minutes on Excel; it's familiar territory for most students.
:::

## Excel pitfalls

Things that look fine in Excel but break in R:

- **Merged cells** — R sees them as one value + empty cells
- **Colors as data** — R can't read cell colors
- **Multiple tables on one sheet** — use `range` to isolate them
- **Formatted numbers** — "1,234" might read as text

::: {.callout-tip}
The best fix is usually to clean the Excel file first, or use `range` to read just the clean part.
:::

::: {.notes}
Students who have worked in labs may recognize these problems. The "colors as data" point usually gets a laugh — some labs really do use cell highlighting to code conditions. The practical takeaway: Excel is a display tool, not a data storage tool. If you receive an Excel file with these issues, the cleanest fix is often to open it in Excel, fix the formatting, and re-save before importing.
:::

## Listing sheet names

```{r}
#| eval: false
# See what sheets are in a file
excel_sheet_names("survey_results.xlsx")
```

Useful when you don't know the file structure yet.

# Other file formats {background-color="#2c3e50"}

## SPSS and SAS files

Psychology labs often have data in SPSS (`.sav`) or SAS formats:

```{r}
#| eval: false
library(haven)

# Read SPSS
spss_data <- read_sav("study.sav")

# Read SAS
sas_data <- read_sas("study.sas")
```

These return tibbles, just like `read_csv()`.

::: {.notes}
Many students will encounter SPSS files in their research labs. The key message is: you don't need SPSS installed to read .sav files. This is a selling point for R — it can read anything. Move through this slide quickly (~1 minute) unless students have specific questions.
:::

## A note about SPSS labels

SPSS files often have **variable labels** and **value labels**:

```{r}
#| eval: false
# Labels are stored as attributes
attr(spss_data$condition, "label")   # "Treatment Condition"
attr(spss_data$condition, "levels")  # 1="Control", 2="Treatment"

# To drop labels and work with plain values:
library(haven)
spss_data <- zap_labels(spss_data)
```

::: {.notes}
This is a "just so you know" slide. Students don't need to master haven labels right now, but they should know that SPSS files carry metadata that can cause unexpected behavior (e.g., a column that looks numeric but is actually a labelled vector). If they hit weird errors after importing a .sav file, zap_labels() is the quick fix. Move on fairly quickly.
:::

## Getting data out of Qualtrics

The most common workflow in psychology:

1. In Qualtrics: **Download** → **CSV** (or use the API — advanced)
2. The export includes metadata rows at the top — skip them:

```{r}
#| eval: false
# Qualtrics CSVs typically have 3 header rows
qualtrics_data <- read_csv("my_survey_export.csv", skip = 2)
```

3. Clean up the automatic column names (they're ugly)
4. Handle the timestamp columns if you need them

::: {.callout-note}
The first two rows after the header contain Qualtrics' own descriptions. `skip = 2` gets past them. Always check with `glimpse()` after importing.
:::

::: {.notes}
This will be the most practically relevant slide for many students. Qualtrics exports are notoriously messy — the first two rows after the header are Qualtrics metadata (question text, import IDs), not data. If you don't skip them, your first two "participants" will be nonsense strings. If time permits, show a real Qualtrics export on screen so students can see the three-row header structure. Spend ~3 minutes here.
:::

# Putting it together {background-color="#2c3e50"}

## A realistic import workflow

```{r}
#| eval: false
# 1. Read the file
raw_data <- read_csv(here("data/qualtrics_export.csv"),
  skip = 2,
  na = c("", "N/A", "-999")
)

# 2. Check what you got
glimpse(raw_data)
problems(raw_data)

# 3. Rename columns
clean_data <- raw_data |>
  rename(
    id = ResponseId,
    age = Q1,
    condition = Q2
  )

# 4. Fix types if needed
clean_data <- clean_data |>
  mutate(
    id = as.character(id),
    age = as.numeric(age)
  )

# 5. Save the cleaned version
write_csv(clean_data, here("data/clean_survey.csv"))
```

::: {.notes}
Walk through this as a complete workflow — this is the template students should internalize. Emphasize the order: read, check, rename, fix types, save. Students often want to skip step 2 (checking), but that's where you catch problems before they propagate. This workflow is essentially what Assignment 3 will ask them to do end to end. Spend ~3 minutes walking through each step.
:::

## Get a head start: Assignment 3

Try this mini-challenge before next class:

```{r}
#| eval: false
# 1. Import the messy CSV from the pair exercise
messy <- read_csv(here("data/messy_survey.csv"), na = c("", "NA", "-999", "N/A"))

# 2. Check for problems
problems(messy)

# 3. Fix at least one column type issue with col_types or
#    readr::parse_number()

# 4. Save your cleaned version
write_csv(clean, here("data/clean_survey.csv"))
```

Assignment 3 will have you do this end to end with real messy data.

::: {.notes}
This is individual work for the remaining class time (~10 minutes). Students who finished the pair exercise quickly can extend their earlier code. If time is short, frame this as homework: "Start this now and finish before next class." Mention that Assignment 3 builds directly on this workflow, so the practice now will pay off.
:::

## Solution

```{r}
#| echo: false
#| eval: false
# Solution to end-of-deck exercise
messy <- read_csv(here("data/messy_survey.csv"),
                  na = c("", "NA", "-999", "N/A"),
                  col_types = cols(
                    id = col_character(),
                    age = col_double(),
                    score = col_double()
                  ))

problems(messy)  # Check what readr flagged

clean <- messy |>
  mutate(age = parse_number(age))

write_csv(clean, here("data/clean_survey.csv"))
```

# Wrapping up {background-color="#2c3e50"}

## The import toolkit

| Function | What it does |
|----------|--------------|
| `read_csv()` | Read CSV files |
| `write_csv()` | Save CSV files |
| `read_excel()` | Read Excel files |
| `read_sav()` | Read SPSS files |
| `problems()` | Check for import issues |
| `col_types` | Control column types |
| `here()` | Build portable file paths |

## Before next class

📖 **Read:**

- [R4DS Ch 28: Quarto](https://r4ds.hadley.nz/quarto)

✅ **Practice:**

- Try importing a CSV or Excel file you have lying around
- Check `problems()` — does it flag anything?
- Practice renaming messy column names

## Key takeaways

1. **`read_csv()` is your default** — it's faster and smarter than `read.csv()`
2. **Use `here()` for file paths** — `here("data/file.csv")` works on any machine
3. **Check your imports** — `glimpse()` and `problems()` are your friends
4. **Name your missing values** — psychology data loves `-999`
5. **Excel is tricky** — use `readxl` and be suspicious of formatting
6. **Save cleaned files** — don't re-clean every time

::: {.notes}
If you need to cut anything from this deck for time, the col_types shortcuts slide and the separate Excel "listing sheet names" slide are the most skippable. The core message students should leave with: always check your imports, and handle missing values at import time.
:::

## The one thing to remember

Your analysis is only as good as your import. Check it before you trust it.

Next time: Quarto

::: {.notes}
Remind students of the reading for next class (R4DS Ch 9: Layers). Next session moves back to ggplot2 with a deeper dive into layers and aesthetics. Preview: "Now that you can import and reshape your own data, we'll learn how to build more sophisticated visualizations with it."
:::
