---
title: "Layers & Aesthetics"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 20, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 7"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)

# Create a simulated psychology dataset we'll use throughout
set.seed(123)
reaction_data <- tibble(
  participant = rep(1:40, each = 2),
  condition = rep(c("Control", "Treatment"), 40),
  rt = c(rnorm(40, mean = 520, sd = 60), rnorm(40, mean = 480, sd = 55)),
  accuracy = c(rbinom(40, 1, 0.82), rbinom(40, 1, 0.88)),
  age_group = rep(c("Young", "Young", "Older", "Older"), 20)
) |>
  mutate(rt = round(rt, 1))
```

# Choosing the right geom {background-color="#2c3e50"}

## Back to the grammar of graphics

We covered the basics in Session 2. Now we go deeper: **matching plot type to data type.**

The wrong geom doesn't just look bad — it can actively mislead.

## Geom selection guide

| Your data | Good geom | Avoid |
|-----------|-----------|-------|
| One continuous variable | `geom_histogram()`, `geom_density()` | `geom_bar()` |
| One categorical variable | `geom_bar()` | `geom_histogram()` |
| Two continuous | `geom_point()`, `geom_smooth()` | — |
| One continuous + one categorical | `geom_boxplot()`, `geom_violin()` | pie charts |
| Two categorical | `geom_count()`, `geom_tile()` | — |
| Change over time | `geom_line()` | `geom_point()` alone |

## One variable: distributions

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
reaction_data |>
  ggplot(aes(x = rt)) +
  geom_histogram(binwidth = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Reaction Times",
    x = "Reaction time (ms)"
  ) +
  theme_minimal(base_size = 14)
```

## histogram vs density

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
library(patchwork)

p1 <- reaction_data |>
  ggplot(aes(x = rt)) +
  geom_histogram(binwidth = 30, fill = "steelblue", color = "white") +
  labs(title = "Histogram", x = "RT (ms)", y = "Count") +
  theme_minimal(base_size = 14)

p2 <- reaction_data |>
  ggplot(aes(x = rt)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Density plot", x = "RT (ms)", y = "Density") +
  theme_minimal(base_size = 14)

p1 + p2
```

Histogram = actual counts. Density = smoothed estimate of the shape.

## One categorical variable

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
reaction_data |>
  ggplot(aes(x = condition)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Observations per Condition",
    x = "Condition",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## One continuous + one categorical

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Reaction Time by Condition",
    x = "Condition",
    y = "Reaction time (ms)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## boxplot vs violin

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5

p1 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Boxplot", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p2 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_violin(alpha = 0.5) +
  labs(title = "Violin plot", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p1 + p2
```

Violin shows the full distribution shape. Boxplot shows summary stats. Both have their place.

# stat_summary() {background-color="#2c3e50"}

## The psychology staple: means + error bars

In psych papers, you'll see this constantly: a bar chart showing group means with error bars. `stat_summary()` is how you make it.

## stat_summary() basics

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
reaction_data |>
  ggplot(aes(x = condition, y = rt)) +
  stat_summary(fun = mean, geom = "bar", fill = "steelblue", width = 0.5) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3) +
  labs(
    title = "Mean Reaction Time by Condition",
    x = "Condition",
    y = "Reaction time (ms)"
  ) +
  theme_minimal(base_size = 14)
```

## Breaking down stat_summary()

```{r}
#| eval: false
# The bar (mean)
stat_summary(fun = mean, geom = "bar")

# The error bars (standard error)
stat_summary(fun.data = mean_se, geom = "errorbar")
```

- `fun` = the function to calculate (mean, median, etc.)
- `fun.data` = a function that returns ymin, y, ymax (like `mean_se`)
- `geom` = what shape to use to show the result

## What are error bars showing?

This matters! Always state it in your figure caption.

| Error bar type | What it means | How to get it |
|---------------|---------------|---------------|
| SE (standard error) | Precision of the mean | `mean_se` |
| SD (standard deviation) | Spread of the data | Write your own |
| 95% CI | Confidence interval | `mean_cl_normal` |

```{r}
#| eval: false
# SE (most common in psych)
stat_summary(fun.data = mean_se, geom = "errorbar")

# 95% CI
stat_summary(fun.data = mean_cl_normal, geom = "errorbar")
```

## Adding individual data points

Error bars alone hide the data. Show the points too:

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "black") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, color = "black") +
  labs(
    title = "Reaction Time by Condition",
    subtitle = "Error bars = SE. Individual points shown.",
    x = "Condition",
    y = "Reaction time (ms)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

Using the `reaction_data` dataset (already in your environment):

1. Create a **bar chart with error bars** showing mean RT by condition
2. Add **individual data points** (jittered) behind the bars
3. Color the bars by condition
4. Add a caption noting that error bars show SE

::: {.callout-tip}
You'll need `stat_summary()` twice — once for the bar, once for the error bars. Look at the examples from the last few slides.
:::

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_jitter(width = 0.15, alpha = 0.4) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.6, width = 0.5) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  labs(
    title = "Mean Reaction Time by Condition",
    x = "Condition",
    y = "Reaction time (ms)",
    caption = "Error bars represent standard error of the mean (SE)."
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

---

## Before we move on

📤 **Upload your code to Canvas** for participation credit. Paste what you have into today's in-class submission — it doesn't need to work perfectly.

# Position & scales {background-color="#2c3e50"}

## Position adjustments

When geoms overlap, position adjustments fix it:

| Position | What it does | When to use |
|----------|--------------|-------------|
| `"dodge"` | Side by side | Grouped bar charts |
| `"stack"` | Stacked on top | Stacked bars |
| `"fill"` | Stacked to 100% | Comparing proportions |
| `"jitter"` | Random wiggle | Overplotted points |

## dodge: grouped bar charts

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
reaction_data |>
  ggplot(aes(x = age_group, y = rt, fill = condition)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge", width = 0.6) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               position = position_dodge(0.6), width = 0.2) +
  labs(
    title = "Reaction Time by Age Group and Condition",
    x = "Age group",
    y = "Mean RT (ms)"
  ) +
  theme_minimal(base_size = 14)
```

## stack and fill

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5

p1 <- reaction_data |>
  ggplot(aes(x = age_group, fill = condition)) +
  geom_bar(position = "stack") +
  labs(title = "Stacked (counts)", x = "Age group", y = "Count") +
  theme_minimal(base_size = 14)

p2 <- reaction_data |>
  ggplot(aes(x = age_group, fill = condition)) +
  geom_bar(position = "fill") +
  labs(title = "Filled (proportions)", x = "Age group", y = "Proportion") +
  theme_minimal(base_size = 14)

p1 + p2
```

## Scales: controlling axes and colors

Scales translate data values into visual properties:

```{r}
#| eval: false
# Control axis range
scale_y_continuous(limits = c(0, 800))

# Control axis breaks (tick marks)
scale_x_continuous(breaks = seq(400, 700, by = 50))

# Set specific colors
scale_fill_manual(values = c("Control" = "lightblue", "Treatment" = "coral"))

# Use a colorblind-friendly palette
scale_fill_viridis_d()
```

## scale_fill_manual()

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Control" = "#3498db", "Treatment" = "#e74c3c")) +
  labs(title = "Custom colors", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## Coordinate systems

```{r}
#| eval: false
# Flip x and y (great for long category labels)
coord_flip()

# Zoom in without dropping data
coord_cartesian(ylim = c(400, 600))
# vs
scale_y_continuous(limits = c(400, 600))  # This DROPS data outside the range!
```

## coord_flip() in action

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
reaction_data |>
  ggplot(aes(y = condition, x = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Flipped axes — better for category labels", y = "Condition", x = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

# Putting it together {background-color="#2c3e50"}

## A polished psychology figure

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 6
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition, color = condition)) +
  geom_jitter(width = 0.2, alpha = 0.35, size = 2) +
  geom_boxplot(alpha = 0.5, width = 0.4, outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "black") +
  scale_fill_manual(values = c("Control" = "#3498db", "Treatment" = "#e74c3c")) +
  scale_color_manual(values = c("Control" = "#2980b9", "Treatment" = "#c0392b")) +
  labs(
    title = "Treatment Reduces Reaction Time",
    subtitle = "Individual data points, box summaries, and mean (diamond) shown",
    x = "Condition",
    y = "Reaction time (ms)",
    caption = "N = 40 per condition. Diamond = mean."
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## Get a head start: Assignment 4

Assignment 4 will ask you to create "bad" and "good" versions of figures. Start experimenting:

1. Take any plot from today and make it **deliberately bad** — wrong geom, missing labels, confusing colors
2. Then fix it. What did you change and why?
3. Try creating the same data with three different geoms. Which one communicates best?

# Wrapping up {background-color="#2c3e50"}

## Today's toolkit

| Tool | What it does |
|------|--------------|
| `geom_histogram()` | Distribution of one continuous variable |
| `geom_density()` | Smooth distribution estimate |
| `geom_boxplot()` | Summary of continuous by categorical |
| `geom_violin()` | Full distribution by categorical |
| `stat_summary()` | Calculate and display summary stats |
| `position = "dodge"` | Side-by-side grouped plots |
| `scale_fill_manual()` | Custom colors |
| `coord_flip()` | Swap axes |

## Before next class

📖 **Read:**

- Supplementary: Visual perception principles (will be posted)
- Optional: Knaflic, *Storytelling with Data*, Ch 1–3

✅ **Practice:**

- Create a bar chart with error bars for a dataset of your choice
- Try boxplot vs violin on the same data
- Experiment with `coord_flip()` and `scale_fill_manual()`

## Key takeaways

1. **Match your geom to your data** — the wrong choice misleads
2. **stat_summary() is powerful** — means, error bars, custom functions
3. **Position adjustments** handle overlap (dodge, stack, fill, jitter)
4. **Scales control** axes, colors, and legends
5. **coord_cartesian() zooms; scale limits drop data** — know the difference

## Questions?

Next time: **Perception & Design**

We'll learn why some visualizations work and others don't — and how to make yours better!
