---
title: "Layers & Aesthetics"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 20, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 7"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)

# Create a simulated psychology dataset we'll use throughout
set.seed(123)
reaction_data <- tibble(
  participant = rep(1:40, each = 2),
  condition = rep(c("Control", "Treatment"), 40),
  rt = c(rnorm(40, mean = 520, sd = 60), rnorm(40, mean = 480, sd = 55)),
  accuracy = c(rbinom(40, 1, 0.82), rbinom(40, 1, 0.88)),
  age_group = rep(c("Young", "Young", "Older", "Older"), 20)
) |>
  mutate(rt = round(rt, 1))
```

# Choosing the right geom {background-color="#2c3e50"}

::: {.notes}
This first section (~25 min total) covers geom selection, distributions, and stat_summary(). The opening comparison is the hook — let it land before moving on.
:::

## Same data, two stories

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.5
#| output-location: slide
#| fig-alt: "Side-by-side comparison of two plots using the same reaction time data. Version A is a plain bar chart of means that hides all variation. Version B overlays jittered individual data points, boxplots, and mean diamonds, revealing the full spread and overlap between Control and Treatment groups."
library(patchwork)

p1 <- ggplot(reaction_data, aes(x = condition, y = rt)) +
  geom_bar(stat = "summary", fun = mean, fill = "grey70") +
  labs(title = "Version A", y = "Mean RT (ms)", x = NULL) +
  theme_minimal(base_size = 14)

p2 <- ggplot(reaction_data, aes(x = condition, y = rt, color = condition)) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 2) +
  geom_boxplot(alpha = 0.3, width = 0.4, outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "black") +
  labs(title = "Version B", y = "Mean RT (ms)", x = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p1 + p2
```

Both use the same data. The difference is **layers** — and the right layers tell the right story.

The wrong geom doesn't just look bad — it can actively mislead.

::: {.notes}
Spend ~2 minutes here. Ask students which version they'd trust more and why. The bar chart hides everything — the spread, outliers, individual data points. Version B tells the full story. This contrast motivates the entire session: choosing the right geom is a design decision, not just a coding one.
:::

## Geom selection guide

| Your data | Good geom | Avoid |
|-----------|-----------|-------|
| One continuous variable | `geom_histogram()`, `geom_density()` | `geom_bar()` |
| One categorical variable | `geom_bar()` | `geom_histogram()` |
| Two continuous | `geom_point()`, `geom_smooth()` | — |
| One continuous + one categorical | `geom_boxplot()`, `geom_violin()` | pie charts |
| Two categorical | `geom_count()`, `geom_tile()` | — |
| Change over time | `geom_line()` | `geom_point()` alone |

::: {.notes}
This table is a reference slide students will come back to. Don't read every row aloud — highlight the "Avoid" column. The most common mistake is using `geom_bar()` for continuous data. Also note that pie charts are absent from the "Good geom" column entirely. Students will ask about pie charts; the short answer is that position along a common axis (bar chart) is easier for the human eye to compare than angle (pie chart).
:::

## One variable: distributions

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Histogram of reaction times in milliseconds with 30ms bins, showing an approximately normal distribution centered around 500ms with a slight right skew."
reaction_data |>
  ggplot(aes(x = rt)) +
  geom_histogram(
    binwidth = 30, 
    fill = "steelblue", 
    color = "white") +
  labs(
    title = "Distribution of Reaction Times",
    x = "Reaction time (ms)"
  ) +
  theme_minimal(base_size = 14)
```

::: {.notes}
Walk through the code line by line. Emphasize `binwidth` — students will forget it and get the default "30 bins" warning. A good rule of thumb: try a few binwidths and see what reveals the shape without being too choppy or too smooth. Students often ask what number to use; there's no single right answer, but for reaction times in the hundreds, 20-30 ms bins are reasonable.
:::

## histogram vs density

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| fig-alt: "Side-by-side comparison of a histogram and density plot of reaction times. The histogram shows discrete count bins while the density plot shows a smooth continuous curve, both revealing a similar roughly normal distribution."
library(patchwork)

p1 <- reaction_data |>
  ggplot(aes(x = rt)) +
  geom_histogram(binwidth = 30, fill = "steelblue", color = "white") +
  labs(title = "Histogram", x = "RT (ms)", y = "Count") +
  theme_minimal(base_size = 14)

p2 <- reaction_data |>
  ggplot(aes(x = rt)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Density plot", x = "RT (ms)", y = "Density") +
  theme_minimal(base_size = 14)

p1 + p2
```

Histogram = actual counts. Density = smoothed estimate of the shape.

::: {.notes}
Quick comparison — don't belabor this. The key point is that density plots are better for comparing distributions across groups (you can overlay them), while histograms are better for seeing exact counts. Students sometimes get confused by the y-axis on density plots; reassure them that the area under the curve sums to 1, but they rarely need to interpret the y-axis value itself.
:::

## One categorical variable

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Bar chart showing the count of observations in each experimental condition, with Control and Treatment groups having equal counts of 40 each."
reaction_data |>
  ggplot(aes(x = condition)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Observations per Condition",
    x = "Condition",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## One continuous + one categorical

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Boxplots comparing reaction time distributions for Control and Treatment conditions. The Control group has a slightly higher median and wider spread than the Treatment group."
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Reaction Time by Condition",
    x = "Condition",
    y = "Reaction time (ms)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## boxplot vs violin

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| fig-alt: "Side-by-side comparison of boxplot and violin plot for reaction time by condition. The boxplot shows median, quartiles, and whiskers, while the violin plot reveals the full distribution shape including where data are most concentrated."

p1 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Boxplot", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p2 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_violin(alpha = 0.5) +
  labs(title = "Violin plot", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p1 + p2
```

Violin shows the full distribution shape. Boxplot shows summary stats. Both have their place.

::: {.notes}
Students may not have seen violin plots before. Point out that a violin is essentially a mirrored density plot — it shows the full distribution shape, which a boxplot hides. Bimodal distributions look obvious in a violin but invisible in a boxplot. Mention that we'll see an even better option (raincloud plots) in the next session.
:::

# stat_summary() {background-color="#2c3e50"}

## The psychology staple: means + error bars

In psych papers, you'll see this constantly: a bar chart showing group means with error bars. `stat_summary()` is how you make it.

::: {.notes}
Transition: "So far we've been plotting raw data. But in psychology, you'll almost always need to show summary statistics — means, error bars, confidence intervals. That's what stat_summary() is for." This is the most psych-relevant section of the deck. Spend ~8 minutes on stat_summary() total. Students will use this constantly in assignments and the final project.
:::

## stat_summary() basics

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Bar chart showing mean reaction time by condition with 95% confidence interval error bars. The Treatment group mean is slightly lower than the Control group, with overlapping error bars."
reaction_data |>
  ggplot(aes(x = condition, y = rt)) +
  stat_summary(
    fun = mean,
    geom = "bar",
    fill = "steelblue", width = 0.5) +
  stat_summary(
    fun.data = mean_cl_normal, 
    geom = "errorbar", 
    width = 0.3) +
  labs(
    title = "Mean Reaction Time by Condition",
    x = "Condition",
    y = "Reaction time (ms)"
  ) +
  theme_minimal(base_size = 14)
```

## Breaking down stat_summary()

```{r}
#| eval: false
# The bar (mean)
stat_summary(fun = mean, geom = "bar")

# The error bars (95% confidence interval)
stat_summary(fun.data = mean_cl_normal, geom = "errorbar")
```

- `fun` = the function to calculate (mean, median, etc.)
- `fun.data` = a function that returns ymin, y, ymax (like `mean_cl_normal`)
- `geom` = what shape to use to show the result

::: {.notes}
The `fun` vs `fun.data` distinction trips students up. Emphasize: `fun` returns one number (e.g., the mean), while `fun.data` returns three (ymin, y, ymax) — which is what you need for error bars. Students don't need to memorize this; they need to know the pattern: one stat_summary() for the bar, another for the error bars.
:::

## What are error bars showing?

This matters! Always state it in your figure caption.

| Error bar type | What it means | How to get it |
|---------------|---------------|---------------|
| SE (standard error) | Precision of the mean | `mean_se` |
| SD (standard deviation) | Spread of the data | Write your own |
| 95% CI | Confidence interval | `mean_cl_normal` |

```{r}
#| eval: false
# SE 
stat_summary(fun.data = mean_se, geom = "errorbar")

# 95% CI (most common in psych)
stat_summary(fun.data = mean_cl_normal, geom = "errorbar")
```

## Why does this matter?

The APA Publication Manual (7th ed.) recommends showing individual data points alongside summary statistics whenever possible.

Bar charts with error bars are ubiquitous in psychology — but they hide the *shape* of the data. Outliers, bimodality, and floor/ceiling effects all disappear behind a rectangle.

::: {.notes}
This is a great moment to connect to the replication crisis discussion from Session 1. One reason problematic results went undetected for so long is that bar-and-error-bar plots hid the underlying data. If reviewers had seen the individual data points, they might have noticed impossible patterns (like the LaCour data). Emphasize that the field is moving toward showing more data, not less.
:::

## Adding individual data points

Error bars alone hide the data. Show the points too:

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Dot plot showing individual reaction times as jittered points for each condition, with black dots marking the group means and error bars showing 95% confidence intervals. Individual variation is clearly visible behind the summary statistics."
reaction_data |>
  ggplot(aes(x = condition, y = rt, color = condition)) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "black") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "black") +
  labs(
    title = "Reaction Time by Condition",
    subtitle = "Error bars = 95%CI. Individual points shown.",
    x = "Condition",
    y = "Reaction time (ms)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

::: {.notes}
Walk through the layering order: jittered points go first (behind), then the summary point and error bars on top. Emphasize that `alpha = 0.4` makes individual points semi-transparent so they don't overwhelm the summary. This is the pattern students should use for most psych figures going forward: show the data AND the summary.
:::

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

::: {.notes}
Give 10 minutes. Circulate actively — the most common issues will be: (1) forgetting to use `stat_summary()` twice (once for bar, once for error bars), (2) putting `geom_jitter()` after the bar so the points are hidden behind it, and (3) not knowing where to put the caption in `labs()`. If pairs finish early, challenge them to add `age_group` as a facet or dodge variable. Reveal the solution after time is up.
:::

Using the `reaction_data` dataset:

1. Create a **bar chart with error bars** showing mean RT by condition
2. Add **individual data points** (jittered) behind the bars
3. Color the bars by condition
4. Add a caption noting that error bars show the 95% CI.

::: {.callout-tip}
You'll need `stat_summary()` twice — once for the bar, once for the error bars. Look at the examples from the last few slides.
:::

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_jitter(width = 0.15, alpha = 0.4) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.6, width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2) +
  labs(
    title = "Mean Reaction Time by Condition",
    x = "Condition",
    y = "Reaction time (ms)",
    caption = "Error bars represent 95% confidence intervals."
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

---

## Before we move on

📤 **Upload your code to Canvas** for participation credit. Paste what you have into today's in-class submission — it doesn't need to work perfectly.

# Position & scales {background-color="#2c3e50"}

::: {.notes}
This second lecture block (~20 min) covers position adjustments, scales, coordinate systems, and the polished figure. This is more about refinement than new geoms — students already know the building blocks, now they learn how to control them precisely.
:::

## Position adjustments

When geoms overlap, position adjustments fix it:

| Position | What it does | When to use |
|----------|--------------|-------------|
| `"dodge"` | Side by side | Grouped bar charts |
| `"stack"` | Stacked on top | Stacked bars |
| `"fill"` | Stacked to 100% | Comparing proportions |
| `"jitter"` | Random wiggle | Overplotted points |

## dodge: grouped bar charts

::: {.notes}
This is the trickiest position adjustment. The key gotcha: when you dodge both the bars and the error bars, they need to dodge by the same width. That's why we use `position_dodge(0.6)` for the error bars to match `width = 0.6` on the bars. If these don't match, the error bars float between the bars. Students will hit this bug — it's a good one to flag proactively.
:::

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Grouped bar chart showing mean reaction time by age group and condition with dodged bars and error bars. Control and Treatment bars are side by side within each age group, allowing direct comparison across both factors."
reaction_data |>
  ggplot(aes(x = age_group, y = rt, fill = condition)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge", width = 0.6) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar",
               position = position_dodge(0.6), width = 0.2) +
  labs(
    title = "Reaction Time by Age Group and Condition",
    x = "Age group",
    y = "Mean RT (ms)"
  ) +
  theme_minimal(base_size = 14)
```

## stack and fill

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| fig-alt: "Side-by-side comparison of stacked and filled bar charts. The stacked chart shows raw counts of Control and Treatment observations by age group, while the filled chart normalizes each bar to 100% to compare proportions directly."

p1 <- reaction_data |>
  ggplot(aes(x = age_group, fill = condition)) +
  geom_bar(position = "stack") +
  labs(title = "Stacked (counts)", x = "Age group", y = "Count") +
  theme_minimal(base_size = 14)

p2 <- reaction_data |>
  ggplot(aes(x = age_group, fill = condition)) +
  geom_bar(position = "fill") +
  labs(title = "Filled (proportions)", x = "Age group", y = "Proportion") +
  theme_minimal(base_size = 14)

p1 + p2
```

## Scales: controlling axes and colors

::: {.notes}
Spend ~3 minutes on this and the next slide together. Students don't need to memorize every scale function — they need to know the naming pattern: `scale_` + the aesthetic (`fill`, `color`, `x`, `y`) + `_` + the type (`manual`, `continuous`, `discrete`, `viridis`). Once they see the pattern, they can guess function names and use autocomplete.
:::

Scales translate data values into visual properties:

```{r}
#| eval: false
# Control axis range
scale_y_continuous(limits = c(0, 800))

# Control axis breaks (tick marks)
scale_x_continuous(breaks = seq(400, 700, by = 50))

# Set specific colors
scale_fill_manual(values = c("Control" = "lightblue", "Treatment" = "coral"))

# Use a colorblind-friendly palette
scale_fill_viridis_d()
```

## scale_fill_manual()

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Boxplots of reaction time by condition using custom colors: blue for Control and red for Treatment, demonstrating how scale_fill_manual() applies specific hex color codes to groups."
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Control" = "#3498db", "Treatment" = "#e74c3c")) +
  labs(title = "Custom colors", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

Use a website like [HTML color codes](https://htmlcolorcodes.com/color-picker/) to find the appropriate HEX code for your color. 

## Coordinate systems

::: {.notes}
The `coord_cartesian()` vs `scale_y_continuous(limits = ...)` distinction is critical and students will confuse them. The key: `coord_cartesian()` zooms the view (like pinching to zoom on a photo), while `scale_y_continuous(limits = ...)` filters the data first and then plots. This means boxplot whiskers, smooth lines, and error bars can all change when you use scale limits because outliers are removed before the calculation. Always use `coord_cartesian()` for zooming.
:::

```{r}
#| eval: false
# Flip x and y (great for long category labels)
coord_flip()

# Zoom in without dropping data
coord_cartesian(ylim = c(400, 600))
# vs
scale_y_continuous(limits = c(400, 600))  # This DROPS data outside the range!
```

## coord_flip() in action

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Horizontal boxplots of reaction time by condition, with categories on the y-axis and reaction time on the x-axis, demonstrating how flipped axes improve readability for categorical labels."
reaction_data |>
  ggplot(aes(y = condition, x = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Flipped axes — better for category labels", y = "Condition", x = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

# Putting it together {background-color="#2c3e50"}

## A polished psychology figure

::: {.notes}
This is the "bringing it all together" slide. Walk through the code slowly, pointing out each layer and why it's there: jittered points for raw data, boxplot for summary, stat_summary diamond for the mean, manual colors for accessibility, and informative labels. This is roughly the quality you'd expect in a journal submission. Students should aim for this level on Assignment 4 and the final project.
:::

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 6
#| fig-alt: "Polished figure combining jittered individual data points, boxplots, and diamond-shaped mean markers for reaction time by condition. Uses custom blue and red colors, informative title and subtitle, and a caption noting sample size and mean indicator."
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition, color = condition)) +
  geom_jitter(width = 0.2, alpha = 0.35, size = 2) +
  geom_boxplot(alpha = 0.5, width = 0.4, outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "black") +
  scale_fill_manual(values = c("Control" = "#3498db", "Treatment" = "#e74c3c")) +
  scale_color_manual(values = c("Control" = "#2980b9", "Treatment" = "#c0392b")) +
  labs(
    title = "Treatment Reduces Reaction Time",
    subtitle = "Individual data points, box summaries, and mean (diamond) shown",
    x = "Condition",
    y = "Reaction time (ms)",
    caption = "N = 40 per condition. Diamond = mean."
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

# Get a head start {background-color="#e67e22"}

## Assignment 4 preview

::: {.notes}
This is the head-start exercise (~10 min remaining). Students don't need to finish this in class — it's a bridge to Assignment 4. Encourage them to have fun making the "bad" version deliberately awful. If students seem finished early, suggest they try adding `facet_wrap(~age_group)` to any of today's plots as a preview of faceting.
:::

Assignment 4 will ask you to create "bad" and "good" versions of figures. Start experimenting:

1. Take any plot from today and make it **deliberately bad** — wrong geom, missing labels, confusing colors
2. Then fix it. What did you change and why?
3. Try creating the same data with three different geoms. Which one communicates best?

# Wrapping up {background-color="#2c3e50"}

## Today's toolkit

| Tool | What it does |
|------|--------------|
| `geom_histogram()` | Distribution of one continuous variable |
| `geom_density()` | Smooth distribution estimate |
| `geom_boxplot()` | Summary of continuous by categorical |
| `geom_violin()` | Full distribution by categorical |
| `stat_summary()` | Calculate and display summary stats |
| `position = "dodge"` | Side-by-side grouped plots |
| `scale_fill_manual()` | Custom colors |
| `coord_flip()` | Swap axes |

## Before next class

📖 **Read:**

- Supplementary: Visual perception principles (will be posted)
- Optional: Knaflic, *Storytelling with Data*, Ch 1–3

✅ **Practice:**

- Create a bar chart with error bars for a dataset of your choice
- Try boxplot vs violin on the same data
- Experiment with `coord_flip()` and `scale_fill_manual()`

## Key takeaways

1. **Match your geom to your data** — the wrong choice misleads
2. **stat_summary() is powerful** — means, error bars, custom functions
3. **Position adjustments** handle overlap (dodge, stack, fill, jitter)
4. **Scales control** axes, colors, and legends
5. **coord_cartesian() zooms; scale limits drop data** — know the difference

## The one thing to remember

The difference between a chart and a figure is intention — every layer should earn its place.

Next time: Perception & Design

::: {.notes}
End on this line and let it land. Next session builds directly on this one — we go from "which geom" to "which color, which theme, which design decisions." Students should have R4DS Ch 9 fresh in their minds and the optional perception reading done before Monday.
:::

