---
title: "Layers & Aesthetics"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 22, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 8"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
    include-in-header:
      text: |
        <link rel="stylesheet" href="toc-nav.css">
    include-after-body:
      text: |
        <script src="toc-nav.js"></script>
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(nycflights13)

# A cleaned, manageable sample of flights we'll use throughout the deck.
set.seed(123)
flights_sample <- flights |>
  filter(!is.na(dep_delay), dep_delay > -30, dep_delay < 180) |>
  mutate(
    time_of_day = if_else(hour < 12, "Morning", "Afternoon/Evening"),
    time_of_day = factor(time_of_day, levels = c("Morning", "Afternoon/Evening"))
  ) |>
  slice_sample(n = 500)
```

## Setup

```{r}
library(tidyverse)
library(nycflights13)
library(showtext)

showtext_auto()  # lets ggplot render Unicode symbols like ◀ ▶

set.seed(422)
some_flights <- flights |>
  filter(carrier %in% c("9E", "B6", "HA")) |>
  filter(!is.na(air_time)) |>
  group_by(carrier) |>
  slice_sample(n = 100)

min_time <- min(some_flights$air_time)
max_time <- max(some_flights$air_time)
```

::: {.notes}
This first section (~25 min total) opens with a layered progression — each slide adds one thing to the previous figure. Walk through them without narrating the code; the code is folded on each slide so students can expand it later if curious. Every function (`stat_summary`, `geom_jitter`, `annotate`, `scale_fill_manual`, etc.) is taught later in the deck. Right now, the point is the visual story: what each layer reveals and what a barer version was hiding.
:::

## Base figure

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "Bar chart of mean air_time by carrier. Three bars of different heights — 9E short, B6 medium, HA much taller."
some_flights |>
  ggplot(aes(x = carrier, y = air_time, fill = carrier)) +
  stat_summary(fun = "mean", geom = "bar")
```

## Remove legend and turn sideways

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "The same bar chart rotated so carriers appear on the left and air_time extends horizontally. Legend removed."
some_flights |>
  ggplot(aes(x = carrier, y = air_time, fill = carrier)) +
  stat_summary(fun = "mean", geom = "bar") +
  guides(fill = "none") +
  coord_flip()
```

## Boxplot

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "Horizontal boxplots of air_time for each of three carriers. 9E and B6 have tight boxes at the low end; HA's box sits far to the right."
some_flights |>
  ggplot(aes(x = carrier, y = air_time, fill = carrier)) +
  geom_boxplot() +
  guides(fill = "none") +
  coord_flip()
```

## Points

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "Scatter plot showing every flight on a single horizontal line per carrier — points overlap heavily because the carrier axis is categorical."
some_flights |>
  ggplot(aes(x = carrier, y = air_time)) +
  geom_point() +
  coord_flip()
```

## Jitter

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "Same plot, but each point now jittered vertically so overlapping observations are visible. Three horizontal clouds, one per carrier."
some_flights |>
  ggplot(aes(x = carrier, y = air_time)) +
  geom_jitter(width = .2) +
  coord_flip()
```

## Colored

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "The jittered dot plot with each point filled by carrier and outlined in black using shape 21."
some_flights |>
  ggplot(aes(x = carrier, y = air_time, fill = carrier)) +
  geom_jitter(width = .2, shape = 21, color = "black", stroke = .4) +
  coord_flip()
```

## Cleaner theme

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "Same plot with theme_minimal() applied — grey background replaced with white, gridlines subtler."
some_flights |>
  ggplot(aes(x = carrier, y = air_time, fill = carrier)) +
  geom_jitter(width = .2, shape = 21, color = "black", stroke = .4) +
  coord_flip() +
  theme_minimal()
```

## Better colors, remove legend

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "Same dot plot with a custom palette — teal for 9E, orange for B6, deep red-brown for HA. Legend removed."
some_flights |>
  ggplot(aes(x = carrier, y = air_time, fill = carrier)) +
  geom_jitter(width = .2, shape = 21, color = "black", stroke = .4) +
  scale_fill_manual(values = c("9E" = "#15616d", "B6" = "#ff7d00", "HA" = "#78290f")) +
  guides(fill = "none") +
  coord_flip() +
  theme_minimal()
```

## Clean up labels

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "Dot plot with a title and the horizontal-axis values suffixed with ' min'. Default x and y labels removed."
some_flights |>
  ggplot(aes(x = carrier, y = air_time, fill = carrier)) +
  geom_jitter(width = .2, shape = 21, color = "black", stroke = .4) +
  scale_y_continuous(labels = scales::label_number(suffix = " min")) +
  scale_fill_manual(values = c("9E" = "#15616d", "B6" = "#ff7d00", "HA" = "#78290f")) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL, title = "Air time by carrier") +
  coord_flip() +
  theme_minimal()
```

## Reference lines at min and max

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "Same dot plot with two vertical dashed reference lines marking the overall shortest and longest flights in the sample."
some_flights |>
  ggplot(aes(x = carrier, y = air_time, fill = carrier)) +
  geom_jitter(width = .2, shape = 21, color = "black", stroke = .4) +
  geom_hline(yintercept = min_time, linetype = "dashed") +
  geom_hline(yintercept = max_time, linetype = "dashed") +
  scale_y_continuous(labels = scales::label_number(suffix = " min")) +
  scale_fill_manual(values = c("9E" = "#15616d", "B6" = "#ff7d00", "HA" = "#78290f")) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL, title = "Air time by carrier") +
  coord_flip() +
  theme_minimal()
```

## Annotate the extremes

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "Same plot with two small arrows and labels at the top — one pointing left to the shortest reference line, one pointing right to the longest."
some_flights |>
  ggplot(aes(x = carrier, y = air_time, fill = carrier)) +
  geom_jitter(width = .2, shape = 21, color = "black", stroke = .4) +
  geom_hline(yintercept = min_time, linetype = "dashed", alpha = .4) +
  geom_hline(yintercept = max_time, linetype = "dashed", alpha = .4) +
  annotate("text", x = 3.5, y = min_time, hjust = 0,
           label = paste0("◀ shortest: ", min_time, " min"),
           color = "grey30", size = 4) +
  annotate("text", x = 3.5, y = max_time, hjust = 1,
           label = paste0("longest: ", max_time, " min ▶"),
           color = "grey30", size = 4) +
  scale_x_discrete(expand = expansion(add = c(0.6, 1))) +
  scale_y_continuous(labels = scales::label_number(suffix = " min")) +
  scale_fill_manual(values = c("9E" = "#15616d", "B6" = "#ff7d00", "HA" = "#78290f")) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL, title = "Air time by carrier") +
  coord_flip(clip = "off") +
  theme_minimal()
```

## Carrier labels inline

```{r}
#| code-fold: true
#| fig-height: 4.5
#| fig-width: 9
#| fig-alt: "Same plot with the carrier tick labels on the left removed and replaced by inline labels sitting just below each row of dots: '9E • Endeavor Air', 'B6 • JetBlue', 'HA • Hawaiian Airlines'."
carrier_labels <- some_flights |>
  group_by(carrier) |>
  summarize(mean_time = mean(air_time)) |>
  mutate(name = c("9E • Endeavor Air", "B6 • JetBlue", "HA • Hawaiian Airlines"))

some_flights |>
  ggplot(aes(x = carrier, y = air_time, fill = carrier)) +
  geom_jitter(width = .2, shape = 21, color = "black", stroke = .4) +
  geom_hline(yintercept = min_time, linetype = "dashed", alpha = .4) +
  geom_hline(yintercept = max_time, linetype = "dashed", alpha = .4) +
  annotate("text", x = 3.5, y = min_time, hjust = 0,
           label = paste0("◀ shortest: ", min_time, " min"),
           color = "grey30", size = 4) +
  annotate("text", x = 3.5, y = max_time, hjust = 1,
           label = paste0("longest: ", max_time, " min ▶"),
           color = "grey30", size = 4) +
  geom_text(data = carrier_labels,
            aes(x = carrier, y = mean_time, label = name),
            inherit.aes = FALSE,
            hjust = 0.5, nudge_x = -0.35,
            fontface = "bold", size = 4.2) +
  geom_text(data = carrier_labels,
            aes(x = carrier, y = mean_time,
                label = paste0("avg: ", round(mean_time), " min")),
            inherit.aes = FALSE,
            hjust = 0.5, nudge_x = -0.6,
            size = 3.5, color = "grey30") +
  scale_x_discrete(expand = expansion(add = c(0.8, 1)), labels = NULL) +
  scale_y_continuous(labels = scales::label_number(suffix = " min")) +
  scale_fill_manual(values = c("9E" = "#15616d", "B6" = "#ff7d00", "HA" = "#78290f")) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL, title = "Air time by carrier") +
  coord_flip(clip = "off") +
  theme_minimal()
```

# Choosing the right geom {background-color="#2c3e50"}


## Geom selection guide

| Your data | Good geom | Avoid |
|-----------|-----------|-------|
| One continuous variable | `geom_histogram()`, `geom_density()` | `geom_bar()` |
| One categorical variable | `geom_bar()` | `geom_histogram()` |
| Two continuous | `geom_point()`, `geom_smooth()` | — |
| One continuous + one categorical | `geom_boxplot()`, `geom_violin()` | pie charts |
| Two categorical | `geom_count()`, `geom_tile()` | — |
| Change over time | `geom_line()` | `geom_point()` alone |

::: {.notes}
This table is a reference slide students will come back to. Don't read every row aloud — highlight the "Avoid" column. The most common mistake is using `geom_bar()` for continuous data. Also note that pie charts are absent from the "Good geom" column entirely. Students will ask about pie charts; the short answer is that position along a common axis (bar chart) is easier for the human eye to compare than angle (pie chart).
:::

## One variable: distributions

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Histogram of departure delays in minutes with 10-minute bins, showing a right-skewed distribution: most flights depart close to on time, with a long tail of delayed flights."
flights_sample |>
  ggplot(aes(x = dep_delay)) +
  geom_histogram(
    binwidth = 10,
    fill = "steelblue",
    color = "white") +
  labs(
    title = "Distribution of Departure Delays",
    x = "Departure delay (min)"
  ) +
  theme_minimal(base_size = 14)
```

::: {.notes}
Walk through the code line by line. Emphasize `binwidth` — students will forget it and get the default "30 bins" warning. A good rule of thumb: try a few binwidths and see what reveals the shape without being too choppy or too smooth. The right skew is a great teaching moment — real-world delay data almost always looks like this. Connect it back to psychology: reaction time distributions look similar.
:::

## histogram vs density

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| fig-alt: "Side-by-side comparison of a histogram and density plot of departure delays. The histogram shows discrete count bins while the density plot shows a smooth continuous curve, both revealing the same right-skewed shape."
library(patchwork)

p1 <- flights_sample |>
  ggplot(aes(x = dep_delay)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(title = "Histogram", x = "Departure delay (min)", y = "Count") +
  theme_minimal(base_size = 14)

p2 <- flights_sample |>
  ggplot(aes(x = dep_delay)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Density plot", x = "Departure delay (min)", y = "Density") +
  theme_minimal(base_size = 14)

p1 + p2
```

Histogram = actual counts. Density = smoothed estimate of the shape.

::: {.notes}
Quick comparison — don't belabor this. The key point is that density plots are better for comparing distributions across groups (you can overlay them), while histograms are better for seeing exact counts. Students sometimes get confused by the y-axis on density plots; reassure them that the area under the curve sums to 1, but they rarely need to interpret the y-axis value itself.
:::

## One categorical variable

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Bar chart showing the count of sampled flights by NYC airport of origin. EWR, JFK, and LGA each have a few hundred flights in the sample."
flights_sample |>
  ggplot(aes(x = origin)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Flights per NYC Airport",
    x = "Origin airport",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## One continuous + one categorical

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Boxplots comparing departure delay distributions across the three NYC airports. Medians are close to zero at all three, with long upper whiskers reflecting the right-skewed tail."
flights_sample |>
  ggplot(aes(x = origin, y = dep_delay, fill = origin)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Departure Delay by Airport",
    x = "Origin airport",
    y = "Departure delay (min)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## boxplot vs violin

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| fig-alt: "Side-by-side comparison of boxplot and violin plot for departure delay by airport. The boxplot shows median, quartiles, and whiskers; the violin reveals the full right-skewed shape, with most density concentrated near zero and a long tail."

p1 <- flights_sample |>
  ggplot(aes(x = origin, y = dep_delay, fill = origin)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Boxplot", x = "Origin", y = "Delay (min)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p2 <- flights_sample |>
  ggplot(aes(x = origin, y = dep_delay, fill = origin)) +
  geom_violin(alpha = 0.5) +
  labs(title = "Violin plot", x = "Origin", y = "Delay (min)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p1 + p2
```

Violin shows the full distribution shape. Boxplot shows summary stats. Both have their place.

::: {.notes}
Students may not have seen violin plots before. Point out that a violin is essentially a mirrored density plot — it shows the full distribution shape, which a boxplot hides. The right-skewed tail of delays is much more obvious in the violin than the boxplot. Mention that we'll see an even better option (raincloud plots) in the next session.
:::

## Showing individual points: `geom_jitter()`

When your x-axis is categorical, `geom_point()` would stack every observation on the same vertical line. `geom_jitter()` adds a small random horizontal wiggle so each point is visible.

```{r}
#| output-location: slide
#| fig-width: 9
#| fig-height: 5
#| fig-alt: "Jittered dot plot of departure delays by NYC airport. Each point is a single flight, with horizontal jitter so overlapping points are visible. All three airports show a dense cluster near zero and a sparse tail of long delays."
flights_sample |>
  ggplot(aes(x = origin, y = dep_delay, color = origin)) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  labs(
    title = "Every Flight, Plotted",
    x = "Origin airport",
    y = "Departure delay (min)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

::: {.notes}
`width = 0.2` keeps points within each category's "lane." `alpha = 0.4` makes them semi-transparent so dense areas look darker. Flag both settings — students will want to tune them. We'll see `geom_jitter()` come back as the "raw data" layer under every summary plot in the rest of the deck.
:::

# stat_summary() {background-color="#2c3e50"}

## The psychology staple: means + error bars

In psych papers, you'll see this constantly: a bar chart showing group means with error bars. `stat_summary()` is how you make it.

::: {.notes}
Transition: "So far we've been plotting raw data. But in psychology, you'll almost always need to show summary statistics — means, error bars, confidence intervals. That's what stat_summary() is for." This is the most psych-relevant section of the deck. Spend ~8 minutes on stat_summary() total. Students will use this constantly in assignments and the final project.
:::

## stat_summary() basics

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Bar chart showing mean departure delay by airport with 95% confidence interval error bars. Means are close to each other across the three airports, with overlapping error bars."
flights_sample |>
  ggplot(aes(x = origin, y = dep_delay)) +
  stat_summary(
    fun = mean,
    geom = "bar",
    fill = "steelblue", width = 0.5) +
  stat_summary(
    fun.data = mean_cl_normal,
    geom = "errorbar",
    width = 0.3) +
  labs(
    title = "Mean Departure Delay by Airport",
    x = "Origin airport",
    y = "Departure delay (min)"
  ) +
  theme_minimal(base_size = 14)
```

## Breaking down stat_summary()

```{r}
#| eval: false
# The bar (mean)
stat_summary(fun = mean, geom = "bar")

# The error bars (95% confidence interval)
stat_summary(fun.data = mean_cl_normal, geom = "errorbar")
```

- `fun` = the function to calculate (mean, median, etc.)
- `fun.data` = a function that returns ymin, y, ymax (like `mean_cl_normal`)
- `geom` = what shape to use to show the result

::: {.notes}
The `fun` vs `fun.data` distinction trips students up. Emphasize: `fun` returns one number (e.g., the mean), while `fun.data` returns three (ymin, y, ymax) — which is what you need for error bars. Students don't need to memorize this; they need to know the pattern: one stat_summary() for the bar, another for the error bars.
:::

## What are error bars showing?

This matters! Always state it in your figure caption.

| Error bar type | What it means | How to get it |
|---------------|---------------|---------------|
| SE (standard error) | Precision of the mean | `mean_se` |
| SD (standard deviation) | Spread of the data | Write your own |
| 95% CI | Confidence interval | `mean_cl_normal` |

```{r}
#| eval: false
# SE
stat_summary(fun.data = mean_se, geom = "errorbar")

# 95% CI (most common in psych)
stat_summary(fun.data = mean_cl_normal, geom = "errorbar")
```

## Why does this matter?

The APA Publication Manual (7th ed.) recommends showing individual data points alongside summary statistics whenever possible.

Bar charts with error bars are ubiquitous in psychology — but they hide the *shape* of the data. Outliers, skew, and bimodality all disappear behind a rectangle.

::: {.notes}
This is a great moment to connect to the replication crisis discussion from Session 1. One reason problematic results went undetected for so long is that bar-and-error-bar plots hid the underlying data. If reviewers had seen the individual data points, they might have noticed impossible patterns (like the LaCour data). The delay data makes the same point — the means look nearly identical, but the underlying distributions are wildly right-skewed.
:::

## Adding individual data points

Error bars alone hide the data. Show the points too:

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Dot plot showing individual flight delays as jittered points for each airport, with black dots marking airport means and error bars showing 95% confidence intervals. Individual variation is clearly visible behind the summary statistics."
flights_sample |>
  ggplot(aes(x = origin, y = dep_delay, color = origin)) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "black") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "black") +
  labs(
    title = "Departure Delay by Airport",
    subtitle = "Error bars = 95% CI. Individual flights shown.",
    x = "Origin airport",
    y = "Departure delay (min)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

::: {.notes}
Walk through the layering order: jittered points go first (behind), then the summary point and error bars on top. Emphasize that `alpha = 0.4` makes individual points semi-transparent so they don't overwhelm the summary. This is the pattern students should use for most psych figures going forward: show the data AND the summary.
:::

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

::: {.notes}
Give 10 minutes. Circulate actively — the most common issues will be: (1) forgetting to use `stat_summary()` twice (once for bar, once for error bars), (2) putting `geom_jitter()` after the bar so the points are hidden behind it, and (3) not knowing where to put the caption in `labs()`. If pairs finish early, challenge them to add `time_of_day` as a facet or fill variable. Reveal the solution after time is up. Students should create their own `flights_sample` using the setup code shown on this slide (or you can share it in the chat).
:::

Using the `flights` dataset (filter to `!is.na(dep_delay)` first):

1. Create a **bar chart with error bars** showing mean `dep_delay` by `origin`
2. Add **individual data points** (jittered) behind the bars
3. Color the bars by `origin`
4. Add a caption noting that error bars show the 95% CI

::: {.callout-tip}
You'll need `stat_summary()` twice — once for the bar, once for the error bars. Look at the examples from the last few slides.
:::

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
flights |>
  filter(!is.na(dep_delay)) |>
  ggplot(aes(x = origin, y = dep_delay, fill = origin)) +
  geom_jitter(width = 0.15, alpha = 0.2) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.6, width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2) +
  labs(
    title = "Mean Departure Delay by NYC Airport",
    x = "Origin airport",
    y = "Departure delay (min)",
    caption = "Error bars represent 95% confidence intervals."
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## Solution: Mean departure delay by airport {background-color="#e67e22"}

```{r}
#| echo: true
#| eval: false
flights |>
  filter(!is.na(dep_delay)) |>
  ggplot(aes(x = origin, y = dep_delay, fill = origin)) +
  geom_jitter(width = 0.15, alpha = 0.2) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.6, width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2) +
  labs(
    title = "Mean Departure Delay by NYC Airport",
    x = "Origin airport",
    y = "Departure delay (min)",
    caption = "Error bars represent 95% confidence intervals."
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## Solution: output {background-color="#e67e22"}

```{r}
#| echo: false
#| eval: true
#| fig-width: 9
#| fig-height: 5
#| fig-alt: "Bar chart of mean departure delay by NYC airport (EWR, JFK, LGA) with 95% confidence interval error bars and jittered individual flight points behind the bars. The three airports have similar means with overlapping confidence intervals."
flights |>
  filter(!is.na(dep_delay)) |>
  ggplot(aes(x = origin, y = dep_delay, fill = origin)) +
  geom_jitter(width = 0.15, alpha = 0.2) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.6, width = 0.5) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2) +
  labs(
    title = "Mean Departure Delay by NYC Airport",
    x = "Origin airport",
    y = "Departure delay (min)",
    caption = "Error bars represent 95% confidence intervals."
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

# Position & scales {background-color="#2c3e50"}

::: {.notes}
This second lecture block (~20 min) covers position adjustments, scales, coordinate systems, and the polished figure. This is more about refinement than new geoms — students already know the building blocks, now they learn how to control them precisely.
:::

## Position adjustments

When geoms overlap, position adjustments fix it:

| Position | What it does | When to use |
|----------|--------------|-------------|
| `"dodge"` | Side by side | Grouped bar charts |
| `"stack"` | Stacked on top | Stacked bars |
| `"fill"` | Stacked to 100% | Comparing proportions |
| `"jitter"` | Random wiggle | Overplotted points |

## dodge: grouped bar charts

::: {.notes}
This is the trickiest position adjustment. The key gotcha: when you dodge both the bars and the error bars, they need to dodge by the same width. That's why we use `position_dodge(0.6)` for the error bars to match `width = 0.6` on the bars. If these don't match, the error bars float between the bars. Students will hit this bug — it's a good one to flag proactively.
:::

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Grouped bar chart showing mean departure delay by airport and time of day with dodged bars and error bars. Morning and Afternoon/Evening bars sit side by side within each airport, showing afternoon delays are consistently higher."
flights_sample |>
  ggplot(aes(x = origin, y = dep_delay, fill = time_of_day)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge", width = 0.6) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar",
               position = position_dodge(0.6), width = 0.2) +
  labs(
    title = "Departure Delay by Airport and Time of Day",
    x = "Origin airport",
    y = "Mean delay (min)",
    fill = "Time of day"
  ) +
  theme_minimal(base_size = 14)
```

## stack and fill

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| fig-alt: "Side-by-side comparison of stacked and filled bar charts. The stacked chart shows raw counts of Morning and Afternoon/Evening flights at each airport; the filled chart normalizes each bar to 100% to compare proportions directly."

p1 <- flights_sample |>
  ggplot(aes(x = origin, fill = time_of_day)) +
  geom_bar(position = "stack") +
  labs(title = "Stacked (counts)", x = "Origin airport", y = "Count", fill = "Time of day") +
  theme_minimal(base_size = 14)

p2 <- flights_sample |>
  ggplot(aes(x = origin, fill = time_of_day)) +
  geom_bar(position = "fill") +
  labs(title = "Filled (proportions)", x = "Origin airport", y = "Proportion", fill = "Time of day") +
  theme_minimal(base_size = 14)

p1 + p2
```

## Scales: controlling axes and colors

::: {.notes}
Spend ~3 minutes on this and the next slide together. Students don't need to memorize every scale function — they need to know the naming pattern: `scale_` + the aesthetic (`fill`, `color`, `x`, `y`) + `_` + the type (`manual`, `continuous`, `discrete`). Once they see the pattern, they can guess function names and use autocomplete.
:::

Scales translate data values into visual properties:

```{r}
#| eval: false
# Control axis breaks (tick marks)
scale_x_continuous(breaks = seq(0, 180, by = 30))

# Set specific colors
scale_fill_manual(values = c("EWR" = "lightblue", "JFK" = "coral", "LGA" = "lightgreen"))
```

## scale_fill_manual()

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Boxplots of departure delay by airport using custom colors: blue for EWR, red for JFK, and yellow for LGA, demonstrating how scale_fill_manual() applies specific hex color codes to groups."
flights_sample |>
  ggplot(aes(x = origin, y = dep_delay, fill = origin)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("EWR" = "#3498db", "JFK" = "#e74c3c", "LGA" = "#f1c40f")) +
  labs(title = "Custom colors", x = "Origin airport", y = "Delay (min)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

Use a website like [HTML color codes](https://htmlcolorcodes.com/color-picker/) to find the appropriate HEX code for your color.

## Coordinate systems

::: {.notes}
The `coord_cartesian()` vs `scale_y_continuous(limits = ...)` distinction is critical and students will confuse them. The key: `coord_cartesian()` zooms the view (like pinching to zoom on a photo), while `scale_y_continuous(limits = ...)` filters the data first and then plots. This means boxplot whiskers, smooth lines, and error bars can all change when you use scale limits because outliers are removed before the calculation. Always use `coord_cartesian()` for zooming.
:::

```{r}
#| eval: false
# Flip x and y (great for long category labels)
coord_flip()

# Zoom in without dropping data
coord_cartesian(ylim = c(-10, 30))
# vs
scale_y_continuous(limits = c(-10, 30))  # This DROPS data outside the range!
```

## coord_flip() in action

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Horizontal boxplots of departure delay by airport, with airport codes on the y-axis and delay on the x-axis, demonstrating how flipped axes improve readability for categorical labels."
flights_sample |>
  ggplot(aes(y = origin, x = dep_delay, fill = origin)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Flipped axes — better for category labels", y = "Origin airport", x = "Delay (min)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

# Putting it together {background-color="#2c3e50"}

## A polished figure

::: {.notes}
This is the "bringing it all together" slide. Walk through the code slowly, pointing out each layer and why it's there: jittered points for raw data, boxplot for summary, stat_summary diamond for the mean, manual colors for distinction, and informative labels. This is roughly the quality you'd expect in a journal submission. Students should aim for this level on Assignment 4 and the final project.
:::

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 6
#| fig-alt: "Polished figure combining jittered individual flight delays, boxplots, and diamond-shaped mean markers for each NYC airport. Uses custom colors, informative title and subtitle, and a caption noting sample size and mean indicator."
flights_sample |>
  ggplot(aes(x = origin, y = dep_delay, fill = origin, color = origin)) +
  geom_jitter(width = 0.2, alpha = 0.35, size = 1.5) +
  # outlier.shape = NA hides boxplot outliers because geom_jitter already shows every point
  geom_boxplot(alpha = 0.5, width = 0.4, outlier.shape = NA) +
  # shape = 18 is a solid diamond — draws attention to the mean
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "black") +
  scale_fill_manual(values = c("EWR" = "#3498db", "JFK" = "#e74c3c", "LGA" = "#f1c40f")) +
  scale_color_manual(values = c("EWR" = "#2980b9", "JFK" = "#c0392b", "LGA" = "#b7950b")) +
  labs(
    title = "Departure Delays Are Similar Across NYC Airports",
    subtitle = "Individual flights, box summaries, and mean (diamond) shown",
    x = "Origin airport",
    y = "Departure delay (min)",
    caption = "N = 500 sampled flights. Diamond = mean."
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

# Get a head start {background-color="#e67e22"}

## Assignment 4 preview

::: {.notes}
This is the head-start exercise (~10 min remaining). Students don't need to finish this in class — it's a bridge to Assignment 4. Encourage them to have fun making the "bad" version deliberately awful. If students seem finished early, suggest they try adding `facet_wrap(~time_of_day)` to any of today's plots as a preview of faceting.
:::

Assignment 4 will ask you to create "bad" and "good" versions of figures. Start experimenting:

1. Take any plot from today and make it **deliberately bad** — wrong geom, missing labels, confusing colors
2. Then fix it. What did you change and why?
3. Try creating the same data with three different geoms. Which one communicates best?

# Wrapping up {background-color="#2c3e50"}

## Today's toolkit

| Tool | What it does |
|------|--------------|
| `geom_histogram()` | Distribution of one continuous variable |
| `geom_density()` | Smooth distribution estimate |
| `geom_boxplot()` | Summary of continuous by categorical |
| `geom_violin()` | Full distribution by categorical |
| `geom_jitter()` | Individual points with horizontal wiggle |
| `stat_summary()` | Calculate and display summary stats |
| `position = "dodge"` | Side-by-side grouped plots |
| `scale_fill_manual()` | Custom colors |
| `coord_flip()` | Swap axes |

## Before next class

📖 **Read:**

- Supplementary: Visual perception principles (will be posted)
- Optional: Knaflic, *Storytelling with Data*, Ch 1–3

✅ **Practice:**

- Create a bar chart with error bars for a dataset of your choice
- Try boxplot vs violin on the same data
- Experiment with `coord_flip()` and `scale_fill_manual()`

## Key takeaways

1. **Match your geom to your data** — the wrong choice misleads
2. **stat_summary() is powerful** — means, error bars, custom functions
3. **Position adjustments** handle overlap (dodge, stack, fill, jitter)
4. **Scales control** axes, colors, and legends
5. **coord_cartesian() zooms; scale limits drop data** — know the difference

## The one thing to remember

The difference between a chart and a figure is intention — every layer should earn its place.

Next time: Perception & Design

::: {.notes}
End on this line and let it land. Next session builds directly on this one — we go from "which geom" to "which color, which theme, which design decisions." Students should have R4DS Ch 9 fresh in their minds and the optional perception reading done before Monday.
:::

