---
title: "Perception & Design"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 22, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 8"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(patchwork)

# Reuse the reaction time dataset
set.seed(123)
reaction_data <- tibble(
  participant = rep(1:40, each = 2),
  condition = rep(c("Control", "Treatment"), 40),
  rt = c(rnorm(40, mean = 520, sd = 60), rnorm(40, mean = 480, sd = 55)),
  accuracy = c(rbinom(40, 1, 0.82), rbinom(40, 1, 0.88)),
  age_group = rep(c("Young", "Young", "Older", "Older"), 20)
) |>
  mutate(rt = round(rt, 1))
```

# How we see data {background-color="#2c3e50"}

::: {.notes}
This first block (~25 min) covers perception, color theory, and decluttering. The opening "what's wrong" slide is meant to be interactive — pause and let students call out problems before revealing the answer. Assignment 3 is due this week and Assignment 4 (Visualization Deep Dive) is assigned today.
:::

## What's wrong with this?

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| fig-alt: "Deliberately poorly designed bar chart with red and green bars, tiny text, yellow background, heavy gridlines, and an uninformative title — used as a 'what's wrong' warm-up exercise."
ggplot(reaction_data, aes(x = condition, y = rt, fill = condition)) +
  geom_bar(stat = "summary", fun = mean) +
  scale_fill_manual(values = c("Control" = "red", "Treatment" = "green")) +
  labs(title = "Results", y = "rt", x = "condition") +
  theme(
    panel.background = element_rect(fill = "lightyellow"),
    panel.grid = element_line(color = "grey50"),
    text = element_text(size = 10)
  )
```

. . .

Red-green palette (colorblind-unfriendly). No informative title. Tiny text. Gridline noise. Bars hiding the data.

Today we learn **why your brain rejects bad figures** — and how to design ones that work.

::: {.notes}
Let the reveal build naturally — use the pause (`. . .`) to give students 30 seconds to identify problems on their own first. Cold-call a few students or ask pairs to share. Most will catch the red-green issue since we mentioned it last session. Fewer will notice the bar chart hiding data or the uninformative title. This sets up the whole session: design isn't decoration, it's communication.
:::

## Your brain processes some visual features before you even think

Some visual features are processed almost instantly — before conscious attention kicks in:

| Attribute | Example |
|-----------|---------|
| **Color** | A red dot among blue dots |
| **Size** | A large circle among small ones |
| **Position** | A point far from the others |
| **Shape** | A triangle among circles |
| **Orientation** | A tilted bar among vertical bars |

These are your tools for drawing the viewer's eye.

::: {.notes}
This connects directly to psych content students may have seen in intro psych or perception courses. Preattentive processing happens in ~250ms — before conscious attention. The practical implication for data viz: if you want something to stand out, use one of these attributes. If you use too many at once, nothing stands out. Spend ~3 minutes here. Students often ask about animation — yes, motion is preattentive too, but we won't use it in static figures.
:::

## Preattentive in action

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
#| fig-alt: "Three panel demonstration of preattentive processing: the left panel uses color to make red points pop among gray ones, the middle panel uses size to make two large points stand out, and the right panel uses position to isolate one outlier far from a cluster."

# Color draws attention
p1 <- tibble(x = 1:10, y = rnorm(10), highlight = c(rep("other", 7), rep("target", 3))) |>
  ggplot(aes(x, y, color = highlight)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("other" = "gray70", "target" = "#e74c3c"), guide = "none") +
  labs(title = "Color") + theme_minimal(base_size = 14)

# Size draws attention
p2 <- tibble(x = 1:10, y = rnorm(10), sz = c(rep(3, 8), 8, 8)) |>
  ggplot(aes(x, y, size = sz)) +
  geom_point(color = "steelblue") +
  scale_size(guide = "none") +
  labs(title = "Size") + theme_minimal(base_size = 14)

# Position draws attention
p3 <- tibble(x = c(rnorm(9, 0, 0.5), 3), y = c(rnorm(9, 0, 0.5), 2.5)) |>
  ggplot(aes(x, y)) +
  geom_point(size = 4, color = "steelblue") +
  labs(title = "Position") + theme_minimal(base_size = 14)

p1 + p2 + p3
```

Your eye goes to the red points, the big points, and the outlier — instantly.

::: {.notes}
Ask students: "Where did your eye go first in each panel?" The answer should be obvious and immediate — that's the whole point. The practical takeaway: use color to highlight your key finding, not to decorate. If everything is colorful, nothing pops.
:::

## Gestalt principles (brief)

The brain groups things automatically:

- **Proximity** — things close together feel like a group
- **Similarity** — things that look alike feel like a group
- **Enclosure** — things inside a border feel like a group

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| fig-alt: "Scatterplot showing two visually distinct clusters of points separated by empty space, demonstrating the Gestalt principle of proximity — nearby points are perceived as belonging to the same group."
# Proximity example
tibble(
  x = c(1, 1.2, 1.1, 3, 3.2, 3.1),
  y = c(1, 1.3, 0.8, 1, 1.2, 0.9),
  group = c("A","A","A","B","B","B")
) |>
  ggplot(aes(x, y, color = group)) +
  geom_point(size = 6) +
  scale_color_manual(values = c("A" = "steelblue", "B" = "coral")) +
  labs(title = "Proximity: two clusters feel like two groups") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none", axis.text = element_blank(), axis.title = element_blank())
```

::: {.notes}
Keep Gestalt brief — 1-2 minutes. Students don't need to memorize the terms, just the intuition: your brain groups things that are close together, look similar, or are enclosed together. This matters for figure design because spacing and visual similarity communicate grouping whether you intend it or not. Facets use proximity; color-coding uses similarity.
:::

# Color theory {background-color="#2c3e50"}

## Three types of color palettes

| Type | When to use | Example |
|------|-------------|---------|
| **Sequential** | One continuous variable (low → high) | Blues, viridis |
| **Diverging** | Values relative to a midpoint | Red–white–blue |
| **Qualitative** | Categorical groups (no order) | Set1, tab10 |

Using the wrong type is one of the most common visualization mistakes.

::: {.notes}
This is a high-value slide — spend ~3 minutes here. The most common mistake: using a qualitative palette for ordered data (like a rainbow for low-to-high), or using a sequential palette for unrelated categories. Ask students: "If you had data on anxiety levels from low to high, which palette type?" (Sequential.) "If you had data on four different therapy types?" (Qualitative.) Test their intuition before moving to examples.
:::

## Sequential palettes

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| fig-alt: "A row of ten tiles colored with the viridis sequential palette, transitioning smoothly from dark purple on the left to bright yellow on the right, illustrating how sequential palettes encode low-to-high values."
tibble(x = 1:10, y = rep(1, 10), val = 1:10) |>
  ggplot(aes(x, y, fill = val)) +
  geom_tile(width = 0.9) +
  scale_fill_viridis_c() +
  labs(title = "viridis (sequential) — great for continuous data") +
  theme_minimal(base_size = 14) +
  theme(axis.text = element_blank(), axis.title = element_blank(), legend.position = "bottom")
```

## Qualitative palettes

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| fig-alt: "A row of eight tiles in distinct pastel colors from the ColorBrewer Set2 qualitative palette, showing how each category gets a visually distinguishable hue with no implied ordering."
tibble(x = 1:8, y = rep(1, 8), cat = letters[1:8]) |>
  ggplot(aes(x, y, fill = cat)) +
  geom_tile(width = 0.9) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Set2 (qualitative) — for categorical groups") +
  theme_minimal(base_size = 14) +
  theme(axis.text = element_blank(), axis.title = element_blank(), legend.position = "bottom")
```

## Colorblind-friendly choices

::: {.notes}
Emphasize that 8% of men is roughly 1 in 12 — in a class of 50, there are likely 2-3 students with some color vision deficiency. This is not an edge case. The practical advice is simple: use viridis or a tested palette, and never rely on color alone (also use shape or labels). Students often ask "how do I check?" — suggest the colorblindness simulator at davidmathlogic.com/colorblind or the `colorblindr` R package.
:::

About 8% of men have some form of color vision deficiency. Red-green is the most common.

```{r}
#| eval: false
# viridis is colorblind-safe AND sequential
scale_fill_viridis_d()   # discrete
scale_fill_viridis_c()   # continuous

# ColorBrewer palettes designed for colorblindness
scale_fill_brewer(palette = "Set2")

# Or set colors manually with safe choices
scale_fill_manual(values = c("#0072B2", "#E69F00", "#009E73"))
# (blue, orange, green — distinguishable for most color vision types)
```

## A bad color choice vs a good one

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| fig-alt: "Side-by-side comparison of two boxplots: the left uses red and green fills that are indistinguishable to colorblind viewers, while the right uses blue and orange fills that are accessible to virtually all viewers."

p1 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Control" = "#ff0000", "Treatment" = "#00aa00")) +
  labs(title = "Red vs Green — bad for colorblind viewers", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p2 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Control" = "#0072B2", "Treatment" = "#E69F00")) +
  labs(title = "Blue vs Orange — works for everyone", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p1 + p2
```

::: {.notes}
Let the side-by-side comparison speak for itself. The blue-orange version on the right is distinguishable to virtually all viewers. You can mention that some journals now require colorblind-accessible figures. The hex codes shown earlier (#0072B2, #E69F00, #009E73) are from the Okabe-Ito palette — a safe default students can copy into any project.
:::

# Decluttering {background-color="#2c3e50"}

## The default is cluttered

::: {.notes}
Transition: "Now that we know what to show, let's talk about what to remove." This section draws heavily on Knaflic's "declutter" philosophy. The before/after comparison here is dramatic — the gray background, heavy gridlines, and redundant legend all add zero information. Students sometimes resist removing gridlines because they think figures look "empty" without them. Remind them: if the reader needs a gridline to read the value, add a direct label instead.
:::

ggplot2's default theme adds a lot of visual noise. Compare:

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| fig-alt: "Side-by-side comparison of the same boxplot using ggplot2's default gray theme on the left versus the cleaner minimal theme on the right, showing how removing the gray background, heavy gridlines, and redundant legend reduces visual clutter."

p1 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot() +
  labs(title = "Default theme (theme_gray)", x = "Condition", y = "RT (ms)")

p2 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot() +
  labs(title = "Cleaned up (theme_minimal)", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p1 + p2
```

## The declutter checklist

::: {.notes}
This is a reference slide students should internalize. Walk through the "Remove" column and ask students to justify each removal — e.g., "Why remove the legend when we have two conditions?" (Because the x-axis already labels them.) The "Keep" column matters too: titles should state a finding, not label the chart type. "Boxplot of RT" is a label; "Treatment reduces reaction time" is a finding.
:::

Ask: does this element help the reader understand the data?

::: {.columns}
::: {.column width="50%"}
**Remove**

| Element | Why |
|---------|-----|
| Gray background | Noise, no info |
| Gridlines (most) | Distraction |
| Redundant legend | X-axis says it |
| Generic axis labels | Add units instead |
:::
::: {.column width="50%"}
**Keep**

| Element | Why |
|---------|-----|
| Title + subtitle | Orients the reader |
| Caption | Source, N, error bar type |
| Meaningful color | Highlights comparisons |
:::
:::

## theme() for fine control

::: {.notes}
Don't try to teach every theme() option — there are dozens. The goal is for students to know that theme() exists and to see the pattern: each element gets a function like `element_blank()` (remove it), `element_text()` (style text), `element_line()` (style lines), or `element_rect()` (style rectangles). They can look up specific options in the ggplot2 documentation. The most common ones they'll use are `legend.position`, `panel.grid`, and `plot.title`.
:::

```{r}
#| eval: false
theme(
  legend.position = "none",                    # Remove legend
  axis.ticks = element_blank(),                # Remove tick marks
  panel.grid = element_blank(),                # Remove gridlines
  plot.title = element_text(face = "bold"),    # Bold the title
  axis.text = element_text(size = 10),         # Font size
  plot.caption = element_text(hjust = 0)       # Left-align caption
)
```

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

::: {.notes}
Give 10 minutes. The cluttered graph on the next slide has at least 7 problems: red-green colors, size mapped to accuracy (binary variable — misleading), no informative title, variable names as axis labels, default gray theme, redundant color and fill, and points plotted behind the boxplot. Circulate and ask pairs to name each problem before they start fixing it. Common student approach: they'll fix the colors first because it's most obvious. Push them to also address the title and labels.
:::

I'll put a cluttered graph on screen. With a partner, **rewrite it** to follow the design principles we just covered.

Remove at least 5 unnecessary elements. Make it tell a clear story.

::: {.callout-tip}
Think about: colors, legend, labels, theme, size mapping, and whether every aesthetic is adding information.
:::

## Your turn: 10 minutes

```{r}
#| fig-alt: "Deliberately cluttered scatterplot with red and green colors, points sized by a binary variable, redundant color and fill mappings, uninformative title, and the default gray theme — students are asked to redesign it."
# The cluttered version — fix this!
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition, color = condition, size = accuracy)) +
  geom_point() +
  geom_boxplot(alpha = 0.3) +
  scale_fill_manual(values = c("Control" = "red", "Treatment" = "green")) +
  scale_color_manual(values = c("Control" = "red", "Treatment" = "green")) +
  labs(x = "condition", y = "rt") +
  ggtitle("data") +
  theme_gray()
```



```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 2, color = "gray50") +
  geom_boxplot(alpha = 0.5, width = 0.4, outlier.shape = NA) +
  scale_fill_manual(values = c("Control" = "#0072B2", "Treatment" = "#E69F00")) +
  labs(
    title = "Treatment Reduces Reaction Time",
    x = "Condition",
    y = "Reaction time (ms)",
    caption = "N = 40 per condition. Points = individual observations."
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid = element_blank())
```

---

## Before we move on

📤 **Upload your code to Canvas** for participation credit. Paste what you have into today's in-class submission — it doesn't need to work perfectly.

# Critiquing bad graphs {background-color="#2c3e50"}

::: {.notes}
This section (~10 min) is the second lecture block. It transitions from "how to build good figures" to "how to spot bad ones." These three examples each illustrate a different category of problem: misleading axes, poor aesthetics, and wrong chart type. Run through them interactively — show the figure, let students diagnose it, then reveal the answer. This builds the critical evaluation skill they'll need for Assignment 4.
:::

## What's wrong here? (1 of 3)

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Bar chart comparing mean reaction times for Control and Treatment conditions where the y-axis starts at 470 instead of 0, making a small ~40 ms difference appear dramatically large."
# Bad: Truncated y-axis exaggerates difference
reaction_data |>
  group_by(condition) |>
  summarize(mean_rt = mean(rt)) |>
  ggplot(aes(x = condition, y = mean_rt, fill = condition)) +
  geom_col() +
  scale_fill_manual(values = c("Control" = "grey60", "Treatment" = "grey40")) +
  coord_cartesian(ylim = c(470, 530)) +
  labs(title = "Treatment Reduces Reaction Time",
       y = "Mean RT (ms)", x = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

. . .

**The problem:** The y-axis starts at 470, not 0. The difference looks massive — but it's only ~40 ms. A truncated axis exaggerates the effect.

::: {.notes}
Students often ask: "But isn't it sometimes okay to not start at zero?" Good question. For bar charts, no — the bar height represents the value, so truncating is always misleading. For line charts or dot plots, it can be acceptable because the position (not the area) encodes the value. This nuance matters and comes up in the storytelling session (Session 16) too.
:::

## What's wrong here? (2 of 3)

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Boxplot with red and green fills, a yellow background, heavy gridlines, variable names as axis labels, and the uninformative title 'Boxplot' — demonstrating multiple common design mistakes at once."
# Bad: Rainbow color + too many gridlines + no informative title
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = age_group)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Young" = "red", "Older" = "green")) +
  labs(title = "Boxplot", y = "rt", x = "condition") +
  theme(
    panel.background = element_rect(fill = "lightyellow"),
    panel.grid.major = element_line(color = "grey50"),
    panel.grid.minor = element_line(color = "grey70")
  )
```

. . .

**The problems:** Red-green palette (colorblind-unfriendly). Title says "Boxplot" (a label, not a finding). Variable names as labels. Distracting background color and gridlines.

## What's wrong here? (3 of 3)

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Pie chart showing reaction times arbitrarily binned into Fast, Medium, and Slow categories, hiding the continuous distribution, individual observations, and any comparison between conditions."
# Bad: Pie chart of continuous data categories
reaction_data |>
  mutate(speed = case_when(
    rt < 450 ~ "Fast",
    rt < 550 ~ "Medium",
    TRUE ~ "Slow"
  )) |>
  count(speed) |>
  ggplot(aes(x = "", y = n, fill = speed)) +
  geom_col(width = 1) +
  coord_polar("y") +
  labs(title = "Speed Categories") +
  theme_void()
```

. . .

**The problems:** Continuous data was binned into arbitrary categories, then displayed as a pie chart. We lost the actual reaction times, the condition comparison, and the ability to see distributions. A histogram or density plot would show far more.

::: {.notes}
This is the strongest example because it shows a cascade of bad decisions: first they binned continuous data (losing information), then they used a pie chart (making comparison harder), and they dropped the condition variable entirely. Each step destroyed information. The lesson: your choice of chart type determines what story you can tell. A pie chart here makes it impossible to compare conditions or see the distribution.
:::

# Get a head start {background-color="#e67e22"}

## Assignment 4 preview

::: {.notes}
Assignment 4 is being assigned today. Briefly outline the three parts so students know what to aim for. Emphasize that the "bad" version should be deliberately bad with specific design violations they can name — not just ugly. They need to articulate what's wrong with it and why the good version is better. The colorblind-accessible version is a separate deliverable, not just "the good version with viridis."
:::

Assignment 4 will ask you to:

1. Create a **"bad" version** of a figure — deliberately violate design principles
2. Create a **"good" version** following what we covered today
3. Create a **colorblind-accessible version**

## Assignment 4 preview

Start experimenting now:

- Take the `reaction_data` dataset
- Make the worst possible version of a figure
- Then make it great
- What did you change?

# Wrapping up {background-color="#2c3e50"}

## Design principles checklist

- [ ] Is the geom appropriate for the data type?
- [ ] Are colors colorblind-friendly?
- [ ] Is the legend necessary? (Remove if redundant)
- [ ] Are axis labels informative (not just variable names)?
- [ ] Is there a title that states the main point?
- [ ] Is there visual clutter to remove?
- [ ] Do error bars have a caption explaining what they show?

## Before next class

📖 **Read:**

- [R4DS Ch 10: Exploratory data analysis](https://r4ds.hadley.nz/eda) (sections 10.1–10.4)

✅ **Practice:**

- Try the "bad vs good" exercise on your own
- Find a graph online and identify ways to improve it

## Key takeaways

1. **Preattentive attributes** guide the eye — use them intentionally
2. **Match your palette type** to your data (sequential, diverging, qualitative)
3. **Design for colorblindness** — always
4. **Declutter ruthlessly** — every element must earn its place
5. **Critical evaluation** — always ask whether each element adds information

## The one thing to remember

You're not designing for yourself. You're designing for a reader who will look at your figure for five seconds.

Next time: Exploratory Data Analysis

::: {.notes}
End on this line — it reframes everything from today. Students naturally design for themselves (they already know what the data shows). The challenge is designing for someone who doesn't. Five seconds is roughly how long a reader spends on a figure before deciding whether to engage. Next session shifts from how to present data to how to explore it (EDA). Remind students that Assignment 3 is due this week and Assignment 4 is now available.
:::

