---
title: "EDA — Variation"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 27, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 9"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(psych)

# Load the Big Five personality dataset — this is our EDA playground
data(bfi)
bfi <- as_tibble(bfi)
```

# What is EDA? {background-color="#2c3e50"}

## Exploratory vs. confirmatory

| | Exploratory (EDA) | Confirmatory |
|---|---|---|
| **Goal** | Discover patterns | Test hypotheses |
| **Attitude** | Curiosity | Rigor |
| **Questions** | Open-ended | Pre-specified |
| **Output** | Visualizations, hunches | p-values, conclusions |

EDA comes **first.** You can't test a hypothesis you didn't notice.

## The EDA mindset

> "EDA is an attitude, not a technique."  
> — John Tukey

- Ask questions
- Answer them by visualizing data
- Use what you learn to ask new questions
- Repeat

There is no single "right" way to do EDA. The goal is **understanding.**

## What we're doing today

Exploring **variation** — what does a single variable look like?

- Distribution shape
- Center and spread
- Outliers and unusual values
- Missing data

Tomorrow: **covariation** — how do variables relate to each other?

## Our dataset: Big Five Personality

```{r}
glimpse(bfi)
```

25 items measuring five personality factors (Agreeableness, Conscientiousness, Extraversion, Neuroticism, Openness) + demographics.

## What's in here

```{r}
#| echo: false
tibble(
  Variable = c("A1–A5", "C1–C5", "E1–E5", "N1–N5", "O1–O5", "gender", "education", "age"),
  Description = c(
    "Agreeableness items (1–6 scale)",
    "Conscientiousness items (1–6 scale)",
    "Extraversion items (1–6 scale)",
    "Neuroticism items (1–6 scale)",
    "Openness items (1–6 scale)",
    "1 = male, 2 = female",
    "1–5 (HS incomplete through graduate)",
    "Age in years"
  )
) |> knitr::kable()
```

# Exploring distributions {background-color="#2c3e50"}

## Start simple: summary statistics

```{r}
bfi |>
  select(age, gender, education) |>
  summary()
```

But summary stats can hide a lot. Always visualize.

## Histograms: the workhorse

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
bfi |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Age",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## Choosing binwidth

The binwidth matters a lot:

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
library(patchwork)

p1 <- bfi |> ggplot(aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "binwidth = 1 (too narrow)", x = "Age") +
  theme_minimal(base_size = 13)

p2 <- bfi |> ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "binwidth = 5 (good)", x = "Age") +
  theme_minimal(base_size = 13)

p3 <- bfi |> ggplot(aes(x = age)) +
  geom_histogram(binwidth = 20, fill = "steelblue", color = "white") +
  labs(title = "binwidth = 20 (too wide)", x = "Age") +
  theme_minimal(base_size = 13)

p1 + p2 + p3
```

There's no "right" answer — try a few and pick the one that shows the shape clearly.

## Density plots: smooth alternative

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
bfi |>
  ggplot(aes(x = age)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(
    title = "Distribution of Age (Density)",
    x = "Age (years)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)
```

## Exploring survey items

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
# All Extraversion items
bfi |>
  select(E1:E5) |>
  pivot_longer(everything(), names_to = "item", values_to = "response") |>
  ggplot(aes(x = response, fill = item)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Extraversion Item Responses",
    x = "Response (1–6)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## Categorical variables

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
bfi |>
  mutate(gender = ifelse(gender == 1, "Male", "Female")) |>
  ggplot(aes(x = gender)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Gender Distribution",
    x = "Gender",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## Counts and proportions

```{r}
# Counts
bfi |>
  mutate(gender = ifelse(gender == 1, "Male", "Female")) |>
  count(gender)
```

```{r}
# Proportions
bfi |>
  mutate(gender = ifelse(gender == 1, "Male", "Female")) |>
  count(gender) |>
  mutate(proportion = round(n / sum(n), 3))
```

# Outliers {background-color="#2c3e50"}

## What are outliers?

Values that are unusual compared to the rest of the data. They might be:

- **Real** — genuinely extreme values (a 95-year-old in a college study)
- **Errors** — data entry mistakes (age = 999)
- **Interesting** — worth investigating

**Never delete an outlier without understanding it.**

## Spotting outliers visually

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
bfi |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Age Distribution — Any Outliers?",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## Spotting outliers with boxplots

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 4
bfi |>
  ggplot(aes(x = age)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Age — Boxplot Shows Outliers as Points",
    x = "Age (years)"
  ) +
  theme_minimal(base_size = 14)
```

Points beyond the whiskers are flagged as outliers by the 1.5×IQR rule.

## Investigating outliers programmatically

```{r}
# Find unusually old or young participants
bfi |>
  filter(age > 60 | age < 18) |>
  select(age, gender, education)
```

## What to do with outliers

| Outlier type | What to do |
|---|---|
| Likely error (age = 999) | Fix or set to NA |
| Real but extreme | Note it, keep it, mention in write-up |
| Suspicious | Investigate further |
| Doesn't affect conclusions | Keep it, move on |

**Document your decisions.** Future you will want to know.

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

Using the `bfi` dataset (already loaded):

1. Plot the distribution of **age** — does it look normal? What shape is it?
2. Check for **outliers** in age. Are any values suspicious?
3. Pick **one personality item** (e.g., E1) and explore its distribution
4. What do you notice? Write down 2 observations.

::: {.callout-tip}
Try both a histogram and a boxplot for age. They show different things. Use `summary()` first to get oriented.
:::

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
# Age distribution
bfi |> ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  theme_minimal()

# Outlier check
bfi |> ggplot(aes(x = age)) + geom_boxplot() + theme_minimal()
bfi |> filter(age > 60)

# One item
bfi |> ggplot(aes(x = E1)) + geom_bar(fill = "steelblue") + theme_minimal()
```

---

## Before we move on

📤 **Upload your code to Canvas** for participation credit. Paste what you have into today's in-class submission — it doesn't need to work perfectly.

# Missing data — a first look {background-color="#2c3e50"}

## Missing data is everywhere

```{r}
# How much missing data do we have?
bfi |>
  summarize(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  arrange(desc(n_missing))
```

## Visualizing missingness

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
bfi |>
  summarize(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  mutate(pct_missing = pct_missing * 100) |>
  filter(pct_missing > 0) |>
  arrange(desc(pct_missing)) |>
  ggplot(aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Percentage of Missing Values by Variable",
    x = "Variable",
    y = "% Missing"
  ) +
  theme_minimal(base_size = 14)
```

## Why it matters for EDA

Missing data can bias your exploration:

- If missingness is related to the variables you're studying, your distributions are wrong
- If certain groups are more likely to have missing data, you might miss important patterns

We'll cover this in depth in Session 14. For now: **always check for missing data before exploring.**

# The EDA workflow {background-color="#2c3e50"}

## A systematic approach

1. **Look at the structure** — `glimpse()`, `summary()`
2. **Check for missing data** — how much? Which variables?
3. **Explore each variable** — histograms, bar charts, summaries
4. **Look for outliers** — boxplots, programmatic checks
5. **Ask questions** — what surprised you? What do you want to know more about?
6. **Repeat** — EDA is iterative

## Putting it together: a mini-EDA

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 6
# Education distribution — what does it look like?
bfi |>
  filter(!is.na(education)) |>
  mutate(education = factor(education, labels = c(
    "HS incomplete", "HS grad", "Some college",
    "College grad", "Graduate degree"
  ))) |>
  ggplot(aes(x = education)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Education Level in the BFI Sample",
    x = "Education level",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Get a head start: Assignment 5

Assignment 5 will have you do a full EDA on a psychology dataset. Start now:

1. Load `bfi` and run `glimpse()` and `summary()`
2. Pick **3 variables** and create distributions for each
3. For each one: What shape is it? Any outliers? Any missing data?
4. Write down **one question** each distribution makes you want to ask

That's the core of EDA — noticing things and getting curious.

# Wrapping up {background-color="#2c3e50"}

## EDA toolkit so far

| Tool | When to use |
|------|-------------|
| `glimpse()` | First look at structure |
| `summary()` | Quick numeric summaries |
| `geom_histogram()` | Continuous distributions |
| `geom_density()` | Smooth distribution shape |
| `geom_bar()` | Categorical distributions |
| `geom_boxplot()` | Outlier detection |
| `count()` | Frequency tables |
| `is.na()` / `sum(is.na())` | Missing data check |

## Before next class

📖 **Read:**

- [R4DS Ch 10: Exploratory data analysis](https://r4ds.hadley.nz/eda) (sections 10.5–10.6)

✅ **Practice:**

- Explore at least 3 variables in `bfi`
- Look for outliers and missing data
- Start generating questions

## Key takeaways

1. **EDA is an attitude** — curiosity, not confirmation
2. **Always visualize** — summary stats hide patterns
3. **Outliers are information** — investigate, don't delete
4. **Check missing data first** — it affects everything
5. **Ask questions, then answer them** — that's the loop

## Questions?

Next time: **EDA — Covariation**

We'll explore relationships between variables — the really interesting stuff!
