---
title: "Perception & Design"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 27, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 9"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
    include-in-header:
      text: |
        <link rel="stylesheet" href="toc-nav.css">
    include-after-body:
      text: |
        <script src="toc-nav.js"></script>
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(patchwork)

# Reuse the reaction time dataset
set.seed(123)
reaction_data <- tibble(
  participant = rep(1:40, each = 2),
  condition = rep(c("Control", "Treatment"), 40),
  rt = c(rnorm(40, mean = 520, sd = 60), rnorm(40, mean = 480, sd = 55)),
  accuracy = c(rbinom(40, 1, 0.82), rbinom(40, 1, 0.88)),
  age_group = rep(c("Young", "Young", "Older", "Older"), 20)
) |>
  mutate(rt = round(rt, 1))
```

# How we see data {background-color="#2c3e50"}

::: {.notes}
This first block (~25 min) covers perception, color theory, and decluttering. The opening "what's wrong" slide is meant to be interactive — pause and let students call out problems before revealing the answer. Assignment 3 is due this week and Assignment 4 (Visualization Deep Dive) is assigned today.
:::

## What's wrong with this?

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| fig-alt: "Deliberately poorly designed bar chart with red and green bars, tiny text, yellow background, heavy gridlines, and an uninformative title — used as a 'what's wrong' warm-up exercise."
ggplot(reaction_data, aes(x = condition, y = rt, fill = condition)) +
  geom_bar(stat = "summary", fun = mean) +
  scale_fill_manual(values = c("Control" = "red", "Treatment" = "green")) +
  labs(title = "Results", y = "rt", x = "condition") +
  theme(
    panel.background = element_rect(fill = "lightyellow"),
    panel.grid = element_line(color = "grey50"),
    text = element_text(size = 10)
  )
```

. . .

Red-green palette (colorblind-unfriendly). No informative title. Tiny text. Gridline noise. Bars hiding the data.

Today we learn **why your brain rejects bad figures** — and how to design ones that work.

::: {.notes}
Let the reveal build naturally — use the pause (`. . .`) to give students 30 seconds to identify problems on their own first. Cold-call a few students or ask pairs to share. Most will catch the red-green issue since we mentioned it last session. Fewer will notice the bar chart hiding data or the uninformative title. This sets up the whole session: design isn't decoration, it's communication.
:::

## Your brain processes some visual features before you even think

Some visual features are processed almost instantly — before conscious attention kicks in:

| Attribute | Example |
|-----------|---------|
| **Color** | A red dot among blue dots |
| **Size** | A large circle among small ones |
| **Position** | A point far from the others |
| **Shape** | A triangle among circles |
| **Orientation** | A tilted bar among vertical bars |

These are your tools for drawing the viewer's eye.

::: {.notes}
This connects directly to psych content students may have seen in intro psych or perception courses. Preattentive processing happens in ~250ms — before conscious attention. The practical implication for data viz: if you want something to stand out, use one of these attributes. If you use too many at once, nothing stands out. Spend ~3 minutes here. Students often ask about animation — yes, motion is preattentive too, but we won't use it in static figures.
:::

## Preattentive in action

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
#| fig-alt: "Three panel demonstration of preattentive processing: the left panel uses color to make red points pop among gray ones, the middle panel uses size to make two large points stand out, and the right panel uses position to isolate one outlier far from a cluster."

# Color draws attention
p1 <- tibble(x = 1:10, y = rnorm(10), highlight = c(rep("other", 7), rep("target", 3))) |>
  ggplot(aes(x, y, color = highlight)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("other" = "gray70", "target" = "#e74c3c"), guide = "none") +
  labs(title = "Color") + theme_minimal(base_size = 14)

# Size draws attention
p2 <- tibble(x = 1:10, y = rnorm(10), sz = c(rep(3, 8), 8, 8)) |>
  ggplot(aes(x, y, size = sz)) +
  geom_point(color = "steelblue") +
  scale_size(guide = "none") +
  labs(title = "Size") + theme_minimal(base_size = 14)

# Position draws attention
p3 <- tibble(x = c(rnorm(9, 0, 0.5), 3), y = c(rnorm(9, 0, 0.5), 2.5)) |>
  ggplot(aes(x, y)) +
  geom_point(size = 4, color = "steelblue") +
  labs(title = "Position") + theme_minimal(base_size = 14)

p1 + p2 + p3
```

Your eye goes to the red points, the big points, and the outlier — instantly.

::: {.notes}
Ask students: "Where did your eye go first in each panel?" The answer should be obvious and immediate — that's the whole point. The practical takeaway: use color to highlight your key finding, not to decorate. If everything is colorful, nothing pops.
:::

## Working memory limits what you can encode

::: {.notes}
This is direct cognitive psychology. Cowan (2001) revised Miller's "7 ± 2" downward — the real capacity is closer to 4 chunks. The visualization implication: each aesthetic you use (color, shape, size) occupies a working memory slot. Eight-category legends fail because readers literally cannot hold all eight color-group mappings in memory while reading the figure. This is the scientific basis for the "max 5 colors" rule of thumb. Spend ~3 minutes here and connect it back to the preattentive slide: just because a viewer *notices* something (preattentive) doesn't mean they can *track* it across categories (working memory).
:::

Research suggests working memory holds roughly **4 items** at a time (Cowan, 2001).

In data visualization, this means:

- Limit categorical groups in a legend to **4–5 maximum**
- Each additional aesthetic (color, shape, size) uses up a "slot"
- If you have 8 conditions, consider **faceting** rather than overlapping

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 3.5
#| fig-alt: "Two side-by-side bar charts: the left has 4 clearly distinguishable groups using a Set2 palette; the right has 8 groups where colors become hard to tell apart, illustrating how working memory limits categorical encoding in figures."
p_wm1 <- tibble(
  group = paste("Group", LETTERS[1:4]),
  value = c(4.2, 3.8, 5.1, 2.9)
) |>
  ggplot(aes(x = group, y = value, fill = group)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "4 groups — manageable", x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p_wm2 <- tibble(
  group = paste("Group", LETTERS[1:8]),
  value = c(4.2, 3.8, 5.1, 2.9, 3.5, 4.8, 2.2, 3.9)
) |>
  ggplot(aes(x = group, y = value, fill = group)) +
  geom_col() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "8 groups — hard to track", x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p_wm1 + p_wm2
```

## Gestalt principles: how the brain groups things

::: {.notes}
Give students a moment to look at each panel before reading the label. The proximity demo is obvious — two clusters, same color, same size. Similarity is more interesting: identical positions, but color alone makes two groups pop out. Enclosure shows up everywhere: annotation boxes, facet borders, legend boxes. All three operate automatically and unconsciously — whether or not you designed them in. This is a 5-minute slide; use the fragment reveals to pace the conversation.
:::

The brain automatically groups visual elements — whether you intend it or not.

::: {.columns}

::: {.column width="33%" .fragment}
**Proximity**

```{r}
#| echo: false
#| fig-height: 2.5
#| fig-width: 3.2
#| fig-alt: "Six gray dots arranged in two tight clusters separated by empty space — demonstrating the Gestalt principle of proximity."
tibble(
  x = c(1, 1.2, 1.1, 3, 3.2, 3.1),
  y = c(1, 1.3, 0.8, 1, 1.2, 0.9)
) |>
  ggplot(aes(x, y)) +
  geom_point(size = 8, color = "steelblue") +
  labs(title = "Close → grouped") +
  theme_void(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

*Nearby items feel like a group*
:::

::: {.column width="33%" .fragment}
**Similarity**

```{r}
#| echo: false
#| fig-height: 2.5
#| fig-width: 3.2
#| fig-alt: "Six dots in a row, alternating blue and orange — demonstrating Gestalt similarity grouping by color."
tibble(
  x = 1:6,
  y = rep(1, 6),
  grp = rep(c("A", "B"), 3)
) |>
  ggplot(aes(x, y, color = grp)) +
  geom_point(size = 8) +
  scale_color_manual(values = c("A" = "steelblue", "B" = "coral")) +
  labs(title = "Same look → grouped") +
  theme_void(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none")
```

*Same appearance = same group*
:::

::: {.column width="33%" .fragment}
**Enclosure**

```{r}
#| echo: false
#| fig-height: 2.5
#| fig-width: 3.2
#| fig-alt: "Two clusters of gray dots, each enclosed by a shaded rectangle — demonstrating the Gestalt principle of enclosure."
tibble(
  x = c(1, 1.2, 1.1, 3, 3.2, 3.1),
  y = c(1, 1.3, 0.8, 1, 1.2, 0.9)
) |>
  ggplot(aes(x, y)) +
  annotate("rect", xmin = 0.7, xmax = 1.5, ymin = 0.5, ymax = 1.6,
           fill = "lightblue", alpha = 0.4, color = "steelblue", linewidth = 0.8) +
  annotate("rect", xmin = 2.7, xmax = 3.5, ymin = 0.6, ymax = 1.5,
           fill = "#fff3cd", alpha = 0.8, color = "goldenrod", linewidth = 0.8) +
  geom_point(size = 8, color = "gray40") +
  labs(title = "Inside border → grouped") +
  theme_void(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

*Inside a border = same group*
:::

:::

## Gestalt principles show up everywhere in ggplot2

::: {.notes}
This is the "so what?" slide for Gestalt. Students often treat these principles as abstract theory — this slide makes them concrete and actionable. The key insight: every ggplot2 function that creates grouping or spatial structure (facets, color scales, annotation boxes) is activating these principles automatically. Good designers make this work for them; bad designers accidentally create misleading groupings. Spend 2–3 minutes. The `annotate("rect")` example is a useful technique they haven't seen yet.
:::

| Principle | How it shows up in your figures |
|-----------|----------------------------------|
| **Proximity** | `facet_wrap()` puts related panels side by side |
| **Similarity** | Color-coding makes same-group points feel connected |
| **Enclosure** | A shaded background or border sets a group apart |
| **Continuity** | Connecting points across time implies a single trend |

These groupings happen whether you design them or not — use them intentionally.

# Color theory {background-color="#2c3e50"}

## Three types of color palettes

| Type | When to use | Example |
|------|-------------|---------|
| **Sequential** | One continuous variable (low → high) | Blues, viridis |
| **Diverging** | Values relative to a midpoint | Red–white–blue |
| **Qualitative** | Categorical groups (no order) | Set1, tab10 |

Using the wrong type is one of the most common visualization mistakes.

::: {.notes}
This is a high-value slide — spend ~3 minutes here. The most common mistake: using a qualitative palette for ordered data (like a rainbow for low-to-high), or using a sequential palette for unrelated categories. Ask students: "If you had data on anxiety levels from low to high, which palette type?" (Sequential.) "If you had data on four different therapy types?" (Qualitative.) Test their intuition before moving to examples.
:::

## First question: `fill` or `color`?

::: {.notes}
This slide introduces the two-family structure. Every scale_ function starts with either `scale_fill_` or `scale_color_` — which one depends on the aesthetic the student mapped in `aes()`. This is a pattern they've seen already with `scale_fill_manual` and `scale_color_manual` in session 8. The new idea is that the same pattern applies to every other palette function. Keep this short — 1–2 minutes — and emphasize it's the same pattern they already know.
:::

Which aesthetic did you map in `aes()`? That tells you which scale family to use.

```{r}
#| eval: false
# Geoms with `fill=`: bars, boxplots, tiles, density
ggplot(data, aes(x, y, fill = group)) + geom_boxplot() +
  scale_fill_manual(values = c(...))     # ← scale_fill_*

# Geoms with `color=`: points, lines
ggplot(data, aes(x, y, color = group)) + geom_point() +
  scale_color_manual(values = c(...))    # ← scale_color_*
```

Same pattern, just swap `fill` for `color` throughout.

## Second question: which palette?

::: {.notes}
Now that they know which family (`fill` vs `color`), they need to pick a palette. This table organizes the choice by data type, which is how students should be thinking about it. The `_c` vs `_d` distinction is the single most common point of confusion — call it out explicitly. Reassure students that `viridis` is already built into ggplot2 — no new package to install. Spend 3 minutes on this slide; students will come back to it as a reference.
:::

Match the palette to your data:

| Your data | Function |
|-----------|----------|
| Continuous numeric (e.g., reaction time) | `scale_fill_viridis_c()` |
| Ordered categories (e.g., dosage levels) | `scale_fill_viridis_d()` |
| Qualitative groups (e.g., therapy type) | `scale_fill_brewer(palette = "Set2")` |
| Diverging from a midpoint (e.g., change scores) | `scale_fill_distiller(palette = "RdBu")` |
| Hand-picked colors | `scale_fill_manual(values = c(...))` |

**`_c` for numeric, `_d` for character/factor.** Same options work for `scale_color_*`.

## Sequential palettes

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| fig-alt: "A row of ten tiles colored with the viridis sequential palette, transitioning smoothly from dark purple on the left to bright yellow on the right, illustrating how sequential palettes encode low-to-high values."
tibble(x = 1:10, y = rep(1, 10), val = 1:10) |>
  ggplot(aes(x, y, fill = val)) +
  geom_tile(width = 0.9) +
  scale_fill_viridis_c() +
  labs(title = "viridis (sequential) — great for continuous data") +
  theme_minimal(base_size = 14) +
  theme(axis.text = element_blank(), axis.title = element_blank(), legend.position = "bottom")
```

## Diverging palettes

::: {.notes}
The key concept here: a diverging palette has a meaningful midpoint, usually white or a neutral color. Values above and below the midpoint get contrasting hues. In psychology, pre–post difference scores are a clean example — zero means "no change," positive means "got worse," negative means "got better." Students often try to use sequential palettes for diverging data — the result is that the negative values and positive values look like they're on the same continuum with no center. Ask: "If I have anxiety change scores from -5 (better) to +5 (worse), which palette type?" (Diverging.) Correlation matrices are another classic use case but we won't cover correlations until Session 17.
:::

Use when values diverge from a meaningful **midpoint** — often 0 or a baseline.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| fig-alt: "A row of ten tiles colored with the RdBu diverging palette, transitioning from dark blue on the left through white in the middle to dark red on the right, illustrating how diverging palettes encode values relative to a midpoint."
tibble(x = 1:10, y = rep(1, 10), val = seq(-1, 1, length.out = 10)) |>
  ggplot(aes(x, y, fill = val)) +
  geom_tile(width = 0.9) +
  scale_fill_distiller(palette = "RdBu") +
  labs(title = "RdBu (diverging) — for data with a meaningful midpoint",
       subtitle = "Psychology examples: pre–post change scores, effect sizes, difference from baseline") +
  theme_minimal(base_size = 14) +
  theme(axis.text = element_blank(), axis.title = element_blank(), legend.position = "bottom")
```

## Qualitative palettes

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| fig-alt: "A row of eight tiles in distinct pastel colors from the ColorBrewer Set2 qualitative palette, showing how each category gets a visually distinguishable hue with no implied ordering."
tibble(x = 1:8, y = rep(1, 8), cat = letters[1:8]) |>
  ggplot(aes(x, y, fill = cat)) +
  geom_tile(width = 0.9) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Set2 (qualitative) — for categorical groups") +
  theme_minimal(base_size = 14) +
  theme(axis.text = element_blank(), axis.title = element_blank(), legend.position = "bottom")
```

## Colorblind-friendly choices

::: {.notes}
Emphasize that 8% of men is roughly 1 in 12 — in a class of 50, there are likely 2-3 students with some color vision deficiency. This is not an edge case. The practical advice is simple: use viridis or a tested palette, and never rely on color alone (also use shape or labels). Students often ask "how do I check?" — suggest the colorblindness simulator at davidmathlogic.com/colorblind or the `colorblindr` R package.
:::

About 8% of men have some form of color vision deficiency. Red-green is the most common.

```{r}
#| eval: false
# viridis is colorblind-safe AND sequential
scale_fill_viridis_d()   # discrete
scale_fill_viridis_c()   # continuous

# ColorBrewer palettes designed for colorblindness
scale_fill_brewer(palette = "Set2")

# Or set colors manually with safe choices
scale_fill_manual(values = c("#0072B2", "#E69F00", "#009E73"))
# (blue, orange, green — distinguishable for most color vision types)
```

## A bad color choice vs a good one

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| fig-alt: "Side-by-side comparison of two boxplots: the left uses red and green fills that are indistinguishable to colorblind viewers, while the right uses blue and orange fills that are accessible to virtually all viewers."

p1 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Control" = "#ff0000", "Treatment" = "#00aa00")) +
  labs(title = "Red vs Green — bad for colorblind viewers", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p2 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Control" = "#0072B2", "Treatment" = "#E69F00")) +
  labs(title = "Blue vs Orange — works for everyone", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p1 + p2
```

::: {.notes}
Let the side-by-side comparison speak for itself. The blue-orange version on the right is distinguishable to virtually all viewers. You can mention that some journals now require colorblind-accessible figures. The hex codes shown earlier (#0072B2, #E69F00, #009E73) are from the Okabe-Ito palette — a safe default students can copy into any project.
:::

## Why cluttered figures fail: cognitive load theory

::: {.notes}
Cognitive load theory (Sweller, 1988) was developed to explain why instructional design matters — and it translates directly to data visualization. The three-type model gives students a vocabulary for critiquing figures: "the gridlines add extraneous load," "the uninformative title doesn't build germane load." Spend ~3 minutes here, then connect forward: the declutter checklist on the next slides is how we operationalize this theory. Students who have taken cognitive psych or educational psych will recognize this framework.
:::

Cognitive load theory (Sweller, 1988) says mental effort has three components:

::: {.columns}
::: {.column width="33%"}
**Intrinsic load**

The complexity of the data itself — you can't reduce it
:::
::: {.column width="33%"}
**Extraneous load**

Clutter, noise, redundant elements — *this is what we control*
:::
::: {.column width="33%"}
**Germane load**

The insight the reader builds — what we're trying to maximize
:::
:::

. . .

Every gray background, redundant legend, and unlabeled variable is **extraneous load** — it taxes working memory without adding information.

**Design goal: minimize extraneous, maximize germane.**

# Decluttering {background-color="#2c3e50"}

## The default is cluttered

::: {.notes}
Transition: "Now that we know what to show, let's talk about what to remove." This section draws heavily on Knaflic's "declutter" philosophy. The before/after comparison here is dramatic — the gray background, heavy gridlines, and redundant legend all add zero information. Students sometimes resist removing gridlines because they think figures look "empty" without them. Remind them: if the reader needs a gridline to read the value, add a direct label instead.
:::

ggplot2's default theme adds a lot of visual noise. Compare:

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| fig-alt: "Side-by-side comparison of the same boxplot using ggplot2's default gray theme on the left versus the cleaner minimal theme on the right, showing how removing the gray background, heavy gridlines, and redundant legend reduces visual clutter."

p1 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot() +
  labs(title = "Default theme (theme_gray)", x = "Condition", y = "RT (ms)")

p2 <- reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_boxplot() +
  labs(title = "Cleaned up (theme_minimal)", x = "Condition", y = "RT (ms)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

p1 + p2
```

## The declutter checklist

::: {.notes}
This is a reference slide students should internalize. Walk through the "Remove" column and ask students to justify each removal — e.g., "Why remove the legend when we have two conditions?" (Because the x-axis already labels them.) The "Keep" column matters too: titles should state a finding, not label the chart type. "Boxplot of RT" is a label; "Treatment reduces reaction time" is a finding.
:::

Ask: does this element help the reader understand the data?

::: {.columns}
::: {.column width="50%"}
**Remove**

| Element | Why |
|---------|-----|
| Gray background | Noise, no info |
| Gridlines (most) | Distraction |
| Redundant legend | X-axis says it |
| Generic axis labels | Add units instead |
:::
::: {.column width="50%"}
**Keep**

| Element | Why |
|---------|-----|
| Title + subtitle | Orients the reader |
| Caption | Source, N, error bar type |
| Meaningful color | Highlights comparisons |
:::
:::

## Meet the `element_*` family

::: {.notes}
Before showing a full theme() call, introduce the vocabulary. Every non-data piece of a ggplot — the title, the axis ticks, the panel background, the gridlines — is one of four kinds of thing: text, a line, a rectangle, or blank (invisible). Once students internalize this, theme() stops feeling random. Spend 2–3 minutes on this slide. Ask: "The axis label — is that text, a line, or a rectangle?" (Text.) "The panel background?" (Rectangle.) "The gridlines?" (Lines.) That builds the mental model.
:::

Every non-data piece of a ggplot is one of four kinds of thing:

| Function | Key arguments | Used for |
|----------|---------------|----------|
| `element_text()` | `size`, `face`, `color`, `hjust` | Titles, axis text, legend text |
| `element_line()` | `color`, `linewidth` | Axis lines, gridlines, ticks |
| `element_rect()` | `fill`, `color` | Panel background, legend box |
| `element_blank()` | — | Removes the element entirely |

## `theme()` puts elements to use

::: {.notes}
Now that students know the four element types, theme() makes sense: you name a piece of the plot on the left side of `=`, and assign an element function on the right to style it. Walk through each line of the example and ask students to identify which element type is being used and why. The `?theme` tip is important — there are ~90 arguments and no one memorizes them all; students should know where to look.
:::

Inside `theme()`, name the piece of the plot, then assign an `element_*` function to style it:

```{r}
#| eval: false
theme(
  plot.title      = element_text(face = "bold", size = 16),  # text → bold, larger
  panel.grid      = element_line(color = "gray90"),           # lines → subtle gray
  plot.background = element_rect(fill = "white"),             # rect → white background
  axis.ticks      = element_blank(),                          # blank → remove ticks
  legend.position = "none"                                    # special: just "none"/"top"/etc.
)
```

Not sure what an argument is called? Run `?theme` — there are ~90 options.

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

::: {.notes}
Give 10 minutes. The cluttered graph on the next slide has at least 7 problems: red-green colors, size mapped to accuracy (binary variable — misleading), no informative title, variable names as axis labels, default gray theme, redundant color and fill, and points plotted behind the boxplot. Circulate and ask pairs to name each problem before they start fixing it. Common student approach: they'll fix the colors first because it's most obvious. Push them to also address the title and labels.
:::

I'll put a cluttered graph on screen. With a partner, **rewrite it** to follow the design principles we just covered.

Remove at least 5 unnecessary elements. Make it tell a clear story.

::: {.callout-tip}
Think about: colors, legend, labels, theme, size mapping, and whether every aesthetic is adding information.
:::

## Your turn: 10 minutes

First, run this to create the dataset:

```{r}
#| code-fold: true
#| code-summary: "Data setup (click to expand)"
set.seed(123)
reaction_data <- tibble(
  participant = rep(1:40, each = 2),
  condition   = rep(c("Control", "Treatment"), 40),
  rt          = c(rnorm(40, mean = 520, sd = 60), rnorm(40, mean = 480, sd = 55)),
  accuracy    = c(rbinom(40, 1, 0.82), rbinom(40, 1, 0.88)),
  age_group   = rep(c("Young", "Young", "Older", "Older"), 20)
) |>
  mutate(rt = round(rt, 1))
```

Now fix this:

```{r}
#| fig-alt: "Deliberately cluttered scatterplot with red and green colors, points sized by a binary variable, redundant color and fill mappings, uninformative title, and the default gray theme — students are asked to redesign it."
# The cluttered version — fix this!
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition, color = condition, size = accuracy)) +
  geom_point() +
  geom_boxplot(alpha = 0.3) +
  scale_fill_manual(values = c("Control" = "red", "Treatment" = "green")) +
  scale_color_manual(values = c("Control" = "red", "Treatment" = "green")) +
  labs(x = "condition", y = "rt", title = "data") +
  theme_gray()
```



```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = condition)) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 2, color = "gray50") +
  geom_boxplot(alpha = 0.5, width = 0.4, outlier.shape = NA) +
  scale_fill_manual(values = c("Control" = "#0072B2", "Treatment" = "#E69F00")) +
  labs(
    title = "Treatment Reduces Reaction Time",
    x = "Condition",
    y = "Reaction time (ms)",
    caption = "N = 40 per condition. Points = individual observations."
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid = element_blank())
```

---

## Before we move on

📤 **[Submit your code on Canvas](https://canvas.uoregon.edu/courses/287793/pages/pair-coding-check-ins)** for participation credit. Paste what you have — it doesn't need to work perfectly.

# Critiquing bad graphs {background-color="#2c3e50"}

::: {.notes}
This section (~10 min) is the second lecture block. It transitions from "how to build good figures" to "how to spot bad ones." These three examples each illustrate a different category of problem: misleading axes, poor aesthetics, and wrong chart type. Run through them interactively — show the figure, let students diagnose it, then reveal the answer. This builds the critical evaluation skill they'll need for Assignment 4.
:::

## What's wrong here? (1 of 3)

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Bar chart comparing mean reaction times for Control and Treatment conditions where the y-axis starts at 478 instead of 0, making a small ~40 ms difference appear dramatically large."
# Bad: Truncated y-axis exaggerates difference
reaction_data |>
  group_by(condition) |>
  summarize(mean_rt = mean(rt)) |>
  ggplot(aes(x = condition, y = mean_rt, fill = condition)) +
  geom_col() +
  scale_fill_manual(values = c("Control" = "grey60", "Treatment" = "grey40")) +
  coord_cartesian(ylim = c(478, 522)) +
  labs(title = "Treatment Reduces Reaction Time",
       y = "Mean RT (ms)", x = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

. . .

**The problem:** The y-axis starts at 478, not 0. The difference looks massive — but it's only ~40 ms. A truncated axis exaggerates the effect.

::: {.notes}
Students often ask: "But isn't it sometimes okay to not start at zero?" Good question. For bar charts, no — the bar height represents the value, so truncating is always misleading. For line charts or dot plots, it can be acceptable because the position (not the area) encodes the value. This nuance matters and comes up in the storytelling session (Session 16) too.
:::

## What's wrong here? (2 of 3)

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Boxplot with red and green fills, a yellow background, heavy gridlines, variable names as axis labels, and the uninformative title 'Boxplot' — demonstrating multiple common design mistakes at once."
# Bad: Rainbow color + too many gridlines + no informative title
reaction_data |>
  ggplot(aes(x = condition, y = rt, fill = age_group)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Young" = "red", "Older" = "green")) +
  labs(title = "Boxplot", y = "rt", x = "condition") +
  theme(
    panel.background = element_rect(fill = "lightyellow"),
    panel.grid.major = element_line(color = "grey50"),
    panel.grid.minor = element_line(color = "grey70")
  )
```

. . .

**The problems:** Red-green palette (colorblind-unfriendly). Title says "Boxplot" (a label, not a finding). Variable names as labels. Distracting background color and gridlines.

## What's wrong here? (3 of 3)

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Pie chart showing reaction times arbitrarily binned into Fast, Medium, and Slow categories, hiding the continuous distribution, individual observations, and any comparison between conditions."
# Bad: Pie chart of continuous data categories
reaction_data |>
  mutate(speed = case_when(
    rt < 450 ~ "Fast",
    rt < 550 ~ "Medium",
    TRUE ~ "Slow"
  )) |>
  count(speed) |>
  ggplot(aes(x = "", y = n, fill = speed)) +
  geom_col(width = 1) +
  coord_polar("y") +
  labs(title = "Speed Categories") +
  theme_void()
```

. . .

**The problems:** Continuous data was binned into arbitrary categories, then displayed as a pie chart. We lost the actual reaction times, the condition comparison, and the ability to see distributions. A histogram or density plot would show far more.

::: {.notes}
This is the strongest example because it shows a cascade of bad decisions: first they binned continuous data (losing information), then they used a pie chart (making comparison harder), and they dropped the condition variable entirely. Each step destroyed information. The lesson: your choice of chart type determines what story you can tell. A pie chart here makes it impossible to compare conditions or see the distribution.
:::

# Get a head start {background-color="#e67e22"}

## Assignment 4 preview

::: {.notes}
Assignment 4 is being assigned today. Briefly outline the three parts so students know what to aim for. Emphasize that the "bad" version should be deliberately bad with specific design violations they can name — not just ugly. They need to articulate what's wrong with it and why the good version is better. The colorblind-accessible version is a separate deliverable, not just "the good version with viridis."
:::

Assignment 4 will ask you to:

1. Create a **"bad" version** of a figure — deliberately violate design principles
2. Create a **"good" version** following what we covered today
3. Create a **colorblind-accessible version**

## Assignment 4 preview

Start experimenting now:

- Take the `reaction_data` dataset
- Make the worst possible version of a figure
- Then make it great
- What did you change?

# Wrapping up {background-color="#2c3e50"}

## Design principles checklist

- [ ] Is the geom appropriate for the data type?
- [ ] Are colors colorblind-friendly?
- [ ] Is the legend necessary? (Remove if redundant)
- [ ] Are axis labels informative (not just variable names)?
- [ ] Is there a title that states the main point?
- [ ] Is there visual clutter to remove?
- [ ] Do error bars have a caption explaining what they show?

## Before next class

📖 **Read:**

- [R4DS Ch 10: Exploratory data analysis](https://r4ds.hadley.nz/eda) (sections 10.1–10.4)

✅ **Practice:**

- Try the "bad vs good" exercise on your own
- Find a graph online and identify ways to improve it

## Key takeaways

1. **Preattentive processing** — your brain detects color, size, and position before conscious thought
2. **Working memory** — limit categories to 4–5; every aesthetic uses a slot
3. **Gestalt principles** — proximity, similarity, and enclosure create grouping automatically
4. **Cognitive load** — minimize extraneous clutter so the insight can land
5. **Palette choice** — match sequential, diverging, or qualitative to your data type
6. **Design for colorblindness** — always; 1 in 12 men has color vision deficiency

These are not design opinions — they are perceptual psychology.

## The one thing to remember

You're not designing for yourself. You're designing for a reader who will look at your figure for five seconds.

Next time: Exploratory Data Analysis

::: {.notes}
End on this line — it reframes everything from today. Students naturally design for themselves (they already know what the data shows). The challenge is designing for someone who doesn't. Five seconds is roughly how long a reader spends on a figure before deciding whether to engage. Next session shifts from how to present data to how to explore it (EDA). Remind students that Assignment 3 is due this week and Assignment 4 is now available.
:::

