---
title: "EDA — Variation"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 29, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 10"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(psych)

# Load the Big Five personality dataset
data(bfi)
bfi <- as_tibble(bfi)

# Load our class survey data
class_survey <- read_csv("../data/class_survey.csv")
```

# What is EDA? {background-color="#2c3e50"}

## Know your data before you test it

Before you run a single statistical test, you should know your data well enough to predict what the results will look like.

. . .

That's what EDA gives you — **no surprises**.

::: {.notes}
Open by asking students: "How many of you have ever run an analysis and been confused by the results?" This is the problem EDA solves. Spend ~2 minutes here. The goal is to frame EDA not as busywork but as something that saves time and prevents mistakes.
:::

## Exploratory vs. confirmatory

| | Exploratory (EDA) | Confirmatory |
|---|---|---|
| **Goal** | Discover patterns | Test hypotheses |
| **Attitude** | Curiosity | Rigor |
| **Questions** | Open-ended | Pre-specified |
| **Output** | Visualizations, hunches | p-values, conclusions |

EDA comes **first.** You can't test a hypothesis you didn't notice.

::: {.notes}
Students sometimes think EDA is optional or that it's what you do when you don't have a hypothesis. Emphasize that even in confirmatory research, EDA is step one. It tells you whether your data even look the way you expect. If they've taken stats, connect to the idea of checking assumptions.
:::

## Good EDA means looking before you test

> "EDA is an attitude, not a technique."  
> — John Tukey

- Ask questions
- Answer them by visualizing data
- Use what you learn to ask new questions
- Repeat

There is no single "right" way to do EDA. The goal is **understanding.**

::: {.notes}
The Tukey quote is worth pausing on. Students who are used to following step-by-step procedures may feel anxious that there's no single recipe. Reassure them: EDA is more like being a detective than following a protocol. The "attitude" framing gives them permission to explore.
:::

## What we're doing today

Exploring **variation** — what does a single variable look like?

- Distribution shape
- Center and spread
- Outliers and unusual values
- Missing data

Tomorrow: **covariation** — how do variables relate to each other?

::: {.notes}
Quick roadmap slide — spend ~30 seconds. This orients students on the two-session arc: today is single variables, next class is relationships between variables.
:::

## Our dataset: Big Five Personality

```{r}
glimpse(bfi)
```

25 items measuring five personality factors (Agreeableness, Conscientiousness, Extraversion, Neuroticism, Openness) + demographics.

::: {.notes}
Walk through the glimpse() output. Point out the variable names, number of rows (~2800), and that some columns are integers (survey items on a 1-6 scale) while age is continuous. Students often ask what A1, C3, etc. stand for — the next slide covers that. If they've taken personality psych, ask what the Big Five are.
:::

## What's in here

```{r}
#| echo: false
tibble(
  Variable = c("A1–A5", "C1–C5", "E1–E5", "N1–N5", "O1–O5", "gender", "education", "age"),
  Description = c(
    "Agreeableness items (1–6 scale)",
    "Conscientiousness items (1–6 scale)",
    "Extraversion items (1–6 scale)",
    "Neuroticism items (1–6 scale)",
    "Openness items (1–6 scale)",
    "1 = male, 2 = female",
    "1–5 (HS incomplete through graduate)",
    "Age in years"
  )
) |> knitr::kable()
```

# Exploring distributions {background-color="#2c3e50"}

## Start simple: summary statistics

```{r}
bfi |>
  select(age, gender, education) |>
  summary()
```

But summary stats can hide a lot. Always visualize.

::: {.notes}
Run this live and walk through the output. Point out the NA counts in the summary — students will see missing data here for the first time. Ask: "What does this summary tell you about education? Is it really a number?" This sets up the idea that summary stats on categorical-coded-as-numeric variables are misleading. Spend ~2 minutes.
:::

## Histograms: the workhorse

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Histogram of participant ages with 5-year bins showing a strong right skew, with most participants in their teens to thirties and a long tail extending past age 60."
bfi |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Age",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

::: {.notes}
This is the first histogram of the session. Walk through the code line by line: the pipe into ggplot, the aes() mapping, and then geom_histogram(). Emphasize that binwidth is a choice, not a default — ggplot will warn them if they don't set it. Ask students to describe the shape: right-skewed, most people are young. Spend ~3 minutes here since this is a core geom.
:::

## Choosing binwidth

The binwidth matters a lot:

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
#| fig-alt: "Three histograms of age side by side comparing binwidths. Binwidth 1 is noisy with erratic spikes, binwidth 5 clearly shows the right-skewed shape, and binwidth 20 oversimplifies into just a few bars that hide the skew."
library(patchwork)

p1 <- bfi |> ggplot(aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "binwidth = 1 (too narrow)", x = "Age") +
  theme_minimal(base_size = 13)

p2 <- bfi |> ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "binwidth = 5 (good)", x = "Age") +
  theme_minimal(base_size = 13)

p3 <- bfi |> ggplot(aes(x = age)) +
  geom_histogram(binwidth = 20, fill = "steelblue", color = "white") +
  labs(title = "binwidth = 20 (too wide)", x = "Age") +
  theme_minimal(base_size = 13)

p1 + p2 + p3
```

There's no "right" answer — try a few and pick the one that shows the shape clearly.

::: {.notes}
Spend ~2 minutes comparing the three panels. Point to specific differences: binwidth=1 shows noisy spikes that are just sampling variability, binwidth=20 hides the right skew entirely. A common student question: "Which binwidth should I use on the assignment?" Answer: try a few, pick the one that shows the overall shape without too much noise. There is no single correct answer.
:::

## Density plots: smooth alternative

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Density plot of participant ages showing a smooth right-skewed curve that peaks around age 20 and gradually tapers off toward age 80."
bfi |>
  ggplot(aes(x = age)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(
    title = "Distribution of Age (Density)",
    x = "Age (years)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)
```

::: {.notes}
Briefly explain that density plots are like a smoothed histogram — the y-axis is density, not count, so the area under the curve sums to 1. Students often ask what the y-axis means; it's fine to say "think of it as relative likelihood" without going deep into probability density. The main advantage: no binwidth decision. The main disadvantage: harder to read exact counts.
:::

## Exploring survey items

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Grouped bar chart of responses to five Extraversion items on a 1 to 6 scale. Items show different response patterns, with some skewing toward higher values and others toward lower values, reflecting reverse-scored items."
# All Extraversion items
bfi |>
  select(E1:E5) |>
  pivot_longer(everything(), names_to = "item", values_to = "response") |>
  ggplot(aes(x = response, fill = item)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Extraversion Item Responses",
    x = "Response (1–6)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

::: {.notes}
This slide introduces pivot_longer in a practical context — students saw it in the tidying session but may need a reminder. The key insight here is that some items look different from others: E1 and E2 might have different distributions because some items are reverse-scored. If a student notices this, it's a great teaching moment about reverse coding in personality measures. Don't dwell on pivot_longer mechanics; the focus is the visualization.
:::

## Categorical variables

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Bar chart of gender distribution showing roughly twice as many female participants as male participants in the BFI sample."
bfi |>
  mutate(gender = ifelse(gender == 1, "Male", "Female")) |>
  ggplot(aes(x = gender)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Gender Distribution",
    x = "Gender",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## Counts and proportions

```{r}
# Counts
bfi |>
  mutate(gender = ifelse(gender == 1, "Male", "Female")) |>
  count(gender)
```

## Counts and proportions

```{r}
# Proportions
bfi |>
  mutate(gender = ifelse(gender == 1, "Male", "Female")) |>
  count(gender) |>
  mutate(proportion = round(n / sum(n), 3))
```

# Outliers {background-color="#2c3e50"}

## What are outliers?

Values that are unusual compared to the rest of the data. They might be:

- **Real** — genuinely extreme values (a 95-year-old in a college study)
- **Errors** — data entry mistakes (age = 999)
- **Interesting** — worth investigating

**Never delete an outlier without understanding it.**

::: {.notes}
This is one of the most important slides in the deck. Spend ~3 minutes here. Students in psych methods courses are often taught to "remove outliers" as a default step. Push back on that: outliers are data, and deleting them is a decision that changes your results. The real question is always "is this value plausible?" A 95-year-old in a college study is real but might not represent your target population. An age of 999 is clearly an error. These are different situations requiring different responses.
:::

## Spotting outliers visually

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Histogram of participant ages with 5-year bins highlighting the right-skewed distribution, with a handful of participants beyond age 60 that could be considered outliers."
bfi |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Age Distribution — Any Outliers?",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## Spotting outliers with boxplots

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 4
#| fig-alt: "Horizontal boxplot of participant ages with the median around age 28, the box spanning roughly 20 to 40, and individual points plotted beyond the upper whisker indicating older outliers."
bfi |>
  ggplot(aes(x = age)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Age — Boxplot Shows Outliers as Points",
    x = "Age (years)"
  ) +
  theme_minimal(base_size = 14)
```

Points beyond the whiskers are flagged as outliers by the 1.5×IQR rule.

::: {.notes}
Briefly explain the 1.5xIQR rule: any point more than 1.5 times the interquartile range above Q3 or below Q1 gets plotted individually. Students don't need to memorize this, just know that boxplots have a built-in outlier detection rule. Transition: "But seeing a dot on a plot isn't enough — let's look at those outliers in the actual data."
:::

## Investigating outliers programmatically

```{r}
# Find unusually old or young participants
bfi |>
  filter(age > 60) |>
  select(age, gender, education)
```

## What to do with outliers

| Outlier type | What to do |
|---|---|
| Likely error (age = 999) | Fix or set to NA |
| Real but extreme | Note it, keep it, mention in write-up |
| Suspicious | Investigate further |
| Doesn't affect conclusions | Keep it, move on |

**Document your decisions.** Future you will want to know.

::: {.notes}
Emphasize the documentation point. In psychology research, you need to justify every data cleaning decision in your methods section. "I removed 3 participants with ages over 80 because the study targeted 18-65 year olds" is fine. "I removed outliers" with no explanation is not. This connects to the reproducibility theme from Session 1.
:::

# Let's explore YOUR data {background-color="#2c3e50"}

## Our class survey

Remember that survey you filled out Week 1? Let's do some EDA on it.

```{r}
glimpse(class_survey)
```

::: {.notes}
Transition: "We've been using the BFI dataset to practice EDA. Now let's see what our own class data looks like." Students perk up when it's their data. Walk through the glimpse output briefly — they should recognize the variable names from the survey they took.
:::

## How many tabs do you have open?

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Histogram of browser tabs open showing a right-skewed distribution, with most students reporting fewer than 20 tabs and a few extreme values stretching the distribution."
ggplot(class_survey, aes(x = tabs_open)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Browser Tabs Open During the Survey",
    x = "Number of tabs",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

::: {.notes}
This variable almost always produces a beautiful right skew with outliers. Ask: "Who's the person with 80+ tabs?" Let students laugh. Then pivot: "Is that an outlier? Is it an error? Or is it just someone who doesn't close tabs?" This reinforces the earlier lesson — outliers are information, not mistakes.
:::

## How does our class sleep?

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Histogram of average hours of sleep per night reported by students, showing the distribution centered around 7 hours."
ggplot(class_survey, aes(x = sleep_hrs)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "How Much Sleep Does Our Class Get?",
    x = "Average hours of sleep per night",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

::: {.notes}
Compare this to the tabs distribution: "Notice how different the shape is. Sleep is roughly symmetric — most people cluster around 7 hours. Tabs was heavily skewed. That's EDA in action — you're noticing things."
:::

## Morning people or night owls?

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Bar chart of chronotype distribution showing counts of students who identify as morning person, evening person, or neither."
ggplot(class_survey, aes(x = chronotype)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Chronotype Distribution in Our Class",
    x = "Chronotype",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

::: {.notes}
A categorical variable from our own data. Ask: "Does this match what you'd expect from a class of college students?" Most college samples skew evening. Transition to the pair coding break.
:::

# Pair coding break {background-color="#e67e22"}

## Your turn: 10 minutes

Using the `penguins` dataset (from the `palmerpenguins` package):

1. Plot the distribution of **flipper_length_mm** — what shape is it?
2. Check for **outliers** in flipper length. Are any values suspicious?

::: {.callout-tip}
Try both a histogram and a boxplot. They show different things. Use `summary()` first to get oriented.
:::

::: {.notes}
Circulate during this exercise. Common issues: students forget to load palmerpenguins, or they get a warning about removed rows (that's the missing data — point it out as a natural segue to the next section). The flipper length distribution is bimodal — this is a great moment to ask "why might there be two peaks?" (Answer: different species have different flipper lengths.) If pairs finish early, suggest they try body_mass_g or bill_length_mm.
:::

```{r}
#| echo: false
#| eval: false
# Solution (reveal after time is up)
library(palmerpenguins)
penguins |> ggplot(aes(x = flipper_length_mm)) +
  geom_histogram(fill = "steelblue", color = "white") +
  theme_minimal()

# Outlier check
penguins |> ggplot(aes(x = flipper_length_mm)) + geom_boxplot() + theme_minimal()
```

---

## Before we move on

📤 **[Submit your code on Canvas](https://canvas.uoregon.edu/courses/287793/pages/pair-coding-check-ins)** for participation credit. Paste what you have — it doesn't need to work perfectly.

# Missing data — a first look {background-color="#2c3e50"}

## Missing data is everywhere

```{r}
# How much missing data do we have?
bfi |>
  summarize(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  arrange(desc(n_missing))
```

::: {.notes}
This code is more advanced than what students have written themselves — across() and the anonymous function syntax may be new. Focus on the output, not the code mechanics. The takeaway: education has the most missing data (~200+ values). Ask: "Why might education be more likely to be missing than personality items?" (People skip demographic questions they find irrelevant or sensitive.)
:::

## Visualizing missingness

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5
#| fig-alt: "Horizontal bar chart showing the percentage of missing values for each variable that has any missingness. Education has the highest rate at around 7 percent, followed by several personality items with lower rates."
bfi |>
  summarize(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  mutate(pct_missing = pct_missing * 100) |>
  filter(pct_missing > 0) |>
  arrange(desc(pct_missing)) |>
  ggplot(aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Percentage of Missing Values by Variable",
    x = "Variable",
    y = "% Missing"
  ) +
  theme_minimal(base_size = 14)
```

## Why it matters for EDA

Missing data can bias your exploration:

- If missingness is related to the variables you're studying, your distributions are wrong
- If certain groups are more likely to have missing data, you might miss important patterns

We'll cover this in depth in Session 14. For now: **always check for missing data before exploring.**

::: {.notes}
Keep this brief — ~1 minute. The point is to plant a seed, not to teach missing data handling. Students just need to know that missing data exists and can bias their exploration. The full treatment comes in Session 14 (Missing Data).
:::

# The EDA workflow {background-color="#2c3e50"}

## A systematic approach

1. **Look at the structure** — `glimpse()`, `summary()`
2. **Check for missing data** — how much? Which variables?
3. **Explore each variable** — histograms, bar charts, summaries
4. **Look for outliers** — boxplots, programmatic checks
5. **Ask questions** — what surprised you? What do you want to know more about?
6. **Repeat** — EDA is iterative

::: {.notes}
This is the "cheat sheet" slide — the workflow students should follow when they sit down with a new dataset. Emphasize that step 6 is the most important: EDA is a loop, not a checklist. Each answer generates new questions. This workflow will be the backbone of Assignment 5. Spend ~2 minutes walking through the steps with the BFI data as a running example.
:::

## Putting it together: a mini-EDA

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 6
#| fig-alt: "Bar chart of education levels in the BFI sample showing the most common category is some college, followed by college graduate and high school graduate, with fewer participants at the extremes of high school incomplete and graduate degree."
# Education distribution — what does it look like?
bfi |>
  filter(!is.na(education)) |>
  mutate(education = factor(education, labels = c(
    "HS incomplete", "HS grad", "Some college",
    "College grad", "Graduate degree"
  ))) |>
  ggplot(aes(x = education)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Education Level in the BFI Sample",
    x = "Education level",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

# Get a head start {background-color="#e67e22"}

## Assignment 5 preview

Assignment 5 will have you do a full EDA on a psychology dataset. Start now:

1. Load `bfi` and run `glimpse()` and `summary()`
2. Pick **3 variables** and create distributions for each
3. For each one: What shape is it? Any outliers? Any missing data?
4. Write down **one question** each distribution makes you want to ask

That's the core of EDA — noticing things and getting curious.

::: {.notes}
Give students ~10 minutes of work time here. Frame this as a head start on Assignment 5, not as something they need to finish now. Circulate and help with code issues. If students are stuck on which 3 variables to pick, suggest one continuous (age), one discrete (a personality item), and one categorical (gender or education).
:::

# Wrapping up {background-color="#2c3e50"}

## EDA toolkit so far

| Tool | When to use |
|------|-------------|
| `glimpse()` | First look at structure |
| `summary()` | Quick numeric summaries |
| `geom_histogram()` | Continuous distributions |
| `geom_density()` | Smooth distribution shape |
| `geom_bar()` | Categorical distributions |
| `geom_boxplot()` | Outlier detection |
| `count()` | Frequency tables |
| `is.na()` / `sum(is.na())` | Missing data check |

## Before next class

📖 **Read:**

- [R4DS Ch 10: Exploratory data analysis](https://r4ds.hadley.nz/eda) (sections 10.5–10.6)

✅ **Practice:**

- Explore at least 3 variables in `bfi`
- Look for outliers and missing data
- Start generating questions

## Key takeaways

1. **EDA is an attitude** — curiosity, not confirmation
2. **Always visualize** — summary stats hide patterns
3. **Outliers are information** — investigate, don't delete
4. **Check missing data first** — it affects everything
5. **Ask questions, then answer them** — that's the loop

## The one thing to remember

EDA isn't a step you finish. It's the habit of looking at your data before you believe anything about it.

Next time: EDA — Covariation

::: {.notes}
End on this line. Remind students that the reading for next time is R4DS Ch 10 sections 10.5-10.6 (covariation). No assignment is due or assigned today, but the Final Project Proposal is due soon — check the schedule. Next session picks up exactly where this one leaves off.
:::
