---
title: "Exploratory Data Analysis: Covariation"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "April 29, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 10"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Setup {visibility="hidden"}

```{r}
#| include: false
library(tidyverse)
library(ggrain)
library(patchwork)
```

# From variation to covariation {background-color="#2c3e50"}

## Recap: Variation

Last session we explored **variation** — how values vary within a single variable:

- Distributions of continuous variables (histograms, density plots)
- Counts of categorical variables (bar charts)
- Identifying typical values and outliers

## Today: Covariation

**Covariation** is the tendency for values of two or more variables to vary together in a related way.

. . .

In psychology, we care deeply about covariation:

- Does anxiety vary with depression?
- Do reaction times vary across different conditions?
- Does age relate to memory performance?

# Categorical + Continuous {background-color="#2c3e50"}

## Example: Mental health by treatment group

```{r}
# Simulated therapy outcome data
therapy_data <- tibble(
  condition = rep(c("Control", "CBT", "Mindfulness"), each = 50),
  depression_post = c(
    rnorm(50, mean = 18, sd = 5),  # Control
    rnorm(50, mean = 12, sd = 5),  # CBT
    rnorm(50, mean = 14, sd = 5)   # Mindfulness
  )
)

```

## Example: Mental health by treatment group

```{r}
glimpse(therapy_data)
```

## Boxplots: The classic choice

```{r}
#| label: boxplot
#| output-location: slide
ggplot(therapy_data, aes(x = condition, y = depression_post)) +
  geom_boxplot() +
  labs(
    title = "Post-treatment depression scores by condition",
    x = "Treatment condition",
    y = "Depression score (BDI-II)"
  ) +
  theme_minimal()
```

## What boxplots show

:::: {.columns}
::: {.column width="40%"}

```{r}
#| ref.label: boxplot
#| echo: false
```

:::

::: {.column width="60%"}
- **Line in middle:** median
- **Box:** 25th to 75th percentile (IQR)
- **Whiskers:** extend to 1.5 × IQR
- **Dots:** outliers beyond whiskers
:::
::::

. . .

Great for comparing distributions, but they hide the actual data points.

## The problem with boxplots

Boxplots summarize, but they hide important information:

- **The actual distribution shape** (is it bimodal? skewed?)
- **Individual data points** (how many observations are there?)
- **The raw data** (where do specific values fall?)

. . .

We can do better.

# Raincloud plots {background-color="#2c3e50"}

## The modern psych visualization

Raincloud plots combine three elements:

  1. **Violin** (distribution shape)
  2. **Boxplot** (summary stats)
  3. **Jittered points** (individual data)

They're increasingly popular in psychology publications because they show **everything**.

Use the `ggrain` package. ([More details here.](https://cran.r-project.org/web/packages/ggrain/vignettes/ggrain.html))

## Building a raincloud

```{r}
#| label: raincloud
#| fig-width: 10
#| fig-height: 6
#| echo: false

library(ggrain)
therapy_data |>
  ggplot(aes(
    x = condition, 
    y = depression_post, 
    fill = condition, 
    color = condition)) +
  # geom_rain creates all parts of your raincloud
  geom_rain(
    alpha = .6,
    # change just the boxplot part
    boxplot.args = list(color = "black")) + 
  # The mean
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "black") +
  scale_fill_manual(
    values = c("Control" = "#0072B2", "CBT" = "#E69F00", "Mindfulness" = "#009E73")) +
  scale_color_manual(
    values = c("Control" = "#0072B2", "CBT" = "#E69F00", "Mindfulness" = "#009E73")) +
  labs(
    title = "Post-Treatment Depression by Condition",
    subtitle = "Diamond = mean. Violin = distribution. Box = IQR. Dots = individual scores.",
    x = "Treatment condition",
    y = "Depression score (BDI-II)"
  ) +
  coord_flip() +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid = element_blank())
```

---

```{r}
#| ref.label: raincloud
#| eval: false
```

## Why rainclouds are great

Compare these three views of the **same data**:

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 7
library(patchwork)

p1 <- ggplot(therapy_data, aes(x = condition, y = depression_post)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Boxplot only", y = "Depression score") +
  theme_minimal()

p2 <- ggplot(therapy_data, aes(x = condition, y = depression_post)) +
  geom_violin(fill = "lightblue") +
  labs(title = "Violin only", y = "Depression score") +
  theme_minimal()

p3 <- ggplot(therapy_data, aes(x = condition, y = depression_post, fill = condition)) +
  geom_rain(alpha = 0.6) +
  scale_fill_manual(values = c("Control" = "#0072B2", "CBT" = "#E69F00", "Mindfulness" = "#009E73")) +
  labs(title = "Raincloud: all the information", y = "Depression score") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none")

p1 + p2 + p3 + plot_layout(ncol = 1)
```

## When to use which

| Plot type | Shows | Best for |
|-----------|-------|----------|
| **Boxplot** | Median, IQR, outliers | Quick comparison, large datasets |
| **Raincloud** | Distribution + summary + raw data | Psychology papers, presentations, publications |
| **Violin + jitter** | Distribution + raw data | Alternative to raincloud, simpler to code |

# Pair coding break {background-color="#e67e22"}

## Your turn: Compare by gender

Using the therapy data, explore whether depression scores differ by gender:

1. Add a `gender` variable to the data (use `sample()` to randomly assign "Male", "Female", "Non-binary")
2. Create a visualization showing depression scores by gender
3. Try at least two different geom types
4. Add appropriate labels

**Time: 10 minutes**

```{r}
#| echo: false
#| eval: false
# SOLUTION
therapy_data_gender <- therapy_data |>
  mutate(gender = sample(c("Male", "Female", "Non-binary"),
                         size = n(),
                         replace = TRUE))

# Boxplot version
ggplot(therapy_data_gender, aes(x = gender, y = depression_post)) +
  geom_boxplot() +
  labs(
    title = "Depression scores by gender",
    x = "Gender",
    y = "Depression score (BDI-II)"
  ) +
  theme_minimal()

# Violin + jitter version
ggplot(therapy_data_gender, aes(x = gender, y = depression_post)) +
  geom_violin() +
  geom_jitter(width = 0.1, alpha = 0.3) +
  labs(
    title = "Depression scores by gender",
    x = "Gender",
    y = "Depression score (BDI-II)"
  ) +
  theme_minimal()
```

# Categorical + Categorical {background-color="#2c3e50"}

## Example: Diagnosis by gender

```{r}
# Simulated diagnostic data
diagnosis_data <- tibble(
  gender = sample(c("Male", "Female", "Non-binary"), 200, replace = TRUE),
  diagnosis = sample(c("Depression", "Anxiety", "Both", "Other"),
                     200, replace = TRUE)
)

head(diagnosis_data)
```

## Option 1: Stacked bar chart

```{r}
#| output-location: slide
ggplot(diagnosis_data, aes(x = gender, fill = diagnosis)) +
  geom_bar() +
  labs(
    title = "Diagnosis by gender",
    x = "Gender",
    y = "Count",
    fill = "Diagnosis"
  ) +
  theme_minimal()
```

## Option 2: Side-by-side bars

```{r}
#| output-location: slide
ggplot(diagnosis_data, aes(x = gender, fill = diagnosis)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Diagnosis by gender",
    x = "Gender",
    y = "Count",
    fill = "Diagnosis"
  ) +
  theme_minimal()
```

## Option 3: Proportions

```{r}
#| output-location: slide
ggplot(diagnosis_data, aes(x = gender, fill = diagnosis)) +
  geom_bar(position = "fill") +
  labs(
    title = "Diagnosis distribution by gender",
    x = "Gender",
    y = "Proportion",
    fill = "Diagnosis"
  ) +
  theme_minimal()
```

## Option 4: geom_count()

Shows the *size* of overlaps:

```{r}
#| output-location: slide
ggplot(diagnosis_data, aes(x = gender, y = diagnosis)) +
  geom_count() +
  labs(
    title = "Diagnosis by gender",
    x = "Gender",
    y = "Diagnosis",
    size = "Count"
  ) +
  theme_minimal()
```

## Option 5: Tile with counts

```{r}
#| output-location: slide
diagnosis_counts <- diagnosis_data |>
  count(gender, diagnosis)

ggplot(diagnosis_counts, aes(x = gender, y = diagnosis, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "Diagnosis by gender",
    x = "Gender",
    y = "Diagnosis",
    fill = "Count"
  ) +
  theme_minimal()
```

# Continuous + Continuous {background-color="#2c3e50"}

## The classic: Scatterplots

```{r}
# Simulated reaction time data
rt_data <- tibble(
  age = runif(100, 18, 70),
  reaction_time = 200 + age * 3 + rnorm(100, 0, 40)
)

glimpse(rt_data)
```

## Basic scatterplot

```{r}
#| output-location: slide
ggplot(rt_data, aes(x = age, y = reaction_time)) +
  geom_point() +
  labs(
    title = "Reaction time by age",
    x = "Age (years)",
    y = "Reaction time (ms)"
  ) +
  theme_minimal()
```

## Add a trend line

```{r}
#| output-location: slide
ggplot(rt_data, aes(x = age, y = reaction_time)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Reaction time increases with age",
    x = "Age (years)",
    y = "Reaction time (ms)"
  ) +
  theme_minimal()
```

## The problem: Overplotting

With lots of data, points overlap and hide the true density:

```{r}
# Lots of data
big_rt_data <- tibble(
  age = runif(5000, 18, 70),
  reaction_time = 200 + age * 3 + rnorm(5000, 0, 40)
)
```

## Overplotting problem demonstrated

```{r}
#| output-location: slide
ggplot(big_rt_data, aes(x = age, y = reaction_time)) +
  geom_point() +
  labs(title = "Hard to see where the data is dense") +
  theme_minimal()
```

## Solution 1: Transparency (alpha)

```{r}
#| output-location: slide
ggplot(big_rt_data, aes(x = age, y = reaction_time)) +
  geom_point(alpha = 0.1) +
  labs(title = "Using alpha = 0.1 to show density") +
  theme_minimal()
```

## Solution 2: geom_bin2d()

```{r}
#| output-location: slide
ggplot(big_rt_data, aes(x = age, y = reaction_time)) +
  geom_bin2d() +
  scale_fill_viridis_c() +
  labs(
    title = "2D bins show density",
    fill = "Count"
  ) +
  theme_minimal()
```

## Solution 3: geom_hex()

```{r}
#| output-location: slide
ggplot(big_rt_data, aes(x = age, y = reaction_time)) +
  geom_hex() +
  scale_fill_viridis_c() +
  labs(
    title = "Hexagonal bins",
    fill = "Count"
  ) +
  theme_minimal()
```

## Correlation coefficient

A single number summary of the linear relationship:

```{r}
cor(rt_data$age, rt_data$reaction_time)
```

. . .

::: {.callout-note}
- **r = 1:** perfect positive relationship
- **r = 0:** no linear relationship
- **r = -1:** perfect negative relationship
:::

. . .

But always plot your data first! (See: Anscombe's Quartet)

# Patterns and models {background-color="#2c3e50"}

## What patterns tell us

When you see covariation, ask:

1. **Could it be coincidence?** (Maybe, especially with small samples)
2. **What's the mechanism?** (How are these variables related?)
3. **Is there a confound?** (Could a third variable explain both?)

. . .

::: {.callout-warning}
**Correlation ≠ Causation**

Covariation suggests a relationship, but doesn't prove one variable *causes* the other.
:::

## Real data example

Let's explore a real dataset: `mpg` (fuel economy data)

```{r}
glimpse(mpg)
```

## Question 1: Does engine size affect fuel economy?

```{r}
#| output-location: slide
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Larger engines are less fuel efficient",
    x = "Engine displacement (liters)",
    y = "Highway MPG"
  ) +
  theme_minimal()
```

## Question 2: Does this vary by car type?

```{r}
#| output-location: slide
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  labs(
    title = "Engine size vs fuel economy by car type",
    x = "Engine displacement (liters)",
    y = "Highway MPG",
    color = "Car type"
  ) +
  theme_minimal()
```

## Question 3: Which car types are most common?

```{r}
#| output-location: slide
ggplot(mpg, aes(y = fct_infreq(class))) +
  geom_bar() +
  labs(
    title = "SUVs are the most common vehicle type",
    x = "Count",
    y = "Car type"
  ) +
  theme_minimal()
```

# End-of-deck exercise {background-color="#e67e22"}

## Your final project proposal

For your final project, you'll need to:

1. Choose a dataset with at least 2-3 variables of interest
2. Formulate 2-3 research questions about relationships in the data
3. Plan visualizations to explore those relationships

**Exercise:** Start exploring potential datasets. Find one that interests you and create 2-3 exploratory visualizations showing different types of covariation (categorical + continuous, continuous + continuous, etc.).

This will form the basis of your proposal, **due today!**

```{r}
#| echo: false
#| eval: false
# SOLUTION (example using palmerpenguins)
library(palmerpenguins)

# Continuous + continuous
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Penguins with longer flippers tend to be heavier",
    x = "Flipper length (mm)",
    y = "Body mass (g)"
  ) +
  theme_minimal()

# Categorical + continuous
ggplot(penguins, aes(x = species, y = bill_length_mm)) +
  geom_violin() +
  geom_jitter(width = 0.1, alpha = 0.3) +
  labs(
    title = "Bill length varies by species",
    x = "Species",
    y = "Bill length (mm)"
  ) +
  theme_minimal()

# Categorical + categorical
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "fill") +
  labs(
    title = "Penguin species are found on different islands",
    x = "Species",
    y = "Proportion",
    fill = "Island"
  ) +
  theme_minimal()
```

# Wrapping up {background-color="#2c3e50"}

## Key takeaways

1. **Covariation = relationships** between variables
2. **Different plot types** for different variable combinations:
   - Categorical + continuous: boxplot, raincloud plots
   - Categorical + categorical: bar charts, tiles, geom_count()
   - Continuous + continuous: scatterplot, 2D bins, hex
3. **Raincloud plots** are the gold standard for psychology — they show distribution, summary stats, and raw data
4. **Watch for overplotting** — use alpha, jitter, or binning
5. **Always visualize first** before computing correlations
6. **Patterns suggest but don't prove** causation

## Before next class

📖 **Read:**

- R4DS Ch 12: Logical vectors
- R4DS Ch 13: Numbers

✅ **Do:**

- Submit Assignment 5
- Submit your Final Project Proposal (due today!)
- Start thinking about how you'll compute scale scores (next session)

## Questions?

See you Monday for data types and scale scoring!
