---
title: "Strings & Factors"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "May 6, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 12"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Setup {visibility="hidden"}

```{r}
#| include: false
library(tidyverse)
```

# Strings: The basics {background-color="#2c3e50"}

## What are strings?

**Strings** are text data — anything in quotes:

```{r}
participant_name <- "Jane Doe"
diagnosis <- "Major Depressive Disorder"
feedback <- "The task was confusing"
```

. . .

In R, the `stringr` package (part of tidyverse) provides tools for working with strings.

All `stringr` functions start with `str_`.

## Creating and combining strings

```{r}
first <- "Jane"
last <- "Doe"

# Combine strings
str_c(first, last)           # No space
str_c(first, last, sep = " ")  # With space
```

. . .

```{r}
# Combine with other text
str_c("Participant: ", first, " ", last)

# Glue is even easier (from the glue package)
library(glue)
glue("Participant: {first} {last}")
```

## String length

```{r}
responses <- c("Yes", "No", "Maybe", "I don't know")

str_length(responses)
```

. . .

Useful for checking free-response data quality:

```{r}
# Flag very short responses
tibble(responses) |>
  mutate(too_short = str_length(responses) < 3)
```

## Changing case

```{r}
messy_data <- c("MALE", "Female", "male", "FEMALE", "Male")

str_to_lower(messy_data)
str_to_upper(messy_data)
str_to_title(messy_data)
```

. . .

Essential for cleaning demographic data!

## Trimming whitespace

Survey data often has extra spaces:

```{r}
messy_responses <- c("  Yes", "No  ", "  Maybe  ")

str_trim(messy_responses)         # Remove leading/trailing
str_squish(messy_responses)       # Also reduce internal spaces
```

## Psychology example: Cleaning demographics

```{r}
survey <- tibble(
  gender = c("  Female", "MALE  ", "female", "Male", "  non-binary")
)

survey |>
  mutate(
    gender_clean = str_to_lower(str_trim(gender))
  )
```

## Detecting patterns

`str_detect()` checks if a pattern is present:

```{r}
feedback <- c(
  "The task was clear",
  "I found it confusing",
  "Very clear instructions",
  "Somewhat confusing"
)

str_detect(feedback, "clear")
```

. . .

Use in `filter()`:

```{r}
tibble(feedback) |>
  filter(str_detect(feedback, "clear"))
```

## Case-insensitive detection

Patterns are case-sensitive by default:

```{r}
str_detect("Clear instructions", "clear")  # FALSE
```

. . .

Make them case-insensitive with `regex(ignore_case = TRUE)`:

```{r}
str_detect("Clear instructions", regex("clear", ignore_case = TRUE))
```

## Replacing text

```{r}
responses <- c("Strongly Disagree", "Disagree", "Strongly Agree", "Agree")

# Replace first match
str_replace(responses, "Strongly ", "Very ")

# Replace all matches
str_replace_all(responses, "Strongly ", "Very ")
```

## Multiple replacements

```{r}
diagnosis <- c("MDD", "GAD", "MDD", "OCD", "GAD")

# Replace multiple patterns at once
str_replace_all(diagnosis, c(
  "MDD" = "Major Depressive Disorder",
  "GAD" = "Generalized Anxiety Disorder",
  "OCD" = "Obsessive-Compulsive Disorder"
))
```

## Extracting parts of strings

```{r}
participant_ids <- c("PSY001", "PSY002", "PSY010", "PSY123")

# Extract substring by position
str_sub(participant_ids, start = 4)  # Get everything after position 3
```

. . .

```{r}
# Extract numbers
str_extract(participant_ids, "\\d+")  # \d+ means "one or more digits"
```

## When you need more: Regular expressions

**Regular expressions (regex)** are powerful pattern matching tools.

. . .

Examples:

- `\\d` matches digits
- `\\s` matches whitespace
- `.` matches any character
- `+` means "one or more"
- `*` means "zero or more"

. . .

::: {.callout-note}
Regex is powerful but complex. For this course, stick to simple patterns. When you need more, check [R4DS Ch 14](https://r4ds.hadley.nz/strings) or [regex101.com](https://regex101.com).
:::

# Pair coding break {background-color="#e67e22"}

## Your turn: Clean text data

You have messy survey responses:

```{r}
messy_survey <- tibble(
  id = 1:5,
  gender = c("  FEMALE", "male  ", "Female", "MALE", "non-binary  "),
  comment = c(
    "Great study!",
    "too long",
    "  Very interesting  ",
    "CONFUSING INSTRUCTIONS",
    "I enjoyed this"
  )
)
```

1. Clean `gender` to lowercase with no extra spaces
2. Clean `comment` to title case with no extra spaces
3. Create a logical column `is_negative` that is TRUE if the comment contains "long" or "confusing" (case-insensitive)
4. Filter to only negative comments

**Time: 10 minutes**

```{r}
#| echo: false
#| eval: false
# SOLUTION
messy_survey |>
  mutate(
    gender_clean = str_to_lower(str_trim(gender)),
    comment_clean = str_to_title(str_trim(comment)),
    is_negative = str_detect(
      comment,
      regex("long|confusing", ignore_case = TRUE)
    )
  ) |>
  filter(is_negative)
```

# Factors {background-color="#2c3e50"}

## What are factors?

**Factors** are R's way of representing categorical data with a fixed set of possible values.

. . .

```{r}
# A character vector
gender_char <- c("Male", "Female", "Female", "Male")

# A factor
gender_fct <- factor(gender_char)
gender_fct
```

. . .

Notice the **Levels** line — those are the possible categories.

## Why use factors?

1. **Memory efficient** — R stores categories once, not repeatedly
2. **Prevent typos** — Can't accidentally add invalid categories
3. **Control order** — Specify the order for plots and tables
4. **Model requirements** — Many statistical models require factors

## Creating factors

```{r}
condition <- c("Control", "Treatment", "Treatment", "Control")

# Let R choose levels (alphabetical)
factor(condition)
```

. . .

```{r}
# Specify levels explicitly
factor(condition, levels = c("Control", "Treatment"))
```

## Order matters!

```{r}
likert <- c("Agree", "Disagree", "Strongly Agree", "Agree", "Strongly Disagree")

# Alphabetical order (default)
factor(likert)
```

. . .

```{r}
# Logical order
factor(likert, levels = c(
  "Strongly Disagree", "Disagree", "Agree", "Strongly Agree"
))
```

## Why order matters: Example

```{r}
#| output-location: slide
survey_data <- tibble(
  response = c("Agree", "Disagree", "Strongly Agree", "Agree",
               "Strongly Disagree", "Agree", "Disagree")
)

# Without factor ordering
ggplot(survey_data, aes(x = response)) +
  geom_bar() +
  labs(title = "Alphabetical order (confusing!)") +
  theme_minimal()
```

## Fixed with factor ordering

```{r}
#| output-location: slide
survey_data <- survey_data |>
  mutate(
    response = factor(response, levels = c(
      "Strongly Disagree", "Disagree", "Agree", "Strongly Agree"
    ))
  )

ggplot(survey_data, aes(x = response)) +
  geom_bar() +
  labs(title = "Logical order (much better!)") +
  theme_minimal()
```

## Forcats: Factor tools

The `forcats` package (part of tidyverse) provides functions for working with factors.

All `forcats` functions start with `fct_`.

Key functions:

- `fct_relevel()` — manually reorder levels
- `fct_reorder()` — reorder by another variable
- `fct_infreq()` — order by frequency
- `fct_recode()` — rename levels
- `fct_collapse()` — combine levels

## fct_relevel(): Manual reordering

```{r}
education <- factor(c("High School", "Bachelor's", "Master's", "High School"))
education
```

. . .

```{r}
# Put in logical order
fct_relevel(education, "High School", "Bachelor's", "Master's")
```

## fct_reorder(): Order by another variable

Extremely useful for plots!

```{r}
therapy <- tibble(
  condition = c("CBT", "Control", "Mindfulness", "Medication"),
  mean_improvement = c(12, 3, 8, 10)
)

therapy |>
  mutate(condition = fct_reorder(condition, mean_improvement))
```

## fct_reorder() in action

```{r}
#| output-location: slide
therapy |>
  mutate(condition = fct_reorder(condition, mean_improvement)) |>
  ggplot(aes(x = mean_improvement, y = condition)) +
  geom_col() +
  labs(
    title = "Treatment effectiveness (ordered by improvement)",
    x = "Mean improvement",
    y = "Treatment condition"
  ) +
  theme_minimal()
```

## fct_infreq(): Order by frequency

```{r}
diagnosis <- c("Depression", "Anxiety", "Depression", "Other",
               "Depression", "Anxiety", "Anxiety", "Depression")

# Order from most to least common
fct_infreq(factor(diagnosis))
```

## fct_infreq() in plots

```{r}
#| output-location: slide
tibble(diagnosis = factor(diagnosis)) |>
  mutate(diagnosis = fct_infreq(diagnosis)) |>
  ggplot(aes(y = diagnosis)) +  # Note: y instead of x to read easily
  geom_bar() +
  labs(
    title = "Most common diagnoses first",
    x = "Count",
    y = "Diagnosis"
  ) +
  theme_minimal()
```

## fct_recode(): Rename levels

```{r}
gender <- factor(c("M", "F", "F", "M", "NB"))

fct_recode(gender,
  "Male" = "M",
  "Female" = "F",
  "Non-binary" = "NB"
)
```

## fct_collapse(): Combine levels

Useful for grouping rare categories:

```{r}
diagnosis <- factor(c("MDD", "GAD", "OCD", "PTSD", "Panic Disorder",
                      "MDD", "GAD", "Social Anxiety"))

fct_collapse(diagnosis,
  Depression = "MDD",
  Anxiety = c("GAD", "OCD", "PTSD", "Panic Disorder", "Social Anxiety")
)
```

## Psychology example: Recoding demographics

```{r}
demo_data <- tibble(
  age_group = c("18-25", "26-35", "18-25", "36-45", "26-35", "46+"),
  education = c("HS", "BA", "BA", "MA", "HS", "PhD")
)

demo_data |>
  mutate(
    # Order age groups logically
    age_group = factor(age_group, levels = c("18-25", "26-35", "36-45", "46+")),

    # Expand education codes
    education = fct_recode(factor(education),
      "High School" = "HS",
      "Bachelor's" = "BA",
      "Master's" = "MA",
      "Doctorate" = "PhD"
    )
  )
```

## Dropping unused levels

After filtering, factors keep old levels:

```{r}
all_diagnoses <- factor(c("Depression", "Anxiety", "Other"))
just_depression <- all_diagnoses[all_diagnoses == "Depression"]
just_depression
```

. . .

Use `fct_drop()` to remove them:

```{r}
fct_drop(just_depression)
```

## Common factor issues

**Problem 1:** Factors created from numbers

```{r}
age_factor <- factor(c(25, 30, 25, 40))
mean(age_factor)  # Error! It's not numeric anymore
```

. . .

**Solution:** Convert back to numeric carefully:

```{r}
as.numeric(as.character(age_factor))  # Correct way
```

## Common factor issues

**Problem 2:** Factors behave differently than strings

```{r}
colors <- factor(c("red", "blue"))

# Can't just add new values
colors[3] <- "green"  # This creates NA!
colors
```

. . .

**Solution:** Convert to character first, or use `fct_expand()` to add levels.

# End-of-deck exercise {background-color="#e67e22"}

## Practice: Clean and visualize survey data

You have messy Likert scale data:

```{r}
likert_data <- tibble(
  question = rep(c("Q1", "Q2", "Q3"), each = 10),
  response = sample(c("strongly agree", "Agree", "NEUTRAL",
                      "disagree", "Strongly Disagree"), 30, replace = TRUE)
)
```

1. Clean the `response` variable to title case
2. Convert `response` to a factor with logical ordering
3. Create a bar chart showing response counts for each question
4. Use `facet_wrap()` to make separate panels for each question
5. Bonus: Use `fct_infreq()` to order responses by overall frequency

```{r}
#| echo: false
#| eval: false
# SOLUTION
likert_data_clean <- likert_data |>
  mutate(
    # Clean to title case
    response = str_to_title(response),

    # Convert to factor with logical order
    response = factor(response, levels = c(
      "Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree"
    ))
  )

# Basic plot
ggplot(likert_data_clean, aes(x = response)) +
  geom_bar() +
  facet_wrap(~question) +
  labs(
    title = "Survey responses by question",
    x = "Response",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Bonus: order by frequency
likert_data_clean |>
  mutate(response = fct_infreq(response)) |>
  ggplot(aes(x = response)) +
  geom_bar() +
  facet_wrap(~question) +
  labs(
    title = "Survey responses (ordered by frequency)",
    x = "Response",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Wrapping up {background-color="#2c3e50"}

## String functions cheat sheet

| Function | Purpose |
|----------|---------|
| `str_c()` | Combine strings |
| `str_length()` | Get string length |
| `str_to_lower()`, `str_to_upper()` | Change case |
| `str_trim()`, `str_squish()` | Remove whitespace |
| `str_detect()` | Find pattern |
| `str_replace()`, `str_replace_all()` | Replace pattern |
| `str_sub()` | Extract substring |

## Factor functions cheat sheet

| Function | Purpose |
|----------|---------|
| `factor()` | Create a factor |
| `fct_relevel()` | Manually reorder levels |
| `fct_reorder()` | Order by another variable |
| `fct_infreq()` | Order by frequency |
| `fct_recode()` | Rename levels |
| `fct_collapse()` | Combine levels |
| `fct_drop()` | Remove unused levels |

## Key takeaways

1. **Strings** are text — use `stringr::str_*()` functions to manipulate
2. **Always clean string data** — case, whitespace, typos
3. **Factors** are categorical data with fixed levels
4. **Factor order matters** for plots and tables
5. **Use forcats** (`fct_*()`) to manipulate factors
6. **Order factors logically** — not alphabetically
7. When in doubt, **check the data type** with `class()` or `glimpse()`

## Before next class

📖 **Read:**

- R4DS Ch 19: Joins

✅ **Do:**

- Submit Assignment 6
- Check your final project data for string/factor issues
- Practice cleaning demographic variables

## Questions?

See you Monday for joins!
