---
title: "Joins"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "May 11, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 13"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Setup {visibility="hidden"}

```{r}
#| include: false
library(tidyverse)

# Toy data for join diagrams (2026 Grammy winners)
artists <- tibble(
  name  = c("Taylor Swift", "Bad Bunny", "Kendrick Lamar"),
  genre = c("Pop", "Latin", "Hip-Hop")
)

grammys_2026 <- tibble(
  name  = c("Bad Bunny", "Kendrick Lamar", "Billie Eilish"),
  award = c("Album of the Year", "Record of the Year", "Song of the Year")
)

# Row-matching diagram for individual join types
plot_join <- function(title, left_active, right_active) {
  left_df <- tibble(
    name  = c("Taylor Swift", "Bad Bunny", "Kendrick Lamar"),
    y     = c(4, 3, 2),
    group = c("left_only", "both", "both")
  ) |> mutate(alpha_val = if_else(left_active, 1, 0.2))

  right_df <- tibble(
    name  = c("Bad Bunny", "Kendrick Lamar", "Billie Eilish"),
    y     = c(3, 2, 1),
    group = c("both", "both", "right_only")
  ) |> mutate(alpha_val = if_else(right_active, 1, 0.2))

  tile_colors <- c(both = "#5B8DB8", left_only = "#E07B54", right_only = "#5BA87A")

  ggplot() +
    geom_tile(data = left_df,
              aes(x = 1, y = y, fill = group, alpha = alpha_val),
              width = 1.7, height = 0.65, color = "white", linewidth = 1.2) +
    geom_text(data = left_df,
              aes(x = 1, y = y, label = name, alpha = alpha_val),
              size = 3, color = "white", fontface = "bold") +
    geom_tile(data = right_df,
              aes(x = 3.2, y = y, fill = group, alpha = alpha_val),
              width = 1.7, height = 0.65, color = "white", linewidth = 1.2) +
    geom_text(data = right_df,
              aes(x = 3.2, y = y, label = name, alpha = alpha_val),
              size = 3, color = "white", fontface = "bold") +
    geom_segment(aes(x = 1.86, xend = 2.34, y = 3, yend = 3),
                 color = "#5B8DB8", linewidth = 0.8, linetype = "dashed") +
    geom_segment(aes(x = 1.86, xend = 2.34, y = 2, yend = 2),
                 color = "#5B8DB8", linewidth = 0.8, linetype = "dashed") +
    annotate("text", x = 1,   y = 4.7, label = "artists",
             fontface = "bold.italic", size = 4, color = "#2c3e50") +
    annotate("text", x = 3.2, y = 4.7, label = "grammys_2026",
             fontface = "bold.italic", size = 4, color = "#2c3e50") +
    scale_fill_manual(values = tile_colors) +
    scale_alpha_identity() +
    coord_cartesian(xlim = c(0, 4.3), ylim = c(0.4, 5.1)) +
    labs(title = title) +
    theme_void() +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold",
                                color = "#2c3e50", margin = margin(b = 6))
    )
}

# Venn diagram for join type summary
make_venn <- function(title, shade = c("inner", "left", "right", "full")) {
  shade <- match.arg(shade)
  r <- 0.9; cx_l <- -0.5; cx_r <- 0.5
  theta <- seq(0, 2 * pi, length.out = 300)
  yi  <- sqrt(r^2 - cx_l^2)       # y-coord of intersection points (~0.748)
  a_l <- atan2(yi, -cx_l)          # angle from left center to intersect (~56°)
  b   <- pi - a_l                  # angle from right center to intersect (~124°)

  lc   <- tibble(x = cx_l + r * cos(theta), y = r * sin(theta))
  rc   <- tibble(x = cx_r + r * cos(theta), y = r * sin(theta))
  lens <- tibble(
    x = c(cx_l + r * cos(seq( a_l, -a_l, length.out = 150)),
          cx_r + r * cos(c(seq(-b, -pi, length.out = 75),
                            seq( pi,  b, length.out = 75)))),
    y = c(         r * sin(seq( a_l, -a_l, length.out = 150)),
                   r * sin(c(seq(-b, -pi, length.out = 75),
                              seq( pi,  b, length.out = 75))))
  )

  fill_col <- "#5B8DB8"
  p <- ggplot() +
    coord_fixed(xlim = c(-1.6, 1.6), ylim = c(-1.2, 1.4)) +
    theme_void() +
    labs(title = title) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold",
                                    size = 11, color = "#2c3e50",
                                    margin = margin(b = 4)))

  if (shade %in% c("left",  "full")) p <- p + geom_polygon(data = lc,   aes(x, y), fill = fill_col, alpha = 0.7)
  if (shade %in% c("right", "full")) p <- p + geom_polygon(data = rc,   aes(x, y), fill = fill_col, alpha = 0.7)
  if (shade == "inner")              p <- p + geom_polygon(data = lens, aes(x, y), fill = fill_col, alpha = 0.9)

  p +
    geom_path(data = lc, aes(x, y), color = "gray40", linewidth = 0.8) +
    geom_path(data = rc, aes(x, y), color = "gray40", linewidth = 0.8) +
    annotate("text", x = cx_l, y = -(r + 0.2),
             label = "artists",      size = 2.8, fontface = "italic", color = "#2c3e50") +
    annotate("text", x = cx_r, y = -(r + 0.2),
             label = "grammys_2026", size = 2.8, fontface = "italic", color = "#2c3e50")
}
```

# Why join data? {background-color="#2c3e50"}

## Three files, one question

Your participant took a survey on Monday and did a cognitive task on Tuesday. The survey data is in one CSV. The task data is in another. Their demographics are in a third.

. . .

You want to ask: *"Did anxious participants respond more slowly?"*

. . .

You can't answer that from any single file. You need **joins** — the tool that connects separate tables through a shared key.

## Two tables, one question

```{r}
# Participant information
participants <- tibble(
  participant_id = 1:4,
  age = c(22, 25, 19, 31),
  condition = c("Control", "Treatment", "Treatment", "Control")
)

# Survey responses (collected separately)
survey <- tibble(
  participant_id = c(1, 2, 3, 5),  # Note: 4 is missing, 5 is extra
  depression = c(12, 18, 10, 25),
  anxiety = c(15, 20, 12, 30)
)
```

## The two tables

:::: {.columns}
::: {.column width="50%"}
**Participants:**
```{r}
participants
```
:::

::: {.column width="50%"}
**Survey:**
```{r}
survey
```
:::
::::

. . .

How do we combine these based on `participant_id`?

# Keys {background-color="#2c3e50"}

## What are keys?

**Keys** are variables that uniquely identify observations and connect tables.

. . .

**Primary key:** Uniquely identifies each observation in a table

- `participant_id` in the participants table
- Each participant appears only once

. . .

**Foreign key:** References the primary key of another table

- `participant_id` in the survey table
- Links back to the participants table

## Checking for unique keys

Before joining, verify that your key truly identifies each row:

```{r}
# Check if participant_id is unique
participants |>
  count(participant_id) |>
  filter(n > 1)
```

. . .

Empty result = good! Each participant appears only once.

## When keys aren't unique

Sometimes keys aren't unique (e.g., longitudinal data):

```{r}
longitudinal <- tibble(
  participant_id = c(1, 1, 2, 2, 3, 3),
  timepoint = c(1, 2, 1, 2, 1, 2),
  depression = c(20, 15, 18, 12, 25, 22)
)
```

. . .

Here, the **combination** of `participant_id` + `timepoint` is the key:

```{r}
longitudinal |>
  count(participant_id, timepoint) |>
  filter(n > 1)
```

# Mutating joins {background-color="#2c3e50"}

## The join family

**Mutating joins** add columns from one table to another:

| Function | Keeps rows from... |
|----------|-------------------|
| `left_join()` | Left table (all rows) |
| `right_join()` | Right table (all rows) |
| `inner_join()` | Both tables (only matches) |
| `full_join()` | Both tables (all rows) |

## left_join(): Most common

Keep **all** rows from the left table, add matching data from the right:

```{r}
participants |>
  left_join(survey, by = "participant_id")
```

. . .

- Participant 4 has no survey data → `NA`
- Participant 5 from survey isn't in participants → excluded

## Visualizing left_join()

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
plot_join(
  "left_join(artists, grammys_2026)",
  left_active  = c(TRUE, TRUE, TRUE),
  right_active = c(TRUE, TRUE, FALSE)
)
```

## right_join(): The mirror

Keep **all** rows from the right table:

```{r}
participants |>
  right_join(survey, by = "participant_id")
```

. . .

Now participant 5 is included (with `NA` for age/condition), but participant 4 is excluded.

## Visualizing right_join()

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
plot_join(
  "right_join(artists, grammys_2026)",
  left_active  = c(FALSE, TRUE, TRUE),
  right_active = c(TRUE, TRUE, TRUE)
)
```

## inner_join(): Only matches

Keep **only** rows that exist in both tables:

```{r}
participants |>
  inner_join(survey, by = "participant_id")
```

. . .

- Participant 4 excluded (no survey data)
- Participant 5 excluded (not in participants)

## Visualizing inner_join()

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
plot_join(
  "inner_join(artists, grammys_2026)",
  left_active  = c(FALSE, TRUE, TRUE),
  right_active = c(TRUE, TRUE, FALSE)
)
```

## full_join(): Everything

Keep **all** rows from both tables:

```{r}
participants |>
  full_join(survey, by = "participant_id")
```

. . .

Every participant appears, with `NA` where data is missing.

## Visualizing full_join()

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
plot_join(
  "full_join(artists, grammys_2026)",
  left_active  = c(TRUE, TRUE, TRUE),
  right_active = c(TRUE, TRUE, TRUE)
)
```

## Join types at a glance

```{r}
#| echo: false
#| layout-ncol: 2
#| fig-width: 4
#| fig-height: 2.8
make_venn("inner_join()", "inner")
make_venn("left_join()",  "left")
make_venn("right_join()", "right")
make_venn("full_join()",  "full")
```

## Which join should I use?

**Most common in psychology: `left_join()`**

. . .

Use when:

- You have a main participant table
- You want to add additional data (surveys, behavioral tasks)
- You want to keep all participants, even those with missing data

. . .

::: {.callout-tip}
Start with `left_join()` unless you have a specific reason to use another type.
:::

## Psychology example: Adding demographics

```{r}
# Task data
task_data <- tibble(
  participant_id = c(1, 2, 3, 1, 2, 3),
  trial = c(1, 1, 1, 2, 2, 2),
  reaction_time = c(450, 380, 520, 430, 395, 510)
)

# Add participant info to each trial
task_data |>
  left_join(participants, by = "participant_id")
```

## Joining by multiple keys

For longitudinal data, join on multiple columns:

```{r}
# Time 1 scores
time1 <- tibble(
  id = c(1, 2, 3),
  time = 1,
  depression = c(20, 18, 25)
)

# Time 2 scores
time2 <- tibble(
  id = c(1, 2, 3),
  time = 2,
  depression = c(15, 12, 22)
)
```

## Combining longitudinal data

```{r}
# Combine into one table
all_times <- bind_rows(time1, time2)

# Add baseline demographics
all_times |>
  left_join(
    participants |> select(participant_id, age, condition),
    by = c("id" = "participant_id")  # Keys have different names
  )
```

# Pair coding break {background-color="#e67e22"}

## Your turn: Combine study data

You have three tables from a therapy study:

```{r}
baseline <- tibble(
  id = 1:5,
  age = c(25, 30, 22, 35, 28),
  baseline_depression = c(22, 25, 18, 30, 20)
)

treatment <- tibble(
  id = c(1, 2, 3, 4),  # Missing id 5
  condition = c("CBT", "Control", "CBT", "Control")
)

followup <- tibble(
  id = c(1, 2, 3, 5),  # Missing id 4, but has id 5
  followup_depression = c(12, 23, 10, 18)
)
```

## Your turn: Combine study data

1. Create a dataset with baseline info and treatment condition (keep all participants)
2. Add followup data (keep all from step 1)
3. How many participants are missing followup data?

**Time: 10 minutes**

```{r}
#| echo: false
#| eval: false
# SOLUTION
# Step 1: Add treatment to baseline
baseline_treatment <- baseline |>
  left_join(treatment, by = "id")

# Step 2: Add followup
complete_data <- baseline_treatment |>
  left_join(followup, by = "id")

complete_data

# Step 3: Count missing followup
sum(is.na(complete_data$followup_depression))
```

# Filtering joins {background-color="#2c3e50"}

## A different purpose

**Filtering joins** don't add columns — they filter rows based on whether matches exist:

- `semi_join()` — Keep rows that **have** a match
- `anti_join()` — Keep rows that **don't have** a match

## semi_join(): Has a match

Keep rows from the left table that exist in the right table:

```{r}
# Which participants completed the survey?
participants |>
  semi_join(survey, by = "participant_id")
```

. . .

Only participants 1, 2, and 3 completed the survey.

## anti_join(): No match

Keep rows from the left table that **don't** exist in the right table:

```{r}
# Which participants didn't complete the survey?
participants |>
  anti_join(survey, by = "participant_id")
```

. . .

Participant 4 didn't complete the survey.

## Psychology use case: Attrition analysis

```{r}
# Who was enrolled
enrolled <- tibble(
  participant_id = 1:10,
  condition = rep(c("Treatment", "Control"), each = 5))

# Who completed
completed <- tibble(
  participant_id = c(1, 2, 3, 4, 7, 8, 9, 10))  # Missing 5 and 6

# Who dropped out?
enrolled |>
  anti_join(completed, by = "participant_id")
```

## Attrition by condition

```{r}
dropout <- enrolled |>
  anti_join(completed, by = "participant_id")

dropout |>
  count(condition)
```

. . .

::: {.callout-warning}
Both dropouts were from the Treatment condition — this could bias results!
:::

# Common join problems {background-color="#2c3e50"}

## Problem 1: Keys with different names

:::: {.columns}
::: {.column width="50%"}
```{r}
demographics <- tibble(
  subj_id = 1:3,
  age = c(25, 30, 22)
)

scores <- tibble(
  participant = c(1, 2, 3),
  depression = c(18, 22, 15)
)
```
:::

::: {.column width="50%"}
**Solution:** Use named vector in `by`:

```{r}
demographics |>
  left_join(scores,
    by = c("subj_id" = "participant"))
```
:::
::::

## Problem 2: Duplicate keys

```{r}
# Duplicate in left table
participants_dup <- tibble(
  id = c(1, 1, 2, 3),  # ID 1 appears twice!
  age = c(25, 25, 30, 22)
)

task_scores <- tibble(
  id = c(1, 2, 3),
  score = c(85, 90, 88)
)
```

## What happens with duplicates?

```{r}
participants_dup |>
  left_join(task_scores, by = "id")
```

. . .

Each duplicate gets matched → creates extra rows!

. . .

::: {.callout-important}
Always check for duplicate keys before joining.
:::

## Problem 3: Missing keys

```{r}
# Some participants have NA for id
messy_data <- tibble(
  id = c(1, 2, NA, 3),
  response = c("Yes", "No", "Yes", "No")
)

demographics <- tibble(
  id = 1:3,
  age = c(25, 30, 22)
)
```

## Joining with NAs

```{r}
messy_data |>
  left_join(demographics, by = "id")
```

. . .

Row with `NA` id can't match → gets `NA` for age.

Fix the data before joining!

## Problem 4: Many-to-many relationships

```{r}
# Multiple trials per participant
trials <- tibble(
  id = c(1, 1, 2, 2),
  trial = c(1, 2, 1, 2),
  rt = c(450, 430, 380, 390)
)

# Multiple sessions per participant
sessions <- tibble(
  id = c(1, 1, 2, 2),
  session = c(1, 2, 1, 2),
  mood = c(5, 6, 4, 5)
)
```

## Many-to-many result

```{r}
trials |>
  left_join(sessions, by = "id")
```

. . .

Creates all combinations — probably not what you want!

**Solution:** Be more specific about what you're joining on, or reshape first.

# Practical psychology examples {background-color="#2c3e50"}

## Example 1: Qualtrics + demographics

```{r}
# Survey responses exported from Qualtrics
qualtrics <- tibble(
  ResponseId = c("R_1", "R_2", "R_3"),
  PHQ9_total = c(12, 18, 8),
  GAD7_total = c(10, 15, 6)
)

# Demographics collected separately
demo <- tibble(
  ResponseId = c("R_1", "R_2", "R_3"),
  age = c(25, 30, 22),
  gender = c("Female", "Male", "Female")
)
```

## Example 1: Result

```{r}
qualtrics |>
  left_join(demo, by = "ResponseId")
```

## Example 2: Pre-post intervention

```{r}
pre_test <- tibble(
  participant = 1:4,
  depression_pre = c(22, 25, 18, 20)
)

post_test <- tibble(
  participant = c(1, 2, 4),  # 3 dropped out
  depression_post = c(12, 23, 15)
)
```

## Example 2: Result

```{r}
# Combine and compute change
pre_test |>
  left_join(post_test, by = "participant") |>
  mutate(change = depression_pre - depression_post)
```

## Example 3: Item-level to scale-level

```{r}
# Individual items
items <- tibble(
  id = rep(1:3, each = 5),
  item = rep(1:5, times = 3),
  response = c(3,4,3,4,3, 2,3,2,3,2, 4,5,4,5,4)
)

# Participant info
participants_info <- tibble(
  id = 1:3,
  condition = c("Control", "Treatment", "Treatment")
)
```

## Example 3: Result

```{r}
# Compute scale scores then join
items |>
  group_by(id) |>
  summarize(scale_mean = mean(response)) |>
  left_join(participants_info, by = "id")
```

# End-of-deck exercise {background-color="#e67e22"}

## Practice: Longitudinal study

You have data from a three-wave longitudinal study:

```{r}
# Baseline demographics
baseline_demo <- tibble(
  pid = 1:6,
  age = c(20, 22, 25, 19, 24, 21),
  condition = rep(c("Intervention", "Control"), each = 3)
)

# Time 1 assessment
time1_scores <- tibble(
  pid = 1:6,
  depression_t1 = c(22, 25, 18, 20, 24, 19)
)

# Time 2 assessment (2 dropouts)
time2_scores <- tibble(
  pid = c(1, 2, 3, 4, 6),  # Missing 5
  depression_t2 = c(15, 24, 12, 18, 17)
)

# Time 3 assessment (1 more dropout)
time3_scores <- tibble(
  pid = c(1, 2, 3, 6),  # Missing 4 and 5
  depression_t3 = c(12, 22, 10, 15)
)
```

## Your tasks

1. Create a complete dataset with all timepoints
2. Compute change scores (T1 to T3)
3. Identify who dropped out at each wave
4. Check if dropout differs by condition

```{r}
#| echo: false
#| eval: false
# SOLUTION

# 1. Complete dataset
complete <- baseline_demo |>
  left_join(time1_scores, by = "pid") |>
  left_join(time2_scores, by = "pid") |>
  left_join(time3_scores, by = "pid")

complete

# 2. Change scores
complete |>
  mutate(change_t1_t3 = depression_t1 - depression_t3)

# 3. Who dropped out at T2?
baseline_demo |>
  anti_join(time2_scores, by = "pid")

# Who dropped out at T3 (but completed T2)?
time2_scores |>
  anti_join(time3_scores, by = "pid")

# 4. Dropout by condition
baseline_demo |>
  anti_join(time3_scores, by = "pid") |>
  count(condition)
```

# Wrapping up {background-color="#2c3e50"}

## Join decision flowchart

:::: {.columns}
::: {.column width="50%"}
Ask yourself:

1. **Do I want to add columns?**
   - Yes → Use a mutating join
   - No → Use a filtering join

2. **Which rows do I want to keep?**
   - All from left → `left_join()`
   - All from right → `right_join()`
   - Only matches → `inner_join()`
   - Everything → `full_join()`
:::

::: {.column width="50%"}
3. **Am I filtering rows?**
   - Keep matches → `semi_join()`
   - Keep non-matches → `anti_join()`
:::
::::

## Key takeaways

1. **Joins combine data** from multiple tables using key variables
2. **Check your keys** for uniqueness before joining
3. **`left_join()` is your workhorse** — keeps all rows from your main table
4. **Filtering joins** help identify completers vs. dropouts
5. **Watch for common problems:** duplicate keys, different key names, NAs
6. **Always inspect** the result to ensure it matches expectations

## Before next class

📖 **Read:**

- R4DS Ch 18: Missing values

✅ **Do:**

- Submit Assignment 6
- Check if your final project needs joins
- Practice joining your own data files

## The one thing to remember

If your data lives in separate files, joins are how you ask a question that spans all of them.

See you Wednesday for handling missing data!
