---
title: "Strings & Factors"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "May 11, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 13"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
    include-in-header:
      text: |
        <link rel="stylesheet" href="toc-nav.css">
    include-after-body:
      text: |
        <script src="toc-nav.js"></script>
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Setup {visibility="hidden"}

```{r}
#| include: false
library(tidyverse)
library(tidytext)

# Load our class survey data
class_survey <- read_csv("../data/class_survey.csv")
```

# The Power Play {background-color="#e67e22"}

## Something new in the team challenge

The top 2 teams on the scoreboard right now get a **one-time strategic choice:**

. . .

**Option A — Boost:** Add **+3 points** to your own team's total

. . .

**Option B — Sabotage:** Remove **2 points** from each of **two other teams** of your choice

. . .

Choices are final. First place picks first.

::: {.notes}
This is the Power Play — a mid-term twist to keep things interesting. Show the scoreboard first (as usual), then reveal this slide. Give teams a minute to react, then have the top 2 teams announce their choices live. If a team is targeted by both, they lose 4 total. Teams can't go below 0. This should take about 5 minutes and generate some fun energy. Don't belabor it — announce, choose, update the scoreboard, move on.
:::

# Strings: The basics {background-color="#2c3e50"}

## How many majors are in this dataset?

```{r}
#| echo: false
tibble(
  id = 1:7,
  major = c("Psychology", "psychology", "Psych", "PSYCHOLOGY", "Biology", "biology ", "Bio")
) |>
  knitr::kable()
```

## How many majors are in this dataset?

R says **7**. You meant **2**.

The problem is inconsistent text — different cases, abbreviations, trailing spaces. Today we learn to fix that.

::: {.notes}
Start by showing the first slide (the table) and asking students "How many majors are in this dataset?" Let them look at the table and answer before revealing that R sees 7 unique values. This is the hook — the gap between what humans see and what R sees motivates the entire session. Spend ~2 minutes here.
:::


## What are strings?

**Strings** are text data — anything in quotes:

```{r}
participant_name <- "Jane Doe"
diagnosis <- "Major Depressive Disorder"
feedback <- "The task was confusing"
```

The `stringr` package (part of tidyverse) gives you tools for working with strings. All `stringr` functions start with `str_`.

::: {.notes}
Emphasize the naming convention: all stringr functions start with str_, so autocomplete in RStudio is very helpful. Students can type str_ and browse available functions. This is a good time to remind them about the tab-completion habit.
:::

## Creating and combining strings

```{r}
first <- "Jane"
last <- "Doe"

# Combine strings
str_c(first, last)           # No space
str_c(first, last, sep = " ")  # With space
```

. . .

```{r}
# Combine with other text
str_c("Participant: ", first, " ", last)

# Glue is even easier (from the glue package)
library(glue)
glue("Participant: {first} {last}")
```

::: {.notes}
Move quickly through str_c() — students will use it occasionally but glue() is more intuitive. Show both so they recognize str_c() in others' code, but recommend glue() for their own work. If students ask: glue is loaded automatically with tidyverse in newer versions, but loading it explicitly does not hurt.
:::

## String length

```{r}
responses <- c("Yes", "No", "Maybe", "I don't know")

str_length(responses)
```

. . .

Useful for checking free-response data quality:

```{r}
# Flag very short responses
tibble(responses) |>
  mutate(too_short = str_length(responses) < 3)
```

## Changing case

```{r}
messy_data <- c("PSYCHOLOGY", "Biology", "psychology", "BIOLOGY", "Psychology")

str_to_lower(messy_data)
str_to_upper(messy_data)
str_to_title(messy_data)
```

. . .

Essential for cleaning survey data!

::: {.notes}
This is one of the most important slides in the deck. str_to_lower() alone solves a huge number of data cleaning problems. Run the code live and ask: "If you had to pick just one of these three functions, which would you use for cleaning survey data?" Answer: str_to_lower(), because it makes everything comparable regardless of how participants typed it.
:::

## Trimming whitespace

Survey data often has extra spaces:

```{r}
messy_responses <- c("  Yes", "No  ", "  Maybe  ")

str_trim(messy_responses)         # Remove leading/trailing
str_squish(messy_responses)       # Also reduce internal spaces
```

::: {.notes}
Students often ask the difference between str_trim() and str_squish(). str_trim() only removes leading and trailing whitespace; str_squish() also collapses multiple internal spaces into one. For survey data, str_trim() is usually sufficient. Trailing spaces from Qualtrics exports are very common and invisible to the eye — this is why str_trim() matters.
:::

## Psychology example: Cleaning survey responses

```{r}
survey <- tibble(
  major = c("  Psychology", "BIOLOGY  ", "psychology", "Biology", "  sociology")
)

survey |>
  mutate(
    major_clean = str_to_lower(str_trim(major))
  )
```

::: {.notes}
This is the payoff slide for the opening example. Walk through it step by step: "First we trim the spaces, then we lowercase everything." Point out the nesting — str_to_lower wraps str_trim. Students could also pipe these with |> but nesting is fine for two functions. After cleaning, ask: "How many unique majors would R see now?" This connects directly back to the hook.
:::

## Detecting patterns

`str_detect()` checks if a pattern is present:

```{r}
feedback <- c(
  "The task was clear",
  "I found it confusing",
  "Very clear instructions",
  "Somewhat confusing"
)

str_detect(feedback, "clear")
```

. . .

Use in `filter()`:

```{r}
tibble(feedback) |>
  filter(str_detect(feedback, "clear"))
```

::: {.notes}
Emphasize the filter() integration — str_detect() returns TRUE/FALSE, which is exactly what filter() needs. This connects back to last session on logicals. Students will use this pattern frequently for things like "find all participants who mentioned anxiety in their open-ended response."
:::

## Case-insensitive detection

Patterns are case-sensitive by default:

```{r}
str_detect("Clear instructions", "clear")  # FALSE
```

. . .

Make them case-insensitive with `regex(ignore_case = TRUE)`:

```{r}
str_detect("Clear instructions", regex("clear", ignore_case = TRUE))
```

::: {.notes}
The regex() wrapper with ignore_case is essential for real survey data where you cannot control capitalization. Students will forget this and wonder why their str_detect() misses matches — remind them that case-sensitivity is the default and they need to opt in to ignoring case.
:::

## Replacing text

```{r}
responses <- c("Strongly Disagree", "Disagree", "Strongly Agree", "Agree")

# Replace first match
str_replace(responses, "Strongly ", "Very ")

# Replace all matches
str_replace_all(responses, "Strongly ", "Very ")
```

## Multiple replacements

```{r}
diagnosis <- c("MDD", "GAD", "MDD", "OCD", "GAD")

# Replace multiple patterns at once
str_replace_all(diagnosis, c(
  "MDD" = "Major Depressive Disorder",
  "GAD" = "Generalized Anxiety Disorder",
  "OCD" = "Obsessive-Compulsive Disorder"
))
```

::: {.notes}
The named vector approach to str_replace_all() is very handy for recoding abbreviations in real data. Emphasize this pattern — students will use it for diagnosis codes, condition labels, and other abbreviation-heavy data. Move fairly quickly through this slide.
:::

## Extracting parts of strings

```{r}
participant_ids <- c("PSY001", "PSY002", "PSY010", "PSY123")

# Extract substring by position
str_sub(participant_ids, start = 4)  # Get everything after position 3
```

. . .

```{r}
# Extract numbers
str_extract(participant_ids, "\\d+")  # \d+ means "one or more digits"
```

## When you need more: Regular expressions

**Regular expressions (regex)** are powerful pattern matching tools.

. . .

Examples:

- `\\d` matches digits
- `\\s` matches whitespace
- `.` matches any character
- `+` means "one or more"
- `*` means "zero or more"

. . .

::: {.callout-note}
Regex is powerful but complex. For this course, stick to simple patterns. When you need more, check [R4DS Ch 14](https://r4ds.hadley.nz/strings) or [regex101.com](https://regex101.com).
:::

::: {.notes}
Do not go deep on regex — this is a "know it exists" slide. Students only need to know that \\d matches digits and that patterns can be more powerful than literal text. If anyone asks for more, point them to regex101.com where they can test patterns interactively. Transition: "You have enough string tools now to clean real data. Let's practice."
:::

# Pair coding break {background-color="#e67e22"}

## Your turn: Clean text data

You have messy survey responses:

```{r}
messy_survey <- tibble(
  id = 1:5,
  major = c("  PSYCHOLOGY", "biology  ", "Psychology", "BIOLOGY", "sociology  "),
  comment = c(
    "Great study!",
    "too long",
    "  Very interesting  ",
    "CONFUSING INSTRUCTIONS",
    "I enjoyed this"
  )
)
```

1. Clean `major` to lowercase with no extra spaces
2. Clean `comment` to title case with no extra spaces
3. Create a logical column `is_negative` that is TRUE if the comment contains "long" or "confusing" (case-insensitive)
4. Filter to only negative comments

**Time: 10 minutes**

::: {.notes}
Give 10 minutes. Common issues to watch for: (1) students forgetting to nest str_to_lower(str_trim(...)) or doing them in the wrong order (order doesn't matter here, but they may think it does), (2) for the is_negative column, students need the pipe (|) inside the regex pattern for "or" — remind them this is different from the logical OR operator, (3) students forgetting regex(ignore_case = TRUE). Walk through the solution afterward, emphasizing the cleaning-then-filtering workflow.
:::

```{r}
#| echo: false
#| eval: false
# SOLUTION
messy_survey |>
  mutate(
    major_clean = str_to_lower(str_trim(major)),
    comment_clean = str_to_title(str_trim(comment)),
    is_negative = str_detect(
      comment,
      regex("long|confusing", ignore_case = TRUE)
    )
  ) |>
  filter(is_negative)
```

---

## Before we move on

📤 **[Submit your code on Canvas](https://canvas.uoregon.edu/courses/287793/pages/pair-coding-check-ins)** for participation credit. Paste what you have — it doesn't need to work perfectly.

# Factors {background-color="#2c3e50"}

## What are factors?

**Factors** are R's way of representing categorical data with a fixed set of possible values.

. . .

```{r}
# A character vector
major_char <- c("Psychology", "Biology", "Biology", "Psychology")

# A factor
major_fct <- factor(major_char)
major_fct
```

. . .

Notice the **Levels** line — those are the possible categories.

::: {.notes}
Run this code live. The key thing students should notice is the "Levels:" line in the output — that is what makes a factor different from a plain character vector. Emphasize that levels define the complete set of valid categories, even if some do not appear in the data. This matters for plots and statistical models.
:::

## Why use factors?

1. **Memory efficient** — R stores categories once, not repeatedly
2. **Prevent typos** — Can't accidentally add invalid categories
3. **Control order** — Specify the order for plots and tables
4. **Model requirements** — Many statistical models require factors

## Creating factors

```{r}
condition <- c("Control", "Treatment", "Treatment", "Control")

# Let R choose levels (alphabetical)
factor(condition)
```

. . .

```{r}
# Specify levels explicitly
factor(condition, levels = c("Control", "Treatment"))
```

::: {.notes}
Emphasize the difference: without specifying levels, R defaults to alphabetical order. With explicit levels, you control the order. For most psychology data, you want explicit levels because alphabetical order rarely makes conceptual sense (e.g., "Control" should come before "Treatment" even though it does alphabetically by coincidence).
:::

## Order matters!

```{r}
likert <- c("Agree", "Disagree", "Strongly Agree", "Agree", "Strongly Disagree")

# Alphabetical order (default)
factor(likert)
```

. . .

```{r}
# Logical order
factor(likert, levels = c(
  "Strongly Disagree", "Disagree", "Agree", "Strongly Agree"
))
```

## Why order matters: Example

::: {.notes}
This two-slide sequence is the most persuasive argument for factors. Show the alphabetical plot first and ask "What is wrong with this chart?" Students will immediately see that the Likert order is scrambled. Then show the fixed version. This aha moment makes the factor concept stick.
:::

```{r}
#| output-location: slide
#| fig-alt: "Bar chart of Likert responses in alphabetical order, making the progression from Disagree to Strongly Agree hard to follow because categories are scrambled."
survey_data <- tibble(
  response = c("Agree", "Disagree", "Strongly Agree", "Agree",
               "Strongly Disagree", "Agree", "Disagree")
)

# Without factor ordering
ggplot(survey_data, aes(x = response)) +
  geom_bar() +
  labs(title = "Alphabetical order (confusing!)") +
  theme_minimal()
```

## Fixed with factor ordering

```{r}
#| output-location: slide
#| fig-alt: "Bar chart of Likert responses reordered from Strongly Disagree to Strongly Agree, showing a clear progression that matches how the scale is meant to be read."
survey_data <- survey_data |>
  mutate(
    response = factor(response, levels = c(
      "Strongly Disagree", "Disagree", "Agree", "Strongly Agree"
    ))
  )

ggplot(survey_data, aes(x = response)) +
  geom_bar() +
  labs(title = "Logical order (much better!)") +
  theme_minimal()
```

## Forcats: Factor tools

The `forcats` package (part of tidyverse) provides functions for working with factors.

All `forcats` functions start with `fct_`.

Key functions:

- `fct_relevel()` — manually reorder levels
- `fct_reorder()` — reorder by another variable
- `fct_infreq()` — order by frequency
- `fct_recode()` — rename levels
- `fct_collapse()` — combine levels

::: {.notes}
Emphasize the naming convention: all forcats functions start with fct_, just like stringr functions start with str_. This makes them easy to find with autocomplete. You do not need to load forcats separately — it is part of tidyverse. Skim the list; each function gets its own slide next.
:::

## fct_relevel(): Manual reordering

```{r}
education <- factor(c("High School", "Bachelor's", "Master's", "High School"))
education
```

. . .

```{r}
# Put in logical order
fct_relevel(education, "High School", "Bachelor's", "Master's")
```

## fct_reorder(): Order by another variable

::: {.notes}
fct_reorder() is arguably the most useful forcats function for data visualization. It reorders factor levels by the values of another variable. In the therapy example, conditions are ordered by mean improvement, making the bar chart immediately readable. Students will use this frequently in assignments and the final project.
:::

Extremely useful for plots!

```{r}
therapy <- tibble(
  condition = c("CBT", "Control", "Mindfulness", "Medication"),
  mean_improvement = c(12, 3, 8, 10)
)

therapy |>
  mutate(condition = fct_reorder(condition, mean_improvement))
```

## fct_reorder() in action

```{r}
#| output-location: slide
#| fig-alt: "Horizontal bar chart showing four therapy conditions ordered by mean improvement, with CBT highest and Control lowest, demonstrating how fct_reorder arranges categories by a numeric variable."
therapy |>
  mutate(condition = fct_reorder(condition, mean_improvement)) |>
  ggplot(aes(x = mean_improvement, y = condition)) +
  geom_col() +
  labs(
    title = "Treatment effectiveness (ordered by improvement)",
    x = "Mean improvement",
    y = "Treatment condition"
  ) +
  theme_minimal()
```

## fct_infreq(): Order by frequency

```{r}
diagnosis <- c("Depression", "Anxiety", "Depression", "Other",
               "Depression", "Anxiety", "Anxiety", "Depression")

# Order from most to least common
fct_infreq(factor(diagnosis))
```

## fct_infreq() in plots

```{r}
#| output-location: slide
#| fig-alt: "Horizontal bar chart of diagnosis categories ordered by frequency, with Depression most common and Other least common, showing how fct_infreq sorts categories from most to least frequent."
tibble(diagnosis = factor(diagnosis)) |>
  mutate(diagnosis = fct_infreq(diagnosis)) |>
  ggplot(aes(y = diagnosis)) +  # Note: y instead of x to read easily
  geom_bar() +
  labs(
    title = "Most common diagnoses first",
    x = "Count",
    y = "Diagnosis"
  ) +
  theme_minimal()
```

## fct_recode(): Rename levels

```{r}
major <- factor(c("Psych", "Bio", "Bio", "Psych", "Soc"))

fct_recode(major,
  "Psychology" = "Psych",
  "Biology" = "Bio",
  "Sociology" = "Soc"
)
```

## fct_collapse(): Combine levels

Useful for grouping rare categories:

```{r}
diagnosis <- factor(c("MDD", "GAD", "OCD", "PTSD", "Panic Disorder",
                      "MDD", "GAD", "Social Anxiety"))

fct_collapse(diagnosis,
  Depression = "MDD",
  Anxiety = c("GAD", "OCD", "PTSD", "Panic Disorder", "Social Anxiety")
)
```

::: {.notes}
fct_collapse() is useful when you have too many levels for meaningful analysis. In psychology, you often collapse rare diagnosis categories into "Other" or group specific disorders into broader categories. Note that OCD is grouped under Anxiety here, which reflects older DSM-IV classification — this could spark a good discussion about how coding decisions reflect theoretical assumptions.
:::

## Psychology example: Recoding demographics

```{r}
demo_data <- tibble(
  age_group = c("18-25", "26-35", "18-25", "36-45", "26-35", "46+"),
  education = c("HS", "BA", "BA", "MA", "HS", "PhD")
)

demo_data |>
  mutate(
    # Order age groups logically
    age_group = factor(age_group, levels = c("18-25", "26-35", "36-45", "46+")),

    # Expand education codes
    education = fct_recode(factor(education),
      "High School" = "HS",
      "Bachelor's" = "BA",
      "Master's" = "MA",
      "Doctorate" = "PhD"
    )
  )
```

## Dropping unused levels

After filtering, factors keep old levels:

```{r}
all_diagnoses <- factor(c("Depression", "Anxiety", "Other"))
just_depression <- all_diagnoses[all_diagnoses == "Depression"]
just_depression
```

. . .

Use `fct_drop()` to remove them:

```{r}
fct_drop(just_depression)
```

::: {.notes}
This is a common gotcha. After filtering, factors remember their old levels, which can create empty bars in plots or confuse statistical models. fct_drop() solves it. Students will encounter this when they subset data and then try to plot — the empty categories show up as blank bars.
:::

## Common factor issues

**Problem 1:** Factors created from numbers

```{r}
age_factor <- factor(c(25, 30, 25, 40))
mean(age_factor)  # Error! It's not numeric anymore
```

. . .

**Solution:** Convert back to numeric carefully:

```{r}
as.numeric(as.character(age_factor))  # Correct way
```

::: {.notes}
This is a classic trap. Students sometimes accidentally convert numeric columns to factors (e.g., when reading CSV files with read.csv instead of read_csv). The key lesson: as.numeric() on a factor gives you the internal level codes (1, 2, 3...), not the original values. You must go through as.character() first. This mistake has caused real errors in published research.
:::

## Common factor issues

**Problem 2:** Factors behave differently than strings

```{r}
colors <- factor(c("red", "blue"))

# Can't just add new values
colors[3] <- "green"  # This creates NA!
colors
```

. . .

**Solution:** Convert to character first, or use `fct_expand()` to add levels.

# Light text analysis {background-color="#2c3e50"}

## What did you say about data?

On the class survey, you described your relationship with data in 1–2 words. Let's see what you said.

```{r}
class_survey |>
  select(data_words) |>
  head(10)
```

::: {.notes}
Transition: "We've been cleaning strings and ordering factors. Now let's do something fun with text data — your text data." Show the raw responses. Students will laugh at some of the answers. This motivates the next step: how do we summarize open-ended text?
:::

## Tokenizing text with tidytext

`unnest_tokens()` splits text into individual words:

```{r}
words <- class_survey |>
  select(data_words) |>
  unnest_tokens(word, data_words)

words
```

Each row is now a single word.

::: {.notes}
Explain: unnest_tokens takes a column of text and turns each word into its own row. The first argument is what to call the new column, the second is the source column. This is the foundation of text analysis — you can't count words until they're in separate rows.
:::

## What words came up most?

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Horizontal bar chart showing the most common words students used to describe their relationship with data, ordered by frequency."
words |>
  anti_join(stop_words, by = "word") |>
  count(word, sort = TRUE) |>
  slice_head(n = 10) |>
  mutate(word = fct_reorder(word, n)) |>
  ggplot(aes(x = n, y = word)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "How Our Class Describes Their Relationship with Data",
    x = "Count",
    y = NULL
  ) +
  theme_minimal(base_size = 14)
```

::: {.notes}
Walk through the pipeline: anti_join removes common filler words ("the", "a", "is"), count tallies each word, and fct_reorder puts them in order for the plot. Ask: "What themes do you see? Are people mostly positive, negative, or neutral?" This connects string skills, factor ordering, and visualization — a nice capstone moment. Point out that fct_reorder here does exactly what they learned earlier in the factors section.
:::

# End-of-deck exercise {background-color="#e67e22"}

## Practice: Clean and visualize survey data

You have messy Likert scale data:

```{r}
likert_data <- tibble(
  question = rep(c("Q1", "Q2", "Q3"), each = 10),
  response = sample(c("strongly agree", "Agree", "NEUTRAL",
                      "disagree", "Strongly Disagree"), 30, replace = TRUE)
)
```

1. Clean the `response` variable to title case
2. Convert `response` to a factor with logical ordering
3. Create a bar chart showing response counts for each question
4. Use `facet_wrap()` to make separate panels for each question
5. Bonus: Use `fct_infreq()` to order responses by overall frequency
6. **A6 preview** — given an `open_responses` tibble with a `response` column, tokenize it and find the most common non-stopwords:

```{r}
#| eval: false
library(tidytext)
open_responses |>
  unnest_tokens(word, response) |>
  anti_join(stop_words, by = "word") |>
  count(word, sort = TRUE)
```

::: {.notes}
Give ~10 minutes. This exercise ties together strings and factors: students must first clean the text (str_to_title), then convert to a factor with proper ordering, then plot. Watch for students who forget to clean before converting to factor, which creates duplicate levels like "agree" and "Agree." The facet_wrap step is review from earlier visualization sessions. If students finish early, the `fct_infreq()` bonus is a good stretch goal, and step 6 previews the tidytext workflow they'll need for A6 Part 5 — emphasize that `unnest_tokens()` + `anti_join(stop_words)` is the standard pattern, no regex required.
:::

```{r}
#| echo: false
#| eval: false
# SOLUTION
likert_data_clean <- likert_data |>
  mutate(
    # Clean to title case
    response = str_to_title(response),

    # Convert to factor with logical order
    response = factor(response, levels = c(
      "Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree"
    ))
  )

# Basic plot
ggplot(likert_data_clean, aes(x = response)) +
  geom_bar() +
  facet_wrap(~question) +
  labs(
    title = "Survey responses by question",
    x = "Response",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Bonus: order by frequency
likert_data_clean |>
  mutate(response = fct_infreq(response)) |>
  ggplot(aes(x = response)) +
  geom_bar() +
  facet_wrap(~question) +
  labs(
    title = "Survey responses (ordered by frequency)",
    x = "Response",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Wrapping up {background-color="#2c3e50"}

## String functions cheat sheet

| Function | Purpose |
|----------|---------|
| `str_c()` | Combine strings |
| `str_length()` | Get string length |
| `str_to_lower()`, `str_to_upper()` | Change case |
| `str_trim()`, `str_squish()` | Remove whitespace |
| `str_detect()` | Find pattern |
| `str_replace()`, `str_replace_all()` | Replace pattern |
| `str_sub()` | Extract substring |

## Factor functions cheat sheet

| Function | Purpose |
|----------|---------|
| `factor()` | Create a factor |
| `fct_relevel()` | Manually reorder levels |
| `fct_reorder()` | Order by another variable |
| `fct_infreq()` | Order by frequency |
| `fct_recode()` | Rename levels |
| `fct_collapse()` | Combine levels |
| `fct_drop()` | Remove unused levels |

## Key takeaways

1. **Strings** are text — use `stringr::str_*()` functions to manipulate
2. **Always clean string data** — case, whitespace, typos
3. **Factors** are categorical data with fixed levels
4. **Factor order matters** for plots and tables
5. **Use forcats** (`fct_*()`) to manipulate factors
6. **Order factors logically** — not alphabetically
7. When in doubt, **check the data type** with `class()` or `glimpse()`

## Before next class

📖 **Read:**

- R4DS Ch 19: Joins

✅ **Do:**

- Submit Assignment 6
- Check your final project data for string/factor issues
- Practice cleaning demographic variables

## The one thing to remember

Messy categories turn into messy results. `str_to_lower()` and `factor()` are your first line of defense.

See you Monday for joins!

::: {.notes}
Remind students that Assignment 6 is now assigned and Assignment 5 is due. The reading for next class is R4DS Ch 19 on joins. Preview that joins let you combine two different datasets — e.g., matching participant demographics to their response data. Encourage students to check their final project data for string and factor issues this week.
:::

