---
title: "Joins"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "May 13, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 14"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
    include-in-header:
      text: |
        <link rel="stylesheet" href="toc-nav.css">
    include-after-body:
      text: |
        <script src="toc-nav.js"></script>
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Setup {visibility="hidden"}

```{r}
#| include: false
library(tidyverse)

# Toy data for join diagrams (2026 Grammy winners)
artists <- tibble(
  name  = c("Taylor Swift", "Bad Bunny", "Kendrick Lamar"),
  genre = c("Pop", "Latin", "Hip-Hop")
)

grammys_2026 <- tibble(
  name  = c("Bad Bunny", "Kendrick Lamar", "Billie Eilish"),
  award = c("Album of the Year", "Record of the Year", "Song of the Year")
)

# Row-matching diagram for individual join types
plot_join <- function(title, left_active, right_active) {
  left_df <- tibble(
    name  = c("Taylor Swift", "Bad Bunny", "Kendrick Lamar"),
    y     = c(4, 3, 2),
    group = c("left_only", "both", "both")
  ) |> mutate(alpha_val = if_else(left_active, 1, 0.2))

  right_df <- tibble(
    name  = c("Bad Bunny", "Kendrick Lamar", "Billie Eilish"),
    y     = c(3, 2, 1),
    group = c("both", "both", "right_only")
  ) |> mutate(alpha_val = if_else(right_active, 1, 0.2))

  tile_colors <- c(both = "#5B8DB8", left_only = "#E07B54", right_only = "#5BA87A")

  ggplot() +
    geom_tile(data = left_df,
              aes(x = 1, y = y, fill = group, alpha = alpha_val),
              width = 1.7, height = 0.65, color = "white", linewidth = 1.2) +
    geom_text(data = left_df,
              aes(x = 1, y = y, label = name, alpha = alpha_val),
              size = 3, color = "white", fontface = "bold") +
    geom_tile(data = right_df,
              aes(x = 3.2, y = y, fill = group, alpha = alpha_val),
              width = 1.7, height = 0.65, color = "white", linewidth = 1.2) +
    geom_text(data = right_df,
              aes(x = 3.2, y = y, label = name, alpha = alpha_val),
              size = 3, color = "white", fontface = "bold") +
    geom_segment(aes(x = 1.86, xend = 2.34, y = 3, yend = 3),
                 color = "#5B8DB8", linewidth = 0.8, linetype = "dashed") +
    geom_segment(aes(x = 1.86, xend = 2.34, y = 2, yend = 2),
                 color = "#5B8DB8", linewidth = 0.8, linetype = "dashed") +
    annotate("text", x = 1,   y = 4.7, label = "artists",
             fontface = "bold.italic", size = 4, color = "#2c3e50") +
    annotate("text", x = 3.2, y = 4.7, label = "grammys_2026",
             fontface = "bold.italic", size = 4, color = "#2c3e50") +
    scale_fill_manual(values = tile_colors) +
    scale_alpha_identity() +
    coord_cartesian(xlim = c(0, 4.3), ylim = c(0.4, 5.1)) +
    labs(title = title) +
    theme_void() +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold",
                                color = "#2c3e50", margin = margin(b = 6))
    )
}

# Venn diagram for join type summary
make_venn <- function(title, shade = c("inner", "left", "right", "full")) {
  shade <- match.arg(shade)
  r <- 0.9; cx_l <- -0.5; cx_r <- 0.5
  theta <- seq(0, 2 * pi, length.out = 300)
  yi  <- sqrt(r^2 - cx_l^2)       # y-coord of intersection points (~0.748)
  a_l <- atan2(yi, -cx_l)          # angle from left center to intersect (~56°)
  b   <- pi - a_l                  # angle from right center to intersect (~124°)

  lc   <- tibble(x = cx_l + r * cos(theta), y = r * sin(theta))
  rc   <- tibble(x = cx_r + r * cos(theta), y = r * sin(theta))
  lens <- tibble(
    x = c(cx_l + r * cos(seq( a_l, -a_l, length.out = 150)),
          cx_r + r * cos(c(seq(-b, -pi, length.out = 75),
                            seq( pi,  b, length.out = 75)))),
    y = c(         r * sin(seq( a_l, -a_l, length.out = 150)),
                   r * sin(c(seq(-b, -pi, length.out = 75),
                              seq( pi,  b, length.out = 75))))
  )

  fill_col <- "#5B8DB8"
  p <- ggplot() +
    coord_fixed(xlim = c(-1.6, 1.6), ylim = c(-1.2, 1.4)) +
    theme_void() +
    labs(title = title) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold",
                                    size = 11, color = "#2c3e50",
                                    margin = margin(b = 4)))

  if (shade %in% c("left",  "full")) p <- p + geom_polygon(data = lc,   aes(x, y), fill = fill_col, alpha = 0.7)
  if (shade %in% c("right", "full")) p <- p + geom_polygon(data = rc,   aes(x, y), fill = fill_col, alpha = 0.7)
  if (shade == "inner")              p <- p + geom_polygon(data = lens, aes(x, y), fill = fill_col, alpha = 0.9)

  p +
    geom_path(data = lc, aes(x, y), color = "gray40", linewidth = 0.8) +
    geom_path(data = rc, aes(x, y), color = "gray40", linewidth = 0.8) +
    annotate("text", x = cx_l, y = -(r + 0.2),
             label = "artists",      size = 2.8, fontface = "italic", color = "#2c3e50") +
    annotate("text", x = cx_r, y = -(r + 0.2),
             label = "grammys_2026", size = 2.8, fontface = "italic", color = "#2c3e50")
}
```

# Why join data? {background-color="#2c3e50"}

## Three files, one question

Your participant took a survey on Monday and did a cognitive task on Tuesday. The survey data is in one CSV. The task data is in another. Their demographics are in a third.

. . .

You want to ask: *"Did anxious participants respond more slowly?"*

. . .

You can't answer that from any single file. You need **joins** — the tool that connects separate tables through a shared key.

::: {.notes}
Open by asking students if they have ever tried to combine two spreadsheets by hand — copying and pasting columns, sorting to line things up. That pain point is the motivation for today. Spend ~3 minutes here. This example (survey + task + demographics) maps directly onto what many of them will face in their final projects.
:::

## Two tables, one question

```{r}
# Participant information
participants <- tibble(
  participant_id = 1:4,
  age = c(22, 25, 19, 31),
  condition = c("Control", "Treatment", "Treatment", "Control")
)

# Survey responses (collected separately)
survey <- tibble(
  participant_id = c(1, 2, 3, 5),  # Note: 4 is missing, 5 is extra
  depression = c(12, 18, 10, 25),
  anxiety = c(15, 20, 12, 30)
)
```

## The two tables

:::: {.columns}
::: {.column width="50%"}
**Participants:**
```{r}
participants
```
:::

::: {.column width="50%"}
**Survey:**
```{r}
survey
```
:::
::::

. . .

How do we combine these based on `participant_id`?

::: {.notes}
Point out the deliberate mismatch: participant 4 is in the left table but not the right, and participant 5 is in the right but not the left. Ask students to predict what should happen to those rows. This asymmetry is the whole reason different join types exist, and it will recur throughout the deck.
:::

# Keys {background-color="#2c3e50"}

## What are keys?

**Keys** are variables that uniquely identify observations and connect tables.

. . .

**Primary key:** Uniquely identifies each observation in a table

- `participant_id` in the participants table
- Each participant appears only once

. . .

**Foreign key:** References the primary key of another table

- `participant_id` in the survey table
- Links back to the participants table

::: {.notes}
Students sometimes confuse keys with row numbers. Emphasize that a key is a meaningful identifier that lives in the data itself, not the row index. A useful analogy: your student ID number follows you across the registrar, financial aid, and library systems — it is the key that links your records across those separate databases.
:::

## Checking for unique keys

Before joining, verify that your key truly identifies each row:

```{r}
# Check if participant_id is unique
participants |>
  count(participant_id) |>
  filter(n > 1)
```

. . .

Empty result = good! Each participant appears only once.

## When keys aren't unique

Sometimes keys aren't unique (e.g., longitudinal data):

```{r}
longitudinal <- tibble(
  participant_id = c(1, 1, 2, 2, 3, 3),
  timepoint = c(1, 2, 1, 2, 1, 2),
  depression = c(20, 15, 18, 12, 25, 22)
)
```

. . .

Here, the **combination** of `participant_id` + `timepoint` is the key:

```{r}
longitudinal |>
  count(participant_id, timepoint) |>
  filter(n > 1)
```

::: {.notes}
This is a common point of confusion. Students think "participant_id is always the key," but in repeated-measures data it is not unique on its own. Ask them: "If you have 3 participants measured at 2 timepoints, how many rows do you have?" This sets up the compound key idea naturally. The count-and-filter pattern for checking uniqueness is worth having them type out — they will use it often.
:::

# Mutating joins {background-color="#2c3e50"}

## The join family

**Mutating joins** add columns from one table to another:

| Function | Keeps rows from... |
|----------|-------------------|
| `left_join()` | Left table (all rows) |
| `right_join()` | Right table (all rows) |
| `inner_join()` | Both tables (only matches) |
| `full_join()` | Both tables (all rows) |

::: {.notes}
Pause here and explain why they are called "mutating" joins — like mutate(), they add new columns. The table is a reference slide; don't belabor it. Tell students you will walk through each type one at a time with the same example so they can see exactly how the output differs. Spend ~1 minute on this overview before moving on.
:::

## left_join(): Most common

Keep **all** rows from the left table, add matching data from the right:

```{r}
participants |>
  left_join(survey, by = "participant_id")
```

. . .

- Participant 4 has no survey data → `NA`
- Participant 5 from survey isn't in participants → excluded

::: {.notes}
This is the most important slide in the first half of the deck. Walk through the output row by row. Ask: "Why does participant 4 have NA for depression and anxiety?" and "Where did participant 5 go?" Students often ask whether the left table is the one before the pipe or the one inside left_join() — clarify that the piped-in table is always the left table. Spend ~3 minutes here.
:::

## Visualizing left_join()

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
#| fig-alt: "Diagram showing left_join keeps all three rows from the left table (artists) and matches two rows from the right table (grammys_2026). The unmatched right-table row (Billie Eilish) is dimmed, indicating it is excluded."
plot_join(
  "left_join(artists, grammys_2026)",
  left_active  = c(TRUE, TRUE, TRUE),
  right_active = c(TRUE, TRUE, FALSE)
)
```

## right_join(): The mirror

Keep **all** rows from the right table:

```{r}
participants |>
  right_join(survey, by = "participant_id")
```

. . .

Now participant 5 is included (with `NA` for age/condition), but participant 4 is excluded.

## Visualizing right_join()

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
#| fig-alt: "Diagram showing right_join keeps all three rows from the right table (grammys_2026) and matches two rows from the left table (artists). The unmatched left-table row (Taylor Swift) is dimmed, indicating it is excluded."
plot_join(
  "right_join(artists, grammys_2026)",
  left_active  = c(FALSE, TRUE, TRUE),
  right_active = c(TRUE, TRUE, TRUE)
)
```

## inner_join(): Only matches

Keep **only** rows that exist in both tables:

```{r}
participants |>
  inner_join(survey, by = "participant_id")
```

. . .

- Participant 4 excluded (no survey data)
- Participant 5 excluded (not in participants)

::: {.notes}
Emphasize that inner_join() silently drops rows that do not match, which means you lose data without a warning. In psychology research this is usually not what you want — you want to know who is missing, not hide them. This is why left_join() is the default recommendation. Students may ask "when would I actually use inner_join?" — a fair answer is when you only want to analyze people who completed both measures and you have already documented the attrition.
:::

## Visualizing inner_join()

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
#| fig-alt: "Diagram showing inner_join keeps only the two rows that match in both tables (Bad Bunny and Kendrick Lamar). Unmatched rows on both sides are dimmed, indicating they are excluded."
plot_join(
  "inner_join(artists, grammys_2026)",
  left_active  = c(FALSE, TRUE, TRUE),
  right_active = c(TRUE, TRUE, FALSE)
)
```

## full_join(): Everything

Keep **all** rows from both tables:

```{r}
participants |>
  full_join(survey, by = "participant_id")
```

. . .

Every participant appears, with `NA` where data is missing.

::: {.notes}
Point out that full_join() gives you the most complete picture — nobody is dropped. But it also means you may have rows with lots of NAs, which connects directly to next session on missing data. This is a nice foreshadowing moment.
:::

## Visualizing full_join()

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
#| fig-alt: "Diagram showing full_join keeps all rows from both tables. All tiles are fully opaque, indicating no rows are excluded regardless of whether they match."
plot_join(
  "full_join(artists, grammys_2026)",
  left_active  = c(TRUE, TRUE, TRUE),
  right_active = c(TRUE, TRUE, TRUE)
)
```

## Join types at a glance

```{r}
#| echo: false
#| layout-ncol: 2
#| fig-width: 4
#| fig-height: 2.8
#| fig-alt:
#|   - "Venn diagram for inner_join showing only the overlapping region shaded, representing rows that exist in both tables."
#|   - "Venn diagram for left_join showing the entire left circle shaded, representing all rows from the left table plus matches from the right."
#|   - "Venn diagram for right_join showing the entire right circle shaded, representing all rows from the right table plus matches from the left."
#|   - "Venn diagram for full_join showing both circles entirely shaded, representing all rows from both tables."
make_venn("inner_join()", "inner")
make_venn("left_join()",  "left")
make_venn("right_join()", "right")
make_venn("full_join()",  "full")
```

## Which join should I use?

**Most common in psychology: `left_join()`**

. . .

Use when:

- You have a main participant table
- You want to add additional data (surveys, behavioral tasks)
- You want to keep all participants, even those with missing data

. . .

::: {.callout-tip}
Start with `left_join()` unless you have a specific reason to use another type.
:::

::: {.notes}
Reinforce this strongly: "If you remember nothing else from today, remember left_join()." In practice, 90% of joins students will do in this course and in research are left joins. The mental model is simple: "I have a main table and I want to stick more columns onto it." Spend ~2 minutes here before transitioning to the psychology example.
:::

## Psychology example: Adding demographics

```{r}
# Task data
task_data <- tibble(
  participant_id = c(1, 2, 3, 1, 2, 3),
  trial = c(1, 1, 1, 2, 2, 2),
  reaction_time = c(450, 380, 520, 430, 395, 510)
)

# Add participant info to each trial
task_data |>
  left_join(participants, by = "participant_id")
```

::: {.notes}
Draw attention to what happened to the row count. The task_data table has 6 rows (3 participants x 2 trials) and the participants table has 4 rows, but the result still has 6 rows. This is because left_join() matched each trial row to the participant's demographics and repeated the demographic info for each trial. Students often expect the result to have more or fewer rows — this is a good moment to check their intuition.
:::

## Joining by multiple keys

For longitudinal data, join on multiple columns:

```{r}
# Time 1 scores
time1 <- tibble(
  id = c(1, 2, 3),
  time = 1,
  depression = c(20, 18, 25)
)

# Time 2 scores
time2 <- tibble(
  id = c(1, 2, 3),
  time = 2,
  depression = c(15, 12, 22)
)
```

## Combining longitudinal data

```{r}
# Combine into one table
all_times <- bind_rows(time1, time2)

# Add baseline demographics
all_times |>
  left_join(
    participants |> select(participant_id, age, condition),
    by = c("id" = "participant_id")  # Keys have different names
  )
```

::: {.notes}
Two things to highlight here. First, this is the first time students see the named-vector syntax for `by` when keys have different names — slow down and explain that the left side of the equals sign is the column name in the left table and the right side is the column name in the right table. Second, note the use of select() inside the join to avoid bringing in columns you do not need. Students often ask "can I just join everything?" — yes, but selecting first keeps your result tidy.
:::

# Pair coding break {background-color="#e67e22"}

## Your turn: Combine study data

You have three tables from a therapy study:

```{r}
baseline <- tibble(
  id = 1:5,
  age = c(25, 30, 22, 35, 28),
  baseline_depression = c(22, 25, 18, 30, 20)
)

treatment <- tibble(
  id = c(1, 2, 3, 4),  # Missing id 5
  condition = c("CBT", "Control", "CBT", "Control")
)

followup <- tibble(
  id = c(1, 2, 3, 5),  # Missing id 4, but has id 5
  followup_depression = c(12, 23, 10, 18)
)
```

## Your turn: Combine study data

1. Create a dataset with baseline info and treatment condition (keep all participants)
2. Add followup data (keep all from step 1)
3. How many participants are missing followup data?

**Time: 10 minutes**

::: {.notes}
Circulate during the exercise. The most common mistake is using inner_join() instead of left_join(), which silently drops participant 5 in step 1. If pairs finish early, ask them to figure out which participants are missing followup data and whether missingness is related to condition. That bridges to the filtering joins section and next week's missing data session. When time is up, walk through the solution and emphasize step 3 — the sum(is.na()) pattern will come back next session.
:::

```{r}
#| echo: false
#| eval: false
# SOLUTION
# Step 1: Add treatment to baseline
baseline_treatment <- baseline |>
  left_join(treatment, by = "id")

# Step 2: Add followup
complete_data <- baseline_treatment |>
  left_join(followup, by = "id")

complete_data

# Step 3: Count missing followup
sum(is.na(complete_data$followup_depression))
```

---

## Before we move on

📤 **[Submit your code on Canvas](https://canvas.uoregon.edu/courses/287793/pages/pair-coding-check-ins)** for participation credit. Paste what you have — it doesn't need to work perfectly.

# Filtering joins {background-color="#2c3e50"}

## A different purpose

**Filtering joins** don't add columns — they filter rows based on whether matches exist:

- `semi_join()` — Keep rows that **have** a match
- `anti_join()` — Keep rows that **don't have** a match

::: {.notes}
Transition cue: "So far every join we have seen adds columns. Filtering joins are different — they only keep or remove rows." Emphasize the contrast: mutating joins change the width of your table, filtering joins change the height. Students sometimes wonder why you would not just use filter() — the answer is that filtering joins let you filter based on whether a match exists in another table, which filter() alone cannot do.
:::

## semi_join(): Has a match

Keep rows from the left table that exist in the right table:

```{r}
# Which participants completed the survey?
participants |>
  semi_join(survey, by = "participant_id")
```

. . .

Only participants 1, 2, and 3 completed the survey.

## anti_join(): No match

Keep rows from the left table that **don't** exist in the right table:

```{r}
# Which participants didn't complete the survey?
participants |>
  anti_join(survey, by = "participant_id")
```

. . .

Participant 4 didn't complete the survey.

## Psychology use case: Attrition analysis

```{r}
# Who was enrolled
enrolled <- tibble(
  participant_id = 1:10,
  condition = rep(c("Treatment", "Control"), each = 5))

# Who completed
completed <- tibble(
  participant_id = c(1, 2, 3, 4, 7, 8, 9, 10))  # Missing 5 and 6

# Who dropped out?
enrolled |>
  anti_join(completed, by = "participant_id")
```

::: {.notes}
This is the "aha" moment for anti_join(). Walk through it slowly: "enrolled has 10 people, completed has 8 — anti_join gives us the 2 who are in enrolled but not in completed." This is exactly what psychologists need for attrition analyses, and students will use it on their final projects if they have longitudinal or multi-session data. Linger on this example — it is more memorable than the abstract definition.
:::

## Attrition by condition

```{r}
dropout <- enrolled |>
  anti_join(completed, by = "participant_id")

dropout |>
  count(condition)
```

. . .

::: {.callout-warning}
Both dropouts were from the Treatment condition — this could bias results!
:::

::: {.notes}
This slide connects back to Session 1's discussion of the replication crisis. If participants drop out of the treatment condition because the treatment is unpleasant, your remaining sample is biased — only the people who tolerated the treatment are left. This is differential attrition and it is a real threat to internal validity. Emphasize: "The code is simple, but the implication for your research is huge."
:::

# Common join problems {background-color="#2c3e50"}

## Problem 1: Keys with different names

:::: {.columns}
::: {.column width="50%"}
```{r}
demographics <- tibble(
  subj_id = 1:3,
  age = c(25, 30, 22)
)

scores <- tibble(
  participant = c(1, 2, 3),
  depression = c(18, 22, 15)
)
```
:::

::: {.column width="50%"}
**Solution:** Use named vector in `by`:

```{r}
demographics |>
  left_join(scores,
    by = c("subj_id" = "participant"))
```
:::
::::

::: {.notes}
This is the single most common join problem students will encounter with real data. Different labs, different Qualtrics surveys, different RA conventions — the ID column is almost never named the same thing across files. Remind them of the syntax: left table name on the left side of the equals sign, right table name on the right. If they forget the `by` argument entirely, R will try to join on all shared column names, which often produces wrong results silently.
:::

## Problem 2: Duplicate keys

```{r}
# Duplicate in left table
participants_dup <- tibble(
  id = c(1, 1, 2, 3),  # ID 1 appears twice!
  age = c(25, 25, 30, 22)
)

task_scores <- tibble(
  id = c(1, 2, 3),
  score = c(85, 90, 88)
)
```

## What happens with duplicates?

```{r}
participants_dup |>
  left_join(task_scores, by = "id")
```

. . .

Each duplicate gets matched → creates extra rows!

. . .

::: {.callout-important}
Always check for duplicate keys before joining.
:::

::: {.notes}
This is the sneakiest bug students will encounter. The join runs without error, but you end up with more rows than expected. Students often do not notice until their summary statistics look wrong. Reinforce the habit: always run the count-and-filter check before joining. Ask: "If your table had 100 participants and after a join it has 105 rows, what happened?" Duplicates.
:::

## Problem 3: Missing keys

```{r}
# Some participants have NA for id
messy_data <- tibble(
  id = c(1, 2, NA, 3),
  response = c("Yes", "No", "Yes", "No")
)

demographics <- tibble(
  id = 1:3,
  age = c(25, 30, 22)
)
```

## Joining with NAs

```{r}
messy_data |>
  left_join(demographics, by = "id")
```

. . .

Row with `NA` id can't match → gets `NA` for age.

Fix the data before joining!

## Problem 4: Many-to-many relationships

```{r}
# Multiple trials per participant
trials <- tibble(
  id = c(1, 1, 2, 2),
  trial = c(1, 2, 1, 2),
  rt = c(450, 430, 380, 390)
)

# Multiple sessions per participant
sessions <- tibble(
  id = c(1, 1, 2, 2),
  session = c(1, 2, 1, 2),
  mood = c(5, 6, 4, 5)
)
```

## Many-to-many result

```{r}
trials |>
  left_join(sessions, by = "id")
```

. . .

Creates all combinations — probably not what you want!

**Solution:** Be more specific about what you're joining on, or reshape first.

::: {.notes}
Many-to-many joins are rare in well-structured data, but students occasionally create them by accident (e.g., joining on participant ID alone when they should also join on session or timepoint). Point out that the result has 8 rows instead of the expected 4 — every trial got matched with every session for the same participant. The fix is usually to add another column to the `by` argument to make the key unique, or to reshape one of the tables first. You do not need to dwell here — just make sure they recognize the symptom (unexpected row explosion).
:::

# Practical psychology examples {background-color="#2c3e50"}

## Example 1: Qualtrics + demographics

```{r}
# Survey responses exported from Qualtrics
qualtrics <- tibble(
  ResponseId = c("R_1", "R_2", "R_3"),
  PHQ9_total = c(12, 18, 8),
  GAD7_total = c(10, 15, 6)
)

# Demographics collected separately
demo <- tibble(
  ResponseId = c("R_1", "R_2", "R_3"),
  age = c(25, 30, 22),
  major = c("Psychology", "Sociology", "Psychology")
)
```

::: {.notes}
Transition cue: "Let me show you three examples of how this works with real psychology workflows." The Qualtrics example is the most immediately relevant for students who are running their own studies. Mention that Qualtrics exports include ResponseId automatically, which makes it a natural join key when you have separate survey blocks exported as different CSVs.
:::

## Example 1: Result

```{r}
qualtrics |>
  left_join(demo, by = "ResponseId")
```

## Example 2: Pre-post intervention

```{r}
pre_test <- tibble(
  participant = 1:4,
  depression_pre = c(22, 25, 18, 20)
)

post_test <- tibble(
  participant = c(1, 2, 4),  # 3 dropped out
  depression_post = c(12, 23, 15)
)
```

## Example 2: Result

```{r}
# Combine and compute change
pre_test |>
  left_join(post_test, by = "participant") |>
  mutate(change = depression_pre - depression_post)
```

## Example 3: Item-level to scale-level

```{r}
# Individual items
items <- tibble(
  id = rep(1:3, each = 5),
  item = rep(1:5, times = 3),
  response = c(3,4,3,4,3, 2,3,2,3,2, 4,5,4,5,4)
)

# Participant info
participants_info <- tibble(
  id = 1:3,
  condition = c("Control", "Treatment", "Treatment")
)
```

## Example 3: Result

```{r}
# Compute scale scores then join
items |>
  group_by(id) |>
  summarize(scale_mean = mean(response)) |>
  left_join(participants_info, by = "id")
```

::: {.notes}
This example shows a common pattern: summarize first, then join. Students often try to join first and then summarize, which can work but is harder to reason about (especially if the join changes the number of rows). Encourage the "reduce then combine" workflow: compute your scale scores, then attach the metadata. This is also a nice callback to the group_by + summarize skills from earlier in the course.
:::

# End-of-deck exercise {background-color="#e67e22"}

## Practice: Longitudinal study

::: {.notes}
This exercise is a head start on Assignment 7 (Joins and Missing Data). It combines multiple join types — left_join for building the dataset, anti_join for identifying dropout — and asks students to think about attrition patterns. Give them ~10 minutes. If they get stuck on task 3 (identifying dropout at each wave), hint that anti_join is the right tool. Task 4 (dropout by condition) is the stretch goal; do not worry if not everyone gets there.
:::

You have data from a three-wave longitudinal study:

```{r}
# Baseline demographics
baseline_demo <- tibble(
  pid = 1:6,
  age = c(20, 22, 25, 19, 24, 21),
  condition = rep(c("Intervention", "Control"), each = 3)
)

# Time 1 assessment
time1_scores <- tibble(
  pid = 1:6,
  depression_t1 = c(22, 25, 18, 20, 24, 19)
)

# Time 2 assessment (2 dropouts)
time2_scores <- tibble(
  pid = c(1, 2, 3, 4, 6),  # Missing 5
  depression_t2 = c(15, 24, 12, 18, 17)
)

# Time 3 assessment (1 more dropout)
time3_scores <- tibble(
  pid = c(1, 2, 3, 6),  # Missing 4 and 5
  depression_t3 = c(12, 22, 10, 15)
)
```

## Your tasks

1. Create a complete dataset with all timepoints
2. Compute change scores (T1 to T3)
3. Identify who dropped out at each wave
4. Check if dropout differs by condition

```{r}
#| echo: false
#| eval: false
# SOLUTION

# 1. Complete dataset
complete <- baseline_demo |>
  left_join(time1_scores, by = "pid") |>
  left_join(time2_scores, by = "pid") |>
  left_join(time3_scores, by = "pid")

complete

# 2. Change scores
complete |>
  mutate(change_t1_t3 = depression_t1 - depression_t3)

# 3. Who dropped out at T2?
baseline_demo |>
  anti_join(time2_scores, by = "pid")

# Who dropped out at T3 (but completed T2)?
time2_scores |>
  anti_join(time3_scores, by = "pid")

# 4. Dropout by condition
baseline_demo |>
  anti_join(time3_scores, by = "pid") |>
  count(condition)
```

# Wrapping up {background-color="#2c3e50"}

## Join decision flowchart

:::: {.columns}
::: {.column width="50%"}
Ask yourself:

1. **Do I want to add columns?**
   - Yes → Use a mutating join
   - No → Use a filtering join

2. **Which rows do I want to keep?**
   - All from left → `left_join()`
   - All from right → `right_join()`
   - Only matches → `inner_join()`
   - Everything → `full_join()`
:::

::: {.column width="50%"}
3. **Am I filtering rows?**
   - Keep matches → `semi_join()`
   - Keep non-matches → `anti_join()`
:::
::::

::: {.notes}
Walk through this flowchart as a decision tree. Consider having students practice with a quick verbal exercise: you describe a scenario and they call out which join to use. For example: "I want to add survey scores to my participant table" (left_join), "I want to find out which participants did NOT complete the follow-up" (anti_join). Two or three of these takes about 2 minutes and solidifies the mental model.
:::

## Key takeaways

1. **Joins combine data** from multiple tables using key variables
2. **Check your keys** for uniqueness before joining
3. **`left_join()` is your workhorse** — keeps all rows from your main table
4. **Filtering joins** help identify completers vs. dropouts
5. **Watch for common problems:** duplicate keys, different key names, NAs
6. **Always inspect** the result to ensure it matches expectations

## Before next class

📖 **Read:**

- R4DS Ch 18: Missing values

✅ **Do:**

- Submit Assignment 6
- Check if your final project needs joins
- Practice joining your own data files

## The one thing to remember

If your data lives in separate files, joins are how you ask a question that spans all of them.

See you Wednesday for handling missing data!

::: {.notes}
End strong on this takeaway. Remind students that Assignment 6 is due before next class, and that they should check whether their final project data will require any joins — if so, now is the time to start thinking about what the key variable is. Next session covers missing data, which pairs naturally with joins since joins often introduce NAs.
:::
