---
title: "Missing Data"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "May 13, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 14"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    code-copy: false
    incremental: false
    include-in-header:
      text: |
        <link rel="stylesheet" href="toc-nav.css">
    include-after-body:
      text: |
        <script src="toc-nav.js"></script>
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Setup {visibility="hidden"}

```{r}
#| include: false
library(tidyverse)
```

# Why missing data matters {background-color="#2c3e50"}

## The data you don't have

In a typical longitudinal psychology study, 30–50% of participants drop out before the final wave.

. . .

If you just delete their data, you might be throwing away the most important part of the story — because **who drops out** is rarely random.

. . .

Today we learn to detect, explore, and handle missing data honestly.

::: {.notes}
**PACING (~80 min total):** Why missing data matters (~5). Types of missing data (~8). Exploring missing data (~10). Handling missing data (~15). Implicit→explicit (~8). Pair coding (~10). Mechanisms + reporting (~12). Multiple imputation (~2). Wrap-up (~5). If running short, cut the multiple imputation section entirely and shorten the MCAR/MAR/MNAR examples — the reporting paragraph at the end is the more important takeaway.

**A6 reminder:** Make sure to flag at the top of class (and again at the end) that Assignment 6 is due Sunday May 17 at 11:59 PM — that's this weekend, so it's the most pressing deadline on their plate.

Ask if anyone ran into NAs while exploring data for their final project. The 30-50% statistic is a real number from attrition research and tends to surprise students. Emphasize the word "honestly" — this session is as much about research ethics and transparency as it is about R functions. Spend ~2 minutes on this opening.
:::

# Types of missing data {background-color="#2c3e50"}

## Explicit missing: NA

**Explicit missing** means you can see the `NA`:

```{r}
survey <- tibble(
  participant = 1:5,
  age = c(25, NA, 30, 22, NA),
  depression = c(12, 18, NA, 10, 15)
)

survey
```

The `NA` values are obvious.

::: {.notes}
Students have seen NA before (it shows up whenever a function can't compute a value, or when data is missing from a CSV), but this is the first time we treat it as a topic in its own right. Make sure they understand that NA means "we do not know" — not zero, not blank, not "no answer." R treats NA differently from any other value, which is why things like `NA == NA` return NA, not TRUE. You do not need to go deep on that quirk, but mention it because students will stumble on it eventually.
:::

## Implicit missing: Rows don't exist

**Implicit missing** means entire rows are absent:

```{r}
appointments <- tibble(
  name = c("Alice", "Bob", "Alice", "Carol"),
  day = c("Mon", "Mon", "Wed", "Wed"),
  attended = c(TRUE, TRUE, TRUE, TRUE)
)

appointments
```

. . .

Who didn't show up? You can't tell because they're not in the data!

::: {.notes}
This distinction (explicit vs. implicit) is one of the most important conceptual takeaways. Ask students: "Bob is not in the Wednesday data. Did he attend and the data was not recorded? Or did he not attend? Or did he cancel?" The data cannot tell us. This is much harder to detect than explicit NAs because there is nothing to see — the row simply does not exist. We will see complete() later in this deck as the main tool for surfacing implicit missingness within a single dataset.
:::

## Why implicit missing matters

In longitudinal studies, missing rows often mean dropout:

```{r}
# Three-wave study
longitudinal <- tibble(
  id = c(1, 1, 1, 2, 2, 3),  # Person 2 missing wave 3, person 3 missing waves 2 and 3
  wave = c(1, 2, 3, 1, 2, 1),
  depression = c(20, 15, 12, 25, 22, 18)
)

longitudinal
```

Person 2 and 3's missing waves are **implicit** — they're not `NA`, they're just absent.

# Exploring missing data {background-color="#2c3e50"}

## Checking for NAs

```{r}
survey <- tibble(
  id = 1:6,
  age = c(25, NA, 30, 22, NA, 28),
  depression = c(12, 18, NA, 10, 15, NA),
  anxiety = c(15, 20, 12, NA, 18, 16)
)

# Check if any NAs exist
any(is.na(survey))
```

. . .

```{r}
# Count total NAs
sum(is.na(survey))
```

::: {.notes}
Walk through these two functions as the first step in any missing-data analysis. The any() check is a quick yes/no diagnostic; sum() tells you the scale of the problem. Point out that is.na() returns TRUE/FALSE, and since R treats TRUE as 1 and FALSE as 0, sum() counts the TRUEs — a useful trick that recurs throughout the tidyverse.
:::

## NAs by column

```{r}
# Count NAs in each column
survey |>
  summarize(
    age_missing = sum(is.na(age)),
    depression_missing = sum(is.na(depression)),
    anxiety_missing = sum(is.na(anxiety))
  )
```

## Better approach: across()

`across(columns, function)` applies the same function to multiple columns at once:

```{r}
survey |>
  summarize(
    across(
      everything(),       # which columns: all of them
      ~ sum(is.na(.x))    # what to do: count NAs (.x = "this column")
    )
  )
```

::: {.notes}
The across() + everything() pattern is essentially new for most students — they only saw it briefly in Session 10 and were told to focus on output, not mechanics. Walk through the anatomy slowly: across() takes (1) which columns and (2) what to do to each. The tilde (~) plus .x is shorthand for "for each column, do this." Spend ~3 minutes on this slide since the syntax is unfamiliar.
:::

## Or as proportions

```{r}
survey |>
  summarize(
    across(
      everything(),
      ~ mean(is.na(.x))
    )
  )
```

::: {.notes}
Swapping sum() for mean() turns counts into proportions — a useful trick that reuses the TRUE = 1 / FALSE = 0 idea. This is especially useful for reporting: "33% of participants were missing age data." Students often ask whether to report counts or proportions; the answer is both, ideally.
:::

## Visualizing missingness

The `naniar` package provides great visualization tools:

```{r}
#| fig-alt: "Heatmap showing the pattern of missing values across all variables in the survey data. Each row is a participant and each column is a variable, with missing cells highlighted to reveal where data gaps occur."
library(naniar)

# Visual summary
vis_miss(survey)
```

::: {.notes}
The naniar package is optional — students do not need to install it for assignments. But the vis_miss() plot is a great diagnostic tool worth showing. Each column is a variable, each row is a participant, and missing cells are colored. This gives an immediate visual sense of where the holes are. If naniar is not installed on your machine, you can skip this live and just show the rendered output.
:::

# Handling missing data {background-color="#2c3e50"}

## Strategy 1: Complete case analysis

**Complete case analysis** (listwise deletion) removes any row with *any* `NA`:

```{r}
survey |>
  drop_na()
```

::: {.notes}
Transition cue: "Now that we know how to find missing data, let's talk about what to do about it." Students often default to this strategy because it is the simplest — the next slide shows why that can be dangerous. (Aside, if students ever see `na.omit()` elsewhere: it's the base R equivalent of `drop_na()` — same behavior.)
:::

## Dangers of complete case analysis

```{r}
# Started with 6 participants
nrow(survey)

# Only 2 complete cases
survey |>
  drop_na() |>
  nrow()
```

. . .

::: {.callout-warning}
You just lost 67% of your data!
:::

::: {.notes}
Let this number land. Going from 6 to 2 participants is dramatic, and it gets worse with real datasets that have many variables — each additional variable with any missingness compounds the loss. In a study with 20 survey items, even a small per-item missing rate can reduce your sample to almost nothing if you use listwise deletion. Ask: "Would you trust a result based on 2 participants?" This motivates the selective dropping approach on the next slide.
:::

## Selective dropping

Only drop rows missing *specific* variables:

```{r}
# Drop only if depression is missing
survey |>
  drop_na(depression)
```

## Drop on multiple variables

```{r}
# Drop only if depression OR anxiety is missing
survey |>
  drop_na(depression, anxiety)
```

## When is dropping okay?

**Dropping is fine when:**

- Missing data is rare (< 5-10%)
- Missingness is truly random
- You have adequate sample size

. . .

**Be cautious when:**

- Missing data is common (> 20%)
- Certain groups have more missingness (systematic bias)
- Sample size is small

::: {.notes}
The 5-10% and 20% thresholds are rules of thumb, not hard cutoffs. The key question is always "does the missingness bias my results?" If depressed people are the ones skipping the depression questionnaire, then even 5% missing data is a problem. Emphasize that the amount of missing data matters less than the pattern. This sets up the MCAR/MAR/MNAR discussion later in the deck.
:::

## Strategy 2: Filling values

Sometimes you can **reasonably fill** missing values:

```{r}
# Carry forward the last observation
time_series <- tibble(
  day = 1:5,
  mood = c(5, NA, NA, 4, 6)
)

time_series |>
  fill(mood)  # Fills downward by default
```

::: {.notes}
fill() is called "last observation carried forward" (LOCF) in clinical research. It assumes the missing values are the same as the most recent non-missing value. This is reasonable for things like demographic variables that do not change between rows (e.g., filling in a participant's age across multiple trials), but more questionable for outcome variables. Students often want to use fill() for everything — the warning slide that follows is important.
:::

## Fill directions

```{r}
# Fill upward
time_series |>
  fill(mood, .direction = "up")
```

## When is filling okay?

**Filling makes sense for:**

- Time series with repeated measures (carry forward last observation)
- Grouping variables that apply to multiple rows
- Values that don't change often

. . .

::: {.callout-warning}
**NEVER fill** when it means making up data you don't have!
:::

## Strategy 3: Replace with a specific value

```{r}
# Replace NAs with a value
survey |>
  mutate(
    age = replace_na(age, 99),  # Code 99 = "no response"
    depression = replace_na(depression, -999)  # Obvious invalid code
  )
```

::: {.notes}
Students often ask "why not just replace NA with the mean?" Replacing with the mean is a form of imputation that shrinks variance and can bias results — it is generally not recommended in modern practice. The placeholder codes shown here (99, -999) are a SPSS-era convention for marking missing data that should not be analyzed. If you use them, you must filter them out before computing any statistics. Generally, leaving values as NA is safer and more transparent.
:::

## Strategy 3: Replace with a specific value

::: {.callout-important}
If you use placeholder codes, **document them clearly** and make sure they can't be mistaken for real data.
:::

## Strategy 4: Leave them as NA

Often the best approach is to **keep NAs** and handle them in analysis:

```{r}
# Most functions have na.rm argument
survey |>
  summarize(
    mean_age = mean(age, na.rm = TRUE),
    mean_depression = mean(depression, na.rm = TRUE)
  )
```

. . .

This is transparent about what data you have.

::: {.notes}
Emphasize that this is often the best default for students at this level. Keeping NAs and using na.rm = TRUE is honest and simple. Remind them that without na.rm = TRUE, mean() will return NA if any values are missing — this is R's way of saying "I cannot give you a trustworthy answer with incomplete data." Students frequently forget na.rm and get confused by NA results; this is the fix.
:::

# Implicit missing → Explicit missing {background-color="#2c3e50"}

## The problem with implicit missing

```{r}
#| code-fold: true
#| 
study_completion <- tibble(
  participant = c(1, 1, 1, 2, 2, 3, 3),
  timepoint = c(1, 2, 3, 1, 2, 1, 3),
  depression = c(20, 15, 12, 25, 22, 18, 14)
)

study_completion
```

. . .

Who's missing which timepoints? Hard to tell.

## complete(): Make implicit missing explicit

```{r}
study_completion |>
  complete(participant, timepoint)
```

. . .

Now we can see: Participant 2 missing timepoint 3, Participant 3 missing timepoint 2.

::: {.notes}
complete() is the key function for this section. It generates every combination of the variables you specify and fills in NA where data is missing. Walk through the output carefully: the original data had 7 rows, but with 3 participants and 3 timepoints there should be 9 rows. The 2 new rows (with NAs for depression) represent the implicit missing data that is now explicit. complete() is the cleanest way to surface implicit missingness within a single dataset — one function generates the grid and exposes the gaps.
:::

## Why this matters

Makes dropout visible for analysis:

```{r}
study_completion |>
  complete(participant, timepoint) |>
  group_by(timepoint) |>
  summarize(
    n_completed = sum(!is.na(depression)),
    n_missing = sum(is.na(depression))
  )
```

## Flagging completion

Common pattern: make implicit explicit, then flag who completed each timepoint:

```{r}
study_completion |>
  complete(participant, timepoint) |>
  mutate(
    completed = if_else(is.na(depression), FALSE, TRUE)
  )
```

This `completed` flag is the building block for attrition analysis — coming up later in the deck.

# Pair coding break {background-color="#e67e22"}

## Your turn: Analyze missing data patterns

You have survey data from a therapy study:

```{r}
therapy_survey <- tibble(
  id = 1:8,
  age = c(25, 30, NA, 22, 28, NA, 35, 26),
  baseline_depression = c(22, 25, 18, 20, 24, 19, NA, 21),
  followup_depression = c(12, 23, NA, 15, NA, NA, NA, 16),
  satisfaction = c(4, 3, NA, 5, 4, NA, NA, 5)
)
```

1. How many participants are missing baseline data? Followup data?
2. How many participants have **complete data** (no NAs anywhere)?
3. Create a version that drops rows missing followup data
4. What percentage of participants completed the followup?

**Time: 10 minutes**

::: {.notes}
Circulate during the exercise. For question 1, students can use sum(is.na()) or the across() approach — either is fine. The most common mistake on question 3 is using drop_na() without specifying the column, which drops too many rows. Question 4 is the most interesting: mean(!is.na(followup_depression)) gives the completion rate directly. When reviewing, connect this to reporting: "You would write something like '50% of participants completed the follow-up assessment.'"
:::

```{r}
#| echo: false
#| eval: false
# SOLUTION

# 1. Count missing by variable
therapy_survey |>
  summarize(
    across(everything(), ~sum(is.na(.x)))
  )

# 2. Complete cases
therapy_survey |>
  drop_na() |>
  nrow()

# 3. Drop if missing followup
therapy_survey |>
  drop_na(followup_depression)

# 4. Completion rate
therapy_survey |>
  summarize(
    completion_rate = mean(!is.na(followup_depression))
  )
```

---

## Before we move on

📤 **[Submit your code on Canvas](https://canvas.uoregon.edu/courses/287793/pages/pair-coding-check-ins)** for participation credit. Paste what you have — it doesn't need to work perfectly.

# Psychology-specific considerations {background-color="#2c3e50"}

## Missing data mechanisms

Statisticians distinguish three types:

1. **MCAR (Missing Completely At Random)** — Missingness unrelated to anything
2. **MAR (Missing At Random)** — Missingness related to observed variables
3. **MNAR (Missing Not At Random)** — Missingness related to the missing value itself

::: {.notes}
This is conceptual — students do not need to memorize the acronyms, but they need to understand the core idea: the reason data is missing affects whether your results are biased. Spend ~5 minutes on this slide and the next two. The names are counterintuitive (especially MAR, which sounds like it should be unproblematic but actually means missingness is related to other measured variables). Use the depression study examples on the next slide to make each type concrete.
:::

## Example: Depression study

**MCAR:** Computer randomly failed to save 5% of responses

- No bias introduced

. . .

**MAR:** Older participants more likely to skip online surveys

- Can account for this by including age as a predictor

. . .

**MNAR:** People with severe depression skip the depression questionnaire

- **This is a problem** — missing values are related to what you're measuring

## Why it matters

- **MCAR:** Complete case analysis is fine (but you lose power)
- **MAR:** More sophisticated methods can help (beyond this course)
- **MNAR:** No easy fix — missing data is fundamentally informative

. . .

::: {.callout-tip}
**Your job:** Always **document and report** how much data is missing and why you think it's missing.
:::

::: {.notes}
The practical takeaway for this course: you cannot statistically test whether data is MNAR (because you do not have the missing values to compare). So the best you can do is (1) document how much is missing, (2) check whether missingness is related to observed variables, and (3) be transparent about your assumptions. For their final projects, students should include a short paragraph about missing data even if they have very little — that is good practice.
:::

## Attrition analysis

In longitudinal studies, always check *who* drops out:

```{r}
study <- tibble(
  id = 1:10,
  condition = rep(c("Treatment", "Control"), each = 5),
  baseline_depression = c(22, 18, 25, 20, 24, 19, 21, 23, 17, 22),
  followup_depression = c(15, 12, NA, NA, NA, 14, 13, 11, 16, 18)
)

study
```

A missing followup score = a dropout.

::: {.notes}
This whole section is about applying the missing-data lens — `is.na(followup_depression)` identifies dropouts directly. Walk through the data: rows 3, 4, 5 are missing followup. Rows 1–5 are Treatment, so all three dropouts are in the Treatment condition. Set up the next slide as: "let's confirm that programmatically."
:::

## Who dropped out?

```{r}
study |>
  filter(is.na(followup_depression))
```

## Dropout by condition

```{r}
study |>
  filter(is.na(followup_depression)) |>
  count(condition)
```

. . .

All dropouts are from the Treatment condition — this could seriously bias results!

::: {.notes}
This is the punchline. If treatment dropouts are systematically different from completers (e.g., the treatment was unpleasant for some subset), then comparing post-treatment scores between conditions is comparing the people who *tolerated* the treatment to the full Control group — apples to oranges. Tie this to the MNAR discussion: this is exactly the kind of pattern that creates serious bias.
:::

## Reporting missing data

In your write-up, report:

1. **How much data is missing** (by variable)
2. **Patterns of missingness** (related to other variables?)
3. **How you handled it** (dropped? kept as NA?)
4. **Potential biases** (who's missing? does it matter?)

. . .

Example:

> "Eight participants (12%) did not complete the follow-up assessment. Dropout was unrelated to baseline depression scores (t = 1.2, p = .24). Analyses used complete case analysis (N = 60)."

::: {.notes}
Read the example paragraph aloud. This is the kind of reporting students should aim for in their final projects — at minimum, the count and percentage of missing data and a statement about how they handled it. They are not expected to run inferential tests for this course; the t-test in the example is just to show what a published version of this paragraph might look like. If a student asks how to test whether dropout is related to a categorical variable like condition (as we just demoed), the standard tool is a chi-squared test of independence — but it is well outside the scope of this course. This connects back to Session 1 and Session 16 themes about transparency and reproducibility.
:::

# Advanced topic: Multiple imputation {background-color="#2c3e50"}

## Beyond this course

More sophisticated approaches exist for handling missing data:

- **Multiple imputation** — create multiple plausible versions of missing data
- **Maximum likelihood** — estimate parameters using all available data
- **Bayesian methods** — incorporate uncertainty about missing values

. . .

Packages in R: `mice`, `Amelia`, `missForest`

. . .

::: {.callout-note}
We won't cover these methods, but know they exist for when you need them in future research!
:::

::: {.notes}
Keep this brief — ~2 minutes. The goal is awareness, not instruction. Students who go on to graduate school will encounter these methods in statistics courses. For now, just knowing the names exist is enough. If a student asks for more detail, suggest they look at the mice package vignette after the course.
:::

# Wrapping up {background-color="#2c3e50"}

## Decision tree for missing data

:::: {.columns}
::: {.column width="50%"}
1. **How much is missing?**
   - < 5%: Usually safe to drop
   - 5-20%: Investigate patterns
   - \> 20%: Be very careful

2. **Why is it missing?**
   - Random: Less concerning
   - Systematic: Potentially biasing
:::

::: {.column width="50%"}
3. **What's your plan?**
   - Drop complete cases?
   - Drop specific variables?
   - Keep as NA and use `na.rm`?
   - Fill (carefully)?

4. **Document everything!**
:::
::::

## Key takeaways

1. **Missing data is normal** in psychology research
2. **Explicit vs implicit** missing — make implicit explicit with `complete()`
3. **Explore patterns** before deciding how to handle
4. **Complete case analysis** (dropping rows) is simple but can lose power
5. **Never make up data** — be transparent about missingness
6. **Document your decisions** — report what's missing and why
7. **Check for bias** — does missingness relate to key variables?

## Functions cheat sheet

| Function | Purpose |
|----------|---------|
| `is.na()` | Check if values are missing |
| `drop_na()` | Remove rows with NAs |
| `replace_na()` | Replace NAs with a value |
| `fill()` | Fill NAs with nearby values |
| `complete()` | Make implicit missing explicit |
| `na.omit()` | Remove rows with NAs (base R) |
| `naniar::vis_miss()` | Visualize missing data |

## Before next class

📖 **Read:**

- R4DS Ch 19: Joins

✅ **Do:**

- ⚠️ **Finish Assignment 6 — due Sunday May 17 at 11:59 PM**
- Start Assignment 7 — the missing-data half is fair game now
- Submit your final project draft (due next Wednesday before class)
- Check your own final project data for missingness

## The one thing to remember

Missing data isn't a problem to solve — it's information about your study. Treat it that way.

See you Wednesday for joins!

::: {.notes}
End on this reframe. Students tend to think of NAs as a nuisance to get rid of. The takeaway is that missing data tells you something — about your study design, your participants, and potentially about your results.

**Three deadlines to call out by name:**
1. **Assignment 6 is due Sunday May 17 at 11:59 PM** — that's the most pressing. Linger on this one — it's only 4 days away.
2. Assignment 7 (Joins + Missing Data) was assigned today, due Sunday May 24. Tell them the missing-data half is doable now; joins comes Wednesday.
3. Final project **draft** (not the final report) is due next Wednesday May 20 before class.

Next session covers joins, which pairs naturally with missing data — joins are a major source of NAs.

The on-your-own exercise that follows is a take-home practice — point students to it but do not work through it in class.
:::

# On your own {background-color="#e67e22"}

## Practice: Longitudinal missing data

::: {.notes}
This is a take-home exercise — students do not need to finish it in class. It focuses entirely on the missing-data tools from today (no joins yet — that's next session). Task 1 uses complete(), tasks 2-3 combine summarize and filter with missing data logic, and task 4 reshapes wide and computes a change score. If they get stuck on task 4 in office hours, remind them: pivot_wider (Session 5) takes them from one-row-per-observation to one-row-per-participant, which is what they need for a wave-1-vs-wave-3 comparison. This exercise is directly relevant to Assignment 7.
:::

You have a three-wave study:

```{r}
longitudinal_study <- tibble(
  participant = c(1, 1, 1, 2, 2, 3, 3, 3, 4, 4),
  wave = c(1, 2, 3, 1, 2, 1, 2, 3, 1, 3),
  depression = c(25, 20, 15, 30, 28, 22, 18, 16, 20, NA),
  anxiety = c(28, 25, 22, 32, NA, 24, 20, 18, 22, 19)
)
```

## Your tasks

1. Make implicit missing waves **explicit** using `complete()`
2. Count how many assessments each participant completed
3. What percentage of participants completed all three waves?
4. Reshape the depression data so each participant has one row with columns for wave 1, 2, and 3. Then compute the wave-1-to-wave-3 change for participants with both timepoints. What's the mean change?

```{r}
#| echo: false
#| eval: false
# SOLUTION

# 1. Make implicit missing explicit
long_complete <- longitudinal_study |>
  complete(participant, wave)

long_complete

# 2. Count assessments per participant
long_complete |>
  group_by(participant) |>
  summarize(
    n_assessments = sum(!is.na(depression))
  )

# 3. Percentage completing all 3 waves
long_complete |>
  group_by(participant) |>
  summarize(
    n_assessments = sum(!is.na(depression))
  ) |>
  summarize(
    pct_complete = mean(n_assessments == 3)
  )

# 4. Reshape wide, then compute wave 1 → wave 3 change
long_complete |>
  select(participant, wave, depression) |>
  pivot_wider(
    names_from = wave,
    values_from = depression,
    names_prefix = "wave_"
  ) |>
  filter(!is.na(wave_1) & !is.na(wave_3)) |>
  mutate(change = wave_1 - wave_3) |>
  summarize(mean_change = mean(change))
```
