---
title: "Missing Data"
subtitle: "PSY 410: Data Science for Psychology"
author: "Dr. Sara Weston"
date: "May 13, 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    footer: "PSY 410 | Session 14"
    chalkboard: true
    code-line-numbers: false
    highlight-style: github
    incremental: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Setup {visibility="hidden"}

```{r}
#| include: false
library(tidyverse)
```

# Why missing data matters {background-color="#2c3e50"}

## Missing data is everywhere in psychology

- Participants skip survey questions
- People drop out of longitudinal studies
- Equipment fails during behavioral tasks
- Participants don't consent to share certain information

. . .

**The problem:** Missing data can bias your results if not handled properly.

## Today's agenda

1. **Types of missing data** — explicit vs implicit
2. **Exploring missingness** — how much? where? patterns?
3. **Handling missing data** — when to drop, when to fill, when to leave alone
4. **Making implicit missing explicit** — the `complete()` function

# Types of missing data {background-color="#2c3e50"}

## Explicit missing: NA

**Explicit missing** means you can see the `NA`:

```{r}
survey <- tibble(
  participant = 1:5,
  age = c(25, NA, 30, 22, NA),
  depression = c(12, 18, NA, 10, 15)
)

survey
```

The `NA` values are obvious.

## Implicit missing: Rows don't exist

**Implicit missing** means entire rows are absent:

```{r}
appointments <- tibble(
  name = c("Alice", "Bob", "Alice", "Carol"),
  day = c("Mon", "Mon", "Wed", "Wed"),
  attended = c(TRUE, TRUE, TRUE, TRUE)
)

appointments
```

. . .

Who didn't show up? You can't tell because they're not in the data!

## Why implicit missing matters

In longitudinal studies, missing rows often mean dropout:

```{r}
# Three-wave study
longitudinal <- tibble(
  id = c(1, 1, 1, 2, 2, 3),  # Person 2 missing wave 3, person 3 missing waves 2 and 3
  wave = c(1, 2, 3, 1, 2, 1),
  depression = c(20, 15, 12, 25, 22, 18)
)

longitudinal
```

Person 2 and 3's missing waves are **implicit** — they're not `NA`, they're just absent.

# Exploring missing data {background-color="#2c3e50"}

## Checking for NAs

```{r}
survey <- tibble(
  id = 1:6,
  age = c(25, NA, 30, 22, NA, 28),
  depression = c(12, 18, NA, 10, 15, NA),
  anxiety = c(15, 20, 12, NA, 18, 16)
)

# Check if any NAs exist
any(is.na(survey))
```

. . .

```{r}
# Count total NAs
sum(is.na(survey))
```

## NAs by column

```{r}
# Count NAs in each column
survey |>
  summarize(
    age_missing = sum(is.na(age)),
    depression_missing = sum(is.na(depression)),
    anxiety_missing = sum(is.na(anxiety))
  )
```

## Better approach: across()

```{r}
survey |>
  summarize(
    across(
      everything(),
      ~sum(is.na(.x))
    )
  )
```

. . .

Or as proportions:

```{r}
survey |>
  summarize(
    across(
      everything(),
      ~mean(is.na(.x))
    )
  )
```

## Which rows have missing data?

```{r}
# Show only rows with any NA
survey |>
  filter(if_any(everything(), is.na))
```

. . .

```{r}
# Show only complete cases
survey |>
  filter(if_all(everything(), ~!is.na(.x)))
```

## Visualizing missingness

The `naniar` package provides great visualization tools:

```{r}
library(naniar)

# Visual summary
vis_miss(survey)
```

## Patterns of missingness

```{r}
# Are certain combinations of variables missing together?
gg_miss_upset(survey)
```

. . .

This shows which combinations of variables are missing in the same rows.

# Handling missing data {background-color="#2c3e50"}

## Strategy 1: Complete case analysis

**Complete case analysis** (listwise deletion) removes any row with *any* `NA`:

```{r}
# Base R way
na.omit(survey)
```

. . .

```{r}
# tidyverse way
survey |>
  drop_na()
```

## Dangers of complete case analysis

```{r}
# Started with 6 participants
nrow(survey)

# Only 2 complete cases
survey |>
  drop_na() |>
  nrow()
```

. . .

::: {.callout-warning}
You just lost 67% of your data!
:::

## Selective dropping

Only drop rows missing *specific* variables:

```{r}
# Drop only if depression is missing
survey |>
  drop_na(depression)
```

. . .

```{r}
# Drop only if depression OR anxiety is missing
survey |>
  drop_na(depression, anxiety)
```

## When is dropping okay?

**Dropping is fine when:**

- Missing data is rare (< 5-10%)
- Missingness is truly random
- You have adequate sample size

. . .

**Be cautious when:**

- Missing data is common (> 20%)
- Certain groups have more missingness (systematic bias)
- Sample size is small

## Strategy 2: Filling values

Sometimes you can **reasonably fill** missing values:

```{r}
# Carry forward the last observation
time_series <- tibble(
  day = 1:5,
  mood = c(5, NA, NA, 4, 6)
)

time_series |>
  fill(mood)  # Fills downward by default
```

## Fill directions

```{r}
# Fill upward
time_series |>
  fill(mood, .direction = "up")
```

. . .

```{r}
# Fill both ways
time_series |>
  fill(mood, .direction = "updown")
```

## When is filling okay?

**Filling makes sense for:**

- Time series with repeated measures (carry forward last observation)
- Grouping variables that apply to multiple rows
- Values that don't change often

. . .

::: {.callout-warning}
**NEVER fill** when it means making up data you don't have!
:::

## Strategy 3: Replace with a specific value

```{r}
# Replace NAs with a value
survey |>
  mutate(
    age = replace_na(age, 99),  # Code 99 = "no response"
    depression = replace_na(depression, -999)  # Obvious invalid code
  )
```

. . .

::: {.callout-important}
If you use placeholder codes, **document them clearly** and make sure they can't be mistaken for real data.
:::

## Strategy 4: Leave them as NA

Often the best approach is to **keep NAs** and handle them in analysis:

```{r}
# Most functions have na.rm argument
survey |>
  summarize(
    mean_age = mean(age, na.rm = TRUE),
    mean_depression = mean(depression, na.rm = TRUE)
  )
```

. . .

This is transparent about what data you have.

# Implicit missing → Explicit missing {background-color="#2c3e50"}

## The problem with implicit missing

```{r}
study_completion <- tibble(
  participant = c(1, 1, 1, 2, 2, 3, 3),
  timepoint = c(1, 2, 3, 1, 2, 1, 3),
  depression = c(20, 15, 12, 25, 22, 18, 14)
)

study_completion
```

. . .

Who's missing which timepoints? Hard to tell.

## complete(): Make implicit missing explicit

```{r}
study_completion |>
  complete(participant, timepoint)
```

. . .

Now we can see: Participant 2 missing timepoint 3, Participant 3 missing timepoint 2.

## Why this matters

Makes dropout visible for analysis:

```{r}
study_completion |>
  complete(participant, timepoint) |>
  group_by(timepoint) |>
  summarize(
    n_completed = sum(!is.na(depression)),
    n_missing = sum(is.na(depression))
  )
```

## Filling after completing

Common pattern: make implicit explicit, then fill with a value:

```{r}
study_completion |>
  complete(participant, timepoint) |>
  mutate(
    completed = if_else(is.na(depression), FALSE, TRUE)
  )
```

# Pair coding break {background-color="#e67e22"}

## Your turn: Analyze missing data patterns

You have survey data from a therapy study:

```{r}
therapy_survey <- tibble(
  id = 1:8,
  age = c(25, 30, NA, 22, 28, NA, 35, 26),
  baseline_depression = c(22, 25, 18, 20, 24, 19, NA, 21),
  followup_depression = c(12, 23, NA, 15, NA, NA, NA, 16),
  satisfaction = c(4, 3, NA, 5, 4, NA, NA, 5)
)
```

1. How many participants are missing baseline data? Followup data?
2. How many participants have **complete data** (no NAs anywhere)?
3. Create a version that drops rows missing followup data
4. What percentage of participants completed the followup?

**Time: 10 minutes**

```{r}
#| echo: false
#| eval: false
# SOLUTION

# 1. Count missing by variable
therapy_survey |>
  summarize(
    across(everything(), ~sum(is.na(.x)))
  )

# 2. Complete cases
therapy_survey |>
  drop_na() |>
  nrow()

# 3. Drop if missing followup
therapy_survey |>
  drop_na(followup_depression)

# 4. Completion rate
therapy_survey |>
  summarize(
    completion_rate = mean(!is.na(followup_depression))
  )
```

# Psychology-specific considerations {background-color="#2c3e50"}

## Missing data mechanisms

Statisticians distinguish three types:

1. **MCAR (Missing Completely At Random)** — Missingness unrelated to anything
2. **MAR (Missing At Random)** — Missingness related to observed variables
3. **MNAR (Missing Not At Random)** — Missingness related to the missing value itself

## Example: Depression study

**MCAR:** Computer randomly failed to save 5% of responses

- No bias introduced

. . .

**MAR:** Older participants more likely to skip online surveys

- Can account for this by including age as a predictor

. . .

**MNAR:** People with severe depression skip the depression questionnaire

- **This is a problem** — missing values are related to what you're measuring

## Why it matters

- **MCAR:** Complete case analysis is fine (but you lose power)
- **MAR:** More sophisticated methods can help (beyond this course)
- **MNAR:** No easy fix — missing data is fundamentally informative

. . .

::: {.callout-tip}
**Your job:** Always **document and report** how much data is missing and why you think it's missing.
:::

## Attrition analysis

In longitudinal studies, always check *who* drops out:

```{r}
baseline <- tibble(
  id = 1:10,
  condition = rep(c("Treatment", "Control"), each = 5),
  baseline_depression = rnorm(10, 20, 5)
)

completers <- tibble(
  id = c(1, 2, 3, 7, 8, 9, 10)  # 4, 5, 6 dropped out
)

# Who dropped out?
baseline |>
  anti_join(completers, by = "id")
```

## Attrition by condition

```{r}
baseline |>
  anti_join(completers, by = "id") |>
  count(condition)
```

. . .

All 3 dropouts from Treatment condition — this could bias results!

## Reporting missing data

In your write-up, report:

1. **How much data is missing** (by variable)
2. **Patterns of missingness** (related to other variables?)
3. **How you handled it** (dropped? kept as NA?)
4. **Potential biases** (who's missing? does it matter?)

. . .

Example:

> "Eight participants (12%) did not complete the follow-up assessment. Dropout was unrelated to baseline depression scores (t = 1.2, p = .24) or treatment condition (χ² = 0.8, p = .37). Analyses used complete case analysis (N = 60)."

# Advanced topic: Multiple imputation {background-color="#2c3e50"}

## Beyond this course

More sophisticated approaches exist for handling missing data:

- **Multiple imputation** — create multiple plausible versions of missing data
- **Maximum likelihood** — estimate parameters using all available data
- **Bayesian methods** — incorporate uncertainty about missing values

. . .

Packages in R: `mice`, `Amelia`, `missForest`

. . .

::: {.callout-note}
We won't cover these methods, but know they exist for when you need them in future research!
:::

# End-of-deck exercise {background-color="#e67e22"}

## Practice: Longitudinal missing data

You have a three-wave study:

```{r}
longitudinal_study <- tibble(
  participant = c(1, 1, 1, 2, 2, 3, 3, 3, 4, 4),
  wave = c(1, 2, 3, 1, 2, 1, 2, 3, 1, 3),
  depression = c(25, 20, 15, 30, 28, 22, 18, 16, 20, NA),
  anxiety = c(28, 25, 22, 32, NA, 24, 20, 18, 22, 19)
)

demographics <- tibble(
  participant = 1:4,
  age = c(22, 30, 25, 28),
  condition = c("Treatment", "Control", "Treatment", "Control")
)
```

## Your tasks

1. Make implicit missing waves **explicit** using `complete()`
2. Join the demographics data
3. Count how many assessments each participant completed
4. Check if completion differs by treatment condition
5. Compute mean depression change (wave 1 to wave 3) for participants with complete data on those waves

```{r}
#| echo: false
#| eval: false
# SOLUTION

# 1. Make implicit missing explicit
long_complete <- longitudinal_study |>
  complete(participant, wave)

long_complete

# 2. Join demographics
long_with_demo <- long_complete |>
  left_join(demographics, by = "participant")

# 3. Count assessments per participant
long_with_demo |>
  group_by(participant) |>
  summarize(
    n_assessments = sum(!is.na(depression))
  )

# 4. Completion by condition
long_with_demo |>
  group_by(participant, condition) |>
  summarize(
    n_complete = sum(!is.na(depression)),
    .groups = "drop"
  ) |>
  group_by(condition) |>
  summarize(
    mean_completed = mean(n_complete)
  )

# 5. Depression change for complete cases
longitudinal_study |>
  filter(wave %in% c(1, 3)) |>
  group_by(participant) |>
  filter(n() == 2 & all(!is.na(depression))) |>  # Both waves present and non-missing
  summarize(
    change = depression[wave == 1] - depression[wave == 3]
  ) |>
  left_join(demographics, by = "participant")
```

# Wrapping up {background-color="#2c3e50"}

## Decision tree for missing data

1. **How much is missing?**
   - < 5%: Usually safe to drop
   - 5-20%: Investigate patterns
   - > 20%: Be very careful

2. **Why is it missing?**
   - Random: Less concerning
   - Systematic: Potentially biasing

3. **What's your plan?**
   - Drop complete cases?
   - Drop specific variables?
   - Keep as NA and use `na.rm`?
   - Fill (carefully)?

4. **Document everything!**

## Key takeaways

1. **Missing data is normal** in psychology research
2. **Explicit vs implicit** missing — make implicit explicit with `complete()`
3. **Explore patterns** before deciding how to handle
4. **Complete case analysis** (dropping rows) is simple but can lose power
5. **Never make up data** — be transparent about missingness
6. **Document your decisions** — report what's missing and why
7. **Check for bias** — does missingness relate to key variables?

## Functions cheat sheet

| Function | Purpose |
|----------|---------|
| `is.na()` | Check if values are missing |
| `drop_na()` | Remove rows with NAs |
| `replace_na()` | Replace NAs with a value |
| `fill()` | Fill NAs with nearby values |
| `complete()` | Make implicit missing explicit |
| `na.omit()` | Remove rows with NAs (base R) |
| `naniar::vis_miss()` | Visualize missing data |

## Before next class

📖 **Read:**

- R4DS Ch 11: Communication

✅ **Do:**

- Submit Assignment 7
- Check your final project for missing data
- Draft your final project report

## Questions?

See you Wednesday for storytelling with data!
