---
title: "Assignment 5 — Answer Key"
subtitle: "Exploratory Data Analysis"
author: "Instructor copy (for grading use)"
date: "2026-05-11"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(psych)

data(bfi)
bfi <- as_tibble(bfi)
```

::: {.callout-note}
## How to use this key

Each task shows a **canonical answer** plus any **other variants to accept**. The rubric is unchanged from the assignment page; this document only clarifies which student approaches should receive full credit.

Grading scale per task: full points = correct; full − 1 = one minor error (typo, formatting); half = one major error; 0 = not attempted or multiple major errors.
:::

::: {.callout-tip}
## Where to be lenient on this assignment

Assignment 5 is the first time students are asked to *interpret* what they see, not just produce code. Many tasks have several valid approaches and even more valid interpretations. As a general rule:

- Accept any **plot type that conveys the same information** (e.g., bar chart vs. histogram for a 1–6 Likert item — both are fine).
- Accept any **threshold** for outliers as long as the student can defend it.
- Accept any **pair of E items** in Part 2; the correlation's direction depends on the pair.
- Don't penalize students for noticing different things in the data — EDA generates different questions for different students. Look for evidence of *engagement*, not adherence to a script.
:::

---

## Part 1 — Exploring distributions (25 pts)

### Task 1.1 — Histograms for age and a personality item (8 pts)

**Canonical answer.**

```{r}
#| fig-width: 8
#| fig-height: 4
#| fig-alt: "Histogram of participant ages with binwidth 5, showing a strong right skew with most participants in their teens through thirties."
ggplot(bfi, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "#2c7fb8", color = "white") +
  labs(x = "Age (years)", y = "Count") +
  theme_minimal()
```

```{r}
#| fig-width: 8
#| fig-height: 4
#| fig-alt: "Histogram of responses to Extraversion item E1 on a 1–6 scale, showing higher response counts at the lower end of the scale because E1 is reverse-scored."
ggplot(bfi, aes(x = E1)) +
  geom_histogram(binwidth = 1, fill = "#2c7fb8", color = "white") +
  labs(x = "E1 response (1–6)", y = "Count") +
  theme_minimal()
```

**What we expect students to notice (any of these earn the interpretation points):**

- Age is **right-skewed** — most respondents are young, with a long tail.
- Personality items are on a 1–6 scale, **roughly symmetric or skewed depending on the item**, often with a mode at 4 or 5.
- Reverse-coded items (E1, E2, A1, C4, C5, O2, O5) skew opposite to the rest of their scale.

**Also accept:**

- `geom_bar()` instead of `geom_histogram()` for the personality item — items are integer-valued, so a bar chart is arguably *more* appropriate. Don't deduct.
- `geom_density()` for age — slightly off-spec ("histograms" was asked) but conveys the same info; leave as full credit and note in feedback.
- Any reasonable binwidth for age (1 to 10 are all defensible). A binwidth of 30+ that flattens the distribution is a minor error.
- Either passing `na.rm = TRUE` or letting ggplot drop NAs with a warning. The warning isn't a real problem.
- Picking any of the 25 personality items, not just E1.

**Common minor issues (−1):**

- No comment / interpretation at all → minor error, since the prompt asks "what do you notice?"
- Forgetting axis labels but otherwise correct.

**Major error (half):**

- Only one of the two required plots.
- Bar chart of `age` with one bar per unique value (no `binwidth`, no histogram) — this misses the point.

### Task 1.2 — Outlier detection in age (8 pts)

**Canonical answer (filter-based).**

```{r}
bfi |>
  filter(age > 60) |>
  count(age) |>
  arrange(desc(age))
```

The instructor reading: ages above ~60 are unusual in this sample (the bulk of respondents are 17–35), but they look real — no `999` or negative values. Worth keeping but flagging in a write-up.

**Also accept any of these approaches:**

```{r}
#| eval: false
# Boxplot-based identification
ggplot(bfi, aes(x = age)) + geom_boxplot()

# 1.5 × IQR rule (the rule the boxplot uses)
q <- quantile(bfi$age, c(0.25, 0.75), na.rm = TRUE)
iqr <- diff(q)
bfi |> filter(age < q[1] - 1.5 * iqr | age > q[2] + 1.5 * iqr)

# Z-score / standard-deviation rule
bfi |> filter(abs(age - mean(age, na.rm = TRUE)) / sd(age, na.rm = TRUE) > 3)

# Just sorting and looking
bfi |> arrange(desc(age)) |> head(20)
bfi |> arrange(age)        |> head(20)

# Implausible-value check (e.g., > 100 or < 13)
bfi |> filter(age > 100 | age < 13)
```

**Be lenient on:**

- The **threshold**. `> 60`, `> 65`, `> 70`, `≥ 50` — any defensible cutoff is fine, as long as the student says *why* (or the choice is obviously reasonable).
- Whether they identify "outliers" or "implausible values" — both framings are correct.
- Whether the student concludes the outliers are real or errors. The right answer is "they look real" but a student who flags them as suspicious and recommends investigation is also correct (that's exactly the EDA mindset).

**Major error (half):**

- Identifies outliers visually but doesn't show *which* ages are outliers (the prompt asks them to do it programmatically — i.e., extract the values, not just plot them).
- Hard-codes a single arbitrary threshold with no rationale and no visual check.

**Not an error:**

- Finding "no outliers" because the student picked a generous threshold like `> 90`. Don't deduct as long as the reasoning is consistent.

### Task 1.3 — Missing data summary (9 pts)

**Canonical answer (the pattern shown in the slides).**

```{r}
bfi |>
  summarize(across(everything(), ~ mean(is.na(.)) * 100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  arrange(desc(pct_missing))
```

The expected reading: `education` has the highest rate (~9.5%), several personality items have 1–2%, and `gender` has 0%.

**Also accept any of the following (all produce the same information):**

```{r}
#| eval: false
# Counts instead of percentages — the prompt says percentage, but if they
# show a count and clearly compute n/2800 somewhere, that's fine
bfi |>
  summarize(across(everything(), ~ sum(is.na(.))))

# colSums + division
colSums(is.na(bfi)) / nrow(bfi) * 100

# sapply / map_dbl
sapply(bfi, function(x) mean(is.na(x))) * 100
map_dbl(bfi, ~ mean(is.na(.))) * 100

# naniar::miss_var_summary() — above and beyond, full credit
naniar::miss_var_summary(bfi)

# A loop / per-column repetition is ugly but valid
# (don't deduct for style, deduct only if it doesn't produce the result)
```

**Be lenient on:**

- **Proportions vs. percentages.** Both convey the same thing. The prompt says "percentage," so a proportion (0.095 vs 9.5%) is at most a minor error.
- **Sorting.** Sorted output is nicer but not required.
- **Format.** Tibble, named vector, bar chart, kable — all fine.

**Major error (half):**

- Only reports missing data for *one* variable instead of all of them.
- Reports the row count of complete cases but never broken down by variable.

**Not an error:**

- Including `gender` and other 0%-missing variables in the table — that's correct, not noise.

---

## Part 2 — Exploring relationships (30 pts)

### Task 2.1 — Correlation between two Extraversion items (10 pts)

**Canonical answer.** The "right" answer depends on which pair the student picks. Reverse-scored items (E1, E2) flip the direction:

| Pair | Expected direction | Approx. r |
|---|---|---|
| E1, E2 | **positive** (both reverse-scored, so they agree) | ~ +0.45 |
| E3, E4 | **positive** (both forward-scored) | ~ +0.45 |
| E1, E3 / E1, E4 / E1, E5 | **negative** (E1 reverse-scored) | ~ −0.40 to −0.55 |
| E2, E3 / E2, E4 / E2, E5 | **negative** | ~ −0.30 to −0.45 |

```{r}
# E1 (reverse-scored "Don't talk a lot") vs. E4 ("Make friends easily")
cor(bfi$E1, bfi$E4, use = "complete.obs")

# E3 vs. E4 (both forward-scored)
cor(bfi$E3, bfi$E4, use = "complete.obs")
```

**What the interpretation should hit:**

- The **direction** (positive or negative) and what it means.
- Whether the **strength** is what they expected. Both items measure Extraversion, so a strong positive (or strong negative, if reverse-coded) is what theory predicts.
- Bonus credit if they explicitly notice that one item is reverse-coded and explains the direction. The hint at the end of the task points to this — students who connect the dots get full credit; students who report a negative correlation as "surprising — they're supposed to measure the same thing" without noticing reverse-coding still get full credit, since the prompt invited that observation.

**Be lenient on:**

- **The pair of items chosen.** Any two E items earn the points.
- The **direction** of the reported correlation, as long as the interpretation matches the number they computed. A student who reports `r = -0.49` between E1 and E4 and says "negatively related" is correct. A student who reports the same number and says "positively related" is wrong.
- Use of `use = "complete.obs"`, `use = "pairwise.complete.obs"`, or `drop_na()` upstream — all three work.

**Common minor issues (−1):**

- Computes the correlation but writes only "they are correlated" without giving direction or strength.
- Reports the correlation rounded to wildly different precision (e.g., "r = 0.0").

**Major error (half):**

- Computes correlation across all 25 items (`cor(bfi)`) without identifying the specific value they're discussing.
- Doesn't pass `use = "complete.obs"` (or equivalent) and reports `NA` as the correlation. (The hint specifically calls this out — but this is one of the most common student mistakes, so apply this leniently the first time and lean toward minor-error if the rest of the work is correct.)

**Not an error:**

- Picking E1 and E2 (both reverse-scored) and reporting a positive correlation as "expected." That's the right read.

### Task 2.2 — Scatterplot with trend line (10 pts)

**Canonical answer.**

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Scatterplot of E1 against E4 with jittered points and a linear trend line, showing a negative relationship between the two items."
ggplot(bfi, aes(x = E1, y = E4)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "E1 response", y = "E4 response") +
  theme_minimal()
```

The visual should match the correlation in 2.1: a negative trend line for E1 vs. E3/E4/E5, a positive one for same-direction pairs.

**Be lenient on:**

- **`geom_point()` vs. `geom_jitter()`.** Likert data overplots heavily — jitter is the better tool, but students who use plain points still see the trend, so plain points earn full credit.
- **Smoother method.** `geom_smooth(method = "lm")` is the canonical answer; default loess is also fine. Both communicate the relationship.
- **`se = TRUE` vs. `se = FALSE`.** Either works.
- **Axis ranges.** Don't deduct for ggplot's default 1–6 limits even if the data clusters in a corner.

**Common minor issues (−1):**

- No trend line at all (the prompt explicitly asks for one) — but if everything else is right, lean toward minor error.
- Plain points with no transparency, so the figure looks like a 6 × 6 grid of dots and you can't see density.

**Major error (half):**

- Scatterplot with a different pair of items than the one used in 2.1, with no acknowledgment.
- No interpretation linking back to the correlation in 2.1 (the prompt asks "does the visual relationship match what the correlation suggested?").

### Task 2.3 — Boxplots by education level (10 pts)

**Canonical answer.**

```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "Boxplots of Extraversion item E4 across five education levels, with education converted to a labeled factor. Distributions look broadly similar across education levels."
bfi |>
  filter(!is.na(education)) |>
  mutate(education = factor(
    education,
    levels = 1:5,
    labels = c("HS incomplete", "HS grad", "Some college", "College grad", "Graduate")
  )) |>
  ggplot(aes(x = education, y = E4)) +
  geom_boxplot(fill = "#2c7fb8", alpha = 0.7) +
  labs(x = "Education", y = "E4 response") +
  theme_minimal()
```

The expected interpretation: distributions look broadly similar across education levels — no large differences in this item by education. (This may vary by item; that's fine.)

**Be lenient on:**

- **The Extraversion item chosen.** Any of E1–E5 is fine.
- **Whether labels are added to the factor.** `factor(education)` (default labels 1–5) shows boxes per level — that's still 5 boxplots, not 1, so it satisfies the task. Adding labels is nicer but not required. Don't deduct.
- **Whether NAs are dropped.** ggplot will drop NAs and emit a warning. That's fine.
- **Whether they use `factor()` or `as.factor()`** — both produce something ggplot will treat as discrete.

**Common minor issues (−1):**

- Single squashed boxplot because the student forgot to convert `education` to a factor. The hint specifically warns against this — minor error, not half-credit, since the *intent* is clearly there.
- No interpretation comment at all.

**Major error (half):**

- Plots the wrong variable (e.g., education on the y-axis or a non-extraversion item).
- Bar chart, scatterplot, or other geom instead of a boxplot — the prompt explicitly asks for boxplots.

---

## Part 3 — Generating questions (20 pts)

### Task 3.1 — Three research questions (8 pts)

**There is no canonical answer here.** Look for questions that are:

1. **Investigable with this data** — i.e., they involve variables actually in the BFI dataset (the 25 personality items, gender, education, age).
2. **Not trivially answered** by what they've already done.
3. **Specific enough to plot or test** — "Are older participants more conscientious?" is investigable. "Why are people the way they are?" is not.

**Examples of strong questions:**

- "Does Conscientiousness increase with age?"
- "Are women higher in Agreeableness than men?"
- "Which Big Five trait shows the most variability across education levels?"
- "Are Neuroticism items more correlated with each other than Openness items?"
- "Is missingness on personality items related to age?"

**Examples of weak questions (still accept, but consider a minor deduction if all three are like this):**

- "What is the average age?" (Already answered by `summary()`.)
- "Are people happy?" (Not measurable in this data.)

**Grading guidance:**

- **Full credit:** three investigable, distinct questions. They do *not* need to all be brilliant — just plausible EDA prompts.
- **Minor (−1):** two strong questions and one weak one, OR a question phrased so vaguely that it would be hard to operationalize.
- **Half:** only one or two questions, or all three questions ask for things the data can't answer.
- **None:** task skipped or non-questions written instead.

### Task 3.2 — Visualization for one question (12 pts)

**Canonical example** (using "Does Conscientiousness increase with age?"):

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Scatterplot of mean Conscientiousness score against age with a smoothed trend line, showing a slight positive relationship between age and conscientiousness."
bfi |>
  mutate(C = rowMeans(across(C1:C5), na.rm = TRUE)) |>
  ggplot(aes(x = age, y = C)) +
  geom_jitter(alpha = 0.2) +
  geom_smooth() +
  labs(x = "Age", y = "Mean Conscientiousness", title = "Conscientiousness by age") +
  theme_minimal()
```

**Grading guidance:**

- The plot should **address the question they posed in 3.1**. Mismatch between question and plot is a major error.
- The plot should be **readable** — labels, sensible geom for the variable types.
- The interpretation comment should **answer the question** they asked, even if the answer is "no clear pattern."

**Be lenient on:**

- **Polish.** This is exploratory — minimal labels and default themes are fine.
- **Whether they compute a composite score** (e.g., mean of C1–C5) or just plot a single item. Both are valid for EDA.
- **Whether the answer is "yes" or "no" or "unclear."** Negative findings are findings.

**Major error (half):**

- Plot doesn't address any of the questions in 3.1.
- No interpretation comment at all (the prompt asks them to "explain your findings").
- Plot uses a wrong geom for the variable types (e.g., scatterplot of two categorical variables).

---

## Part 4 — EDA report (15 pts)

The 1-page (300–400 word) write-up. Look for **all four sections**:

1. **The data** — brief description (BFI, 2,800 participants, 25 items, demographics).
2. **Key findings** — 2–3 patterns from their work above. Should be specific (e.g., "E1 and E4 correlate at −0.49") rather than vague ("there are correlations").
3. **Data quality issues** — should mention missing data on `education` and possibly age outliers.
4. **Next steps** — at least one concrete follow-up.

**Grading guidance:**

| Quality | Pts | Looks like |
|---|---|---|
| Full | 15 | All four sections present, within 300–400 words, references specific findings from earlier tasks, written as narrative (not bullet points or code). |
| −1 | 13–14 | All four sections, but slightly under or over word count, OR one section is one sentence. Treat as a single minor issue. |
| Half | 7–8 | One section missing, OR written as a list of bullets without narrative, OR doesn't reference the actual findings (generic boilerplate). |
| None | 0 | Report missing or fewer than 100 words. |

**Be lenient on:**

- **Word count.** 250 or 450 words is fine if all four sections are addressed substantively. The 300–400 range is a target, not a hard rule.
- **Style.** Some students write tightly, some discursively. As long as the four sections are there and the content is real.
- **Whether the "key findings" repeat earlier tasks.** That's the point of a summary — synthesis, not novelty.

**Major error (half):**

- Bulleted answers under each header with no narrative — the prompt explicitly says "narrative text (not code comments)."
- Generic statements that don't reference *their* analysis (e.g., "Missing data is common in psychology research" without naming `education` or any item).

---

## QMD file knits without errors (10 pts)

Same criterion as Assignment 4. The file should render to HTML.

| Level | Pts | Looks like |
|---|---|---|
| Full | 10 | The submitted `.qmd` renders to HTML without errors. |
| −1 | 9 | Renders, but with one easily-fixed warning or a single broken inline reference. |
| Half | 5 | Renders only after manually setting `eval: false` on one chunk, OR the rendered HTML is missing one task's output. |
| None | 0 | Does not render at all; no HTML file submitted; or the HTML doesn't match the submitted `.qmd`. |

A student whose Task 2.1 code is wrong but who set `#| eval: false` on it so the file renders should still get Full here.

---

## PSY 510 — Graduate extension (preliminary analysis write-up)

Replaces Part 4 for grad students. Five sections required:

1. **Dataset description** (2–3 sentences)
2. **Data quality and exclusions** (1–2 sentences)
3. **Descriptive statistics** — table of means, SDs, ranges for ≥ 5 variables
4. **Key patterns** — 1–2 paragraphs, references to figures by number
5. **Analysis plan** — 1 paragraph

| Level | Pts | Looks like |
|---|---|---|
| Full | 15 | All five sections present in order, within 500–700 words, includes a `kable()` table for descriptives, references figures by number, reads as continuous prose. |
| −1 | 13–14 | All five sections but slightly off word count, OR the descriptives table uses `print()` rather than `kable()`. |
| Half | 7–8 | One section missing, OR no descriptives table at all, OR figures referenced only by code chunk label rather than figure number. |
| None | 0 | Write-up missing or fewer than 200 words. |

**Be lenient on:**

- Whether they include all 5 of their "key" variables in the descriptives table or pick a different 5+. Either is fine.
- Variable selection for the descriptives table — any 5 reasonable variables count.
- Slightly more or less than 500–700 words — same logic as the undergrad report.

**Major error (half):**

- Reads as a list of answers rather than a coherent piece of writing — the prompt explicitly asks for the latter.
- No interpretation in "Key patterns" — just raw output dumped under a heading.
